\documentclass{book}
\usepackage{makeidx}
\makeindex
\pagestyle{plain} \linespread{1}
\begin{document}
\frontmatter



Pride and \textbf{Prejudice}\index{prejudice}\stepcounter{footnote}\footnotetext{
\textbf{prejudice} \emph{n.}  (PREH-joo-diss)
an opinion that is preconceived and (usually) unfavorable.}

by Jane Austen


\mainmatter

\chapter{1}


It is a truth universally acknowledged, that a single man in
possession of a good fortune, must be in want of a wife.

However little known the feelings or views of such a man may
be on his first entering a neighbourhood, this truth is so well
fixed in the minds of the surrounding families, that he is considered
the rightful property of some one or other of their daughters.

"My dear Mr. Bennet," said his lady to him one day, "have you
heard that Netherfield Park is \textbf{let}\index{let}\stepcounter{footnote}\footnotetext{
\textbf{let} \emph{v.}  (LEHT)
 grant use or occupation of under terms of a contract.} at last?"

Mr. Bennet replied that he had not.

"But it is," returned she; "for Mrs. Long has just been here, and
she told me all about it."

Mr. Bennet made no answer.

"Do you not want to know who has taken it?" cried his wife
impatiently.

"\emph{you} want to tell me, and I have no objection to hearing it."

This was invitation enough.

"Why, my dear, you must know, Mrs. Long says that Netherfield
is taken by a young man of large fortune from the north of
England; that he came down on Monday in a \textbf{chaise}\index{chaise}\stepcounter{footnote}\footnotetext{
\textbf{chaise} \emph{n.}  (SHAYZ)
 a two-wheeled carriage drawn by one horse, often with a collapsible hood.} and four to
see the place, and was so much delighted with it, that he agreed
with Mr. Morris immediately; that he is to take possession
before \textbf{Michaelmas}\index{michaelmas}\stepcounter{footnote}\footnotetext{
\textbf{Michaelmas} \emph{n.}  (MIH-kuhl-muhs)
A Christian feast honoring the archangel Michael.}, and some of his servants are to be in the
house by the end of next week."

"What is his name?"

"Bingley."

"Is he married or single?"

"Oh!  Single, my dear, to be sure!  A single man of large
fortune; four or five thousand a year.  What a fine thing for our
girls!"

"How so?  How can it affect them?"

"My dear Mr. Bennet," replied his wife, "how can you be so
tiresome!  You must know that I am thinking of his marrying
one of them."

"Is that his design in settling here?"

"Design!  Nonsense, how can you talk so!  But it is very likely
that he \emph{may} fall in love with one of them, and therefore you
must visit him as soon as he comes."

"I see no occasion for that.  You and the girls may go, or you
may send them by themselves, which perhaps will be still
better, for as you are as handsome as any of them, Mr. Bingley
may like you the best of the party."

"My dear, you flatter me.  I certainly \emph{have} had my share of
beauty, but I do not pretend to be anything extraordinary now.
When a woman has five grown-up daughters, she ought to give
over thinking of her own beauty."

"In such cases, a woman has not often much beauty to think of."

"But, my dear, you must indeed go and see Mr. Bingley when
he comes into the neighbourhood."

"It is more than I engage for, I assure you."

"But consider your daughters.  Only think what an establishment
it would be for one of them.  Sir William and Lady Lucas are
determined to go, merely on that account, for in general, you
know, they visit no newcomers.  Indeed you must go, for it will
be impossible for \emph{us} to visit him if you do not."

"You are \textbf{over-scrupulous}\index{over-scrupulous}\stepcounter{footnote}\footnotetext{
\textbf{} \emph{}  ()
}, surely.  I dare say Mr. Bingley will
be very glad to see you; and I will send a few lines by you to
assure him of my hearty consent to his marrying whichever he
chooses of the girls; though I must throw in a good word for
my little Lizzy."

"I desire you will do no such thing.  Lizzy is not a bit better
than the others; and I am sure she is not half so handsome as
Jane, nor half so good-humoured as Lydia.  But you are always
giving \emph{her} the preference."

"They have none of them much to recommend them," replied he;
"they are all silly and ignorant like other girls; but Lizzy
has something more of quickness than her sisters."

"Mr. Bennet, how \emph{can} you abuse your own children in such a
way?  You take delight in vexing\index{vexing} me.  You have no compassion
for my poor nerves."

"You mistake me, my dear.  I have a high respect for your
nerves.  They are my old friends.  I have heard you mention
them with consideration these last twenty years at least."

Mr. Bennet was so odd a mixture of quick parts, \textbf{sarcastic}\index{sarcastic}\stepcounter{footnote}\footnotetext{
\textbf{sarcastic} \emph{adj.}  (sahr-KAS-tik)
expressing ridicule that wounds; marked by sarcasm.} humour,
reserve, and \textbf{caprice}\index{caprice}\stepcounter{footnote}\footnotetext{
\textbf{caprice} \emph{n.}  (ka-PREES)
A whimsical change of mind; an impulsive desire.}, that the experience of three-and-twenty
years had been insufficient to make his wife understand his
character.  \emph{her} mind was less difficult to develop.  She was a
woman of mean understanding, little information, and uncertain
temper.  When she was discontented, she fancied herself nervous.
The business of her life was to get her daughters married; its
\textbf{solace}\index{solace}\stepcounter{footnote}\footnotetext{
\textbf{solace} \emph{n.} \textbf{-ed}  (SOH-lihs)
the act of comforting in distress; consolation.} was visiting and news.



\chapter{2}


Mr. Bennet was among the earliest of those who waited on Mr.
Bingley.  He had always intended to visit him, though to the last
always assuring his wife that he should not go; and till the
evening after the visit was paid she had no knowledge of it.
It was then disclosed in the following manner.  Observing his
second daughter employed in trimming a hat, he suddenly
\textbf{addressed}\index{addressed}\stepcounter{footnote}\footnotetext{
\textbf{} \emph{}  ()
} her with:

"I hope Mr. Bingley will like it, Lizzy."

"We are not in a way to know \emph{what} Mr. Bingley likes," said
her mother resentfully, "since we are not to visit."

"But you forget, mamma," said Elizabeth, "that we shall meet
him at the assemblies, and that Mrs. Long promised to introduce
him."

"I do not believe Mrs. Long will do any such thing.  She has two
nieces of her own.  She is a selfish, \textbf{hypocritical}\index{hypocritical}\stepcounter{footnote}\footnotetext{
\textbf{hypocritical} \emph{adj.}  (hih-poh-KRIH-tih-kuhl)
professing feelings or virtues one does not have.} woman, and I
have no opinion of her."

"No more have I," said Mr. Bennet; "and I am glad to find that
you do not depend on her serving you."

Mrs. Bennet \textbf{deigned}\index{deigned}\stepcounter{footnote}\footnotetext{
\textbf{deign} \emph{v.} \textbf{-ed}  (DAYN)
 To condescend to give or bestow; to grant.} not to make any reply, but, unable to
contain herself, began scolding one of her daughters.

"Don't keep coughing so, Kitty, for Heaven's sake!  Have a little
compassion on my nerves.  You tear them to pieces."

"Kitty has no \textbf{discretion}\index{discretion}\stepcounter{footnote}\footnotetext{
\textbf{discretion} \emph{n.}  (dihs-KREH-shun)
knowing how to avoid embarrassment or distress; tact} in her coughs," said her father; "she
times them ill."

"I do not cough for my own amusement," replied Kitty fretfully\index{fretfully}.
"When is your next ball to be, Lizzy?"

"To-morrow \textbf{fortnight}\index{fortnight}\stepcounter{footnote}\footnotetext{
\textbf{fortnight} \emph{n.}  (FORT-nite)
a period of fourteen consecutive days.}."

"Aye, so it is," cried her mother, "and Mrs. Long does not come
back till the day before; so it will be impossible for her to
introduce him, for she will not know him herself."

"Then, my dear, you may have the advantage of your friend, and
introduce Mr. Bingley to \emph{her}."

"Impossible, Mr. Bennet, impossible, when I am not acquainted
with him myself; how can you be so teasing?"

"I honour your \textbf{circumspection}\index{circumspection}\stepcounter{footnote}\footnotetext{
\textbf{circumspection} \emph{n.}  (ser-kuhm-SPEHK-shun)
knowing how to avoid embarrassment or distress; prudent.}.  A fortnight\index{fortnight}'s acquaintance is
certainly very little.  One cannot know what a man really is by
the end of a fortnight\index{fortnight}.  But if \emph{we} do not venture somebody else
will; and after all, Mrs. Long and her daughters must stand their
chance; and, therefore, as she will think it an act of kindness,
if you decline the office, I will take it on myself."

The girls stared at their father.  Mrs. Bennet said only,
"Nonsense, nonsense!"

"What can be the meaning of that \textbf{emphatic}\index{emphatic}\stepcounter{footnote}\footnotetext{
\textbf{emphatic} \emph{adj.} \textbf{-ly}  (ehm-FAT-tik)
forceful and definite in expression or action.} exclamation?" cried
he.  "Do you consider the forms of introduction, and the stress
that is laid on them, as nonsense?  I cannot quite agree with
you \emph{there}.  What say you, Mary?  For you are a young lady of
deep reflection, I know, and read great books and make \textbf{extracts}\index{extracts}\stepcounter{footnote}\footnotetext{
\textbf{extract} \emph{tr.v.}  (EHX-trakt)
To draw or pull out; a thing in concentrate form.}."

Mary wished to say something sensible, but knew not how.

"While Mary is adjusting her ideas," he continued, "let\index{let} us return
to Mr. Bingley."

"I am sick of Mr. Bingley," cried his wife.

"I am sorry to hear \emph{that}; but why did not you tell me that
before?  If I had known as much this morning I certainly would
not have called on him.  It is very unlucky; but as I have
actually paid the visit, we cannot escape the acquaintance now."

The astonishment of the ladies was just what he wished; that of
Mrs. Bennet perhaps surpassing the rest; though, when the first
\textbf{tumult}\index{tumult}\stepcounter{footnote}\footnotetext{
\textbf{tumult} \emph{n.}  (TUH-muhlt)
 a state of commotion and loud confused noise.} of joy was over, she began to declare that it was what she
had expected all the while.

"How good it was in you, my dear Mr. Bennet!  But I knew I should
persuade you at last.  I was sure you loved your girls too well
to neglect such an acquaintance.  Well, how pleased I am! and it
is such a good joke, too, that you should have gone this morning
and never said a word about it till now."

"Now, Kitty, you may cough as much as you choose," said Mr.
Bennet; and, as he spoke, he left the room, \textbf{fatigued}\index{fatigued}\stepcounter{footnote}\footnotetext{
\textbf{} \emph{}  ()
} with the
\textbf{raptures}\index{raptures}\stepcounter{footnote}\footnotetext{
\textbf{rapture} \emph{n.}  (RAP-cher)
extacy; bliss, especially from a religious experience.} of his wife.

"What an excellent father you have, girls!" said she, when the
door was shut.  "I do not know how you will ever make him
\textbf{amends}\index{amends}\stepcounter{footnote}\footnotetext{
\textbf{amends} \emph{pl.n.}  (uh-MEHNDZ)
financial or other compensation for an injury or insult.} for his kindness; or me, either, for that matter.  At our
time of life it is not so pleasant, I can tell you, to be making
new acquaintances every day; but for your sakes, we would do
anything.  Lydia, my love, though you \emph{are} the youngest, I dare
say Mr. Bingley will dance with you at the next ball."

"Oh!" said Lydia \textbf{stoutly}\index{stoutly}\stepcounter{footnote}\footnotetext{
\textbf{stout} \emph{adj.} \textbf{-ly}  (STOWT)
 in a resolute manner; having determination or boldness.}, "I am not afraid; for though I \emph{am} the
youngest, I'm the tallest."

The rest of the evening was spent in \textbf{conjecturing}\index{conjecturing}\stepcounter{footnote}\footnotetext{
\textbf{conjecture} \emph{n.} \textbf{-ed}  (kun-JEK-tyoor)
Inference from incomplete evidence; a logical guess.} how soon he
would return Mr. Bennet's visit, and determining when they
should ask him to dinner.



\chapter{3}


Not all that Mrs. Bennet, however, with the assistance of her
five daughters, could ask on the subject\index{subject}, was sufficient to draw
from her husband any satisfactory description of Mr. Bingley.
They attacked him in various ways--with barefaced questions,
\textbf{ingenious}\index{ingenious}\stepcounter{footnote}\footnotetext{
\textbf{ingenious} \emph{adj.}  (ihn-JEEN-yuhs)
marked by brilliant intellect, imagination, skill, or creativity.} \textbf{suppositions}\index{suppositions}\stepcounter{footnote}\footnotetext{
\textbf{supposition} \emph{n.}  (suhp-puh-ZIH-shun)
an opinion based on incomplete evidence.}, and distant \textbf{surmises}\index{surmises}\stepcounter{footnote}\footnotetext{
\textbf{surmise} \emph{v.}  (ser-MYZ)
 To infer something based on inconclusive evidence.}; but he \textbf{eluded}\index{eluded}\stepcounter{footnote}\footnotetext{
\textbf{elude} \emph{tr.v.} \textbf{-ed}  (ee-LOOD)
 To evade or escape from persuers; to avoid capture.} the
skill of them all, and they were at last obliged to accept the
second-hand intelligence of their neighbour, Lady Lucas.  Her
report was highly favourable.  Sir William had been delighted
with him.  He was quite young, wonderfully handsome, extremely
agreeable, and, to crown the whole, he meant to be at the next
assembly with a large party.  Nothing could be more delightful!
To be fond of dancing was a certain step towards falling in love;
and very lively hopes of Mr. Bingley's heart were entertained.

"If I can but see one of my daughters happily settled at
Netherfield," said Mrs. Bennet to her husband, "and all the
others equally well married, I shall have nothing to wish for."

In a few days Mr. Bingley returned Mr. Bennet's visit, and sat
about ten \textbf{minutes}\index{minutes}\stepcounter{footnote}\footnotetext{
\textbf{} \emph{}  ()
} with him in his library.  He had entertained
hopes of being admitted to a sight of the young ladies, of
whose beauty he had heard much; but he saw only the father.
The ladies were somewhat more fortunate, for they had the
advantage of \textbf{ascertaining}\index{ascertaining}\stepcounter{footnote}\footnotetext{
\textbf{ascertain} \emph{tr.v.} \textbf{-ing}  (as-ser-TANE)
to understand, discover, or clarify with certainty.} from an upper window that he wore
a blue coat, and rode a black horse.

An invitation to dinner was soon afterwards \textbf{dispatched}\index{dispatched}\stepcounter{footnote}\footnotetext{
\textbf{despatch} \emph{v.} \textbf{-ed}  (dihs-PATCH)
to send off or away; same as 'dispatch'.}; and
already had Mrs. Bennet planned the courses that were to do
credit to her housekeeping, when an answer arrived which
\textbf{deferred}\index{deferred}\stepcounter{footnote}\footnotetext{
\textbf{defer} \emph{v.} \textbf{-ed}  (dih-FER)
 To postpone or put off.} it all.  Mr. Bingley was obliged to be in town the
following day, and, consequently, unable to accept the honour
of their invitation, etc.  Mrs. Bennet was quite \textbf{disconcerted}\index{disconcerted}\stepcounter{footnote}\footnotetext{
\textbf{disconcert} \emph{tr.v.} \textbf{-ed}  (dihs-kuhn-SER-ted)
to upset one's self-possession or self-bearing; to alarm.}.
She could not imagine what business he could have in town so
soon after his arrival in Hertfordshire; and she began to fear
that he might be always flying about from one place to another,
and never settled at Netherfield as he ought to be.  Lady Lucas
quieted her fears a little by starting the idea of his being gone
to London only to get a large party for the ball; and a report
soon followed that Mr. Bingley was to bring twelve ladies and
seven gentlemen with him to the assembly.  The girls \textbf{grieved}\index{grieved}\stepcounter{footnote}\footnotetext{
\textbf{grieve} \emph{v.} \textbf{-ed}  (GREEV)
}
over such a number of ladies, but were comforted the day
before the ball by hearing, that instead of twelve he brought
only six with him from London--his five sisters and a cousin.
And when the party entered the assembly room it consisted of
only five altogether--Mr. Bingley, his two sisters, the husband
of the eldest, and another young man.

Mr. Bingley was good-looking and gentlemanlike; he had a pleasant
\textbf{countenance}\index{countenance}\stepcounter{footnote}\footnotetext{
\textbf{countenance} \emph{n.}  (KOUN-teh-nance)
facial appearance or expression; to give permission.}, and easy, \textbf{unaffected}\index{unaffected}\stepcounter{footnote}\footnotetext{
\textbf{unaffected} \emph{adj.} \textbf{-ly}  (uhn-uh-FEK-tihd)
free or artificiality; sincere and genuine.} manners.  His sisters were fine
women, with an air of decided \textbf{fashion}\index{fashion}\stepcounter{footnote}\footnotetext{
\textbf{fashion} \emph{n.}  (FA-shun)
 to make out of components, often in an improvising manner.}.  His brother-in-law, Mr.
Hurst, merely looked the gentleman; but his friend Mr. Darcy soon
drew the attention of the room by his fine, tall person, handsome
features, noble \textbf{mien}\index{mien}\stepcounter{footnote}\footnotetext{
\textbf{mien} \emph{n.}  (MEE-yehn)
 bearing of mannor or conduct; dignified.}, and the report which was in general
\textbf{circulation}\index{circulation}\stepcounter{footnote}\footnotetext{
\textbf{circulation} \emph{n.}  (ser-kyoo-LAY-shun)
opening to public discussion and debate; distribution of a publication.} within five minutes\index{minutes} after his entrance, of his having
ten thousand a year.  The gentlemen pronounced him to be a fine
figure of a man, the ladies declared he was much handsomer than
Mr. Bingley, and he was looked at with great admiration for about
half the evening, till his manners gave a disgust which turned
the tide of his popularity; for he was discovered to be proud;
to be above his company, and above being pleased; and not all his
large estate in Derbyshire could then save him from having a most
forbidding, disagreeable countenance\index{countenance}, and being unworthy to be
compared with his friend.

Mr. Bingley had soon made himself acquainted with all the
\textbf{principal}\index{principal}\stepcounter{footnote}\footnotetext{
\textbf{principal} \emph{adj.} \textbf{-ly}  (PRIHN-sih-puhl)
initial money invested, not the income derived from it.} people in the room; he was lively and unreserved,
danced every dance, was angry that the ball closed so early,
and talked of giving one himself at Netherfield.  Such \textbf{amiable}\index{amiable}\stepcounter{footnote}\footnotetext{
\textbf{amiable} \emph{adj.}  (AY-myuh-bl)
good-natured and likable.}
qualities must speak for themselves.  What a contrast between
him and his friend!  Mr. Darcy danced only once with Mrs. Hurst
and once with Miss Bingley, declined being introduced to any
other lady, and spent the rest of the evening in walking about
the room, speaking occasionally to one of his own party.  His
character was decided.  He was the proudest, most disagreeable
man in the world, and everybody hoped that he would never come
there again.  Amongst the most violent against him was Mrs.
Bennet, whose dislike of his general behaviour was sharpened
into particular \textbf{resentment}\index{resentment}\stepcounter{footnote}\footnotetext{
\textbf{resentment} \emph{n.}  (rih-ZEHNT-mehnt)
a feeling of deep and bitter anger or ill-will.} by his having slighted one of her
daughters.

Elizabeth Bennet had been obliged, by the \textbf{scarcity}\index{scarcity}\stepcounter{footnote}\footnotetext{
\textbf{scarcity} \emph{n.}  (SKAIR-sih-tee)
a small and inadequate amount.} of gentlemen,
to sit down for two dances; and during part of that time,
Mr. Darcy had been standing near enough for her to hear a
conversation between him and Mr. Bingley, who came from the
dance for a few minutes\index{minutes}, to press his friend to join it.

"Come, Darcy," said he, "I must have you dance.  I hate to see
you standing about by yourself in this stupid manner.  You had
much better dance."

"I certainly shall not.  You know how I detest it, unless I am
particularly acquainted with my partner.  At such an assembly as
this it would be \textbf{insupportable}\index{insupportable}\stepcounter{footnote}\footnotetext{
\textbf{insupportable} \emph{adj.}  (ihn-suhp-POR-tuh-bl)
incapable of being justified or explained; intolerable.}.  Your sisters are engaged, and
there is not another woman in the room whom it would not be a
punishment to me to stand up with."

"I would not be so \textbf{fastidious}\index{fastidious}\stepcounter{footnote}\footnotetext{
\textbf{fastidious} \emph{adj.}  (fas-TIH-dee-yus)
careful attention to detail; overly concerned with cleanliness.} as you are," cried Mr. Bingley,
"for a kingdom!  Upon my honour, I never met with so many
pleasant girls in my life as I have this evening; and there are
several of them you see uncommonly pretty."

"\emph{you} are dancing with the only handsome girl in the room,"
said Mr. Darcy, looking at the eldest Miss Bennet.

"Oh! She is the most beautiful creature I ever beheld!  But
there is one of her sisters sitting down just behind you, who is
very pretty, and I dare say very agreeable.  Do let\index{let} me ask my
partner to introduce you."

"Which do you mean?" and turning round he looked for a
moment at Elizabeth, till catching her eye, he withdrew his own
and coldly said: "She is \textbf{tolerable}\index{tolerable}\stepcounter{footnote}\footnotetext{
\textbf{tolerable} \emph{adj.} \textbf{-ly}  (TAW-ler-uh-bl)
able to be tolerated or endured.}, but not handsome enough to
tempt \emph{me}; I am in no humour at present to give \textbf{consequence}\index{consequence}\stepcounter{footnote}\footnotetext{
\textbf{consequence} \emph{n.}  (KAHN-sih-kwense)
the natural or unavoidable result of an action.}
to young ladies who are slighted by other men.  You had better
return to your partner and enjoy her smiles, for you are wasting
your time with me."

Mr. Bingley followed his advice.  Mr. Darcy walked off; and
Elizabeth remained with no very \textbf{cordial}\index{cordial}\stepcounter{footnote}\footnotetext{
\textbf{cordial} \emph{adj.} \textbf{-ly}  (KOR-juhl)
friendly, warm, or sincere; a strong and highly flavored sweet liquor usually drunk after a meal.} feelings toward him.
She told the story, however, with great spirit among her friends;
for she had a lively, playful \textbf{disposition}\index{disposition}\stepcounter{footnote}\footnotetext{
\textbf{disposition} \emph{n.}  (dih-puh-ZIH-shun)
One's usual mood; temperament.}, which delighted in
anything ridiculous.

The evening altogether passed off pleasantly to the whole
family.  Mrs. Bennet had seen her eldest daughter much
admired by the Netherfield party.  Mr. Bingley had danced with
her twice, and she had been distinguished by his sisters.  Jane
was as much gratified by this as her mother could be, though in
a quieter way.  Elizabeth felt Jane's pleasure.  Mary had heard
herself mentioned to Miss Bingley as the most accomplished
girl in the neighbourhood; and Catherine and Lydia had been
fortunate enough never to be without partners, which was all
that they had yet learnt to care for at a ball.  They returned,
therefore, in good spirits to Longbourn, the village where they
lived, and of which they were the principal\index{principal} inhabitants.  They
found Mr. Bennet still up.  With a book he was regardless of
time; and on the present occasion he had a good deal of
curiosity as to the events of an evening which had raised such
splendid expectations.  He had rather hoped that his wife's
views on the stranger would be disappointed; but he soon
found out that he had a different story to hear.

"Oh! my dear Mr. Bennet," as she entered the room, "we have
had a most delightful evening, a most excellent ball.  I wish you
had been there.  Jane was so admired, nothing could be like it.
Everybody said how well she looked; and Mr. Bingley thought
her quite beautiful, and danced with her twice!  Only think of
\emph{that}, my dear; he actually danced with her twice! and she was
the only creature in the room that he asked a second time.
First of all, he asked Miss Lucas.  I was so \textbf{vexed}\index{vexed}\stepcounter{footnote}\footnotetext{
\textbf{vex} \emph{tr.v.} \textbf{-ed}  (VEKS)
 to trouble or bother, especially with petty annoyances.} to see him
stand up with her!  But, however, he did not admire her at all;
indeed, nobody can, you know; and he seemed quite struck with
Jane as she was going down the dance.  So he inquired who she
was, and got introduced, and asked her for the two next.  Then
the two third he danced with Miss King, and the two fourth with
Maria Lucas, and the two fifth with Jane again, and the two
sixth with Lizzy, and the \emph{\textbf{boulanger}\index{boulanger}\stepcounter{footnote}\footnotetext{
\textbf{Boulanger} \emph{1}  (boo-lawn-ZHAY)
French music teacher of several modern American composers, including Virgil Thomson and Aaron Copland.}}--"

"If he had had any compassion for \emph{me}," cried her husband
impatiently, "he would not have danced half so much!  For God's
sake, say no more of his partners.  O that he had sprained
his ankle in the first place!"

"Oh! my dear, I am quite delighted with him.  He is so
excessively handsome!  And his sisters are charming women.
I never in my life saw anything more elegant than their dresses.
I dare say the lace upon Mrs. Hurst's gown--"

Here she was interrupted again.  Mr. Bennet protested against
any description of finery.  She was therefore obliged to seek
another branch of the subject\index{subject}, and related, with much bitterness
of spirit and some exaggeration, the shocking rudeness of Mr.
Darcy.

"But I can assure you," she added, "that Lizzy does not lose
much by not suiting \emph{his} fancy; for he is a most disagreeable,
horrid man, not at all worth pleasing.  So high and so \textbf{conceited}\index{conceited}\stepcounter{footnote}\footnotetext{
\textbf{conceit} \emph{n.} \textbf{-ed}  (kuhn-SEE-tehd)
 false pride; an exaggerated sense of self-importance.}
that there was no enduring him!  He walked here, and he walked
there, fancying himself so very great!  Not handsome enough to
dance with!  I wish you had been there, my dear, to have given
him one of your set-downs.  I quite detest the man."



\chapter{4}


When Jane and Elizabeth were alone, the former, who had been
cautious in her praise of Mr. Bingley before, expressed to her
sister just how very much she admired him.

"He is just what a young man ought to be," said she, "sensible,
good-humoured, lively; and I never saw such happy manners!--so
much ease, with such perfect good breeding!"

"He is also handsome," replied Elizabeth, "which a young man
ought likewise to be, if he possibly can.  His character is thereby
complete."

"I was very much flattered by his asking me to dance a second
time.  I did not expect such a compliment."

"Did not you? I did for you.  But that is one great difference
between us.  Compliments always take \emph{you} by surprise, and
\emph{me} never.  What could be more natural than his asking you
again?  He could not help seeing that you were about five times
as pretty as every other woman in the room.  No thanks to his
\textbf{gallantry}\index{gallantry}\stepcounter{footnote}\footnotetext{
\textbf{gallantry} \emph{n.}  (GAL-luhn-tree)
heroic courage in face of danger; courtesy towards women.} for that.  Well, he certainly is very agreeable, and I
give you leave to like him.  You have liked many a stupider
person."

"Dear Lizzy!"

"Oh! you are a great deal too \textbf{apt}\index{apt}\stepcounter{footnote}\footnotetext{
\textbf{apartment} \emph{n.}  (uh-PART-mehnt)
}, you know, to like people in
general.  You never see a fault in anybody.  All the world are
good and agreeable in your eyes.  I never heard you speak ill of
a human being in your life."

"I would not wish to be hasty in \textbf{censuring}\index{censuring}\stepcounter{footnote}\footnotetext{
\textbf{censure} \emph{n.} \textbf{-ed}  (SEHN-shoor)
 A strong expression of criticism or disapproval.} anyone; but I always
speak what I think."

"I know you do; and it is \emph{that} which makes the wonder.  With \emph{your}
good sense, to be so honestly blind to the \textbf{follies}\index{follies}\stepcounter{footnote}\footnotetext{
\textbf{folly} \emph{n.}  (FAW-lee)
 a foolish act lacking judgement or foresight.} and nonsense
of others!  \textbf{Affectation}\index{affectation}\stepcounter{footnote}\footnotetext{
\textbf{affectation} \emph{n.}  (af-fek-TAY-shun)
a deliberate pretense or exaggerated display.} of \textbf{candour}\index{candour}\stepcounter{footnote}\footnotetext{
\textbf{} \emph{}  ()
} is common enough--one meets
with it everywhere.  But to be \textbf{candid}\index{candid}\stepcounter{footnote}\footnotetext{
\textbf{candid} \emph{adj.}  (KAN-did)
blunt directness in manner or speech; without subtlety or evasion.} without \textbf{ostentation}\index{ostentation}\stepcounter{footnote}\footnotetext{
\textbf{ostentation} \emph{n.}  (ahs-tehn-TAY-shun)
a pretentious, showy, or vulgar display meant to impress others.} or
design--to take the good of everybody's character and make it
still better, and say nothing of the bad--belongs to you alone.
And so you like this man's sisters, too, do you?  Their manners
are not equal to his."

"Certainly not--at first.  But they are very pleasing women when
you converse with them.  Miss Bingley is to live with her
brother, and keep his house; and I am much mistaken if we shall
not find a very charming neighbour in her."

Elizabeth listened in silence, but was not convinced; their
behaviour at the assembly had not been calculated to please in
general; and with more quickness of observation and less \textbf{pliancy}\index{pliancy}\stepcounter{footnote}\footnotetext{
\textbf{pliant} \emph{adj.}  (PLY-ant)
Easily bent; easily adaptable.}
of temper than her sister, and with a judgement too \textbf{unassailed}\index{unassailed}\stepcounter{footnote}\footnotetext{
\textbf{} \emph{}  ()
} by
any attention to herself, she was very little \textbf{disposed}\index{disposed}\stepcounter{footnote}\footnotetext{
\textbf{dispose} \emph{v.} \textbf{-ed}  (dihs-POHZ)
 a personal inclination (usually followed by `to').} to approve
them.  They were in fact very fine ladies; not deficient in good
humour when they were pleased, nor in the power of making
themselves agreeable when they chose it, but proud and
conceited\index{conceited}.  They were rather handsome, had been educated in
one of the first \textbf{private}\index{private}\stepcounter{footnote}\footnotetext{
\textbf{private} \emph{adj.}  (PRY-viht)
 an enlisted soldier of the lowest rank.} \textbf{seminaries}\index{seminaries}\stepcounter{footnote}\footnotetext{
\textbf{seminary} \emph{n.}  (SEH-mih-nair-ee)
A school, especially a one for religious instruction.} in town, had a fortune of
twenty thousand pounds, were in the habit of spending more
than they ought, and of associating with people of rank, and
were therefore in every respect entitled to think well of
themselves, and meanly of others.  They were of a respectable
family in the north of England; a circumstance more deeply
impressed on their memories than that their brother's fortune
and their own had been acquired by trade.

Mr. Bingley inherited property to the amount of nearly a
hundred thousand pounds from his father, who had intended to
purchase an estate, but did not live to do it.  Mr. Bingley
intended it likewise, and sometimes made choice of his county;
but as he was now provided with a good house and the liberty of
a manor, it was doubtful to many of those who best knew the
easiness of his temper, whether he might not spend the
remainder of his days at Netherfield, and leave the next
generation to purchase.

His sisters were anxious for his having an estate of his own; but,
though he was now only established as a tenant, Miss Bingley
was by no means unwilling to preside at his table--nor was Mrs.
Hurst, who had married a man of more fashion\index{fashion} than fortune, less
disposed\index{disposed} to consider his house as her home when it suited her.
Mr. Bingley had not been of age two years, when he was tempted
by an accidental recommendation to look at Netherfield House.
He did look at it, and into it for half-an-hour--was pleased with
the situation and the principal\index{principal} rooms, satisfied with what the
owner said in its praise, and took it immediately.

Between him and Darcy there was a very steady\index{steady} friendship, in
spite of great opposition of character.  Bingley was \textbf{endeared}\index{endeared}\stepcounter{footnote}\footnotetext{
\textbf{endear} \emph{tr.v.} \textbf{-ed}  (ehn-DEER)
} to
Darcy by the easiness, openness, and \textbf{ductility}\index{ductility}\stepcounter{footnote}\footnotetext{
\textbf{ductile} \emph{adj.}  (DUHK-tihl)
Easily formed into wire or hammered into thin sheets.} of his temper,
though no disposition\index{disposition} could offer a greater contrast to his own,
and though with his own he never appeared dissatisfied.  On the
strength of Darcy's regard, Bingley had the firmest reliance, and
of his judgement the highest opinion.  In understanding, Darcy
was the superior\index{superior}.  Bingley was by no means deficient, but Darcy
was clever.  He was at the same time \textbf{haughty}\index{haughty}\stepcounter{footnote}\footnotetext{
\textbf{haughty} \emph{adj.}  (HAWT-tee)
showing arrogant superiority to and disdain of those one views as inferior.}, reserved, and
fastidious\index{fastidious}, and his manners, though well-bred, were not inviting.
In that respect his friend had greatly the advantage.  Bingley was
sure of being liked wherever he appeared, Darcy was continually
giving offense.

The manner in which they spoke of the Meryton assembly was
sufficiently characteristic.  Bingley had never met with more
pleasant people or prettier girls in his life; everybody had been
most kind and attentive to him; there had been no formality, no
stiffness; he had soon felt acquainted with all the room; and, as
to Miss Bennet, he could not conceive an angel more beautiful.
Darcy, on the contrary, had seen a collection of people in whom
there was little beauty and no fashion\index{fashion}, for none of whom he had
felt the smallest interest, and from none received either attention
or pleasure.  Miss Bennet he acknowledged to be pretty, but she
smiled too much.

Mrs. Hurst and her sister allowed it to be so--but still they
admired her and liked her, and pronounced her to be a sweet
girl, and one whom they would not object to know more of.
Miss Bennet was therefore established as a sweet girl, and their
brother felt authorized by such commendation\index{commendation} to think of her as
he chose.



\chapter{5}


Within a short walk of Longbourn lived a family with whom
the Bennets were particularly intimate.  Sir William Lucas
had been formerly in trade in Meryton, where he had made a
tolerable\index{tolerable} fortune, and risen to the honour of knighthood by an
address\index{address} to the king during his \textbf{mayoralty}\index{mayoralty}\stepcounter{footnote}\footnotetext{
\textbf{mayoralty} \emph{n.}  (MAY-or-uhl-tee)
the position of mayor.}.  The distinction had
perhaps been felt too strongly.  It had given him a disgust
to his business, and to his residence in a small market town;
and, in quitting them both, he had removed with his family
to a house about a mile from Meryton, denominated from that
period Lucas Lodge, where he could think with pleasure of his
own importance, and, \textbf{unshackled}\index{unshackled}\stepcounter{footnote}\footnotetext{
\textbf{unshackle} \emph{tr.v.} \textbf{-ed}  (uhn-SHA-kl)
To free from shackles or chains.} by business, occupy himself
\textbf{solely}\index{solely}\stepcounter{footnote}\footnotetext{
\textbf{solely} \emph{adv.}  (SOAL-lee)
 alone; without any others involved.} in being civil to all the world.  For, though elated by his
rank, it did not render\index{render} him \textbf{supercilious}\index{supercilious}\stepcounter{footnote}\footnotetext{
\textbf{supercilious} \emph{adj.} \textbf{-ness}  (soo-per-SIH-lih-yus)
feeling or expressing contempt or disdain.}; on the contrary, he was
all attention to everybody.  By nature inoffensive, friendly, and
obliging, his presentation at St. James's had made him courteous.

Lady Lucas was a very good kind of woman, not too clever to
be a valuable neighbour to Mrs. Bennet.  They had several
children.  The eldest of them, a sensible, intelligent young
woman, about twenty-seven, was Elizabeth's intimate friend.

That the Miss Lucases and the Miss Bennets should meet to
talk over a ball was absolutely necessary; and the morning after
the assembly brought the former to Longbourn to hear and to
communicate.

"\emph{you} began the evening well, Charlotte," said Mrs. Bennet with
civil self-command to Miss Lucas.  "\emph{you} were Mr. Bingley's
first choice."

"Yes; but he seemed to like his second better."

"Oh! you mean Jane, I suppose, because he danced with her
twice.  To be sure that \emph{did} seem as if he admired her--indeed
I rather believe he \emph{did}--I heard something about it--but I
hardly know what--something about Mr. Robinson."

"Perhaps you mean what I overheard between him and Mr. Robinson;
did not I mention it to you?  Mr. Robinson's asking him how he
liked our Meryton assemblies, and whether he did not think there
were a great many pretty women in the room, and \emph{which} he thought
the prettiest? and his answering immediately to the last
question: 'Oh! the eldest Miss Bennet, beyond a doubt; there
cannot be two opinions on that point.'"

"Upon my word!  Well, that is very decided indeed--that does
seem as if--but, however, it may all come to nothing, you know."

"\emph{my} overhearings were more to the purpose than \emph{yours}, Eliza,"
said Charlotte.  "Mr. Darcy is not so well worth listening to
as his friend, is he?--poor Eliza!--to be only just \emph{tolerable\index{tolerable}}."

"I beg you would not put it into Lizzy's head to be vexed\index{vexed} by
his ill-treatment, for he is such a disagreeable man, that it
would be quite a misfortune to be liked by him.  Mrs. Long
told me last night that he sat close to her for half-an-hour
without once opening his lips."

"Are you quite sure, ma'am?--is not there a little mistake?"
said Jane.  "I certainly saw Mr. Darcy speaking to her."

"Aye--because she asked him at last how he liked Netherfield,
and he could not help answering her; but she said he seemed
quite angry at being spoke to."

"Miss Bingley told me," said Jane, "that he never speaks much,
unless among his intimate acquaintances.  With \emph{them} he is
remarkably agreeable."

"I do not believe a word of it, my dear.  If he had been so very
agreeable, he would have talked to Mrs. Long.  But I can guess
how it was; everybody says that he is eat up with pride, and I
dare say he had heard somehow that Mrs. Long does not keep
a carriage, and had come to the ball in a \textbf{hack}\index{hack}\stepcounter{footnote}\footnotetext{
\textbf{hack} \emph{v.}  (HAK)
 cough spasmodically.} chaise\index{chaise}."

"I do not mind his not talking to Mrs. Long," said Miss Lucas,
"but I wish he had danced with Eliza."

"Another time, Lizzy," said her mother, "I would not dance
with \emph{him}, if I were you."

"I believe, ma'am, I may safely promise you \emph{never} to dance
with him."

"His pride," said Miss Lucas, "does not offend \emph{me} so much as
pride often does, because there is an excuse for it.  One cannot
wonder that so very fine a young man, with family, fortune,
everything in his favour, should think highly of himself.  If I
may so express it, he has a \emph{right} to be proud."

"That is very true," replied Elizabeth, "and I could easily
forgive \emph{his} pride, if he had not mortified \emph{mine}."

"Pride," observed Mary, who \textbf{piqued}\index{piqued}\stepcounter{footnote}\footnotetext{
\textbf{pique} \emph{n.} \textbf{-ed}  (PEEK)
 a feeling of wounded pride.} herself upon the \textbf{solidity}\index{solidity}\stepcounter{footnote}\footnotetext{
\textbf{solidity} \emph{n.}  (soh-LIH-dih-tee)
the state of being solid.}
of her reflections, "is a very common failing, I believe.  By
all that I have ever read, I am convinced that it is very common
indeed; that human nature is particularly prone to it, and
that there are very few of us who do not cherish a feeling of
XXXself-\textbf{complacency}\index{complacency}\stepcounter{footnote}\footnotetext{
\textbf{complacency} \emph{n.}  (kuhm-PLAY-suhn-see)
a feeling of ease or satisfaction, often causing unawareness of danger.} XXX on the score of some quality or other, real
or imaginary.  \textbf{Vanity}\index{vanity}\stepcounter{footnote}\footnotetext{
\textbf{vanity} \emph{n.}  (VA-nih-tee)
} and pride are different things, though
the words are often used synonymously.  A person may be proud
without being vain.  Pride relates more to our opinion of
ourselves, vanity\index{vanity} to what we would have others think of us."

"If I were as rich as Mr. Darcy," cried a young Lucas, who
came with his sisters, "I should not care how proud I was.  I
would keep a pack of foxhounds, and drink a bottle of wine a
day."

"Then you would drink a great deal more than you ought," said
Mrs. Bennet; "and if I were to see you at it, I should take away
your bottle directly."

The boy protested that she should not; she continued to declare
that she would, and the argument ended only with the visit.



\chapter{6}


The ladies of Longbourn soon waited on those of Netherfield.
The visit was soon returned in due form.  Miss Bennet's
pleasing manners grew on the goodwill of Mrs. Hurst and Miss
Bingley; and though the mother was found to be \textbf{intolerable}\index{intolerable}\stepcounter{footnote}\footnotetext{
\textbf{intolerable} \emph{adj.}  (ihn-TAW-ler-uh-bl)
impossible to tolerate or endure; completely unacceptable.},
and the younger sisters not worth speaking to, a wish of
being better acquainted with \emph{them} was expressed towards
the two eldest.  By Jane, this attention was received with the
greatest pleasure, but Elizabeth still saw superciliousness in
their treatment of everybody, hardly excepting even her sister,
and could not like them; though their kindness to Jane, such as it
was, had a value as arising in all probability from the influence
of their brother's admiration.  It was generally evident
whenever they met, that he \emph{did} admire her and to \emph{her} it was
equally evident that Jane was yielding to the preference which
she had begun to entertain for him from the first, and was in a
way to be very much in love; but she considered with pleasure
that it was not likely to be discovered by the world in general,
since Jane united, with great strength of feeling, a \textbf{composure}\index{composure}\stepcounter{footnote}\footnotetext{
\textbf{composure} \emph{n.}  (kuhm-POH-zher)
steadiness of mind under stress.}
of temper and a uniform cheerfulness of manner which would
guard her from the suspicions of the \textbf{impertinent}\index{impertinent}\stepcounter{footnote}\footnotetext{
\textbf{impertinent} \emph{adj.}  (im-PER-tih-nent)
improperly forward or bold.}.  She
mentioned this to her friend Miss Lucas.

"It may perhaps be pleasant," replied Charlotte, "to be able to
impose\index{impose} on the public in such a case; but it is sometimes a
disadvantage to be so very guarded.  If a woman conceals her
affection with the same skill from the object of it, she may lose
the opportunity of fixing him; and it will then be but poor
\textbf{consolation}\index{consolation}\stepcounter{footnote}\footnotetext{
\textbf{consolation} \emph{n.}  (kahn-soh-LAY-shun)
the act of consoling; giving relief to someone in affliction.} to believe the world equally in the dark.  There is
so much of gratitude or vanity\index{vanity} in almost every attachment, that
it is not safe to leave any to itself.  We can all \emph{begin} freely--a
slight preference is natural enough; but there are very few of us
who have heart enough to be really in love without encouragement.
In nine cases out of ten a women had better show \emph{more} affection
than she feels.  Bingley likes your sister undoubtedly; but he
may never do more than like her, if she does not help him on."

"But she does help him on, as much as her nature will allow.
If I can \textbf{perceive}\index{perceive}\stepcounter{footnote}\footnotetext{
\textbf{perceive} \emph{tr.v.}  (per-SEEV)
to become aware of through the senses.} her regard for him, he must be a \textbf{simpleton}\index{simpleton}\stepcounter{footnote}\footnotetext{
\textbf{simpleton} \emph{n.}  (SIM-puhl-tuhn)
a person lacking intelligence or common sense.},
indeed, not to discover it too."

"Remember, Eliza, that he does not know Jane's disposition\index{disposition} as
you do."

"But if a woman is \textbf{partial}\index{partial}\stepcounter{footnote}\footnotetext{
\textbf{partial} \emph{adj.}  (PAR-shuhl)
having a strong preference or liking for.} to a man, and does not \textbf{endeavour}\index{endeavour}\stepcounter{footnote}\footnotetext{
\textbf{endeavour} \emph{n.}  (ehn-DEH-ver)
a project or undertaking; to try; variant of endeavor.} to
conceal it, he must find it out."

"Perhaps he must, if he sees enough of her.  But, though
Bingley and Jane meet tolerably often, it is never for many
hours together; and, as they always see each other in large
mixed parties, it is impossible that every moment should be
employed in conversing together.  Jane should therefore make
the most of every half-hour in which she can command his
attention.  When she is secure of him, there will be more leisure
for falling in love as much as she chooses."

"Your plan is a good one," replied Elizabeth, "where nothing is
in question but the desire of being well married, and if I were
determined to get a rich husband, or any husband, I dare say I
should adopt it.  But these are not Jane's feelings; she is not
acting by design.  As yet, she cannot even be certain of the
degree of her own regard nor of its reasonableness.  She has
known him only a fortnight\index{fortnight}.  She danced four dances with him
at Meryton; she saw him one morning at his own house, and
has since dined with him in company four times.  This is not
quite enough to make her understand his character."

"Not as you represent it.  Had she merely \emph{dined} with him, she
might only have discovered whether he had a good appetite; but
you must remember that four evenings have also been spent
together--and four evenings may do a great deal."

"Yes; these four evenings have enabled them to ascertain\index{ascertain} that
they both like Vingt-un better than Commerce; but with respect
to any other leading characteristic, I do not imagine that much
has been unfolded."

"Well," said Charlotte, "I wish Jane success with all my heart;
and if she were married to him to-morrow, I should think she
had as good a chance of happiness as if she were to be studying
his character for a twelvemonth.  Happiness in marriage is
entirely  a matter of chance.  If the dispositions\index{dispositions} of the parties
are ever so well known to each other or ever so similar beforehand,
it does not advance their \textbf{felicity}\index{felicity}\stepcounter{footnote}\footnotetext{
\textbf{felicity} \emph{n.}  (fuh-LIH-sih-tee)
state of positive emotions ranging from contentment to intense joy.} in the least.  They always
continue to grow sufficiently unlike afterwards to have their
share of \textbf{vexation}\index{vexation}\stepcounter{footnote}\footnotetext{
\textbf{vexation} \emph{n.}  (vex-SAY-shun)
something or someone that causes anxiety; a source of unhappiness.}; and it is better to know as little as possible
of the defects of the person with whom you are to pass your life."

"You make me laugh, Charlotte; but it is not sound.  You know
it is not sound, and that you would never act in this way
yourself."

Occupied in observing Mr. Bingley's attentions to her sister,
Elizabeth was far from suspecting that she was herself becoming
an object of some interest in the eyes of his friend.  Mr. Darcy
had at first scarcely allowed her to be pretty; he had looked at
her without admiration at the ball; and when they next met, he
looked at her only to criticise.  But no sooner had he made it
clear to himself and his friends that she hardly had a good feature
in her face, than he began to find it was \textbf{rendered}\index{rendered}\stepcounter{footnote}\footnotetext{
\textbf{render} \emph{tr.v.} \textbf{-ed}  (REHN-der)
} uncommonly
intelligent by the beautiful expression of her dark eyes.  To this
discovery succeeded some others equally \textbf{mortifying}\index{mortifying}\stepcounter{footnote}\footnotetext{
\textbf{mortify} \emph{v.} \textbf{-ing}  (MAWR-tih-fy)
 to make someone feel great shame or humiliation.}.  Though he
had detected with a critical eye more than one failure of perfect
symmetry in her form, he was forced to acknowledge her figure
to be light and pleasing; and in spite of his asserting that her
manners were not those of the fashionable world, he was caught
by their easy playfulness.  Of this she was perfectly unaware;
to her he was only the man who made himself agreeable nowhere,
and who had not thought her handsome enough to dance with.

He began to wish to know more of her, and as a step towards
conversing with her himself, attended to her conversation with
others.  His doing so drew her notice.  It was at Sir William
Lucas's, where a large party were assembled.

"What does Mr. Darcy mean," said she to Charlotte, "by
listening to my conversation with Colonel Forster?"

"That is a question which Mr. Darcy only can answer."

"But if he does it any more I shall certainly let\index{let} him know that I
see what he is about.  He has a very \textbf{satirical}\index{satirical}\stepcounter{footnote}\footnotetext{
\textbf{satirical} \emph{adj.}  (sa-TEE-rih-kuhl)
exposing human folly to ridicule; relating to satire.} eye, and if I do not
begin by being impertinent\index{impertinent} myself, I shall soon grow afraid of
him."

On his approaching them soon afterwards, though without
seeming to have any intention of speaking, Miss Lucas defied
her friend to mention such a subject\index{subject} to him; which immediately
\textbf{provoking}\index{provoking}\stepcounter{footnote}\footnotetext{
\textbf{provoke} \emph{tr.v.} \textbf{-ing}  (pruh-VOHK)
to intentionally provoke anger or ill-will.} Elizabeth to do it, she turned to him and said:

"Did you not think, Mr. Darcy, that I expressed myself
uncommonly well just now, when I was teasing Colonel Forster
to give us a ball at Meryton?"

"With great energy; but it is always a subject\index{subject} which makes a lady
energetic."

"You are \textbf{severe}\index{severe}\stepcounter{footnote}\footnotetext{
\textbf{severe} \emph{adj.}  (suh-VEER)
 very bad in degree or extent.} on us."

"It will be \emph{her} turn soon to be teased," said Miss Lucas.  "I
am going to open the instrument, Eliza, and you know what
follows."

"You are a very strange creature by way of a friend!--always
wanting me to play and sing before anybody and everybody!
If my vanity\index{vanity} had taken a musical turn, you would have been
\textbf{invaluable}\index{invaluable}\stepcounter{footnote}\footnotetext{
\textbf{invaluable} \emph{adj.}  (ihn-VAL-yoo-uh-bl)
priceless; having incalculable monetary worth.}; but as it is, I would really rather not sit down
before those who must be in the habit of hearing the very best
performers."  On Miss Lucas's persevering, however, she added,
"Very well, if it must be so, it must."  And gravely\index{gravely} glancing at
Mr. Darcy, "There is a fine old saying, which everybody here is of
course familiar with: 'Keep your breath to cool your porridge';
and I shall keep mine to swell my song."

Her performance was pleasing, though by no means \textbf{capital}\index{capital}\stepcounter{footnote}\footnotetext{
\textbf{capital} \emph{n.}  (KA-pih-tuhl)
assets available for use in the production of further assets; money currently in position (not credit).}.
After a song or two, and before she could reply to the \textbf{entreaties}\index{entreaties}\stepcounter{footnote}\footnotetext{
\textbf{entreaty} \emph{n.}  (ehn-TREE-tee)
a sincere plea or request.}
of several that she would sing again, she was eagerly succeeded
at the instrument by her sister Mary, who having, in consequence\index{consequence}
of being the only plain one in the family, worked hard for
knowledge and accomplishments, was always impatient for
display.

Mary had neither genius nor taste; and though vanity\index{vanity} had given
her \textbf{application}\index{application}\stepcounter{footnote}\footnotetext{
\textbf{application} \emph{n.}  (ap-lih-KAY-shun)
a formal request for assistance, employment, or admission.}, it had given her likewise a \textbf{pedantic}\index{pedantic}\stepcounter{footnote}\footnotetext{
\textbf{pedantic} \emph{adj.}  (puh-DAN-tihk)
marked by a narrow, often trivial display of book learning.} air and
conceited\index{conceited} manner, which would have injured a higher degree of
excellence than she had reached.  Elizabeth, easy and unaffected\index{unaffected},
had been listened to with much more pleasure, though not
playing half so well; and Mary, at the end of a long \textbf{concerto}\index{concerto}\stepcounter{footnote}\footnotetext{
\textbf{concerto} \emph{n.}  (kawn-CHER-to)
a musical composition for orchestra and a soloist.},
was glad to purchase praise and gratitude by Scotch and Irish
airs, at the request of her younger sisters, who, with some of the
Lucases, and two or three officers, joined eagerly in dancing at
one end of the room.

Mr. Darcy stood near them in silent \textbf{indignation}\index{indignation}\stepcounter{footnote}\footnotetext{
\textbf{indignation} \emph{n.}  (ihn-dig-NAY-shun)
a feeling of righteous anger.} at such a \textbf{mode}\index{mode}\stepcounter{footnote}\footnotetext{
\textbf{mode} \emph{n.}  (MOAD)
}
of passing the evening, to the exclusion of all conversation, and
was too much \textbf{engrossed}\index{engrossed}\stepcounter{footnote}\footnotetext{
\textbf{engross} \emph{tr.v.} \textbf{-ed}  (ehn-GROHSS)
to enclose or envelop completely, as if by swallowing; to fully occupy.} by his thoughts to perceive\index{perceive} that Sir
William Lucas was his neighbour, till Sir William thus began:

"What a charming amusement for young people this is, Mr. Darcy!
There is nothing like dancing after all.  I consider it as one
of the first refinements of polished society."

"Certainly, sir; and it has the advantage also of being in \textbf{vogue}\index{vogue}\stepcounter{footnote}\footnotetext{
\textbf{vogue} \emph{n.}  (VOAG)
 the popular taste, style, or fashion at a given time.}
amongst the less polished societies of the world.  Every savage
can dance."

Sir William only smiled.  "Your friend performs delightfully," he
continued after a pause, on seeing Bingley join the group; "and I
doubt not that you are an \textbf{adept}\index{adept}\stepcounter{footnote}\footnotetext{
\textbf{adept} \emph{adj.}  (ad-DEPT)
 highly skilled; someone highly skilled.} in the science yourself, Mr.
Darcy."

"You saw me dance at Meryton, I believe, sir."

"Yes, indeed, and received no inconsiderable pleasure from the
sight.  Do you often dance at St. James's?"

"Never, sir."

"Do you not think it would be a proper compliment to the
place?"

"It is a compliment which I never pay to any place if I can
avoid it."

"You have a house in town, I conclude?"

Mr. Darcy bowed.

"I had once had some thought of fixing in town myself--for I am
fond of superior\index{superior} society; but I did not feel quite certain that the
air of London would agree with Lady Lucas."

He paused in hopes of an answer; but his companion was not
disposed\index{disposed} to make any; and Elizabeth at that instant moving
towards them, he was struck with the action of doing a very
gallant thing, and called out to her:

"My dear Miss Eliza, why are you not dancing?  Mr. Darcy, you
must allow me to present this young lady to you as a very
desirable partner.  You cannot refuse to dance, I am sure when
so much beauty is before you."  And, taking her hand, he would
have given it to Mr. Darcy who, though extremely surprised,
was not unwilling to receive it, when she instantly drew back,
and said with some discomposure to Sir William:

"Indeed, sir, I have not the least intention of dancing.  I entreat\index{entreat}
you not to suppose that I moved this way in order to beg for a
partner."

Mr. Darcy, with \textbf{grave}\index{grave}\stepcounter{footnote}\footnotetext{
\textbf{grave} \emph{adj.} \textbf{-ly}  (GRAYV)
 of great gravity or crucial import; requiring serious thought.} \textbf{propriety}\index{propriety}\stepcounter{footnote}\footnotetext{
\textbf{propriety} \emph{n.}  (pruh-PRY-eh-tee)
correct or appropriate behavior.}, requested to be allowed the
honour of her hand, but in vain.  Elizabeth was determined; nor
did Sir William at all shake her purpose by his attempt at
persuasion.

"You excel so much in the dance, Miss Eliza, that it is cruel to
deny me the happiness of seeing you; and though this gentleman
dislikes the amusement in general, he can have no objection, I
am sure, to oblige us for one half-hour."

"Mr. Darcy is all politeness," said Elizabeth, smiling.

"He is, indeed; but, considering the \textbf{inducement}\index{inducement}\stepcounter{footnote}\footnotetext{
\textbf{inducement} \emph{n.}  (ihn-DOOS-muhnt)
a positive motivational influence.}, my dear Miss
Eliza, we cannot wonder at his \textbf{complaisance}\index{complaisance}\stepcounter{footnote}\footnotetext{
\textbf{complaisance} \emph{n.}  (kuhm-PLAY-sense)
a willingness or tendency to yield to the will of others; amiability.}--for who would
object to such a partner?"

Elizabeth looked archly\index{archly}, and turned away.  Her resistance had
not injured her with the gentleman, and he was thinking of her
with some complacency\index{complacency}, when thus \textbf{accosted}\index{accosted}\stepcounter{footnote}\footnotetext{
\textbf{accost} \emph{tr.v.} \textbf{-ed}  (uh-KAWST)
to approach and verbally attack, affront, or make demand of.} by Miss Bingley:

"I can guess the subject\index{subject} of your \textbf{reverie}\index{reverie}\stepcounter{footnote}\footnotetext{
\textbf{reverie} \emph{n.}  (REHV-er-ee)
absent-minded or fanciful musing while awake; daydreaming.}."

"I should imagine not."

"You are considering how insupportable\index{insupportable} it would be to pass many
evenings in this manner--in such society; and indeed I am quite
of your opinion.  I was never more annoyed!  The \textbf{insipidity}\index{insipidity}\stepcounter{footnote}\footnotetext{
\textbf{insipid} \emph{adj.} \textbf{-ity}  (ihn-SIH-pihd)
 Lacking flavor or taste; trite, dull, unoriginal, or hackneyed.}, and
yet the noise--the nothingness, and yet the self-importance of all
those people!  What would I give to hear your \textbf{strictures}\index{strictures}\stepcounter{footnote}\footnotetext{
\textbf{stricture} \emph{n.}  (STRIHK-cher)
A restriction, limit, or rule.} on them!"

"You conjecture\index{conjecture} is totally wrong, I assure you.  My mind was
more agreeably engaged.  I have been meditating on the very
great pleasure which a pair of fine eyes in the face of a pretty
woman can bestow\index{bestow}."

Miss Bingley immediately fixed her eyes on his face, and desired
he would tell her what lady had the credit of inspiring such
reflections.  Mr. Darcy replied with great \textbf{intrepidity}\index{intrepidity}\stepcounter{footnote}\footnotetext{
\textbf{intrepid} \emph{adj.} \textbf{-ity}  (ihn-TREH-pihd)
resolute courageousness.}:

"Miss Elizabeth Bennet."

"Miss Elizabeth Bennet!" repeated Miss Bingley.  "I am all
astonishment.  How long has she been such a favourite?--and
pray, when am I to wish you joy?"

"That is exactly the question which I expected you to ask.  A
lady's imagination is very rapid\index{rapid}; it jumps from admiration to
love, from love to \textbf{matrimony}\index{matrimony}\stepcounter{footnote}\footnotetext{
\textbf{matrimony} \emph{n.}  (MAT-rih-moh-nee)
The state of being married.}, in a moment.  I knew you would
be wishing me joy."

"Nay, if you are serious about it, I shall consider the matter is
absolutely settled.  You will be having a charming mother-in-law,
indeed; and, of course, she will always be at Pemberley with you."

He listened to her with perfect \textbf{indifference}\index{indifference}\stepcounter{footnote}\footnotetext{
\textbf{indifference} \emph{n.}  (ihn-DIF-fer-ense)
the trait of lacking enthusiasm or interest; apathy.} while she chose to
entertain herself in this manner; and as his composure\index{composure} convinced
her that all was safe, her wit flowed long.



\chapter{7}


Mr. Bennet's property consisted almost entirely in an estate of
two thousand a year, which, unfortunately for his daughters, was
\textbf{entailed}\index{entailed}\stepcounter{footnote}\footnotetext{
\textbf{entail} \emph{tr.v.} \textbf{-ed}  (ehn-TAYL)
 To have as a necessary clause, condition, or consequence; to include.}, in \textbf{default}\index{default}\stepcounter{footnote}\footnotetext{
\textbf{default} \emph{n.}  (dih-FAWLT)
failing to meet a financial obligation.} of \textbf{heirs}\index{heirs}\stepcounter{footnote}\footnotetext{
\textbf{heir} \emph{n.}  (AIR)
 a person who inherits.} male, on a distant relation; and their
mother's fortune, though \textbf{ample}\index{ample}\stepcounter{footnote}\footnotetext{
\textbf{ample} \emph{adj.} \textbf{-ly}  (AM-puhl)
 more than enough in size, scope, or capacity.} for her situation in life, could
but ill supply the deficiency of his.  Her father had been an
attorney in Meryton, and had left her four thousand pounds.

She had a sister married to a Mr. Phillips, who had been a clerk
to their father and succeeded him in the business, and a brother
settled in London in a respectable line of trade.

The village of Longbourn was only one mile from Meryton; a
most convenient distance for the young ladies, who were usually
tempted thither three or four times a week, to pay their duty to
their aunt and to a \textbf{milliner}\index{milliner}\stepcounter{footnote}\footnotetext{
\textbf{milliner} \emph{n.}  (MIHL-lih-ner)
someone who makes and sells hats.}'s shop just over the way.  The two
youngest of the family, Catherine and Lydia, were particularly
frequent in these attentions; their minds were more vacant than
their sisters', and when nothing better offered, a walk to
Meryton was necessary to amuse their morning hours and
furnish conversation for the evening; and however bare of news
the country in general might be, they always \textbf{contrived}\index{contrived}\stepcounter{footnote}\footnotetext{
\textbf{} \emph{}  ()
} to learn
some from their aunt.  At present, indeed, they were well
supplied both with news and happiness by the recent arrival of
a \textbf{militia}\index{militia}\stepcounter{footnote}\footnotetext{
\textbf{militia} \emph{n.}  (muh-LIH-shuh)
 an army composed of civilians, usually for defense only.} \textbf{regiment}\index{regiment}\stepcounter{footnote}\footnotetext{
\textbf{regiment} \emph{n.}  (REH-jih-muhnt)
army unit smaller than a division.} in the neighbourhood; it was to remain the
whole winter, and Meryton was the headquarters.

Their visits to Mrs. Phillips were now productive of the most
interesting intelligence.  Every day added something to their
knowledge of the officers' names and connections.  Their
lodgings were not long a secret, and at length they began to
know the officers themselves.  Mr. Phillips visited them all, and
this opened to his nieces a store of felicity\index{felicity} unknown before.
They could talk of nothing but officers; and Mr. Bingley's large
fortune, the mention of which gave animation to their mother,
was worthless in their eyes when opposed to the regimentals of
an \textbf{ensign}\index{ensign}\stepcounter{footnote}\footnotetext{
\textbf{ensign} \emph{n.}  (EHN-sihn)
an emblem displaying as symbol of nationality, often with insignia of a specific military unit.}.

After listening one morning to their \textbf{effusions}\index{effusions}\stepcounter{footnote}\footnotetext{
\textbf{effusion} \emph{n.}  (ehf-FYOO-zhun)
an unrestrained expression of emotion; an outburst.} on this subject\index{subject}, Mr.
Bennet coolly observed:

"From all that I can collect by your manner of talking, you must
be two of the silliest girls in the country.  I have suspected it
some time, but I am now convinced."

Catherine was disconcerted\index{disconcerted}, and made no answer; but Lydia,
with perfect indifference\index{indifference}, continued to express her admiration of
Captain Carter, and her hope of seeing him in the course of the
day, as he was going the next morning to London.

"I am astonished, my dear," said Mrs. Bennet, "that you should
be so ready to think your own children silly.  If I wished to think
\textbf{slightingly}\index{slightingly}\stepcounter{footnote}\footnotetext{
\textbf{slighting} \emph{adj.} \textbf{-ly}  (SLY-ting)
in a disparaging manner; belittling.} of anybody's children, it should not be of my own,
however."

"If my children are silly, I must hope to be always sensible of it."

"Yes--but as it happens, they are all of them very clever."

"This is the only point, I flatter myself, on which we do not
agree.  I had hoped that our \textbf{sentiments}\index{sentiments}\stepcounter{footnote}\footnotetext{
\textbf{sentiment} \emph{n.}  (SEHN-tih-mehnt)
A thought or attitude based mostly on emotion.} coincided in every
particular, but I must so far differ from you as to think our two
youngest daughters uncommonly foolish."

"My dear Mr. Bennet, you must not expect such girls to have
the sense of their father and mother.  When they get to our age, I
dare say they will not think about officers any more than we do.
I remember the time when I liked a red coat myself very well--and,
indeed, so I do still at my heart; and if a smart young colonel,
with five or six thousand a year, should want one of my girls I
shall not say nay to him; and I thought Colonel Forster looked
very becoming the other night at Sir William's in his regimentals."

"Mamma," cried Lydia, "my aunt says that Colonel Forster and
Captain Carter do not go so often to Miss Watson's as they did
when they first came; she sees them now very often standing in
Clarke's library."

Mrs. Bennet was prevented replying by the entrance of the
footman with a note for Miss Bennet; it came from Netherfield,
and the servant waited for an answer.  Mrs. Bennet's eyes
sparkled with pleasure, and she was eagerly calling out, while
her daughter read,

"Well, Jane, who is it from?  What is it about?  What does he
say?  Well, Jane, make haste and tell us; make haste, my love."

"It is from Miss Bingley," said Jane, and then read it aloud.

"\emph{my} \emph{dear} \emph{friend},--

"If you are not so compassionate as to dine to-day with Louisa
and me, we shall be in danger of hating each other for the rest
of our lives, for a whole day's \textbf{tete-a-tete}\index{tete-a-tete}\stepcounter{footnote}\footnotetext{
\textbf{tte--tte} \emph{adv.}  ()
a private conversation between two people.} between two women
can never end without a quarrel.  Come as soon as you can on
receipt of this.  My brother and the gentlemen are to dine with
the officers.--Yours ever,

"\emph{caroline} \emph{bingley}"

"With the officers!" cried Lydia.  "I wonder my aunt did not tell
us of \emph{that}."

"Dining out," said Mrs. Bennet, "that is very unlucky."

"Can I have the carriage?" said Jane.

"No, my dear, you had better go on horseback, because it seems
likely to rain; and then you must stay all night."

"That would be a good scheme," said Elizabeth, "if you were
sure that they would not offer to send her home."

"Oh! but the gentlemen will have Mr. Bingley's chaise\index{chaise} to go to
Meryton, and the Hursts have no horses to theirs."

"I had much rather go in the coach."

"But, my dear, your father cannot spare the horses, I am sure.
They are wanted in the farm, Mr. Bennet, are they not?"

"They are wanted in the farm much oftener than I can get them."

"But if you have got them to-day," said Elizabeth, "my mother's
purpose will be answered."

She did at last \textbf{extort}\index{extort}\stepcounter{footnote}\footnotetext{
\textbf{extort} \emph{tr.v.} \textbf{-ed}  (ehx-TORT)
obtain by coercion or intimidation.} from her father an acknowledgment that
the horses were engaged.  Jane was therefore obliged to go on
horseback, and her mother attended her to the door with many
cheerful \textbf{prognostics}\index{prognostics}\stepcounter{footnote}\footnotetext{
\textbf{prognostic} \emph{adj.}  (prahg-NAH-stik)
a prediction, forcast, or indication as to the future.} of a bad day.  Her hopes were answered;
Jane had not been gone long before it rained hard.  Her sisters
were uneasy for her, but her mother was delighted.  The rain
continued the whole evening without \textbf{intermission}\index{intermission}\stepcounter{footnote}\footnotetext{
\textbf{intermission} \emph{n.}  (ihn-ter-MIH-shun)
a time of temporarily suspended activity.}; Jane certainly
could not come back.

"This was a lucky idea of mine, indeed!" said Mrs. Bennet more
than once, as if the credit of making it rain were all her own.  Till
the next morning, however, she was not aware of all the felicity\index{felicity}
of her \textbf{contrivance}\index{contrivance}\stepcounter{footnote}\footnotetext{
\textbf{contrivance} \emph{n.}  (kuhn-TRY-vuhns)
a device that is useful for a particular job; any improvised arrangement for temporary use.}.  Breakfast was scarcely over when a servant
from Netherfield brought the following note for Elizabeth:

"\emph{my} \emph{dearest} \emph{lizzy},--

"I find myself very unwell this morning, which, I suppose, is to
be \textbf{imputed}\index{imputed}\stepcounter{footnote}\footnotetext{
\textbf{impute} \emph{tr.v.} \textbf{-ing}  (im-PYOOT)
 to assign fault or responsibility to.} to my getting wet through yesterday.  My kind friends
will not hear of my returning till I am better.  They insist also
on my seeing Mr. Jones--therefore do not be alarmed if you should
hear of his having been to me--and, excepting a sore throat and
headache, there is not much the matter with me.--Yours, etc."

"Well, my dear," said Mr. Bennet, when Elizabeth had read the
note aloud, "if your daughter should have a dangerous fit of
illness--if she should die, it would be a comfort to know that it
was all in pursuit of Mr. Bingley, and under your orders."

"Oh! I am not afraid of her dying.  People do not die of little
\textbf{trifling}\index{trifling}\stepcounter{footnote}\footnotetext{
\textbf{trifling} \emph{n.}  (TRIHL-ling)
not worth considering because of little value or importance.} colds.  She will be taken good care of.  As long as she
stays there, it is all very well.  I would go an see her if I could
have the carriage."

Elizabeth, feeling really anxious, was determined to go to her,
though the carriage was not to be had; and as she was no
horsewoman, walking was her only alternative.  She declared her
\textbf{resolution}\index{resolution}\stepcounter{footnote}\footnotetext{
\textbf{resolution} \emph{n.}  (reh-zuh-LOO-shun)
}.

"How can you be so silly," cried her mother, "as to think of such
a thing, in all this dirt!  You will not be fit to be seen when you
get there."

"I shall be very fit to see Jane--which is all I want."

"Is this a hint to me, Lizzy," said her father, "to send for
the horses?"

"No, indeed, I do not wish to avoid the walk.  The distance is
nothing when one has a motive; only three miles.  I shall be back
by dinner."

"I admire the activity of your \textbf{benevolence}\index{benevolence}\stepcounter{footnote}\footnotetext{
\textbf{benevolence} \emph{n.}  (buh-NEH-vuh-lenss)
an inclination toward kind or charitable acts.}," observed Mary, "but
every impulse of feeling should be guided by reason; and, in my
opinion, \textbf{exertion}\index{exertion}\stepcounter{footnote}\footnotetext{
\textbf{exertion} \emph{n.}  (ehg-ZER-shun)
strenuous use of physical or mental energy; hard work.} should always be in proportion to what is
required."

"We will go as far as Meryton with you," said Catherine and
Lydia.  Elizabeth accepted their company, and the three young
ladies set off together.

"If we make haste," said Lydia, as they walked along, "perhaps
we may see something of Captain Carter before he goes."

In Meryton they parted; the two youngest repaired to the lodgings of
one of the officers' wives, and Elizabeth continued her walk alone,
crossing field after field at a quick pace, jumping over \textbf{stiles}\index{stiles}\stepcounter{footnote}\footnotetext{
\textbf{stile} \emph{n.}  (STY-uhl)
 A vertical part of a panel or frame, as in a door or window sash.}
and springing over puddles with impatient activity, and finding
herself at last within view of the house, with weary ankles, dirty
stockings, and a face glowing with the warmth of exercise.

She was shown into the breakfast-parlour, where all but Jane
were assembled, and where her appearance created a great deal
of surprise.  That she should have walked three miles so early
in the day, in such dirty weather, and by herself, was almost
incredible to Mrs. Hurst and Miss Bingley; and Elizabeth was
convinced that they held her in \textbf{contempt}\index{contempt}\stepcounter{footnote}\footnotetext{
\textbf{contempt} \emph{n.}  (kuhn-TEHMPT)
open disrespect for a person or thing.} for it.  She was
received, however, very politely by them; and in their brother's
manners there was something better than politeness; there was
good humour and kindness.  Mr. Darcy said very little, and Mr.
Hurst nothing at all.  The former was divided between admiration
of the brilliancy which exercise had given to her complexion,
and doubt as to the occasion's justifying her coming so far
alone.  The latter was thinking only of his breakfast.

Her inquiries after her sister were not very favourably answered.
Miss Bennet had slept ill, and though up, was very feverish, and
not well enough to leave her room.  Elizabeth was glad to be
taken to her immediately; and Jane, who had only been withheld
by the fear of giving alarm or inconvenience from expressing in
her note how much she longed for such a visit, was delighted at
her entrance.  She was not equal, however, to much conversation,
and when Miss Bingley left them together, could attempt little
besides expressions of gratitude for the extraordinary kindness
she was treated with.  Elizabeth silently attended her.

When breakfast was over they were joined by the sisters; and
Elizabeth began to like them herself, when she saw how much
affection and \textbf{solicitude}\index{solicitude}\stepcounter{footnote}\footnotetext{
\textbf{solicitude} \emph{n.}  (saw-LIH-sih-tood)
care or concern; excessive concern.} they showed for Jane.  The \textbf{apothecary}\index{apothecary}\stepcounter{footnote}\footnotetext{
\textbf{apothecary} \emph{n.}  (uh-PAH-theh-KA-ree)
a health professional who prepares and dispenses drugs; a pharmacist.}
came, and having examined his patient, said, as might be
supposed, that she had caught a violent cold, and that they must
endeavour\index{endeavour} to get the better of it; advised her to return to bed,
and promised her some \textbf{draughts}\index{draughts}\stepcounter{footnote}\footnotetext{
\textbf{draught} \emph{n.}  (draft)
 same as draft.}.  The advice was followed
\textbf{readily}\index{readily}\stepcounter{footnote}\footnotetext{
\textbf{readily} \emph{adv.}  (REH-dih-lee)
 without difficulty; immediately.}, for the feverish symptoms increased, and her head ached
\textbf{acutely}\index{acutely}\stepcounter{footnote}\footnotetext{
\textbf{acute} \emph{adj.} \textbf{-ly}  (uh-KYOOT)
 having a rapid onset}.  Elizabeth did not quit her room for a moment; nor were
the other ladies often absent; the gentlemen being out, they had,
in fact, nothing to do elsewhere.

When the clock struck three, Elizabeth felt that she must go, and
very unwillingly said so.  Miss Bingley offered her the carriage,
and she only wanted a little pressing to accept it, when Jane
testified such concern in parting with her, that Miss Bingley was
obliged to convert the offer of the chaise\index{chaise} to an invitation to
remain at Netherfield for the present.  Elizabeth most thankfully
consented, and a servant was dispatched to Longbourn to
acquaint the family with her stay and bring back a supply of
clothes.



\chapter{8}


At five o'clock the two ladies retired to dress, and at half-past
six Elizabeth was \textbf{summoned}\index{summoned}\stepcounter{footnote}\footnotetext{
\textbf{summon} \emph{tr.v.} \textbf{-ed}  (SUHM-muhn)
} to dinner.  To the civil inquiries
which then poured in, and amongst which she had the pleasure
of distinguishing the much superior\index{superior} solicitude\index{solicitude} of Mr. Bingley's,
she could not make a very favourable answer.  Jane was by no
means better.  The sisters, on hearing this, repeated three or four
times how much they were grieved\index{grieved}, how shocking it was to have
a bad cold, and how excessively they disliked being ill
themselves; and then thought no more of the matter: and their
indifference\index{indifference} towards Jane when not immediately before them
restored Elizabeth to the enjoyment of all her former dislike.

Their brother, indeed, was the only one of the party whom she
could regard with any complacency\index{complacency}.  His anxiety for Jane was
evident, and his attentions to herself most pleasing, and
they prevented her feeling herself so much an intruder as she
believed she was considered by the others.  She had very little
notice from any but him.  Miss Bingley was engrossed\index{engrossed} by Mr.
Darcy, her sister scarcely less so; and as for Mr. Hurst, by
whom Elizabeth sat, he was an \textbf{indolent}\index{indolent}\stepcounter{footnote}\footnotetext{
\textbf{indolent} \emph{adj.}  (IHN-doh-lehnt)
lazy as a lifestyle.} man, who lived only to
eat, drink, and play at cards; who, when he found her to prefer
a plain dish to a \textbf{ragout}\index{ragout}\stepcounter{footnote}\footnotetext{
\textbf{ragout} \emph{n.}  (ra-GOOH)
well-seasoned meat or fish stew with vegetables.}, had nothing to say to her.

When dinner was over, she returned directly to Jane, and Miss
Bingley began abusing her as soon as she was out of the room.
Her manners were pronounced to be very bad indeed, a mixture
of pride and \textbf{impertinence}\index{impertinence}\stepcounter{footnote}\footnotetext{
\textbf{impertinence} \emph{n.}  (im-PER-tih-nense)
the trait of being rude and insolent; inappropriate playfullness.}; she had no conversation, no style, no
beauty.  Mrs. Hurst thought the same, and added:

"She has nothing, in short, to recommend her, but being an
excellent walker.  I shall never forget her appearance this
morning.  She really looked almost wild."

"She did, indeed, Louisa.  I could hardly keep my countenance\index{countenance}.
Very \textbf{nonsensical}\index{nonsensical}\stepcounter{footnote}\footnotetext{
\textbf{nonsensical} \emph{adj.}  (nahn-SEHN-sih-kuhl)
so lacking in intelligence or good sense as to be laughable.} to come at all!  Why must \emph{she} be scampering
about the country, because her sister had a cold?  Her hair, so
untidy, so blowsy!"

"Yes, and her petticoat; I hope you saw her petticoat, six inches
deep in mud, I am absolutely certain; and the gown which had
been let\index{let} down to hide it not doing its office."

"Your picture may be very exact, Louisa," said Bingley; "but
this was all lost upon me.  I thought Miss Elizabeth Bennet
looked remarkably well when she came into the room this
morning.  Her dirty petticoat quite escaped my notice."

"\emph{you} observed it, Mr. Darcy, I am sure," said Miss Bingley;
"and I am inclined to think that you would not wish to see
\emph{your} sister make such an exhibition."

"Certainly not."

"To walk three miles, or four miles, or five miles, or whatever it
is, above her ankles in dirt, and alone, quite alone!  What could
she mean by it?  It seems to me to show an \textbf{abominable}\index{abominable}\stepcounter{footnote}\footnotetext{
\textbf{abominable} \emph{adj.} \textbf{-ly}  (uh-BAH-mih-nuh-bl)
unequivocally detestable; morally repugnant.} sort of
conceited\index{conceited} independence, a most country-town indifference\index{indifference} to
\textbf{decorum}\index{decorum}\stepcounter{footnote}\footnotetext{
\textbf{decorum} \emph{n.}  (deh-KAW-ruhm)
propriety in manners or conduct.}."

"It shows an affection for her sister that is very pleasing," said
Bingley.

"I am afraid, Mr. Darcy," observed Miss Bingley in a half
whisper, "that this adventure has rather affected your
admiration of her fine eyes."

"Not at all," he replied; "they were brightened by the exercise."
A short pause followed this speech, and Mrs. Hurst began again:

"I have a excessive regard for Miss Jane Bennet, she is really
a very sweet girl, and I wish with all my heart she were well
settled.  But with such a father and mother, and such low
connections, I am afraid there is no chance of it."

"I think I have heard you say that their uncle is an attorney on
Meryton."

"Yes; and they have another, who lives somewhere near Cheapside."

"That is capital\index{capital}," added her sister, and they both laughed \textbf{heartily}\index{heartily}\stepcounter{footnote}\footnotetext{
\textbf{heartily} \emph{adv.}  (HAR-tih-lee)
}.

"If they had uncles enough to fill \emph{all} Cheapside," cried
Bingley, "it would not make them one jot less agreeable."

"But it must very materially lessen their chance of marrying men
of any consideration in the world," replied Darcy.

To this speech Bingley made no answer; but his sisters gave it
their hearty \textbf{assent}\index{assent}\stepcounter{footnote}\footnotetext{
\textbf{assent} \emph{intr.v.} \textbf{-ed}  (as-SENT)
To agree with a proposal or statement.}, and indulged their \textbf{mirth}\index{mirth}\stepcounter{footnote}\footnotetext{
\textbf{mirth} \emph{n.}  (MERTH)
 great merriment, laughing, and gaiety.} for some time at the
expense of their dear friend's \textbf{vulgar}\index{vulgar}\stepcounter{footnote}\footnotetext{
\textbf{vulgar} \emph{adj.}  (VUHL-ger)
 tastelessly indecent; lacking refinement or cultivation.} relations.

With a renewal of tenderness, however, they returned to her
room on leaving the dining-parlour, and sat with her till
summoned\index{summoned} to coffee.  She was still very poorly, and Elizabeth
would not quit her at all, till late in the evening, when she had
the comfort of seeing her sleep, and when it seemed to her rather
right than pleasant that she should go downstairs herself.  On
entering the drawing-room she found the whole party at \textbf{loo}\index{loo}\stepcounter{footnote}\footnotetext{
\textbf{loo} \emph{n.}  (LOOH)
a toilet in England.}, and
was immediately invited to join them; but suspecting them to be
playing high she declined it, and making her sister the excuse,
said she would amuse herself for the short time she could stay
below, with a book.  Mr. Hurst looked at her with astonishment.

"Do you prefer reading to cards?" said he; "that is rather
\textbf{singular}\index{singular}\stepcounter{footnote}\footnotetext{
\textbf{singular} \emph{adj.}  (SING-yoo-ler)
the single one of its kind; individual.}."

"Miss Eliza Bennet," said Miss Bingley, "despises cards.  She is
a great reader, and has no pleasure in anything else."

"I deserve neither such praise nor such censure\index{censure}," cried Elizabeth;
"I am \emph{not} a great reader, and I have pleasure in many things."

"In nursing your sister I am sure you have pleasure," said Bingley;
"and I hope it will be soon increased by seeing her quite well."

Elizabeth thanked him from her heart, and then walked towards
the table where a few books were lying.  He immediately offered
to fetch her others--all that his library afforded.

"And I wish my collection were larger for your benefit and my
own credit; but I am an idle fellow, and though I have not many,
I have more than I ever looked into."

Elizabeth assured him that she could suit herself perfectly with
those in the room.

"I am astonished," said Miss Bingley, "that my father should
have left so small a collection of books.  What a delightful library
you have at Pemberley, Mr. Darcy!"

"It ought to be good," he replied, "it has been the work of many
generations."

"And then you have added so much to it yourself, you are
always buying books."

"I cannot comprehend\index{comprehend} the neglect of a family library in such days
as these."

"Neglect!  I am sure you neglect nothing that can add to the
beauties of that noble place.  Charles, when you build \emph{your}
house, I wish it may be half as delightful as Pemberley."

"I wish it may."

"But I would really advise you to make your purchase in that
neighbourhood, and take Pemberley for a kind of model.  There
is not a finer county in England than Derbyshire."

"With all my heart; I will buy Pemberley itself if Darcy will
sell it."

"I am talking of possibilities, Charles."

"Upon my word, Caroline, I should think it more possible to get
Pemberley by purchase than by imitation."

Elizabeth was so much caught with what passed, as to leave her
very little attention for her book; and soon laying it \textbf{wholly}\index{wholly}\stepcounter{footnote}\footnotetext{
\textbf{wholly} \emph{adv.}  (HOAL-lee)
}
aside, she drew near the card-table, and stationed herself
between Mr. Bingley and his eldest sister, to observe the game.

"Is Miss Darcy much grown since the spring?" said Miss
Bingley; "will she be as tall as I am?"

"I think she will.  She is now about Miss Elizabeth Bennet's
height, or rather taller."

"How I long to see her again!  I never met with anybody who
delighted me so much.  Such a countenance\index{countenance}, such manners!  And
so extremely accomplished for her age!  Her performance on the
\textbf{pianoforte}\index{pianoforte}\stepcounter{footnote}\footnotetext{
\textbf{pianoforte} \emph{n.}  (pih-ah-noh-FOR-tay)
a musical instrument somewhat resembling a harpsichord.} is exquisite."

"It is amazing to me," said Bingley, "how young ladies can have
patience to be so very accomplished as they all are."

"All young ladies accomplished!  My dear Charles, what do you mean?"

"Yes, all of them, I think.  They all paint tables, cover screens,
and net purses.  I scarcely know anyone who cannot do all this,
and I am sure I never heard a young lady spoken of for the first
time, without being informed that she was very accomplished."

"Your list of the common extent of accomplishments," said Darcy,
"has too much truth.  The word is applied to many a woman who
deserves it no otherwise than by netting a purse or covering
a screen.  But I am very far from agreeing with you in your
estimation of ladies in general.  I cannot boast of knowing
more than half-a-dozen, in the whole range of my acquaintance,
that are really accomplished."

"Nor I, I am sure," said Miss Bingley.

"Then," observed Elizabeth, "you must comprehend\index{comprehend} a great deal
in your idea of an accomplished woman."

"Yes, I do comprehend\index{comprehend} a great deal in it."

"Oh! certainly," cried his faithful assistant, "no one can be really
esteemed accomplished who does not greatly surpass what is
usually met with.  A woman must have a thorough knowledge of
music, singing, drawing, dancing, and the modern languages, to
deserve the word; and besides all this, she must possess a certain
something in her air and manner of walking, the tone of her
voice, her address\index{address} and expressions, or the word will be but
half-deserved."

"All this she must possess," added Darcy, "and to all this she
must yet add something more substantial, in the improvement of
her mind by extensive reading."

"I am no longer surprised at your knowing \emph{only} six accomplished
women.  I rather wonder now at your knowing \emph{any}."

"Are you so severe\index{severe} upon your own sex as to doubt the possibility
of all this?"

"I never saw such a woman.  I never saw such capacity, and
taste, and application\index{application}, and elegance, as you describe united."

Mrs. Hurst and Miss Bingley both cried out against the \textbf{injustice}\index{injustice}\stepcounter{footnote}\footnotetext{
\textbf{injustice} \emph{n.}  (ihn-JUHS-tis)
a Violation of a person's rights; an unjust act.}
of her implied doubt, and were both protesting that they knew
many women who answered this description, when Mr. Hurst
called them to order, with bitter complaints of their inattention
to what was going \textbf{forward}\index{forward}\stepcounter{footnote}\footnotetext{
\textbf{forward} \emph{adj.}  (FOR-werd)
 lacking restraint or modesty.}.  As all conversation was thereby at
an end, Elizabeth soon afterwards left the room.

"Elizabeth Bennet," said Miss Bingley, when the door was
closed on her, "is one of those young ladies who seek to
recommend themselves to the other sex by undervaluing their
own; and with many men, I dare say, it succeeds.  But, in my
opinion, it is a \textbf{paltry}\index{paltry}\stepcounter{footnote}\footnotetext{
\textbf{paltry} \emph{adj.}  (PAWL-tree)
 not worth considering; contemptibly small in amount.} device, a very mean art."

"Undoubtedly," replied Darcy, to whom this remark was chiefly
addressed\index{addressed}, "there is a meanness in \emph{all} the arts which ladies
sometimes \textbf{condescend}\index{condescend}\stepcounter{footnote}\footnotetext{
\textbf{condescend} \emph{intr.v.} \textbf{-ed}  (kahn-duh-SEND)
do something considered beneath one's dignity; to lower oneself.} to employ for captivation.  Whatever
bears \textbf{affinity}\index{affinity}\stepcounter{footnote}\footnotetext{
\textbf{affinity} \emph{n.}  (af-FIH-nih-tee)
a natural attraction or feeling of kinship.} to \textbf{cunning}\index{cunning}\stepcounter{footnote}\footnotetext{
\textbf{cunning} \emph{adj.}  (KUHN-ning)
shrewdness as shown by being skilled in deception.} is despicable."

Miss Bingley was not so entirely satisfied with this reply as to
continue the subject\index{subject}.

Elizabeth joined them again only to say that her sister was worse,
and that she could not leave her.  Bingley urged Mr. Jones being
sent for immediately; while his sisters, convinced that no country
advice could be of any service, recommended an express to town for
one of the most \textbf{eminent}\index{eminent}\stepcounter{footnote}\footnotetext{
\textbf{eminent} \emph{adj.}  (EHM-mih-nehnt)
standing above others in quality or position.} physicians.  This she would not hear of;
but she was not so unwilling to comply\index{comply} with their brother's
proposal; and it was settled that Mr. Jones should be sent for
early in the morning, if Miss Bennet were not decidedly better.
Bingley was quite uncomfortable; his sisters declared that they
were miserable.  They solaced\index{solaced} their \textbf{wretchedness}\index{wretchedness}\stepcounter{footnote}\footnotetext{
\textbf{wretched} \emph{adj.} \textbf{-ness}  (RETCH-ehd)
a state affliction or misfortune; miserable.}, however, by
\textbf{duets}\index{duets}\stepcounter{footnote}\footnotetext{
\textbf{} \emph{}  ()
} after supper, while he could find no better relief to his
feelings than by giving his housekeeper directions that every
attention might be paid to the sick lady and her sister.



\chapter{9}


Elizabeth passed the chief of the night in her sister's room, and
in the morning had the pleasure of being able to send a tolerable\index{tolerable}
answer to the inquiries which she very early received from Mr.
Bingley by a housemaid, and some time afterwards from the two
elegant ladies who waited on his sisters.  In spite of this
\textbf{amendment}\index{amendment}\stepcounter{footnote}\footnotetext{
\textbf{amendment} \emph{n.}  (uh-MEHND-muhnt)
a statement that revises, adds to, or improves a proposal or document.}, however, she requested to have a note sent to Longbourn,
desiring her mother to visit Jane, and form her own judgement of
her situation.  The note was immediately dispatched, and its
contents as quickly complied with.  Mrs. Bennet, accompanied by
her two youngest girls, reached Netherfield soon after the family
breakfast.

Had she found Jane in any apparent danger, Mrs. Bennet would
have been very miserable; but being satisfied on seeing her that
her illness was not alarming, she had no wish of her recovering
immediately, as her restoration to health would probably remove
her from Netherfield.  She would not listen, therefore, to her
daughter's proposal of being carried home; neither did the
apothecary\index{apothecary}, who arrived about the same time, think it at all
advisable.  After sitting a little while with Jane, on Miss
Bingley's appearance and invitation, the mother and three
daughter all attended her into the breakfast parlour.  Bingley met
them with hopes that Mrs. Bennet had not found Miss Bennet
worse than she expected.

"Indeed I have, sir," was her answer.  "She is a great deal too
ill to be moved.  Mr. Jones says we must not think of moving her.
We must trespass a little longer on your kindness."

"Removed!" cried Bingley.  "It must not be thought of.  My
sister, I am sure, will not hear of her removal."

"You may depend upon it, Madam," said Miss Bingley, with cold
\textbf{civility}\index{civility}\stepcounter{footnote}\footnotetext{
\textbf{civility} \emph{n.}  (sih-VIL-ih-tee)
formal or perfunctory politeness.}, "that Miss Bennet will receive every possible attention
while she remains with us."

Mrs. Bennet was \textbf{profuse}\index{profuse}\stepcounter{footnote}\footnotetext{
\textbf{profuse} \emph{adj.}  (proh-FYOOS)
produced or growing in extreme abundance; plentiful.} in her acknowledgments.

"I am sure," she added, "if it was not for such good friends I do
not know what would become of her, for she is very ill indeed,
and suffers a vast\index{vast} deal, though with the greatest patience in the
world, which is always the way with her, for she has, without
exception, the sweetest temper I have ever met with.  I often tell
my other girls they are nothing to \emph{her}.  You have a sweet room
here, Mr. Bingley, and a charming \textbf{prospect}\index{prospect}\stepcounter{footnote}\footnotetext{
\textbf{prospect} \emph{n.}  (PRAHSS-pehkt)
the possibility of future success.} over the gravel walk.
I do not know a place in the country that is equal to Netherfield.
You will not think of quitting it in a hurry, I hope, though you
have but a short lease."

"Whatever I do is done in a hurry," replied he; "and therefore if I
should resolve to quit Netherfield, I should probably be off in
five minutes\index{minutes}.  At present, however, I consider myself as quite
fixed here."

"That is exactly what I should have supposed of you," said
Elizabeth.

"You begin to comprehend\index{comprehend} me, do you?" cried he, turning
towards her.

"Oh! yes--I understand you perfectly."

"I wish I might take this for a compliment; but to be so easily
seen through I am afraid is pitiful."

"That is as it happens.  It does not follow that a deep, intricate
character is more or less \textbf{estimable}\index{estimable}\stepcounter{footnote}\footnotetext{
\textbf{estimable} \emph{adj.}  (EHS-tih-muh-bl)
may be computed or estimated.} than such a one as yours."

"Lizzy," cried her mother, "remember where you are, and do not
run on in the wild manner that you are suffered to do at home."

"I did not know before," continued Bingley immediately, "that
your were a studier of character.  It must be an amusing study."

"Yes, but intricate characters are the \emph{most} amusing.  They
have at least that advantage."

"The country," said Darcy, "can in general supply but a few
subjects\index{subjects} for such a study.  In a country neighbourhood you move
in a very confined and unvarying society."

"But people themselves alter so much, that there is something
new to be observed in them for ever."

"Yes, indeed," cried Mrs. Bennet, offended by his manner of
mentioning a country neighbourhood.  "I assure you there is
quite as much of \emph{that} going on in the country as in town."

Everybody was surprised, and Darcy, after looking at her for a
moment, turned silently away.  Mrs. Bennet, who fancied she
had gained a complete victory over him, continued her \textbf{triumph}\index{triumph}\stepcounter{footnote}\footnotetext{
\textbf{triumph} \emph{intr.v.}  (TRY-uhmf)
a successful ending of a struggle or contest; a win.}.

"I cannot see that London has any great advantage over the
country, for my part, except the shops and public places.  The
country is a vast\index{vast} deal pleasanter, is it not, Mr. Bingley?"

"When I am in the country," he replied, "I never wish to leave it;
and when I am in town it is pretty much the same.  They have
each their advantages, and I can be equally happy in either."

"Aye--that is because you have the right disposition\index{disposition}.  But that
gentleman," looking at Darcy, "seemed to think the country was
nothing at all."

"Indeed, Mamma, you are mistaken," said Elizabeth, blushing for
her mother.  "You quite mistook Mr. Darcy.  He only meant that
there was not such a variety of people to be met with in the
country as in the town, which you must acknowledge to be
true."

"Certainly, my dear, nobody said there were; but as to not
meeting with many people in this neighbourhood, I believe
there are few neighbourhoods larger.  I know we dine with
four-and-twenty families."

Nothing but concern for Elizabeth could enable Bingley to keep
his countenance\index{countenance}.  His sister was less delicate, and directed her
eyes towards Mr. Darcy with a very expressive smile.  Elizabeth,
for the sake of saying something that might turn her mother's
thoughts, now asked her if Charlotte Lucas had been at
Longbourn since \emph{her} coming away.

"Yes, she called yesterday with her father.  What an agreeable
man Sir William is, Mr. Bingley, is not he?  So much the man of
fashion\index{fashion}!  So \textbf{genteel}\index{genteel}\stepcounter{footnote}\footnotetext{
\textbf{genteel} \emph{adj.}  (jehn-TEEL)
marked by refinement in taste and manners; well-bred and polite.} and easy!  He had always something to say
to everybody.  \emph{that} is my idea of good breeding; and those
persons who fancy themselves very important, and never open
their mouths, quite mistake the matter."

"Did Charlotte dine with you?"

"No, she would go home.  I fancy she was wanted about the
mince-pies.  For my part, Mr. Bingley, I always keep servants
that can do their own work; \emph{my} daughters are brought up very
differently.  But everybody is to judge for themselves, and the
Lucases are a very good sort of girls, I assure you.  It is a pity
they are not handsome!  Not that I think Charlotte so \emph{very}
plain--but then she is our particular friend."

"She seems a very pleasant young woman."

"Oh! dear, yes; but you must own she is very plain.  Lady Lucas
herself has often said so, and envied me Jane's beauty.  I do not
like to boast of my own child, but to be sure, Jane--one does
not often see anybody better looking.  It is what everybody says.
I do not trust my own \textbf{partiality}\index{partiality}\stepcounter{footnote}\footnotetext{
\textbf{partiality} \emph{n.}  (par-shee-AL-ih-tee)
a predisposition to like something.}.  When she was only fifteen,
there was a man at my brother Gardiner's in town so much in
love with her that my sister-in-law was sure he would make her
an offer before we came away.  But, however, he did not.
Perhaps he thought her too young.  However, he wrote some
verses on her, and very pretty they were."

"And so ended his affection," said Elizabeth impatiently.  "There
has been many a one, I fancy, overcome in the same way.  I
wonder who first discovered the \textbf{efficacy}\index{efficacy}\stepcounter{footnote}\footnotetext{
\textbf{efficacy} \emph{n.}  (EHF-fih-kuh-see)
capacity or power to produce a desired effect or result.} of poetry in driving
away love!"

"I have been used to consider poetry as the \emph{food} of love," said
Darcy.

"Of  a fine, stout\index{stout}, healthy love it may.  Everything nourishes
what is strong already.  But if it be only a slight, thin sort of
\textbf{inclination}\index{inclination}\stepcounter{footnote}\footnotetext{
\textbf{inclination} \emph{n.}  (ihn-klih-NAY-shun)
an attitude that favors one alternative over others.}, I am convinced that one good sonnet will starve it
entirely away."

Darcy only smiled; and the general pause which ensued made
Elizabeth tremble lest her mother should be exposing herself
again.  She longed to speak, but could think of nothing to say;
and after a short silence Mrs. Bennet began repeating her thanks
to Mr. Bingley for his kindness to Jane, with an apology for
troubling him also with Lizzy.  Mr. Bingley was unaffectedly\index{unaffectedly}
civil in his answer, and forced his younger sister to be civil
also, and say what the occasion required.  She performed her
part indeed without much graciousness, but Mrs. Bennet was
satisfied, and soon afterwards ordered her carriage.  Upon this
signal, the youngest of her daughters put herself forward\index{forward}.  The
two girls had been whispering to each other during the whole
visit, and the result of it was, that the youngest should tax
Mr. Bingley with having promised on his first coming into the
country to give a ball at Netherfield.

Lydia was a stout\index{stout}, well-grown girl of fifteen, with a fine
complexion and good-humoured countenance\index{countenance}; a favourite with her
mother, whose affection had brought her into public at an early
age.  She had high animal spirits, and a sort of natural
self-consequence\index{consequence}, which the attention of the officers, to whom
her uncle's good dinners, and her own easy manners recommended
her, had increased into assurance.  She was very equal,
therefore, to address\index{address} Mr. Bingley on the subject\index{subject} of the ball, and
abruptly reminded him of his promise; adding, that it would be
the most shameful thing in the world if he did not keep it.  His
answer to this sudden attack was delightful to their mother's ear:

"I am perfectly ready, I assure you, to keep my engagement; and
when your sister is recovered, you shall, if you please, name the
very day of the ball.  But you would not wish to be dancing
when she is ill."

Lydia declared herself satisfied.  "Oh! yes--it would be much
better to wait till Jane was well, and by that time most likely
Captain Carter would be at Meryton again.  And when you have
given \emph{your} ball," she added, "I shall insist on their giving one
also.  I shall tell Colonel Forster it will be quite a shame if he
does not."

Mrs. Bennet and her daughters then departed, and Elizabeth
returned instantly to Jane, leaving her own and her relations'
behaviour to the remarks of the two ladies and Mr. Darcy; the
latter of whom, however, could not be \textbf{prevailed}\index{prevailed}\stepcounter{footnote}\footnotetext{
\textbf{prevail} \emph{intr.v.} \textbf{-ed}  (prih-VAYL)
to be stronger or able to subdue; to win over or convince.} on to join in
their censure\index{censure} of \emph{her}, in spite of all Miss Bingley's \textbf{witticisms}\index{witticisms}\stepcounter{footnote}\footnotetext{
\textbf{witticism} \emph{n.}  (WIT-tih-sizm)
A witty, clever, or amusing remark or saying.} on
\emph{fine} \emph{eyes}.



\chapter{10}


The day passed much as the day before had done.  Mrs. Hurst
and Miss Bingley had spent some hours of the morning with the
\textbf{invalid}\index{invalid}\stepcounter{footnote}\footnotetext{
\textbf{invalid} \emph{n.}  (IHN-vuh-lihd)
someone rendered non-functional by disability or other condition.}, who continued, though slowly, to mend; and in the
evening Elizabeth joined their party in the drawing-room.  The
loo\index{loo}-table, however, did not appear.  Mr. Darcy was writing, and
Miss Bingley, seated near him, was watching the progress of his
letter and repeatedly calling off his attention by messages to
his sister.  Mr. Hurst and Mr. Bingley were at \textbf{piquet}\index{piquet}\stepcounter{footnote}\footnotetext{
\textbf{piquet} \emph{n.}  (pih-KEHT)
 a card game for two players using a reduced pack of 32 cards.}, and Mrs.
Hurst was observing their game.

Elizabeth took up some needlework, and was sufficiently
amused in attending to what passed between Darcy and his
companion.  The \textbf{perpetual}\index{perpetual}\stepcounter{footnote}\footnotetext{
\textbf{perpetual} \emph{adj.} \textbf{-ly}  (per-PEHT-yoo-uhl)
continuing forever or indefinitely.} \textbf{commendations}\index{commendations}\stepcounter{footnote}\footnotetext{
\textbf{commendation} \emph{n.}  (kah-men-DAY-shun)
An act of praise, bestowing honor and public recognition.} of the lady, either on
his handwriting, or on the evenness of his lines, or on the length
of his letter, with the perfect unconcern with which her praises
were received, formed a curious \textbf{dialogue}\index{dialogue}\stepcounter{footnote}\footnotetext{
\textbf{dialog} \emph{n.} \textbf{-ue}  (DIE-uh-log)
a conversation between two people.}, and was exactly in
union with her opinion of each.

"How delighted Miss Darcy will be to receive such a letter!"

He made no answer.

"You write uncommonly fast."

"You are mistaken.  I write rather slowly."

"How many letters you must have occasion to write in the
course of a year!  Letters of business, too!  How \textbf{odious}\index{odious}\stepcounter{footnote}\footnotetext{
\textbf{odious} \emph{adj.}  (OH-dih-yus)
utterly detestable and hateful.} I should
think them!"

"It is fortunate, then, that they fall to my lot instead of yours."

"Pray tell your sister that I long to see her."

"I have already told her so once, by your desire."

"I am afraid you do not like your pen.  Let\index{Let} me mend it for you.
I mend pens remarkably well."

"Thank you--but I always mend my own."

"How can you contrive\index{contrive} to write so even?"

He was silent.

"Tell your sister I am delighted to hear of her improvement on
the harp; and pray let\index{let} her know that I am quite in raptures\index{raptures} with
her beautiful little design for a table, and I think it infinitely
superior\index{superior} to Miss Grantley's."

"Will you give me leave to defer\index{defer} your raptures\index{raptures} till I write again?
At present I have not room to do them justice."

"Oh! it is of no consequence\index{consequence}.  I shall see her in January.  But do
you always write such charming long letters to her, Mr. Darcy?"

"They are generally long; but whether always charming it is not
for me to determine."

"It is a rule with me, that a person who can write a long letter
with ease, cannot write ill."

"That will not do for a compliment to Darcy, Caroline," cried
her brother, "because he does \emph{not} write with ease.  He studies
too much for words of four syllables.  Do not you, Darcy?"

"My style of writing is very different from yours."

"Oh!" cried Miss Bingley, "Charles writes in the most careless
way imaginable.  He leaves out half his words, and blots the
rest."

"My ideas flow so rapidly\index{rapidly} that I have not time to express
them--by which means my letters sometimes convey no ideas
at all to my correspondents."

"Your \textbf{humility}\index{humility}\stepcounter{footnote}\footnotetext{
\textbf{humility} \emph{n.}  (hyoo-MIL-lih-tee)
the quality of being humble or modest; without pride or arrogance.}, Mr. Bingley," said Elizabeth, "must \textbf{disarm}\index{disarm}\stepcounter{footnote}\footnotetext{
\textbf{disarm} \emph{v.}  (dihs-ARM)
remove offensive capability from; to take weapons away from.}
\textbf{reproof}\index{reproof}\stepcounter{footnote}\footnotetext{
\textbf{reproof} \emph{n.}  (rih-PROOF)
an expression of criticism and censure; to scold or rebuke.}."

"Nothing is more deceitful," said Darcy, "than the appearance of
humility\index{humility}.  It is often only carelessness of opinion, and sometimes
an indirect boast."

"And which of the two do you call \emph{my} little recent piece of
modesty?"

"The indirect boast; for you are really proud of your defects in
writing, because you consider them as proceeding from a
\textbf{rapidity}\index{rapidity}\stepcounter{footnote}\footnotetext{
\textbf{rapid} \emph{adj.} \textbf{-ity}  (RAP-pihd)
 in quick succession or flow.} of thought and carelessness of \textbf{execution}\index{execution}\stepcounter{footnote}\footnotetext{
\textbf{execution} \emph{n.}  (ehx-seh-KYOO-shun)
completion of an order or plan.}, which, if not
estimable\index{estimable}, you think at least highly interesting.  The power of
doing anything with quickness is always prized much by the
possessor, and often without any attention to the imperfection of
the performance.  When you told Mrs. Bennet this morning that
if you ever resolved upon quitting Netherfield you should be
gone in five minutes\index{minutes}, you meant it to be a sort of \textbf{panegyric}\index{panegyric}\stepcounter{footnote}\footnotetext{
\textbf{panegyric} \emph{n.}  (pan-uh-JEE-rihk)
formal public expression of praise.}, of
compliment to yourself--and yet what is there so very \textbf{laudable}\index{laudable}\stepcounter{footnote}\footnotetext{
\textbf{laudable} \emph{adj.}  (LAW-duh-bl)
worthy of high praise.}
in a \textbf{precipitance}\index{precipitance}\stepcounter{footnote}\footnotetext{
\textbf{precipitancy} \emph{n.}  (prih-SIH-pih-tan-see)
the quality of happening with headlong haste or without warning.} which must leave very necessary business
undone, and can be of no real advantage to yourself or anyone
else?"

"Nay," cried Bingley, "this is too much, to remember at night all
the foolish things that were said in the morning.  And yet, upon
my honour, I believe what I said of myself to be true, and I
believe it at this moment.  At least, therefore, I did not assume
the character of needless precipitance merely to show off before
the ladies."

"I dare say you believed it; but I am by no means convinced that
you would be gone with such \textbf{celerity}\index{celerity}\stepcounter{footnote}\footnotetext{
\textbf{celerity} \emph{n.}  (suh-LEHR-rih-tee)
speedy rapid action.}.  Your conduct would be
quite as dependent on chance as that of any man I know; and if,
as you were mounting your horse, a friend were to say,  'Bingley,
you had better stay till next week,' you would probably do it,
you would probably not go--and at another word, might stay a
month."

"You have only proved by this," cried Elizabeth, "that Mr.
Bingley did not do justice to his own disposition\index{disposition}.  You have
shown him off now much more than he did himself."

"I am exceedingly gratified," said Bingley, "by your converting
what my friend says into a compliment on the sweetness of my
temper.  But I am afraid you are giving it a turn which that
gentleman did by no means intend; for he would certainly think
better of me, if under such a circumstance I were to give a flat
denial, and ride off as fast as I could."

"Would Mr. Darcy then consider the \textbf{rashness}\index{rashness}\stepcounter{footnote}\footnotetext{
\textbf{rash} \emph{adj.} \textbf{-ness}  (RASH)
 haste or recklessness in face of danger.} of your original
intentions as \textbf{atoned}\index{atoned}\stepcounter{footnote}\footnotetext{
\textbf{atone} \emph{v.} \textbf{-ed}  (uh-TOAN)
 to correct or make right a sin or transgression.} for by your \textbf{obstinacy}\index{obstinacy}\stepcounter{footnote}\footnotetext{
\textbf{obstinacy} \emph{n.}  (AHB-stih-nuh-see)
resolute adherence to one's own ideas; stubborn.} in \textbf{adhering}\index{adhering}\stepcounter{footnote}\footnotetext{
\textbf{adhere} \emph{v.} \textbf{-ing}  (ad-HEER)
to stick to (like glue); to be resolute in an idea or ideology.} to it?"

"Upon my word, I cannot exactly explain the matter; Darcy must
speak for himself."

"You expect me to account for opinions which you choose to
call mine, but which I have never acknowledged.  Allowing the
case, however, to stand according to your representation, you
must remember, Miss Bennet, that the friend who is supposed to
desire his return to the house, and the delay of his plan, has
merely desired it, asked it without offering one argument in
favour of its propriety\index{propriety}."

"To yield readily\index{readily}--easily--to the \emph{persuasion} of a friend is
no merit with you."

"To yield without \textbf{conviction}\index{conviction}\stepcounter{footnote}\footnotetext{
\textbf{conviction} \emph{n.}  (kun-VIK-shun)
an unshakable belief not needing proof or evidence.} is no compliment to the understanding
of either."

"You appear to me, Mr. Darcy, to allow nothing for the
influence of friendship and affection.  A regard for the requester
would often make one readily\index{readily} yield to a request, without waiting
for arguments to reason one into it.  I am not particularly
speaking of such a case as you have supposed about Mr.
Bingley.  We may as well wait, perhaps, till the circumstance
occurs before we discuss the discretion\index{discretion} of his behaviour
thereupon.  But in general and ordinary cases between friend and
friend, where one of them is desired by the other to change a
resolution\index{resolution} of no very great moment, should you think ill of that
person for \textbf{complying}\index{complying}\stepcounter{footnote}\footnotetext{
\textbf{comply} \emph{intr.v.} \textbf{-ing}  (kuhm-PLY)
to follow another person's order; to acquiesce or act in accord.} with the desire, without waiting to be
argued into it?"

"Will it not be advisable, before we proceed on this subject\index{subject}, to
arrange with rather more precision the degree of importance
which is to \textbf{appertain}\index{appertain}\stepcounter{footnote}\footnotetext{
\textbf{appertain} \emph{intr.v.}  (ap-per-TAYN)
be a part or attribute of.} to this request, as well as the degree of
intimacy \textbf{subsisting}\index{subsisting}\stepcounter{footnote}\footnotetext{
\textbf{subsist} \emph{v.} \textbf{-ed}  (suhb-SIST)
} between the parties?"

"By all means," cried Bingley; "let\index{let} us hear all the particulars,
not forgetting their \textbf{comparative}\index{comparative}\stepcounter{footnote}\footnotetext{
\textbf{comparative} \emph{adj.}  (kum-PAIR-ruh-tihv)
something's significance in relation to something else.} height and size; for that will
have more weight in the argument, Miss Bennet, than you may be
aware of.  I assure you, that if Darcy were not such a great tall
fellow, in comparison with myself, I should not pay him half so
much \textbf{deference}\index{deference}\stepcounter{footnote}\footnotetext{
\textbf{deference} \emph{n.}  (DEH-fer-ense)
a tendency to yield to the will of others.}.  I declare I do not know a more awful object
than Darcy, on particular occasions, and in particular places; at
his own house especially, and of a Sunday evening, when he has
nothing to do."

Mr. Darcy smiled; but Elizabeth thought she could perceive\index{perceive} that
he was rather offended, and therefore checked her laugh.  Miss
Bingley warmly resented the \textbf{indignity}\index{indignity}\stepcounter{footnote}\footnotetext{
\textbf{indignity} \emph{n.}  (ihn-DIHG-nih-tee)
insulting, humiliating, or degrading treatment.} he had received, in an
\textbf{expostulation}\index{expostulation}\stepcounter{footnote}\footnotetext{
\textbf{expostulate} \emph{intr.v.}  (ex-PAHS-tyoo-layt)
an expression of opposition to a course of action.} with her brother for talking such nonsense.

"I see your design, Bingley," said his friend.  "You dislike an
argument, and want to silence this."

"Perhaps I do.  Arguments are too much like disputes.  If you and
Miss Bennet will defer\index{defer} yours till I am out of the room, I shall
be very thankful; and then you may say whatever you like of me."

"What you ask," said Elizabeth, "is no sacrifice on my side; and
Mr. Darcy had much better finish his letter."

Mr. Darcy took her advice, and did finish his letter.

When that business was over, he applied to Miss Bingley and
Elizabeth for an indulgence of some music.  Miss Bingley moved
with some \textbf{alacrity}\index{alacrity}\stepcounter{footnote}\footnotetext{
\textbf{alacrity} \emph{n.}  (uh-LAK-rih-tee)
liveliness and eagerness.} to the pianoforte\index{pianoforte}; and, after a polite request
that Elizabeth would lead the way which the other as politely
and more earnestly\index{earnestly} negatived, she seated herself.

Mrs. Hurst sang with her sister, and while they were thus
employed, Elizabeth could not help observing, as she turned
over some music-books that lay on the instrument, how frequently
Mr. Darcy's eyes were fixed on her.  She hardly knew how to
suppose that she could be an object of admiration to so great a
man; and yet that he should look at her because he disliked her,
was still more strange.  She could only imagine, however, at last
that she drew his notice because there was something more wrong
and \textbf{reprehensible}\index{reprehensible}\stepcounter{footnote}\footnotetext{
\textbf{reprehensible} \emph{adj.}  (reh-pree-HEN-sih-bl)
deserving severe rebuke or censure.}, according to his ideas of right, than in any
other person present.  The supposition\index{supposition} did not pain her.  She
liked him too little to care for his \textbf{approbation}\index{approbation}\stepcounter{footnote}\footnotetext{
\textbf{approbation} \emph{n.}  (ap-proh-BAY-shun)
official recognition or approval.}.

After playing some Italian songs, Miss Bingley varied the charm
by a lively Scotch air; and soon afterwards Mr. Darcy, drawing
near Elizabeth, said to her:

"Do not you feel a great inclination\index{inclination}, Miss Bennet, to seize such
an opportunity of dancing a \textbf{reel}\index{reel}\stepcounter{footnote}\footnotetext{
\textbf{reel} \emph{n.}  (REEL)
 a lively dance of Scottish highlanders; marked by circular moves and gliding steps.}?"

She smiled, but made no answer.  He repeated the question, with
some surprise at her silence.

"Oh!" said she, "I heard you before, but I could not immediately
determine what to say in reply.  You wanted me, I know, to say
'Yes,' that you might have the pleasure of despising my taste;
but I always delight in overthrowing those kind of schemes,
and cheating a person of their \textbf{premeditated}\index{premeditated}\stepcounter{footnote}\footnotetext{
\textbf{premeditate} \emph{v.} \textbf{-ed}  (prih-MEHD-ih-tayt)
to preconsider or plan in advance.} contempt\index{contempt}.  I have,
therefore, made up my mind to tell you, that I do not want to
dance a reel\index{reel} at all--and now despise me if you dare."

"Indeed I do not dare."

Elizabeth, having rather expected to \textbf{affront}\index{affront}\stepcounter{footnote}\footnotetext{
\textbf{affront} \emph{tr.v.} \textbf{-ed}  (af-FRUHNT)
a deliberately offensive act or statement.} him, was amazed at
his gallantry\index{gallantry}; but there was a mixture of sweetness and \textbf{archness}\index{archness}\stepcounter{footnote}\footnotetext{
\textbf{arch} \emph{n.} \textbf{-ness}  (ARCH)
}
in her manner which made it difficult for her to affront\index{affront} anybody;
and Darcy had never been so \textbf{bewitched}\index{bewitched}\stepcounter{footnote}\footnotetext{
\textbf{bewitch} \emph{tr.v.} \textbf{-ed}  (bee-WICH)
to put under a spell as if by magic.} by any woman as he
was by her.  He really believed, that were it not for the
\textbf{inferiority}\index{inferiority}\stepcounter{footnote}\footnotetext{
\textbf{inferior} \emph{adj.} \textbf{-ity}  (ihn-fih-ree-AW-rih-tee)
lower in rank, importance, or quality.} of her connections, he should be in some danger.

Miss Bingley saw, or suspected enough to be jealous; and her
great anxiety for the recovery of her dear friend Jane received
some assistance from her desire of getting rid of Elizabeth.

She often tried to provoke\index{provoke} Darcy into disliking her guest, by
talking of their supposed marriage, and planning his happiness in
such an \textbf{alliance}\index{alliance}\stepcounter{footnote}\footnotetext{
\textbf{alliance} \emph{n.}  (uh-LIE-anse)
an organization of people (or countries) in a pact or treaty.}.

"I hope," said she, as they were walking together in the
\textbf{shrubbery}\index{shrubbery}\stepcounter{footnote}\footnotetext{
\textbf{shrubbery} \emph{n.}  (SHRUH-ber-ee)
a collection of shrubs or bushes growing together.} the next day, "you will give your mother-in-law a few
hints, when this desirable event takes place, as to the advantage
of holding her tongue; and if you can \textbf{compass}\index{compass}\stepcounter{footnote}\footnotetext{
\textbf{compass} \emph{n.}  (KAHM-pass)
 an area in which something operates or has power or control.} it, do sure the
younger girls of running after officers.  And, if I may mention so
delicate a subject\index{subject}, endeavour\index{endeavour} to check that little something,
bordering on conceit\index{conceit} and impertinence\index{impertinence}, which your lady
possesses."

"Have you anything else to propose for my \textbf{domestic}\index{domestic}\stepcounter{footnote}\footnotetext{
\textbf{domestic} \emph{adj.}  (duh-MEHS-tihk)
relating to the home or family; a household servant.} felicity\index{felicity}?"

"Oh! yes.  Do let\index{let} the portraits of your uncle and aunt Phillips be
placed in the gallery at Pemberley.  Put them next to your
great-uncle the judge.  They are in the same profession, you
know, only in different lines.  As for your Elizabeth's picture, you
must not have it taken, for what painter could do justice to those
beautiful eyes?"

"It would not be easy, indeed, to catch their expression, but their
colour and shape, and the eyelashes, so remarkably fine, might
be copied."

At that moment they were met from another walk by Mrs. Hurst
and Elizabeth herself.

"I did not know that you intended to walk," said Miss Bingley,
in some confusion, lest they had been overheard.

"You used us abominably ill," answered Mrs. Hurst, "running
away without telling us that you were coming out."

Then taking the disengaged arm of Mr. Darcy, she left Elizabeth
to walk by herself.  The path just admitted three.  Mr. Darcy felt
their rudeness, and immediately said:

"This walk is not wide enough for our party.  We had better go
into the avenue."

But Elizabeth, who had not the least inclination\index{inclination} to remain with
them, laughingly answered:

"No, no; stay where you are.  You are charmingly grouped, and
appear to uncommon advantage.  The \textbf{picturesque}\index{picturesque}\stepcounter{footnote}\footnotetext{
\textbf{picturesque} \emph{adj.}  (pik-choo-RESK)
suitable for a picture; pretty as a picture.} would be
spoilt by admitting a fourth.  Good-bye."

She then ran gaily off, rejoicing as she \textbf{rambled}\index{rambled}\stepcounter{footnote}\footnotetext{
\textbf{ramble} \emph{intr.v.} \textbf{-ed}  (RAM-bl)
 to move or speak aimlessly; to drone.} about, in the
hope of being at home again in a day or two.  Jane was already
so much recovered as to intend leaving her room for a couple of
hours that evening.



\chapter{11}


When the ladies removed after dinner, Elizabeth ran up to her
sister, and seeing her well guarded from cold, attended her into
the drawing-room, where she was welcomed by her two friends
with many professions of pleasure; and Elizabeth had never seen
them so agreeable as they were during the hour which passed
before the gentlemen appeared.  Their powers of conversation
were considerable.  They could describe an entertainment with
accuracy, relate an \textbf{anecdote}\index{anecdote}\stepcounter{footnote}\footnotetext{
\textbf{anecdote} \emph{n.}  (AN-nehk-doat)
short account of an incident (especially biographical).} with humour, and laugh at their
acquaintance with spirit.

But when the gentlemen entered, Jane was no longer the first
object; Miss Bingley's eyes were instantly turned toward Darcy,
and she had something to say to him before he had advanced
many steps.  He addressed\index{addressed} himself to Miss Bennet, with a polite
congratulation; Mr. Hurst also made her a slight bow, and said
he was "very glad;" but \textbf{diffuseness}\index{diffuseness}\stepcounter{footnote}\footnotetext{
\textbf{diffuse} \emph{v.} \textbf{-ed}  (dih-FYOOS)
 spread out through a wide area or large volume.} and warmth remained for
Bingley's \textbf{salutation}\index{salutation}\stepcounter{footnote}\footnotetext{
\textbf{} \emph{}  ()
}.  He was full of joy and attention.  The first
half-hour was spent in piling up the fire, lest she should suffer
from the change of room; and she removed at his desire to the
other side of the fireplace, that she might be further from the
door.  He then sat down by her, and talked scarcely to anyone
else.  Elizabeth, at work in the opposite corner, saw it all with
great delight.

When tea was over, Mr. Hurst reminded his sister-in-law of the
card-table--but in vain.  She had obtained private\index{private} intelligence
that Mr. Darcy did not wish for cards; and Mr. Hurst soon found
even his open \textbf{petition}\index{petition}\stepcounter{footnote}\footnotetext{
\textbf{petition} \emph{n.}  (puh-TIH-shun)
a formal request of an authority.} rejected.  She assured him that no one
intended to play, and the silence of the whole party on the
subject\index{subject} seemed to justify her.  Mr. Hurst had therefore nothing
to do, but to stretch himself on one of the sofas and go to
sleep.  Darcy took up a book; Miss Bingley did the same; and
Mrs. Hurst, principally\index{principally} occupied in playing with her bracelets
and rings, joined now and then in her brother's conversation
with Miss Bennet.

Miss Bingley's attention was quite as much engaged in watching
Mr. Darcy's progress through \emph{his} book, as in reading her own;
and she was perpetually\index{perpetually} either making some inquiry, or looking
at his page.  She could not win him, however, to any conversation;
he merely answered her question, and read on.  At length, quite
exhausted by the attempt to be amused with her own book, which
she had only chosen because it was the second volume of his,
she gave a great yawn and said, "How pleasant it is to spend an
evening in this way!  I declare after all there is no enjoyment
like reading!  How much sooner one tires of anything than of a
book!  When I have a house of my own, I shall be miserable if
I have not an excellent library."

No one made any reply.  She then yawned again, threw aside her
book, and cast her eyes round the room in quest for some
amusement; when hearing her brother mentioning a ball to Miss
Bennet, she turned suddenly towards him and said:

"By the bye, Charles, are you really serious in meditating a dance
at Netherfield?  I would advise you, before you determine on it,
to consult the wishes of the present party; I am much mistaken if
there are not some among us to whom a ball would be rather a
punishment than a pleasure."

"If you mean Darcy," cried her brother, "he may go to bed, if he
chooses, before it begins--but as for the ball, it is quite a settled
thing; and as soon as Nicholls has made white soup enough, I
shall send round my cards."

"I should like balls infinitely better," she replied, "if they
were carried on in a different manner; but there is something
\textbf{insufferably}\index{insufferably}\stepcounter{footnote}\footnotetext{
\textbf{insufferable} \emph{adj.} \textbf{-ly}  (ihn-SUHF-fer-uh-bl)
hard to endure, intolerable.} \textbf{tedious}\index{tedious}\stepcounter{footnote}\footnotetext{
\textbf{tedious} \emph{adj.}  (TEE-dee-yuhs)
 so lacking in interest as to cause mental weariness.} in the usual process of such a meeting.  It
would surely be much more rational\index{rational} if conversation instead of
dancing were made the order of the day."

"Much more rational\index{rational}, my dear Caroline, I dare say, but it would
not be near so much like a ball."

Miss Bingley made no answer, and soon afterwards she got up
and walked about the room.  Her figure was elegant, and she
walked well; but Darcy, at whom it was all aimed, was still
inflexibly \textbf{studious}\index{studious}\stepcounter{footnote}\footnotetext{
\textbf{studious} \emph{adj.} \textbf{-ly}  (STOO-dih-uhs)
marked by care and effort; diligent in study.}.  In the desperation of her feelings, she
resolved on one effort more, and, turning to Elizabeth, said:

"Miss Eliza Bennet, let\index{let} me persuade you to follow my example,
and take a turn about the room.  I assure you it is very
refreshing after sitting so long in one \textbf{attitude}\index{attitude}\stepcounter{footnote}\footnotetext{
\textbf{attitude} \emph{n.}  (AT-tih-tood)
}."

Elizabeth was surprised, but agreed to it immediately.  Miss
Bingley succeeded no less in the real object of her civility\index{civility};
Mr. Darcy looked up.  He was as much awake to the \textbf{novelty}\index{novelty}\stepcounter{footnote}\footnotetext{
\textbf{novelty} \emph{n.}  (NAH-vuhl-tee)
 originality by being new and surprising.} of
attention in that quarter as Elizabeth herself could be, and
\textbf{unconsciously}\index{unconsciously}\stepcounter{footnote}\footnotetext{
\textbf{unconscious} \emph{adj.} \textbf{-ly}  (uhn-KAHN-shus)
without awareness; lacking sensory perception; not conscious.} closed his book.  He was directly invited to join
their party, but he declined it, observing that he could imagine
but two motives for their choosing to walk up and down the
room together, with either of which motives his joining them
would interfere.  "What could he mean?  She was dying to know
what could be his meaning?"--and asked Elizabeth whether she
could at all understand him?

"Not at all," was her answer; "but depend upon it, he means to
be severe\index{severe} on us, and our surest way of disappointing him will be
to ask nothing about it."

Miss Bingley, however, was incapable of disappointing Mr.
Darcy in anything, and \textbf{persevered}\index{persevered}\stepcounter{footnote}\footnotetext{
\textbf{persevere} \emph{intr.v.} \textbf{-ly}  (per-suh-VEER)
} therefore in requiring an
explanation of his two motives.

"I have not the smallest objection to explaining them," said he,
as soon as she allowed him to speak.  "You either choose this
method of passing the evening because you are in each other's
\textbf{confidence}\index{confidence}\stepcounter{footnote}\footnotetext{
\textbf{confidence} \emph{n.}  (KAHN-fih-dense)
a feeling of trust (in someone or something).}, and have secret affairs to discuss, or because you are
conscious that your figures appear to the greatest advantage in
walking; if the first, I would be completely in your way, and if
the second, I can admire you much better as I sit by the fire."

"Oh! shocking!" cried Miss Bingley.  "I never heard anything so
abominable\index{abominable}.  How shall we punish him for such a speech?"

"Nothing so easy, if you have but the inclination\index{inclination}," said Elizabeth.
"We can all \textbf{plague}\index{plague}\stepcounter{footnote}\footnotetext{
\textbf{plague} \emph{n.}  (PLAYG)
a serious and widespread disease epidemic, often seen as divine retribution.} and punish one another.  Tease him--laugh
at him.  Intimate as you are, you must know how it is to be
done."

"But upon my honour, I do \emph{not}.  I do assure you that my
intimacy has not yet taught me \emph{that}.  Tease calmness of
manner and presence of mind!  No, no--feel he may defy us
there.  And as to laughter, we will not expose ourselves, if you
please, by attempting to laugh without a subject\index{subject}.  Mr. Darcy may
hug himself."

"Mr. Darcy is not to be laughed at!" cried Elizabeth.  "That is
an uncommon advantage, and uncommon I hope it will continue, for
it would be a great loss to \emph{me} to have many such acquaintances.
I dearly love a laugh."

"Miss Bingley," said he, "has given me more credit than can be.
The wisest and the best of men--nay, the wisest and best of their
actions--may be rendered\index{rendered} ridiculous by a person whose first
object in life is a joke."

"Certainly," replied Elizabeth--"there are such people, but I
hope I am not one of \emph{them}.  I hope I never ridicule what is
wise and good.  Follies and nonsense, \textbf{whims}\index{whims}\stepcounter{footnote}\footnotetext{
\textbf{} \emph{}  ()
} and
\textbf{inconsistencies}\index{inconsistencies}\stepcounter{footnote}\footnotetext{
\textbf{inconsistency} \emph{n.}  (ihn-kuhn-SIS-tehn-see)
}, \emph{do} divert\index{divert} me, I own, and I laugh at them
whenever I can.  But these, I suppose, are precisely what you
are without."

"Perhaps that is not possible for anyone.  But it has been the
study of my life to avoid those weaknesses which often expose a
strong understanding to ridicule."

"Such as vanity\index{vanity} and pride."

"Yes, vanity\index{vanity} is a weakness indeed.  But pride--where there is a
real \textbf{superiority}\index{superiority}\stepcounter{footnote}\footnotetext{
\textbf{superior} \emph{adj.} \textbf{-ity}  (soo-PEER-rih-yer)
better than something or someone else in quality or rank.} of mind, pride will be always under good
\textbf{regulation}\index{regulation}\stepcounter{footnote}\footnotetext{
\textbf{regulation} \emph{n.}  (rehg-yoo-LAY-shun)
an authoritative rule or law.}."

Elizabeth turned away to hide a smile.

"Your examination of Mr. Darcy is over, I presume," said Miss
Bingley; "and pray what is the result?"

"I am perfectly convinced by it that Mr. Darcy has no defect.
He owns it himself without disguise."

"No," said Darcy, "I have made no such \textbf{pretension}\index{pretension}\stepcounter{footnote}\footnotetext{
\textbf{pretension} \emph{n.}  (prih-TEHN-shun)
a false or unsupportable pretext or quality.}.  I have
faults enough, but they are not, I hope, of understanding.
My temper I dare not \textbf{vouch}\index{vouch}\stepcounter{footnote}\footnotetext{
\textbf{vouch} \emph{v.}  (VOUCH)
 to give personal assurance; to guarantee.} for.  It is, I believe, too little
yielding--certainly too little for the convenience of the world.
I cannot forget the follies and \textbf{vices}\index{vices}\stepcounter{footnote}\footnotetext{
\textbf{} \emph{}  ()
} of other so soon as I ought,
nor their offenses against myself.  My feelings are not puffed
about with every attempt to move them.  My temper would perhaps be
called resentful.  My good opinion once lost, is lost forever."

"\emph{that} is a failing indeed!" cried Elizabeth.  "\textbf{Implacable}\index{implacable}\stepcounter{footnote}\footnotetext{
\textbf{implacable} \emph{adj.}  (im-PLA-kuh-bl)
impossible to appease, sooth, or placate.}
resentment\index{resentment} \emph{is} a shade in a character.  But you have chosen your
fault well.  I really cannot \emph{laugh} at it.  You are safe from me."

"There is, I believe, in every disposition\index{disposition} a tendency to some
particular evil--a natural defect, which not even the best
education can overcome."

"And \emph{your} defect is to hate everybody."

"And yours," he replied with a smile, "is willfully to
misunderstand them."

"Do let\index{let} us have a little music," cried Miss Bingley, tired of a
conversation in which she had no share.  "Louisa, you will not
mind my waking Mr. Hurst?"

Her sister had not the smallest objection, and the pianoforte\index{pianoforte} was
opened; and Darcy, after a few moments' \textbf{recollection}\index{recollection}\stepcounter{footnote}\footnotetext{
\textbf{recollection} \emph{n.}  (reh-kuhl-LEHK-shun)
something recalled to the mind.}, was not
sorry for it.  He began to feel the danger of paying Elizabeth too
much attention.



\chapter{12}


In consequence\index{consequence} of an agreement between the sisters, Elizabeth
wrote the next morning to their mother, to beg that the carriage
might be sent for them in the course of the day.  But Mrs.
Bennet, who had calculated on her daughters remaining at
Netherfield till the following Tuesday, which would exactly
finish Jane's week, could not bring herself to receive them with
pleasure before.  Her answer, therefore, was not \textbf{propitious}\index{propitious}\stepcounter{footnote}\footnotetext{
\textbf{propitious} \emph{adj.}  (proh-PIH-shus)
presenting favorable circumstances.}, at
least not to Elizabeth's wishes, for she was impatient to get
home.  Mrs. Bennet sent them word that they could not possibly
have the carriage before Tuesday; and in her postscript it was
added, that if Mr. Bingley and his sister pressed them to stay
longer, she could spare them very well.  Against staying longer,
however, Elizabeth was positively resolved--nor did she much
expect it would be asked; and fearful, on the contrary, as being
considered as intruding themselves needlessly long, she urged
Jane to borrow Mr. Bingley's carriage immediately, and at
length it was settled that their original design of leaving
Netherfield that morning should be mentioned, and the request
made.

The communication excited many professions of concern; and
enough was said of wishing them to stay at least till the
following day to work on Jane; and till the morrow their going
was deferred.  Miss Bingley was then sorry that she had
proposed the delay, for her jealousy and dislike of one sister
much exceeded her affection for the other.

The master of the house heard with real sorrow that they were
to go so soon, and repeatedly tried to persuade Miss Bennet that
it would not be safe for her--that she was not enough recovered;
but Jane was firm where she felt herself to be right.

To Mr. Darcy it was welcome intelligence--Elizabeth had been
at Netherfield long enough.  She attracted him more than he
liked--and Miss Bingley was \textbf{uncivil}\index{uncivil}\stepcounter{footnote}\footnotetext{
\textbf{uncivil} \emph{adj.}  (uhn-SIV-ihl)
lacking decency or good manners.} to \emph{her}, and more teasing
than usual to himself.  He wisely resolved to be particularly
careful that no sign of admiration should \emph{now} escape him,
nothing that could elevate her with the hope of influencing his
felicity\index{felicity}; sensible that if such an idea had been suggested,
his behaviour during the last day must have material weight in
confirming or crushing it.  Steady\index{Steady} to his purpose, he scarcely
spoke ten words to her through the whole of Saturday, and
though they were at one time left by themselves for half-an-hour,
he adhered\index{adhered} most \textbf{conscientiously}\index{conscientiously}\stepcounter{footnote}\footnotetext{
\textbf{conscientious} \emph{adj.} \textbf{-ly}  (kahn-shee-EHN-shus)
careful or meticulous.} to his book, and would not
even look at her.

On Sunday, after morning service, the separation, so agreeable
to almost all, took place.  Miss Bingley's civility\index{civility} to Elizabeth
increased at last very rapidly\index{rapidly}, as well as her affection for Jane;
and when they parted, after assuring the latter of the pleasure
it would always give her to see her either at Longbourn or
Netherfield, and embracing her most tenderly, she even shook
hands with the former.  Elizabeth took leave of the whole party
in the liveliest of spirits.

They were not welcomed home very cordially\index{cordially} by their mother.
Mrs. Bennet wondered at their coming, and thought them very
wrong to give so much trouble, and was sure Jane would have
caught cold again.  But their father, though very \textbf{laconic}\index{laconic}\stepcounter{footnote}\footnotetext{
\textbf{laconic} \emph{adj.}  (lak-KAW-nihk)
 brief and to the point; concise.} in his
expressions of pleasure, was really glad to see them; he had felt
their importance in the family circle.  The evening conversation,
when they were all assembled, had lost much of its animation,
and almost all its sense by the absence of Jane and Elizabeth.

They found Mary, as usual, deep in the study of thorough-bass
and human nature; and had some extracts\index{extracts} to admire, and some
new observations of threadbare \textbf{morality}\index{morality}\stepcounter{footnote}\footnotetext{
\textbf{morality} \emph{n.}  (maw-RAL-lih-tee)
acting in line with right or good conduct.} to listen to.  Catherine
and Lydia had information for them of a different sort.  Much
had been done and much had been said in the regiment\index{regiment} since the
preceding Wednesday; several of the officers had dined lately
with their uncle, a private\index{private} had been flogged, and it had actually
been hinted that Colonel Forster was going to be married.



\chapter{13}


"I hope, my dear," said Mr. Bennet to his wife, as they were
at breakfast the next morning, "that you have ordered a good
dinner to-day, because I have reason to expect an addition to
our family party."

"Who do you mean, my dear?  I know of nobody that is coming,
I am sure, unless Charlotte Lucas should happen to call in--and
I hope \emph{my} dinners are good enough for her.  I do not believe
she often sees such at home."

"The person of whom I speak is a gentleman, and a stranger."

Mrs. Bennet's eyes sparkled.  "A gentleman and a stranger!  It is
Mr. Bingley, I am sure!  Well, I am sure I shall be extremely glad
to see Mr. Bingley.  But--good Lord! how unlucky!  There is not
a bit of fish to be got to-day.  Lydia, my love, ring the bell--I
must speak to Hill this moment."

"It is \emph{not} Mr. Bingley," said her husband; "it is a person whom
I never saw in the whole course of my life."

This roused a general astonishment; and he had the pleasure of
being eagerly questioned by his wife and his five daughters at
once.

After amusing himself some time with their curiosity, he thus
explained:

"About a month ago I received this letter; and about a fortnight\index{fortnight}
ago I answered it, for I thought it a case of some \textbf{delicacy}\index{delicacy}\stepcounter{footnote}\footnotetext{
\textbf{delicacy} \emph{n.}  (DEH-lih-kuh-see)
something considered choice to eat, often rare.}, and
requiring early attention.  It is from my cousin, Mr. Collins, who,
when I am dead, may turn you all out of this house as soon as he
pleases."

"Oh! my dear," cried his wife, "I cannot bear to hear that
mentioned.  Pray do not talk of that odious\index{odious} man.  I do think it is
the hardest thing in the world, that your estate should be entailed\index{entailed}
away from your own children; and I am sure, if I had been you, I
should have tried long ago to do something or other about it."

Jane and Elizabeth tried to explain to her the nature of an entail\index{entail}.
They had often attempted to do it before, but it was a subject\index{subject}
on which Mrs. Bennet was beyond the reach of reason, and she
continued to \textbf{rail}\index{rail}\stepcounter{footnote}\footnotetext{
\textbf{rail} \emph{n.}  (RAYL)
 to complain bitterly; to criticize severely.} bitterly against the cruelty of settling an
estate away from a family of five daughters, in favour of a man
whom nobody cared anything about.

"It certainly is a most \textbf{iniquitous}\index{iniquitous}\stepcounter{footnote}\footnotetext{
\textbf{iniquitous} \emph{adj.}  (ih-NIK-wih-tuhs)
wicked; violating the rights of others; nefarious; criminal.} affair," said Mr. Bennet,
"and nothing can clear Mr. Collins from the guilt of inheriting
Longbourn.  But if you will listen to his letter, you may perhaps
be a little softened by his manner of expressing himself."

"No, that I am sure I shall not; and I think it is very impertinent\index{impertinent}
of him to write to you at all, and very hypocritical\index{hypocritical}.  I hate such
false friends.  Why could he not keep on quarreling with you, as
his father did before him?"

"Why, indeed; he does seem to have had some \textbf{filial}\index{filial}\stepcounter{footnote}\footnotetext{
\textbf{filial} \emph{adj.}  (FIH-lee-uhl)
 relating to a son or daughtor.} \textbf{scruples}\index{scruples}\stepcounter{footnote}\footnotetext{
\textbf{scruple} \emph{n.} \textbf{-ed}  (SKROO-puhl)
 a moral, principal, or ethic that guides a person's actions.} on
that head, as you will hear."

"Hunsford, near Westerham, Kent,
15th October.

"Dear Sir,--

"The disagreement subsisting\index{subsisting} between yourself and my late
honoured father always gave me much uneasiness, and since I
have had the misfortune to lose him, I have frequently wished
to heal the \textbf{breach}\index{breach}\stepcounter{footnote}\footnotetext{
\textbf{breach} \emph{n.}  (BREECH)
an opening (especially a gap in a dike or fortification).}; but for some time I was kept back by my own
doubts, fearing lest it might seem disrespectful to his memory
for me to be on good terms with anyone with whom it had always
pleased him to be at variance.--'There, Mrs. Bennet.'--My
mind, however, is now made up on the subject\index{subject}, for having
received \textbf{ordination}\index{ordination}\stepcounter{footnote}\footnotetext{
\textbf{ordination} \emph{n.}  (or-dih-NAY-shun)
formal acceptance into the clergy.} at Easter, I have been so fortunate as to
be distinguished by the \textbf{patronage}\index{patronage}\stepcounter{footnote}\footnotetext{
\textbf{patronage} \emph{n.}  (PAT-ruh-nehdge)
customers seen as collective body.} of the Right Honourable Lady
Catherine de Bourgh, widow of Sir Lewis de Bourgh, whose
\textbf{bounty}\index{bounty}\stepcounter{footnote}\footnotetext{
\textbf{bounty} \emph{n.}  (BOUN-tee)
} and \textbf{beneficence}\index{beneficence}\stepcounter{footnote}\footnotetext{
\textbf{beneficence} \emph{n.}  (beh-NIF-fih-sense)
quality of being kind, good, or charitable.} has preferred me to the valuable \textbf{rectory}\index{rectory}\stepcounter{footnote}\footnotetext{
\textbf{rectory} \emph{n.}  (REHK-ter-ree)
the official residence of a parson provided by a church.}
of this \textbf{parish}\index{parish}\stepcounter{footnote}\footnotetext{
\textbf{parish} \emph{n.}  (PA-rish)
 a local church community.}, where it shall be my \textbf{earnest}\index{earnest}\stepcounter{footnote}\footnotetext{
\textbf{earnest} \emph{adj.} \textbf{-ly}  (ER-nehst)
 sincerity or firm belief in the validity of one's opinions.} endeavour\index{endeavour} to \textbf{demean}\index{demean}\stepcounter{footnote}\footnotetext{
\textbf{demean} \emph{tr.v.}  (dih-MEEN)
 to debase in character; to disgrace or dishonor.}
myself with grateful respect towards her ladyship, and be ever
ready to perform those rites and ceremonies which are instituted
by the Church of England.  As a clergyman,  moreover, I feel it
my duty to promote and establish the blessing of peace in all
families within in the reach of my influence; and on these
grounds I flatter myself that my present \textbf{overtures}\index{overtures}\stepcounter{footnote}\footnotetext{
\textbf{overture} \emph{n.}  (OH-ver-choor)
A musical introduction to a larger work, such as an opera.} are highly
\textbf{commendable}\index{commendable}\stepcounter{footnote}\footnotetext{
\textbf{commend} \emph{tr.v.} \textbf{-ed}  (kuh-MEND)
 to praise as admirable; to reccomend.}, and that the circumstance of my being next in the
entail\index{entail} of Longbourn estate will be kindly overlooked on your
side, and not lead you to reject the offered olive-branch.
I cannot be otherwise than concerned at being the means of
injuring your amiable\index{amiable} daughters, and beg leave to apologise for
it, as well as to assure you of my readiness to make them every
possible amends\index{amends}--but of this hereafter.  If you should have
no objection to receive me into your house, I propose myself
the satisfaction of waiting on you and your family, Monday,
November 18th, by four o'clock, and shall probably trespass on
your hospitality till the Saturday se'ennight following, which I
can do without any inconvenience, as Lady Catherine is far from
objecting to my occasional absence on a Sunday, provided that
some other clergyman is engaged to do the duty of the day.--I
remain, dear sir, with respectful compliments to your lady and
daughters, your well-wisher and friend,

"\emph{william} \emph{collins}"

"At four o'clock, therefore, we may expect this peace-making
gentleman," said Mr. Bennet, as he folded up the letter.  "He
seems to be a most conscientious\index{conscientious} and polite young man, upon
my word, and I doubt not will prove a valuable acquaintance,
especially if Lady Catherine should be so indulgent as to let\index{let}
him come to us again."

"There is some sense in what he says about the girls, however,
and if he is disposed\index{disposed} to make them any amends\index{amends}, I shall not be
the person to discourage him."

"Though it is difficult," said Jane, "to guess in what way he can
mean to make us the \textbf{atonement}\index{atonement}\stepcounter{footnote}\footnotetext{
\textbf{} \emph{}  ()
} he thinks our due, the wish is
certainly to his credit."

Elizabeth was chiefly struck by his extraordinary deference\index{deference} for
Lady Catherine, and his kind intention of christening, marrying,
and burying his parishioners whenever it were required.

"He must be an \textbf{oddity}\index{oddity}\stepcounter{footnote}\footnotetext{
\textbf{oddity} \emph{n.}  (AHD-dih-tee)
something unusual, perhaps worthy of collecting.}, I think," said she.  "I cannot make him
out.--There is something very \textbf{pompous}\index{pompous}\stepcounter{footnote}\footnotetext{
\textbf{pompous} \emph{adj.}  (PAHM-puhs)
puffed up with vanity or exaggerated dignity.} in his style.--And what
can he mean by apologising for being next in the entail\index{entail}?--We
cannot suppose he would help it if he could.--Could he be a
sensible man, sir?"

"No, my dear, I think not.  I have great hopes of finding him
quite the reverse.  There is a mixture of \textbf{servility}\index{servility}\stepcounter{footnote}\footnotetext{
\textbf{servile} \emph{adj.}  (SER-vyl)
 abject or slavish submissiveness.} and
self-importance in his letter, which promises well.  I am
impatient to see him."

"In point of \textbf{composition}\index{composition}\stepcounter{footnote}\footnotetext{
\textbf{composition} \emph{n.}  (kahm-puh-ZIH-shun)
}," said Mary, "the letter does not seem
defective.  The idea of the olive-branch perhaps is not wholly\index{wholly}
new, yet I think it is well expressed."

To Catherine and Lydia, neither the letter nor its writer were
in any degree interesting.  It was next to impossible that their
cousin should come in a scarlet coat, and it was now some
weeks since they had received pleasure from the society of a
man in any other colour.  As for their mother, Mr. Collins's
letter had done away much of her ill-will, and she was preparing
to see him with a degree of composure\index{composure} which astonished her
husband and daughters.

Mr. Collins was punctual\index{punctual} to his time, and was received with
great politeness by the whole family.  Mr. Bennet indeed said
little; but the ladies were ready enough to talk, and Mr. Collins
seemed neither in need of encouragement, nor inclined to be
silent himself.  He was a tall, heavy-looking young man of
five-and-twenty.  His air was grave\index{grave} and \textbf{stately}\index{stately}\stepcounter{footnote}\footnotetext{
\textbf{stately} \emph{adj.}  (STAYT-lee)
 refined or imposing in manner or appearance; dignified.}, and his
manners were very formal.  He had not been long seated before
he complimented Mrs. Bennet on having so fine a family of
daughters; said he had heard much of their beauty, but that in
this instance fame had fallen short of the truth; and added,
that he did not doubt her seeing them all in due time disposed\index{disposed}
of in marriage.  This gallantry\index{gallantry} was not much to the taste of
some of his hearers; but Mrs. Bennet, who quarreled with no
compliments, answered most readily\index{readily}.

"You are very kind, I am sure; and I wish with all my heart it
may prove so, for else they will be \textbf{destitute}\index{destitute}\stepcounter{footnote}\footnotetext{
\textbf{destitute} \emph{adj.}  (DEH-stih-toot)
poor enough to need help from others.} enough.  Things are
settled so oddly."

"You allude\index{allude}, perhaps, to the entail\index{entail} of this estate."

"Ah! sir, I do indeed.  It is a \textbf{grievous}\index{grievous}\stepcounter{footnote}\footnotetext{
\textbf{grievous} \emph{adj.}  (GREE-vuhs)
shockingly brutal or cruel; causing great anguish.} affair to my poor girls,
you must confess.  Not that I mean to find fault with \emph{you}, for
such things I know are all chance in this world.  There is no
knowing how estates will go when once they come to be entailed\index{entailed}."

"I am very sensible, madam, of the hardship to my fair cousins,
and could say much on the subject\index{subject}, but that I am cautious of
appearing forward\index{forward} and \textbf{precipitate}\index{precipitate}\stepcounter{footnote}\footnotetext{
\textbf{precipitate} \emph{v.}  (prih-SIH-pih-tayt)
to bring about abruptly.}.  But I can assure the young
ladies that I come prepared to admire them.  At present I will
not say more; but, perhaps, when we are better acquainted--"

He was interrupted by a \textbf{summons}\index{summons}\stepcounter{footnote}\footnotetext{
\textbf{summons} \emph{n.}  (SUHM-muhnz)
a call in an official matter, like to appear in court.}\index{summons\index{summons}} to dinner; and the girls smiled
on each other.  They were not the only objects of Mr. Collins's
admiration.  The hall, the dining-room, and all its furniture,
were examined and praised; and his commendation\index{commendation} of everything
would have touched Mrs. Bennet's heart, but for the mortifying\index{mortifying}
supposition\index{supposition} of his viewing it all as his own future property.
The dinner too in its turn was highly admired; and he begged to
know to which of his fair cousins the excellency of its cooking
was owing.  But he was set right there by Mrs. Bennet, who
assured him with some \textbf{asperity}\index{asperity}\stepcounter{footnote}\footnotetext{
\textbf{asperity} \emph{n.}  (as-SPEHR-ih-tee)
} that they were very well able to
keep a good cook, and that her daughters had nothing to do in
the kitchen.  He begged pardon for having displeased her.  In a
softened tone she declared herself not at all offended; but he
continued to apologise for about a quarter of an hour.



\chapter{14}


During dinner, Mr. Bennet scarcely spoke at all; but when the
servants were withdrawn, he thought it time to have some
conversation with his guest, and therefore started a subject\index{subject} in
which he expected him to shine, by observing that he seemed
very fortunate in his \textbf{patroness}\index{patroness}\stepcounter{footnote}\footnotetext{
\textbf{patroness} \emph{n.}  (PAY-truh-nehss)
a woman who is a sponsor or benefactor.}.  Lady Catherine de Bourgh's
attention to his wishes, and consideration for his comfort,
appeared very remarkable.  Mr. Bennet could not have chosen
better.  Mr. Collins was \textbf{eloquent}\index{eloquent}\stepcounter{footnote}\footnotetext{
\textbf{eloquent} \emph{adj.}  (EH-loh-kwehnt)
expressing oneself readily, clearly, and effectively.} in her praise.  The subject\index{subject}
elevated him to more than usual \textbf{solemnity}\index{solemnity}\stepcounter{footnote}\footnotetext{
\textbf{solemnity} \emph{n.}  (suh-LEHM-niH-tee)
the trait of dignified seriousness.} of manner, and with a
most important \textbf{aspect}\index{aspect}\stepcounter{footnote}\footnotetext{
\textbf{aspect} \emph{n.}  (ASS-pehkt)
} he protested that "he had never in his life
witnessed such behaviour in a person of rank--such \textbf{affability}\index{affability}\stepcounter{footnote}\footnotetext{
\textbf{affable} \emph{adj.}  (AF-fuh-bl)
 friendly and approachable; easy to talk to.}
and \textbf{condescension}\index{condescension}\stepcounter{footnote}\footnotetext{
\textbf{condescension} \emph{n.}  (kahn-dih-SEN-shun)
a disrespecetful communication that patronizes the recipient.}, as he had himself experienced from Lady
Catherine.  She had been graciously pleased to approve of both
of the \textbf{discourses}\index{discourses}\stepcounter{footnote}\footnotetext{
\textbf{} \emph{}  ()
} which he had already had the honour of
preaching before her.  She had also asked him twice to dine at
Rosings, and had sent for him only the Saturday before, to make
up her pool of \textbf{quadrille}\index{quadrille}\stepcounter{footnote}\footnotetext{
\textbf{quadrille} \emph{n.}  (kwah-DRIHL)
} in the evening.  Lady Catherine was
reckoned proud by many people he knew, but \emph{he} had never
seen anything but affability in her.  She had always spoken to
him as she would to any other gentleman; she made not the
smallest objection to his joining in the society of the
neighbourhood nor to his leaving the parish\index{parish} occasionally for a
week or two, to visit his relations.  She had even condescended\index{condescended}
to advise him to marry as soon as he could, provided he chose
with discretion\index{discretion}; and had once paid him a visit in his humble
parsonage, where she had perfectly approved all the alterations
he had been making, and had even \textbf{vouchsafed}\index{vouchsafed}\stepcounter{footnote}\footnotetext{
\textbf{vouchsafe} \emph{tr.v.} \textbf{-ed}  (VOUCH-sayfe)
to grant or agree to in a condescending manner.} to suggest some
herself--some shelves in the closet upstairs."

"That is all very proper and civil, I am sure," said Mrs. Bennet,
"and I dare say she is a very agreeable woman.  It is a pity that
great ladies in general are not more like her.  Does she live near
you, sir?"

"The garden in which stands my humble \textbf{abode}\index{abode}\stepcounter{footnote}\footnotetext{
\textbf{abode} \emph{v.}  (uh-BOAD)
 a person's legal place of residence.} is separated only
by a lane from Rosings Park, her ladyship's residence."

"I think you said she was a widow, sir?  Has she any family?"

"She has only one daughter, the heiress of Rosings, and of very
extensive property."

"Ah!" said Mrs. Bennet, shaking her head, "then she is better off
than many girls.  And what sort of young lady is she?  Is she
handsome?"

"She is a most charming young lady indeed.  Lady Catherine
herself says that, in point of true beauty, Miss de Bourgh is far
superior\index{superior} to the handsomest of her sex, because there is that in
her features which marks the young lady of distinguished birth.
She is unfortunately of a sickly \textbf{constitution}\index{constitution}\stepcounter{footnote}\footnotetext{
\textbf{} \emph{}  ()
}, which has prevented
her from making that progress in many accomplishments which
she could not have otherwise failed of, as I am informed by the
lady who superintended\index{superintended} her education, and who still resides with
them.  But she is perfectly amiable\index{amiable}, and often condescends\index{condescends} to
drive by my humble abode\index{abode} in her little \textbf{phaeton}\index{phaeton}\stepcounter{footnote}\footnotetext{
\textbf{phaeton} \emph{n.}  (FEE-tuhn)
 a four-wheeled open carriage, usually drawn by two horses.} and ponies."

"Has she been presented?  I do not remember her name among
the ladies at court."

"Her \textbf{indifferent}\index{indifferent}\stepcounter{footnote}\footnotetext{
\textbf{indifferent} \emph{adj.}  (ihn-DIF-fer-uhnt)
(one with) a lack of interest; apathetic.} state of health unhappily prevents her being
in town; and by that means, as I told Lady Catherine one day,
has deprived the British court of its brightest ornaments.
Her ladyship seemed pleased with the idea; and you may imagine
that I am happy on every occasion to offer those little
delicate compliments which are always acceptable to ladies.
I have more than once observed to Lady Catherine, that her
charming daughter seemed born to be a duchess, and that the
most elevated rank, instead of giving her consequence\index{consequence}, would
be adorned by her.  These are the kind of little things which
please her ladyship, and it is a sort of attention which I
conceive myself \textbf{peculiarly}\index{peculiarly}\stepcounter{footnote}\footnotetext{
\textbf{peculiar} \emph{adj.} \textbf{-ly}  (puh-KYOOL-yer)
strange, eccentric, or odd.} bound to pay."

"You judge very properly," said Mr. Bennet, "and it is happy for
you that you possess the talent of flattering with delicacy\index{delicacy}.  May I
ask whether these pleasing attentions proceed from the impulse
of the moment, or are the result of previous study?"

"They arise chiefly from what is passing at the time, and though
I sometimes amuse myself with suggesting and arranging such
little elegant compliments as may be adapted to ordinary occasions,
I always wish to give them as unstudied an air as possible."

Mr. Bennet's expectations were fully answered.  His cousin was
as absurd as he had hoped, and he listened to him with the
keenest enjoyment, maintaining at the same time the most
\textbf{resolute}\index{resolute}\stepcounter{footnote}\footnotetext{
\textbf{resolute} \emph{adj.} \textbf{-ly}  (REH-zoh-loot)
firm in purpose or belief; determined.} composure\index{composure} of countenance\index{countenance}, and, except in an occasional
glance at Elizabeth, requiring no partner in his pleasure.

By tea-time, however, the dose had been enough, and Mr.
Bennet was glad to take his guest into the drawing-room again,
and, when tea was over, glad to invite him to read aloud to the
ladies.  Mr. Collins readily\index{readily} assented\index{assented}, and a book was produced;
but, on beholding it (for everything announced it to be from a
circulating library), he started back, and begging pardon,
protested that he never read novels.  Kitty stared at him, and
Lydia exclaimed.  Other books were produced, and after some
\textbf{deliberation}\index{deliberation}\stepcounter{footnote}\footnotetext{
\textbf{deliberation} \emph{n.}  (dih-lib-uh-RAY-shun)
careful consideration.} he chose Fordyce's Sermons.  Lydia gaped as he
opened the volume, and before he had, with very \textbf{monotonous}\index{monotonous}\stepcounter{footnote}\footnotetext{
\textbf{monotonous} \emph{adj.}  (muh-NAHT-nuhs)
tediously repetitious or lacking in variety; an unvarying tone or sound.}
solemnity\index{solemnity}, read three pages, she interrupted him with:

"Do you know, mamma, that my uncle Phillips talks of turning
away Richard; and if he does, Colonel Forster will hire him.  My
aunt told me so herself on Saturday.  I shall walk to Meryton
to-morrow to hear more about it, and to ask when Mr. Denny
comes back from town."

Lydia was bid by her two eldest sisters to hold her tongue; but
Mr. Collins, much offended, laid aside his book, and said:

"I have often observed how little young ladies are interested by
books of a serious stamp, though written solely\index{solely} for their benefit.
It amazes me, I confess; for, certainly, there can be nothing so
advantageous to them as instruction.  But I will no longer
\textbf{importune}\index{importune}\stepcounter{footnote}\footnotetext{
\textbf{importune} \emph{v.} \textbf{-ed}  (im-por-TOON)
to beg persistently and urgently.} my young cousin."

Then turning to Mr. Bennet, he offered himself as his \textbf{antagonist}\index{antagonist}\stepcounter{footnote}\footnotetext{
\textbf{antagonist} \emph{n.}  (an-TAG-guh-nist)
someone who offers opposition; an adversary.}
at backgammon.  Mr. Bennet accepted the challenge, observing
that he acted very wisely in leaving the girls to their own trifling\index{trifling}
amusements.  Mrs. Bennet and her daughters apologised most
civilly for Lydia's interruption, and promised that it should not
occur again, if he would resume his book; but Mr. Collins, after
assuring them that he bore his young cousin no ill-will, and
should never resent her behaviour as any affront\index{affront}, seated himself
at another table with Mr. Bennet, and prepared for backgammon.



\chapter{15}


Mr. Collins was not a sensible man, and the deficiency of nature
had been but little assisted by education or society; the greatest
part of his life having been spent under the guidance of an
\textbf{illiterate}\index{illiterate}\stepcounter{footnote}\footnotetext{
\textbf{illiterate} \emph{adj.}  (ihl-LIH-ter-ruht)
not able to read or write .} and \textbf{miserly}\index{miserly}\stepcounter{footnote}\footnotetext{
\textbf{miserly} \emph{adj.}  (MY-zer-lee)
avaricious or lacking generosity.} father; and though he belonged to one of
the universities, he had merely kept the necessary terms, without
forming at it any useful acquaintance.  The \textbf{subjection}\index{subjection}\stepcounter{footnote}\footnotetext{
\textbf{subject} \emph{adj.} \textbf{-ion}  (suhb-JEHKT)
} in which
his father had brought him up had given him originally great
humility\index{humility} of manner; but it was now a good deal counteracted by
the self-conceit\index{conceit} of a weak head, living in retirement, and the
consequential feelings of early and unexpected prosperity.  A
fortunate chance had recommended him to Lady Catherine de
Bourgh when the living of Hunsford was vacant; and the respect
which he felt for her high rank, and his \textbf{veneration}\index{veneration}\stepcounter{footnote}\footnotetext{
\textbf{veneration} \emph{n.}  (veh-ner-RAY-shun)
a profound awe inspired by a deity.} for her as his
patroness\index{patroness}, mingling with a very good opinion of himself, of his
authority as a clergyman, and his right as a \textbf{rector}\index{rector}\stepcounter{footnote}\footnotetext{
\textbf{rector} \emph{n.}  (REHK-ter)
 A clergyman presiding over a parish in the Protestant Episcopal Church.}, made him
altogether a mixture of pride and \textbf{obsequiousness}\index{obsequiousness}\stepcounter{footnote}\footnotetext{
\textbf{obsequious} \emph{adj.} \textbf{-ness}  (ahb-SEE-kwih-uhs)
showing servile compliance; fawning.},
self-importance and humility\index{humility}.

Having now a good house and a very sufficient income, he
intended to marry; and in seeking a \textbf{reconciliation}\index{reconciliation}\stepcounter{footnote}\footnotetext{
\textbf{reconciliation} \emph{n.}  (reh-kuhn-sih-lih-AY-shun)
reestablishing friendly relations.} with the
Longbourn family he had a wife in view, as he meant to choose
one of the daughters, if he found them as handsome and amiable\index{amiable}
as they were represented by common report.  This was his plan
of amends\index{amends}--of atonement--for inheriting their father's estate;
and he thought it an excellent one, full of eligibility and
suitableness, and excessively generous and \textbf{disinterested}\index{disinterested}\stepcounter{footnote}\footnotetext{
\textbf{disinterested} \emph{adj.}  (dihs-IHN-tris-tehd)
} on his
own part.

His plan did not vary on seeing them.  Miss Bennet's lovely face
confirmed his views, and established all his strictest \textbf{notions}\index{notions}\stepcounter{footnote}\footnotetext{
\textbf{} \emph{}  ()
} of
what was due to \textbf{seniority}\index{seniority}\stepcounter{footnote}\footnotetext{
\textbf{seniority} \emph{n.}  (seen-YAW-rih-tee)
higher rank than others by reason of longer service.}; and for the first evening \emph{she} was his
settled choice.  The next morning, however, made an alteration;
for in a quarter of an hour's tete-a-tete with Mrs. Bennet before
breakfast, a conversation beginning with his parsonage-house,
and leading naturally to the \textbf{avowal}\index{avowal}\stepcounter{footnote}\footnotetext{
\textbf{avowal} \emph{n.}  (uh-VOW-uhl)
 a sincere acknowledgment or assertion.} of his hopes, that a mistress
might be found for it at Longbourn, produced from her, amid
very \textbf{complaisant}\index{complaisant}\stepcounter{footnote}\footnotetext{
\textbf{complaisant} \emph{adj.}  (kuhm-PLAY-zant)
} smiles and general encouragement, a caution
against the very Jane he had fixed on.  "As to her \emph{younger}
daughters, she could not take upon her to say--she could not
positively answer--but she did not \emph{know} of any \textbf{prepossession}\index{prepossession}\stepcounter{footnote}\footnotetext{
\textbf{prepossession} \emph{n.}  (pree-puh-ZEH-shun)
an opinion formed without adequate evidence.};
her \emph{eldest} daughter, she must just mention--she felt it
\textbf{incumbent}\index{incumbent}\stepcounter{footnote}\footnotetext{
\textbf{incumbent} \emph{adj.}  (ihn-KUHM-behnt)
} on her to hint, was likely to be very soon engaged."

Mr. Collins had only to change from Jane to Elizabeth--and it
was soon done--done while Mrs. Bennet was stirring the fire.
Elizabeth, equally next to Jane in birth and beauty, succeeded
her of course.

Mrs. Bennet treasured up the hint, and trusted that she might
soon have two daughters married; and the man whom she could
not bear to speak of the day before was now high in her good
graces.

Lydia's intention of walking to Meryton was not forgotten;
every sister except Mary agreed to go with her; and Mr. Collins
was to attend them, at the request of Mr. Bennet, who was most
anxious to get rid of him, and have his library to himself; for
thither Mr. Collins had followed him after breakfast; and there he
would continue, \textbf{nominally}\index{nominally}\stepcounter{footnote}\footnotetext{
\textbf{nominal} \emph{adj.} \textbf{-ly}  (NAH-mih-nuhl)
 in name only.} engaged with one of the largest \textbf{folios}\index{folios}\stepcounter{footnote}\footnotetext{
\textbf{folio} \emph{n.}  (FOH-lee-oh)
}
in the collection, but really talking to Mr. Bennet, with little
\textbf{cessation}\index{cessation}\stepcounter{footnote}\footnotetext{
\textbf{cessation} \emph{n.}  (sehs-SAY-shun)
a stopping.}, of his house and garden at Hunsford.  Such doings
\textbf{discomposed}\index{discomposed}\stepcounter{footnote}\footnotetext{
\textbf{discompose} \emph{tr.v.} \textbf{-ed}  (dihs-kuhm-POAZ)
To disturb a person's composure; to ruffle or disturb.} Mr. Bennet exceedingly.  In his library he had been
always sure of leisure and \textbf{tranquillity}\index{tranquillity}\stepcounter{footnote}\footnotetext{
\textbf{tranquillity} \emph{n.}  (tran-KWIHL-lih-tee)
a serene state of peace and quiet.}; and though prepared, as
he told Elizabeth, to meet with folly\index{folly} and conceit\index{conceit} in every other
room of the house, he was used to be free from them there; his
civility\index{civility}, therefore, was most prompt in inviting Mr. Collins to
join his daughters in their walk; and Mr. Collins, being in fact
much better fitted for a walker than a reader, was extremely
pleased to close his large book, and go.

In pompous\index{pompous} nothings on his side, and civil assents\index{assents} on that of
his cousins, their time passed till they entered Meryton.  The
attention of the younger ones was then no longer to be gained by
him.  Their eyes were immediately wandering up in the street in
quest of the officers, and nothing less than a very smart bonnet
indeed, or a really new muslin in a shop window, could recall
them.

But the attention of every lady was soon caught by a young man,
whom they had never seen before, of most gentlemanlike
appearance, walking with another officer on the other side of the
way.  The officer was the very Mr. Denny concerning whose
return from London Lydia came to inquire, and he bowed as
they passed.  All were struck with the stranger's air, all
wondered who he could be; and Kitty and Lydia, determined if
possible to find out, led the way across the street, under \textbf{pretense}\index{pretense}\stepcounter{footnote}\footnotetext{
\textbf{pretense} \emph{n.}  (PREE-tehnse)
pretending with intention to deceive.}
of wanting something in an opposite shop, and fortunately had
just gained the pavement when the two gentlemen, turning back,
had reached the same spot.  Mr. Denny addressed\index{addressed} them directly,
and \textbf{entreated}\index{entreated}\stepcounter{footnote}\footnotetext{
\textbf{entreat} \emph{v.} \textbf{-ed}  (ehn-TREET)
 to make a forthright, sincere request of.} permission to introduce his friend, Mr. Wickham,
who had returned with him the day before from town, and he
was happy to say had accepted a commission in their \textbf{corps}\index{corps}\stepcounter{footnote}\footnotetext{
\textbf{corps} \emph{n.}  (KOR)
 an army unit usually consisting of two or more divisions.}.
This was exactly as it should be; for the young man wanted only
regimentals to make him completely charming.  His appearance
was greatly in his favour; he had all the best part of beauty, a
fine countenance\index{countenance}, a good figure, and very pleasing address\index{address}.  The
introduction was followed up on his side by a happy readiness of
conversation--a readiness at the same time perfectly correct and
\textbf{unassuming}\index{unassuming}\stepcounter{footnote}\footnotetext{
\textbf{unassuming} \emph{adj.}  (uhn-uh-SOO-ming)
not arrogant or presuming; modest in conduct.}; and the whole party were still standing and talking
together very agreeably, when the sound of horses drew their
notice, and Darcy and Bingley were seen riding down the street.
On distinguishing the ladies of the group, the two gentlemen
came directly towards them, and began the usual civilities.
Bingley was the principal\index{principal} spokesman, and Miss Bennet the
principal\index{principal} object.  He was then, he said, on his way to Longbourn
on purpose to inquire after her.  Mr. Darcy \textbf{corroborated}\index{corroborated}\stepcounter{footnote}\footnotetext{
\textbf{corroborate} \emph{tr.v.}  (ker-RAH-ber-ayt)
to support or establish by evidence or proof.} it with
a bow, and was beginning to determine not to fix his eyes on
Elizabeth, when they were suddenly arrested by the sight of the
stranger, and Elizabeth happening to see the countenance\index{countenance} of
both as they looked at each other, was all astonishment at the
effect of the meeting.  Both changed colour, one looked white,
the other red.  Mr. Wickham, after a few moments, touched his
hat--a salutation which Mr. Darcy just deigned\index{deigned} to return.  What
could be the meaning of it?  It was impossible to imagine; it was
impossible not to long to know.

In another minute\index{minute}, Mr. Bingley, but without seeming to have
noticed what passed, took leave and rode on with his friend.

Mr. Denny and Mr. Wickham walked with the young ladies to
the door of Mr. Phillip's house, and then made their bows, in
spite of Miss Lydia's pressing entreaties that they should come
in, and even in spite of Mrs. Phillips's throwing up the parlour
window and loudly seconding the invitation.

Mrs. Phillips was always glad to see her nieces; and the two
eldest, from their recent absence, were particularly welcome, and
she was eagerly expressing her surprise at their sudden return
home, which, as their own carriage had not fetched them, she
should have known nothing about, if she had not happened to
see Mr. Jones's shop-boy in the street, who had told her that
they were not to send any more draughts\index{draughts} to Netherfield because
the Miss Bennets were come away, when her civility\index{civility} was
claimed towards Mr. Collins by Jane's introduction of him.  She
received him with her very best politeness, which he returned
with as much more, apologising for his intrusion, without any
previous acquaintance with her, which he could not help
flattering himself, however, might be justified by his relationship
to the young ladies who introduced him to her notice.  Mrs.
Phillips was quite \textbf{awed}\index{awed}\stepcounter{footnote}\footnotetext{
\textbf{awe} \emph{n.} \textbf{-ed}  (AW)
 a state inspired by fearful wonderment or reverence.} by such an excess of good breeding; but
her \textbf{contemplation}\index{contemplation}\stepcounter{footnote}\footnotetext{
\textbf{contemplation} \emph{n.}  (kahn-tehm-LAY-shun)
a deep thinking session about a particular topic; a long consideration.} of one stranger was soon put to an end by
exclamations and inquiries about the other; of whom, however,
she could only tell her nieces what they already knew, that Mr.
Denny had brought him from London, and that he was to have a
lieutenant's commission in the ----shire.  She had been watching
him the last hour, she said, as he walked up and down the street,
and had Mr. Wickham appeared, Kitty and Lydia would certainly
have continued the occupation, but unluckily no one passed
windows now except a few of the officers, who, in comparison
with the stranger, were become "stupid, disagreeable fellows."
Some of them were to dine with the Phillipses the next day, and
their aunt promised to make her husband call on Mr. Wickham,
and give him an invitation also, if the family from Longbourn
would come in the evening.  This was agreed to, and Mrs.
Phillips protested that they would have a nice comfortable noisy
game of lottery tickets, and a little bit of hot supper afterwards.
The prospect\index{prospect} of such delights was very cheering, and they parted
in mutual good spirits.  Mr. Collins repeated his apologies in
quitting the room, and was assured with unwearying civility\index{civility} that
they were perfectly needless.

As they walked home, Elizabeth related to Jane what she had
seen pass between the two gentlemen; but though Jane would
have defended either or both, had they appeared to be in the
wrong, she could no more explain such behaviour than her sister.

Mr. Collins on his return highly gratified Mrs. Bennet by
admiring Mrs. Phillips's manners and politeness.  He protested
that, except Lady Catherine and her daughter, he had never seen
a more elegant woman; for she had not only received him with
the utmost civility\index{civility}, but even pointedly included him in her
invitation for the next evening, although utterly unknown to her
before.  Something, he supposed, might be attributed to his
connection with them, but yet he had never met with so much
attention in the whole course of his life.



\chapter{16}


As no objection was made to the young people's engagement
with their aunt, and all Mr. Collins's scruples\index{scruples} of leaving Mr.
and Mrs. Bennet for a single evening during his visit were most
\textbf{steadily}\index{steadily}\stepcounter{footnote}\footnotetext{
\textbf{steady} \emph{adj.}  (STEH-dee)
fixed or unmoving in place or nature; consistent or reliable.} resisted, the coach conveyed him and his five cousins
at a suitable hour to Meryton; and the girls had the pleasure of
hearing, as they entered the drawing-room, that Mr. Wickham
had accepted their uncle's invitation, and was then in the house.

When this information was given, and they had all taken their
seats, Mr. Collins was at leisure to look around him and admire,
and he was so much struck with the size and furniture of the
apartment\index{apartment}, that he declared he might almost have supposed
himself in the small summer breakfast parlour at Rosings; a
comparison that did not at first convey much gratification; but
when Mrs. Phillips understood from him what Rosings was, and
who was its \textbf{proprietor}\index{proprietor}\stepcounter{footnote}\footnotetext{
\textbf{proprietor} \emph{n.}  (proh-PRY-eht-tor)
someone who owns (is legal possessor of) a business.}--when she had listened to the description
of only one of Lady Catherine's drawing-rooms, and found that
the chimney-piece alone had cost eight hundred pounds, she felt
all the force of the compliment, and would hardly have resented
a comparison with the housekeeper's room.

In describing to her all the \textbf{grandeur}\index{grandeur}\stepcounter{footnote}\footnotetext{
\textbf{grandeur} \emph{n.}  (GRAN-dyoor)
the quality of being magnificent, splendid, or grand.} of Lady Catherine and her
mansion, with occasional \textbf{digressions}\index{digressions}\stepcounter{footnote}\footnotetext{
\textbf{digression} \emph{n.}  (dy-GREH-shun)
an act of wandering from the main path or subject matter.} in praise of his own humble
abode\index{abode}, and the improvements it was receiving, he was happily
employed until the gentlemen joined them; and he found in
Mrs. Phillips a very attentive listener, whose opinion of
his consequence\index{consequence} increased with what she heard, and who was
resolving to \textbf{retail}\index{retail}\stepcounter{footnote}\footnotetext{
\textbf{retail} \emph{n.}  (REE-tayl)
 the selling of goods directly to consumers, usually in small quantities.} it all among her neighbours as soon as she
could.  To the girls, who could not listen to their cousin,
and who had nothing to do but to wish for an instrument, and
examine their own indifferent\index{indifferent} imitations of china on the
mantelpiece, the interval of waiting appeared very long.  It was
over at last, however.  The gentlemen did approach, and when
Mr. Wickham walked into the room, Elizabeth felt that she had
neither been seeing him before, nor thinking of him since, with
the smallest degree of unreasonable admiration.  The officers of
the ----shire were in general a very creditable, gentlemanlike
set, and the best of them were of the present party; but Mr.
Wickham was as far beyond them all in person, countenance\index{countenance}, air,
and walk, as \emph{they} were superior\index{superior} to the broad-faced, stuffy
uncle Phillips, breathing port wine, who followed them into
the room.

Mr. Wickham was the happy man towards whom almost every
female eye was turned, and Elizabeth was the happy woman by
whom he finally seated himself; and the agreeable manner in
which he immediately fell into conversation, though it was only
on its being a wet night, made her feel that the commonest,
dullest, most threadbare topic might be rendered\index{rendered} interesting by
the skill of the speaker.

With such rivals for the notice of the fair as Mr. Wickham and
the officers, Mr. Collins seemed to sink into insignificance; to
the young ladies he certainly was nothing; but he had still at
intervals a kind listener in Mrs. Phillips, and was by her
watchfulness, most abundantly supplied with coffee and muffin.
When the card-tables were placed, he had the opportunity of
obliging her in turn, by sitting down to \textbf{whist}\index{whist}\stepcounter{footnote}\footnotetext{
\textbf{whist} \emph{n.}  (HWIHST)
 a card game for four people, played in partnetships, using 52 cards.}.

"I know little of the game at present," said he, "but I shall be
glad to improve myself, for in my situation in life--" Mrs. Phillips
was very glad for his \textbf{compliance}\index{compliance}\stepcounter{footnote}\footnotetext{
\textbf{compliance} \emph{n.}  (kuhm-PLY-ans)
the act of submitting; acquiescence.}, but could not wait for his
reason.

Mr. Wickham did not play at whist\index{whist}, and with ready delight was he
received at the other table between Elizabeth and Lydia.  At first
there seemed danger of Lydia's engrossing\index{engrossing} him entirely, for she
was a most determined talker; but being likewise extremely fond
of lottery tickets, she soon grew too much interested in the
game, too eager in making bets and exclaiming after prizes to
have attention for anyone in particular.  Allowing for the
common demands of the game, Mr. Wickham was therefore at leisure
to talk to Elizabeth, and she was very willing to hear him,
though what she chiefly wished to hear she could not hope to be
told--the history of his acquaintance with Mr. Darcy.  She dared
not even mention that gentleman.  Her curiosity, however, was
unexpectedly relieved.  Mr. Wickham began the subject\index{subject} himself.
He inquired how far Netherfield was from Meryton; and, after
receiving her answer, asked in a hesitating manner how long
Mr. Darcy had been staying there.

"About a month," said Elizabeth; and then, unwilling to let\index{let} the
subject\index{subject} drop, added, "He is a man of very large property in
Derbyshire, I understand."

"Yes," replied Mr. Wickham; "his estate there is a noble one.
A clear ten thousand per \textbf{annum}\index{annum}\stepcounter{footnote}\footnotetext{
\textbf{} \emph{}  ()
}.  You could not have met with a
person more capable of giving you certain information on that
head than myself, for I have been connected with his family in
a particular manner from my infancy."

Elizabeth could not but look surprised.

"You may well be surprised, Miss Bennet, at such an \textbf{assertion}\index{assertion}\stepcounter{footnote}\footnotetext{
\textbf{assertion} \emph{n.}  (uh-SER-shun)
a declaration made as if no supporting evidence were necessary.},
after seeing, as you probably might, the very cold manner of our
meeting yesterday.  Are you much acquainted with Mr. Darcy?"

"As much as I ever wish to be," cried Elizabeth very warmly.
"I have spent four days in the same house with him, and I think
him very disagreeable."

"I have no right to give \emph{my} opinion," said Wickham, "as to his
being agreeable or otherwise.  I am not qualified to form one.  I
have known him too long and too well to be a fair judge.  It is
impossible for \emph{me} to be \textbf{impartial}\index{impartial}\stepcounter{footnote}\footnotetext{
\textbf{impartial} \emph{adj.} \textbf{-ity}  (ihm-PAR-shuhl)
unprejudiced; free from undue bias or preconceived opinions.}.  But I believe your opinion
of him would in general astonish--and perhaps you would not
express it quite so strongly anywhere else.  Here you are in your
own family."

"Upon my word, I say no more \emph{here} than I might say in any
house in the neighbourhood, except Netherfield.  He is not at all
liked in Hertfordshire.  Everybody is disgusted with his pride.
You will not find him more favourably spoken of by anyone."

"I cannot pretend to be sorry," said Wickham, after a short
interruption, "that he or that any man should not be estimated
beyond their deserts; but with \emph{him} I believe it does not often
happen.  The world is blinded by his fortune and consequence\index{consequence},
or frightened by his high and imposing manners, and sees him
only as he chooses to be seen."

"I should take him, even on \emph{my} slight acquaintance, to be an
ill-tempered man."  Wickham only shook his head.

"I wonder," said he, at the next opportunity of speaking,
"whether he is likely to be in this country much longer."

"I do not at all know; but I \emph{heard} nothing of his going away
when I was at Netherfield.  I hope your plans in favour of the
----shire will not be affected by his being in the neighbourhood."

"Oh! no--it is not for \emph{me} to be driven away by Mr. Darcy.  If
\emph{he} wishes to avoid seeing \emph{me}, he must go.  We are not on
friendly terms, and it always gives me pain to meet him, but I
have no reason for avoiding \emph{him} but what I might proclaim
before all the world, a sense of very great ill-usage, and most
painful regrets at his being what he is.  His father, Miss Bennet,
the late Mr. Darcy, was one of the best men that ever breathed,
and the truest friend I ever had; and I can never be in company
with this Mr. Darcy without being grieved\index{grieved} to the soul by a
thousand tender recollections\index{recollections}.  His behaviour to myself has been
\textbf{scandalous}\index{scandalous}\stepcounter{footnote}\footnotetext{
\textbf{scandalous} \emph{adj.}  (SKAN-duh-luhs)
giving offense to moral sensibilities and injurious to reputation.}; but I \textbf{verily}\index{verily}\stepcounter{footnote}\footnotetext{
\textbf{verily} \emph{adv.}  (VEH-rih-lee)
 (archaic) truly or confidentially.} believe I could forgive him anything and
everything, rather than his disappointing the hopes and
disgracing the memory of his father."

Elizabeth found the interest of the subject\index{subject} increase, and listened
with all her heart; but the delicacy\index{delicacy} of it prevented further inquiry.

Mr. Wickham began to speak on more general topics, Meryton,
the neighbourhood, the society, appearing highly pleased with all
that he had yet seen, and speaking of the latter with gentle but
very intelligible gallantry\index{gallantry}.

"It was the prospect\index{prospect} of constant society, and good society," he
added, "which was my chief inducement\index{inducement} to enter the ----shire.
I knew it to be a most respectable, agreeable corps\index{corps}, and my
friend Denny tempted me further by his account of their
present quarters, and the very great attentions and excellent
acquaintances Meryton had \textbf{procured}\index{procured}\stepcounter{footnote}\footnotetext{
\textbf{procure} \emph{v.} \textbf{-ed,-ing}  (proh-KYOOR)
to obtain or acquire.} them.  Society, I own, is
necessary to me.  I have been a disappointed man, and my spirits
will not bear \textbf{solitude}\index{solitude}\stepcounter{footnote}\footnotetext{
\textbf{solitude} \emph{n.}  (SAW-lih-tood)
the state of being alone or isolated from others.}.  I \emph{must} have employment and society.
A military life is not what I was intended for, but circumstances
have now made it eligible.  The church \emph{ought} to have been
my profession--I was brought up for the church, and I should at
this time have been in possession of a most valuable living, had
it pleased the gentleman we were speaking of just now."

"Indeed!"

"Yes--the late Mr. Darcy bequeathed me the next presentation
of the best living in his gift.  He was my godfather, and
excessively attached to me.  I cannot do justice to his kindness.
He meant to provide for me amply, and thought he had done it;
but when the living fell, it was given elsewhere."

"Good heavens!" cried Elizabeth; "but how could \emph{that} be?
How could his will be disregarded?  Why did you not seek legal
\textbf{redress}\index{redress}\stepcounter{footnote}\footnotetext{
\textbf{redress} \emph{tr.v.}  (rih-DREHSS)
to make reparations or amends for.}?"

"There was just such an informality in the terms of the \textbf{bequest}\index{bequest}\stepcounter{footnote}\footnotetext{
\textbf{bequest} \emph{n.}  (bih-KWEHST)
(legal) a gift of personal property by will.}
as to give me no hope from law.  A man of honour could not have
doubted the intention, but Mr. Darcy chose to doubt it--or to
treat it as a merely conditional recommendation, and to assert that
I had forfeited all claim to it by extravagance, \textbf{imprudence}\index{imprudence}\stepcounter{footnote}\footnotetext{
\textbf{imprudence} \emph{n.}  (ihm-PROO-dense)
a lack of caution in practical affairs}--in
short anything or nothing.  Certain it is, that the living became
vacant two years ago, exactly as I was of an age to hold it, and
that it was given to another man; and no less certain is it, that
I cannot accuse myself of having really done anything to deserve
to lose it.  I have a warm, unguarded temper, and I may have
spoken my opinion \emph{of} him, and \emph{to} him, too freely.  I can recall
nothing worse.  But the fact is, that we are very different sort
of men, and that he hates me."

"This is quite shocking!  He deserves to be publicly disgraced."

"Some time or other he \emph{will} be--but it shall not be by \emph{me}.
Till I can forget his father, I can never defy or expose \emph{him}."

Elizabeth honoured him for such feelings, and thought him
handsomer than ever as he expressed them.

"But what," said she, after a pause, "can have been his motive?
What can have \textbf{induced}\index{induced}\stepcounter{footnote}\footnotetext{
\textbf{induce} \emph{tr.v.} \textbf{-ed}  (ihn-DOOS)
to cause or bring about; to influence or persuade.} him to behave so cruelly?"

"A thorough, determined dislike of me--a dislike which I cannot
but attribute in some measure to jealousy.  Had the late Mr.
Darcy liked me less, his son might have \textbf{borne}\index{borne}\stepcounter{footnote}\footnotetext{
\textbf{borne} \emph{v.}  (BORN)
} with me better;
but his father's uncommon attachment to me irritated him, I
believe, very early in life.  He had not a temper to bear the sort of
competition in which we stood--the sort of preference which
was often given me."

"I had not thought Mr. Darcy so bad as this--though I have
never liked him.  I had not thought so very ill of him.  I had
supposed him to be despising his fellow-creatures in general, but
did not suspect him of descending to such \textbf{malicious}\index{malicious}\stepcounter{footnote}\footnotetext{
\textbf{malicious} \emph{adj.}  (muh-LIH-shuhs)
wishing evil to others; deliberately harmful.} revenge,
such injustice\index{injustice}, such \textbf{inhumanity}\index{inhumanity}\stepcounter{footnote}\footnotetext{
\textbf{inhumanity} \emph{n.}  (ihn-hyoo-MAN-nih-tee)
an act of atrocious cruelty.} as this."

After a few minutes\index{minutes}' reflection, however, she continued,
"I \emph{do} remember his boasting one day, at Netherfield, of the
implacability of his resentments\index{resentments}, of his having an unforgiving
temper.  His disposition\index{disposition} must be dreadful."

"I will not trust myself on the subject\index{subject}," replied Wickham; "I
can hardly be just to him."

Elizabeth was again deep in thought, and after a time exclaimed,
"To treat in such a manner the \textbf{godson}\index{godson}\stepcounter{footnote}\footnotetext{
\textbf{godson} \emph{n.}  (GAHD-suhn)
A male godchild.}, the friend, the favourite
of his father!"  She could have added, "A young man, too,
like \emph{you}, whose very countenance\index{countenance} may vouch\index{vouch} for your being
amiable\index{amiable}"--but she contented herself with, "and one, too, who
had probably been his companion from childhood, connected
together, as I think you said, in the closest manner!"

"We were born in the same parish\index{parish}, within the same park; the
greatest part of our youth was passed together; inmates of the
same house, sharing the same amusements, objects of the same
parental care.  \emph{my} father began life in the profession which
your uncle, Mr. Phillips, appears to do so much credit to--but
he gave up everything to be of use to the late Mr. Darcy and
devoted all his time to the care of the Pemberley property.
He was most highly esteemed by Mr. Darcy, a most intimate,
confidential friend.  Mr. Darcy often acknowledged himself
to be under the greatest obligations to my father's active
\textbf{superintendence}\index{superintendence}\stepcounter{footnote}\footnotetext{
\textbf{superintend} \emph{tr.v.} \textbf{-ence}  (soo-per-ihn-TEHND)
To oversee and direct; to supervise}, and when, immediately before my father's
death, Mr. Darcy gave him a voluntary promise of providing for
me, I am convinced that he felt it to be as much a debt of
gratitude to \emph{him}, as of his affection to myself."

"How strange!" cried Elizabeth.  "How abominable\index{abominable}!  I wonder
that the very pride of this Mr. Darcy has not made him just to
you!  If from no better motive, that he should not have been too
proud to be dishonest--for dishonesty I must call it."

"It \emph{is} wonderful," replied Wickham, "for almost all his actions
may be traced to pride; and pride had often been his best friend.
It has connected him nearer with \textbf{virtue}\index{virtue}\stepcounter{footnote}\footnotetext{
\textbf{virtue} \emph{n.}  (VER-choo)
morality with respect to sexual relations.} than with any other
feeling.  But we are none of us consistent, and in his behaviour
to me there were stronger impulses even than pride."

"Can such abominable\index{abominable} pride as his have ever done him good?"

"Yes.  It has often led him to be \textbf{liberal}\index{liberal}\stepcounter{footnote}\footnotetext{
\textbf{} \emph{}  ()
} and generous, to give his
money freely, to display hospitality, to assist his tenants, and
relieve the poor.  Family pride, and \emph{filial\index{filial}} pride--for he is very
proud of what his father was--have done this.  Not to appear to
disgrace his family, to \textbf{degenerate}\index{degenerate}\stepcounter{footnote}\footnotetext{
\textbf{degenerate} \emph{adj.}  (dih-JEHN-ner-eht)
a person who deviates from acceptable norms, especially in sexual behavior.} from the popular qualities, or
lose the influence of the Pemberley House, is a powerful motive.
He has also \emph{brotherly} pride, which, with \emph{some} brotherly
affection, makes him a very kind and careful guardian of his
sister, and you will hear him generally cried up as the most
attentive and best of brothers."

"What sort of girl is Miss Darcy?"

He shook his head.  "I wish I could call her amiable\index{amiable}.  It gives
me pain to speak ill of a Darcy.  But she is too much like her
brother--very, very proud.  As a child, she was affectionate
and pleasing, and extremely fond of me; and I have devoted hours
and hours to her amusement.  But she is nothing to me now.
She is a handsome girl, about fifteen or sixteen, and, I
understand, highly accomplished.  Since her father's death,
her home has been London, where a lady lives with her, and
superintends\index{superintends} her education."

After many pauses and many trials of other subjects\index{subjects}, Elizabeth
could not help reverting once more to the first, and saying:

"I am astonished at his intimacy with Mr. Bingley!  How can Mr.
Bingley, who seems good humour itself, and is, I really believe,
truly amiable\index{amiable}, be in friendship with such a man?  How can they
suit each other?  Do you know Mr. Bingley?"

"Not at all."

"He is a sweet-tempered, amiable\index{amiable}, charming man.  He cannot
know what Mr. Darcy is."

"Probably not; but Mr. Darcy can please where he chooses.  He
does not want abilities.  He can be a conversible companion if he
thinks it worth his while.  Among those who are at all his equals
in consequence\index{consequence}, he is a very different man from what he is to the
less prosperous.  His pride never deserts him; but with the rich
he is liberal-minded, just, sincere, rational\index{rational}, honourable, and
perhaps agreeable--allowing something for fortune and figure."

The whist\index{whist} party soon afterwards breaking up, the players
gathered round the other table and Mr. Collins took his station
between his cousin Elizabeth and Mrs. Phillips.  The usual
inquiries as to his success was made by the latter.  It had not
been very great; he had lost every point; but when Mrs. Phillips
began to express her concern thereupon, he assured her with
much earnest\index{earnest} \textbf{gravity}\index{gravity}\stepcounter{footnote}\footnotetext{
\textbf{gravity} \emph{n.}  (GRA-vih-tee)
 a serious and solemn manner.} that it was not of the least importance, that
he considered the money as a mere \textbf{trifle}\index{trifle}\stepcounter{footnote}\footnotetext{
\textbf{} \emph{}  ()
}, and begged that she
would not make herself uneasy.

"I know very well, madam," said he, "that when persons sit down
to a card-table, they must take their chances of these things, and
happily I am not in such circumstances as to make five shillings
any object.  There are undoubtedly many who could not say the
same, but thanks to Lady Catherine de Bourgh, I am removed
far beyond the necessity of regarding little matters."

Mr. Wickham's attention was caught; and after observing Mr.
Collins for a few moments, he asked Elizabeth in a low voice
whether her relation was very intimately acquainted with the
family of de Bourgh.

"Lady Catherine de Bourgh," she replied, "has very lately given
him a living.  I hardly know how Mr. Collins was first
introduced to her notice, but he certainly has not known her
long."

"You know of course that Lady Catherine de Bourgh and Lady
Anne Darcy were sisters; consequently that she is aunt to the
present Mr. Darcy."

"No, indeed, I did not.  I knew nothing at all of Lady Catherine's
connections.  I never heard of her existence till the day before
yesterday."

"Her daughter, Miss de Bourgh, will have a very large fortune,
and it is believed that she and her cousin will unite the two
estates."

This information made Elizabeth smile, as she thought of poor
Miss Bingley.  Vain indeed must be all her attentions, vain and
useless her affection for his sister and her praise of himself,
if he were already self-destined for another.

"Mr. Collins," said she, "speaks highly both of Lady Catherine
and her daughter; but from some particulars that he has related
of her ladyship, I suspect his gratitude misleads him, and that in
spite of her being his patroness\index{patroness}, she is an arrogant, conceited\index{conceited}
woman."

"I believe her to be both in a great degree," replied Wickham;
"I have not seen her for many years, but I very well remember that
I never liked her, and that her manners were \textbf{dictatorial}\index{dictatorial}\stepcounter{footnote}\footnotetext{
\textbf{dictatorial} \emph{adj.}  (dihk-tuh-TAW-ih-uhl)
domineering; characteristic of an absolute ruler.} and
\textbf{insolent}\index{insolent}\stepcounter{footnote}\footnotetext{
\textbf{insolent} \emph{adj.}  (IHN-soh-lehnt)
marked by casual, arrogant disrespect.}.  She has the reputation of being remarkably sensible
and clever; but I rather believe she derives part of her abilities
from her rank and fortune, part from her \textbf{authoritative}\index{authoritative}\stepcounter{footnote}\footnotetext{
\textbf{authoritative} \emph{adj.}  (aw-thaw-rih-TAY-tihv)
of recognized authority or excellence.} manner,
and the rest from the pride for her nephew, who chooses that
everyone connected with him should have an understanding of
the first class."

Elizabeth allowed that he had given a very rational\index{rational} account of it,
and they continued talking together, with mutual satisfaction
till supper put an end to cards, and gave the rest of the ladies
their share of Mr. Wickham's attentions.  There could be no
conversation in the noise of Mrs. Phillips's supper party, but
his manners recommended him to everybody.  Whatever he said,
was said well; and whatever he did, done gracefully.  Elizabeth
went away with her head full of him.  She could think of nothing
but of Mr. Wickham, and of what he had told her, all the way
home; but there was not time for her even to mention his name
as they went, for neither Lydia nor Mr. Collins were once silent.
Lydia talked \textbf{incessantly}\index{incessantly}\stepcounter{footnote}\footnotetext{
\textbf{incessant} \emph{adj.} \textbf{-ly}  (ihn-SEHS-suhnt)
constant without interruption.} of lottery tickets, of the fish she had
lost and the fish she had won; and Mr. Collins in describing the
civility\index{civility} of Mr. and Mrs. Phillips, protesting that he did not in
the least regard his losses at whist\index{whist}, \textbf{enumerating}\index{enumerating}\stepcounter{footnote}\footnotetext{
\textbf{enumerate} \emph{tr.v.}  (ee-NYOO-mer-ayt)
to list or count off item by item.} all the dishes
at supper, and repeatedly fearing that he crowded his cousins,
had more to say than he could well manage before the carriage
stopped at Longbourn House.



\chapter{17}


Elizabeth related to Jane the next day what had passed between
Mr. Wickham and herself.  Jane listened with astonishment and
concern; she knew not how to believe that Mr. Darcy could be
so unworthy of Mr. Bingley's regard; and yet, it was not in her
nature to question the \textbf{veracity}\index{veracity}\stepcounter{footnote}\footnotetext{
\textbf{veracity} \emph{n.}  (ver-RAS-sih-tee)
truthfulness.} of a young man of such amiable\index{amiable}
appearance as Wickham.  The possibility of his having endured
such unkindness, was enough to interest all her tender feelings;
and nothing remained therefore to be done, but to think well of
them both, to defend the conduct of each, and throw into the
account of accident or mistake whatever could not be otherwise
explained.

"They have both," said she, "been deceived, I dare say, in some
way or other, of which we can form no idea.  Interested people
have perhaps misrepresented each to the other.  It is, in short,
impossible for us to conjecture\index{conjecture} the causes or circumstances
which may have \textbf{alienated}\index{alienated}\stepcounter{footnote}\footnotetext{
\textbf{alienate} \emph{tr.v.} \textbf{-ed}  (AY-lee-uh-nayt)
to estrange; to disassociate with.} them, without actual blame on either
side."

"Very true, indeed; and now, my dear Jane, what have you got
to say on behalf of the interested people who have probably been
concerned in the business?  Do clear \emph{them} too, or we shall be
obliged to think ill of somebody."

"Laugh as much as you choose, but you will not laugh me out of
my opinion.  My dearest Lizzy, do but consider in what a
disgraceful light it places Mr. Darcy, to be treating his father's
favourite in such a manner, one whom his father had promised to
provide for.  It is impossible.  No man of common humanity, no
man who had any value for his character, could be capable of it.
Can his most intimate friends be so excessively deceived in him?
Oh! no."

"I can much more easily believe Mr. Bingley's being \textbf{imposed}\index{imposed}\stepcounter{footnote}\footnotetext{
\textbf{impose} \emph{v.} \textbf{-ed}  (ihm-POAZ)
to establish as obligatory.} on,
than that Mr. Wickham should invent such a history of himself
as he gave me last night; names, facts, everything mentioned
without ceremony.  If it be not so, let\index{let} Mr. Darcy contradict it.
Besides, there was truth in his looks."

"It is difficult indeed--it is distressing.  One does not know what
to think."

"I beg your pardon; one knows exactly what to think."

But Jane could think with certainty on only one point--that Mr.
Bingley, if he \emph{had} been imposed\index{imposed} on, would have much to suffer
when the affair became public.

The two young ladies were summoned\index{summoned} from the shrubbery\index{shrubbery},
where this conversation passed, by the arrival of the very
persons of whom they had been speaking; Mr. Bingley and his
sisters came to give their personal invitation for the
long-expected ball at Netherfield, which was fixed for the
following Tuesday.  The two ladies were delighted to see their
dear friend again, called it an age since they had met, and
repeatedly asked what she had been doing with herself since
their separation.  To the rest of the family they paid little
attention; avoiding Mrs. Bennet as much as possible, saying not
much to Elizabeth, and nothing at all to the others.  They were
soon gone again, rising from their seats with an activity which
took their brother by surprise, and hurrying off as if eager to
escape from Mrs. Bennet's civilities.

The prospect\index{prospect} of the Netherfield ball was extremely agreeable to
every female of the family.  Mrs. Bennet chose to consider it as
given in compliment to her eldest daughter, and was particularly
flattered by receiving the invitation from Mr. Bingley himself,
instead of a \textbf{ceremonious}\index{ceremonious}\stepcounter{footnote}\footnotetext{
\textbf{ceremonious} \emph{adj.}  (sehr-rih-MOH-nih-uhs)
rigidly formal or bound by convention or ritual.} card.  Jane pictured to herself a happy
evening in the society of her two friends, and the attentions of
her brother; and Elizabeth thought with pleasure of dancing a
great deal with Mr. Wickham, and of seeing a confirmation of
everything in Mr. Darcy's look and behavior.  The happiness
anticipated by Catherine and Lydia depended less on any single
event, or any particular person, for though they each, like
Elizabeth, meant to dance half the evening with Mr. Wickham,
he was by no means the only partner who could satisfy them, and
a ball was, at any rate, a ball.  And even Mary could assure her
family that she had no \textbf{disinclination}\index{disinclination}\stepcounter{footnote}\footnotetext{
\textbf{disinclination} \emph{n.}  (dihs-IN-klih-NAY-shun)
a certain degree of unwillingness; reluctance.} for it.

"While I can have my mornings to myself," said she, "it is
enough--I think it is no sacrifice to join occasionally in evening
engagements.  Society has claims on us all; and I profess myself
one of those who consider intervals of recreation and amusement
as desirable for everybody."

Elizabeth's spirits were so high on this occasion, that though she
did not often speak unnecessarily to Mr. Collins, she could not
help asking him whether he intended to accept Mr. Bingley's
invitation, and if he did, whether he would think it proper to join
in the evening's amusement; and she was rather surprised to find
that he entertained no scruple\index{scruple} whatever on that head, and was
very far from dreading a rebuke either from the Archbishop, or
Lady Catherine de Bourgh, by venturing to dance.

"I am by no means of the opinion, I assure you," said he, "that
a ball of this kind, given by a young man of character, to
respectable people, can have any evil tendency; and I am so
far from objecting to dancing myself, that I shall hope to be
honoured with the hands of all my fair cousins in the course of
the evening; and I take this opportunity of \textbf{soliciting}\index{soliciting}\stepcounter{footnote}\footnotetext{
\textbf{solicit} \emph{v.} \textbf{-ation}  (suh-LIH-sit)
} yours, Miss
Elizabeth, for the two first dances especially, a preference which
I trust my cousin Jane will attribute to the right cause, and not
to any disrespect for her."

Elizabeth felt herself completely taken in.  She had fully
proposed being engaged by Mr. Wickham for those very dances;
and to have Mr. Collins instead! her liveliness had never been
worse timed.  There was no help for it, however.  Mr. Wickham's
happiness and her own were \textbf{perforce}\index{perforce}\stepcounter{footnote}\footnotetext{
\textbf{perforce} \emph{adv.}  (per-FORS)
by necessity; by force of circumstance.} delayed a little longer,
and Mr. Collins's proposal accepted with as good a grace as she
could.  She was not the better pleased with his gallantry\index{gallantry} from
the idea it suggested of something more.  It now first struck
her, that \emph{she} was selected from among her sisters as worthy of
being mistress of Hunsford Parsonage, and of assisting to form
a quadrille\index{quadrille} table at Rosings, in the absence of more eligible
visitors.  The idea soon reached to conviction\index{conviction}, as she observed
his increasing civilities toward herself, and heard his
frequent attempt at a compliment on her wit and \textbf{vivacity}\index{vivacity}\stepcounter{footnote}\footnotetext{
\textbf{vivacity} \emph{n.}  (vih-VAS-sih-tee)
characterized by high spirits and animation.}; and
though more astonished than gratified herself by this effect
of her charms, it was not long before her mother gave her to
understand that the probability of their marriage was extremely
agreeable to \emph{her}.  Elizabeth, however, did not choose to take
the hint, being well aware that a serious dispute must be the
consequence\index{consequence} of any reply.  Mr. Collins might never make the
offer, and till he did, it was useless to quarrel about him.

If there had not been a Netherfield ball to prepare for and talk
of, the younger Miss Bennets would have been in a very pitiable
state at this time, for from the day of the invitation, to the day
of the ball, there was such a \textbf{succession}\index{succession}\stepcounter{footnote}\footnotetext{
\textbf{succession} \emph{n.}  (suhk-SEH-shun)
the action of following in order.} of rain as prevented their
walking to Meryton once.  No aunt, no officers, no news could
be sought after--the very shoe-roses for Netherfield were got
by \textbf{proxy}\index{proxy}\stepcounter{footnote}\footnotetext{
\textbf{proxy} \emph{n.}  (PRAHK-see)
 a person authorized to act for another; an agent.}.  Even Elizabeth might have found some trial of her
patience in weather which totally suspended the improvement of
her acquaintance with Mr. Wickham; and nothing less than a
dance on Tuesday, could have made such a Friday, Saturday,
Sunday, and Monday \textbf{endurable}\index{endurable}\stepcounter{footnote}\footnotetext{
\textbf{endurable} \emph{adj.}  (ehn-DOOR-uh-bl)
capable of being tolerated although unpleasant.} to Kitty and Lydia.



\chapter{18}


Till Elizabeth entered the drawing-room at Netherfield, and
looked in vain for Mr. Wickham among the cluster of red coats
there assembled, a doubt of his being present had never occurred
to her.  The certainty of meeting him had not been checked by
any of those recollections\index{recollections} that might not unreasonably have
alarmed her.  She had dressed with more than usual care, and
prepared in the highest spirits for the conquest of all that
remained unsubdued of his heart, trusting that it was not more
than might be won in the course of the evening.  But in an
instant arose the dreadful suspicion of his being purposely
omitted for Mr. Darcy's pleasure in the Bingleys' invitation
to the officers; and though this was not exactly the case, the
absolute fact of his absence was pronounced by his friend Denny,
to whom Lydia eagerly applied, and who told them that Wickham
had been obliged to go to town on business the day before, and
was not yet returned; adding, with a significant smile, "I do not
imagine his business would have called him away just now, if he
had not wanted to avoid a certain gentleman here."

This part of his intelligence, though unheard by Lydia, was
caught by Elizabeth, and, as it assured her that Darcy was not
less answerable for Wickham's absence than if her first surmise\index{surmise}
had been just, every feeling of displeasure against the former
was so sharpened by immediate disappointment, that she could
hardly reply with tolerable\index{tolerable} civility\index{civility} to the polite inquiries
which he directly afterwards approached to make.  Attendance,
\textbf{forbearance}\index{forbearance}\stepcounter{footnote}\footnotetext{
\textbf{forbearance} \emph{n.}  (for-BAIR-ense)
good-natured tolerance of delay or incompetence.}, patience with Darcy, was injury to Wickham.  She
was resolved against any sort of conversation with him, and
turned away with a degree of ill-humour which she could not
wholly\index{wholly} \textbf{surmount}\index{surmount}\stepcounter{footnote}\footnotetext{
\textbf{surmount} \emph{tr.v.}  (ser-MOWNT)
To overcome.} even in speaking to Mr. Bingley, whose blind
partiality\index{partiality} provoked\index{provoked} her.

But Elizabeth was not formed for ill-humour; and though every
prospect\index{prospect} of her own was destroyed for the evening, it could not
dwell long on her spirits; and having told all her griefs to
Charlotte Lucas, whom she had not seen for a week, she was
soon able to make a voluntary transition to the oddities of her
cousin, and to point him out to her particular notice.  The first
two dances, however, brought a return of distress; they were
dances of \textbf{mortification}\index{mortification}\stepcounter{footnote}\footnotetext{
\textbf{mortification} \emph{n.}  (mor-tih-fih-KAY-shun)
strong feelings of embarrassment.}.  Mr. Collins, awkward and \textbf{solemn}\index{solemn}\stepcounter{footnote}\footnotetext{
\textbf{solemn} \emph{adj.}  (SAH-lehm)
 dignified and somber in manner or character.},
apologising instead of attending, and often moving wrong
without being aware of it, gave her all the shame and misery
which a disagreeable partner for a couple of dances can give.
The moment of her release from him was \textbf{ecstasy}\index{ecstasy}\stepcounter{footnote}\footnotetext{
\textbf{ecstasy} \emph{n.}  (EHX-tuh-see)
 a state of elated bliss.}.

She danced next with an officer, and had the refreshment of
talking of Wickham, and of hearing that he was universally liked.
When those dances were over, she returned to Charlotte Lucas,
and was in conversation with her, when she found herself
suddenly addressed\index{addressed} by Mr. Darcy who took her so much by surprise
in his application\index{application} for her hand, that, without knowing what she
did, she accepted him.  He walked away again immediately, and
she was left to \textbf{fret}\index{fret}\stepcounter{footnote}\footnotetext{
\textbf{fret} \emph{n.}  (FREHT)
 agitation resulting from active worry.} over her own want of presence of mind;
Charlotte tried to \textbf{console}\index{console}\stepcounter{footnote}\footnotetext{
\textbf{console} \emph{tr.v.}  (kahn-SOHL)
To give comfort or solace to.} her:

"I dare say you will find him very agreeable."

"Heaven forbid!  \emph{that} would be the greatest misfortune of all!
To find a man agreeable whom one is determined to hate!  Do not
wish me such an evil."

When the dancing recommenced, however, and Darcy approached to
claim her hand, Charlotte could not help cautioning her in a
whisper, not to be a simpleton\index{simpleton}, and allow her fancy for Wickham
to make her appear unpleasant in the eyes of a man ten times his
consequence\index{consequence}.  Elizabeth made no answer, and took her place in
the set, amazed at the dignity to which she was arrived in being
allowed to stand opposite to Mr. Darcy, and reading in her
neighbours' looks, their equal amazement in beholding it.  They
stood for some time without speaking a word; and she began to
imagine that their silence was to last through the two dances,
and at first was resolved not to break it; till suddenly
fancying that it would be the greater punishment to her partner
to oblige him to talk, she made some slight observation on the
dance.  He replied, and was again silent.  After a pause of
some minutes\index{minutes}, she addressed\index{addressed} him a second time with:--"It is
\emph{your} turn to say something now, Mr. Darcy.  I talked about
the dance, and \emph{you} ought to make some sort of remark on the
size of the room, or the number of couples."

He smiled, and assured her that whatever she wished him to say
should be said.

"Very well.  That reply will do for the present.  Perhaps by and
by I may observe that private\index{private} balls are much pleasanter than
public ones.  But \emph{now} we may be silent."

"Do you talk by rule, then, while you are dancing?"

"Sometimes.  One must speak a little, you know.  It would look
odd to be entirely silent for half an hour together; and yet for
the advantage of \emph{some}, conversation ought to be so arranged, as
that they may have the trouble of saying as little as possible."

"Are you consulting your own feelings in the present case, or do
you imagine that you are gratifying mine?"

"Both," replied Elizabeth archly\index{archly}; "for I have always seen a great
similarity in the turn of our minds.  We are each of an unsocial,
\textbf{taciturn}\index{taciturn}\stepcounter{footnote}\footnotetext{
\textbf{taciturn} \emph{adj.}  (TAS-sih-tern)
habitually reserved and uncommunicative; untalkative.} disposition\index{disposition}, unwilling to speak, unless we expect to say
something that will amaze the whole room, and be handed down
to \textbf{posterity}\index{posterity}\stepcounter{footnote}\footnotetext{
\textbf{posterity} \emph{n.}  (paws-TEHR-ih-tee)
Future generations.} with all the \textbf{eclat}\index{eclat}\stepcounter{footnote}\footnotetext{
\textbf{clat} \emph{n.}  ()
} of a \textbf{proverb}\index{proverb}\stepcounter{footnote}\footnotetext{
\textbf{proverb} \emph{n.}  (PRAH-verb)
 A short saying that expresses a basic truth, value, or fact.}."

"This is no very striking resemblance of your own character,
I am sure," said he.  "How near it may be to \emph{mine}, I cannot
pretend to say.  \emph{you} think it a faithful portrait undoubtedly."

"I must not decide on my own performance."

He made no answer, and they were again silent till they had gone
down the dance, when he asked her if she and her sisters did not
very often walk to Meryton.  She answered in the \textbf{affirmative}\index{affirmative}\stepcounter{footnote}\footnotetext{
\textbf{affirmative} \emph{adj.}  (uh-FER-muh-tihv)
},
and, unable to resist the temptation, added, "When you met us
there the other day, we had just been forming a new acquaintance."

The effect was immediate.  A deeper shade of \textbf{hauteur}\index{hauteur}\stepcounter{footnote}\footnotetext{
\textbf{hauteur} \emph{n.}  (haw-TER)
Arrogance or haughtiness.} overspread
his features, but he said not a word, and Elizabeth, though
blaming herself for her own weakness, could not go on.  At
length Darcy spoke, and in a constrained manner said, "Mr.
Wickham is blessed with such happy manners as may ensure his
\emph{making} friends--whether he may be equally capable of \emph{retaining}
them, is less certain."

"He has been so unlucky as to lose \emph{your} friendship," replied
Elizabeth with \textbf{emphasis}\index{emphasis}\stepcounter{footnote}\footnotetext{
\textbf{emphasis} \emph{n.}  (EHM-fuh-sihs)
special importance or significance.}, "and in a manner which he is likely to
suffer from all his life."

Darcy made no answer, and seemed desirous of changing the
subject\index{subject}.  At that moment, Sir William Lucas appeared close to
them, meaning to pass through the set to the other side of the
room; but on perceiving Mr. Darcy, he stopped with a bow of
superior\index{superior} courtesy to compliment him on his dancing and his
partner.

"I have been most highly gratified indeed, my dear sir.  Such
very superior\index{superior} dancing is not often seen.  It is evident that you
belong to the first circles.  Allow me to say, however, that your
fair partner does not disgrace you, and that I must hope to have
this pleasure often repeated, especially when a certain desirable
event, my dear Eliza (glancing at her sister and Bingley) shall
take place.  What congratulations will then flow in!  I appeal to
Mr. Darcy:--but let\index{let} me not interrupt you, sir.  You will not
thank me for detaining you from the bewitching\index{bewitching} converse of that
young lady, whose bright eyes are also \textbf{upbraiding}\index{upbraiding}\stepcounter{footnote}\footnotetext{
\textbf{upbraid} \emph{tr.v.} \textbf{-ed}  (uhp-BRAYD)
 to give a severe scolding.} me."

The latter part of this address\index{address} was scarcely heard by Darcy;
but Sir William's allusion\index{allusion} to his friend seemed to strike him
forcibly, and his eyes were directed with a very serious
expression towards Bingley and Jane, who were dancing together.
Recovering himself, however, shortly, he turned to his partner,
and said, "Sir William's interruption has made me forget what
we were talking of."

"I do not think we were speaking at all.  Sir William could not
have interrupted two people in the room who had less to say for
themselves.  We have tried two or three subjects\index{subjects} already without
success, and what we are to talk of next I cannot imagine."

"What think you of books?" said he, smiling.

"Books--oh! no.  I am sure we never read the same, or not with
the same feelings."

"I am sorry you think so; but if that be the case, there can at
least be no want of subject\index{subject}.  We may compare our different
opinions."

"No--I cannot talk of books in a ball-room; my head is always
full of something else."

"The \emph{present} always occupies you in such scenes--does it?"
said he, with a look of doubt.

"Yes, always," she replied, without knowing what she said,
for her thoughts had wandered far from the subject\index{subject}, as soon
afterwards appeared by her suddenly exclaiming, "I remember
hearing you once say, Mr. Darcy, that you hardly ever forgave,
that you resentment\index{resentment} once created was unappeasable.  You are
very cautious, I suppose, as to its \emph{being} \emph{created}."

"I am," said he, with a firm voice.

"And never allow yourself to be blinded by prejudice\index{prejudice}?"

"I hope not."

"It is particularly incumbent\index{incumbent} on those who never change their
opinion, to be secure of judging properly at first."

"May I ask to what these questions tend?"

"Merely to the illustration of \emph{your} character," said she,
\textbf{endeavouring}\index{endeavouring}\stepcounter{footnote}\footnotetext{
\textbf{} \emph{}  ()
}\index{endeavouring} to shake off her gravity\index{gravity}.  "I am trying to make
it out."

"And what is your success?"

She shook her head.  "I do not get on at all.  I hear such
different accounts of you as puzzle me exceedingly."

"I can readily\index{readily} believe," answered he gravely\index{gravely}, "that reports may
vary greatly with respect to me; and I could wish, Miss Bennet,
that you were not to sketch my character at the present moment,
as there is reason to fear that the performance would reflect no
credit on either."

"But if I do not take your likeness now, I may never have
another opportunity."

"I would by no means suspend any pleasure of yours," he coldly
replied.  She said no more, and they went down the other dance
and parted in silence; and on each side dissatisfied, though not
to an equal degree, for in Darcy's breast there was a tolerable\index{tolerable}
powerful feeling towards her, which soon procured\index{procured} her pardon,
and directed all his anger against another.

They had not long separated, when Miss Bingley came towards
her, and with an expression of civil disdain accosted\index{accosted} her:

"So, Miss Eliza, I hear you are quite delighted with George
Wickham!  Your sister has been talking to me about him, and
asking me a thousand questions; and I find that the young man
quite forgot to tell you, among his other communication, that
he was the son of old Wickham, the late Mr. Darcy's \textbf{steward}\index{steward}\stepcounter{footnote}\footnotetext{
\textbf{steward} \emph{n.}  (STYOO-werd)
someone who manages property or other affairs for someone else.}.
Let\index{Let} me recommend you, however, as a friend, not to give \textbf{implicit}\index{implicit}\stepcounter{footnote}\footnotetext{
\textbf{implicit} \emph{adj.} \textbf{-ness}  (ihm-PLIH-siht)
implied; inherent in the nature of something.}
confidence\index{confidence} to all his assertions\index{assertions}; for as to Mr. Darcy's using
him ill, it is perfectly false; for, on the contrary, he has
always been remarkably kind to him, though George Wickham has
treated Mr. Darcy in a most \textbf{infamous}\index{infamous}\stepcounter{footnote}\footnotetext{
\textbf{infamous} \emph{adj.}  (IHN-fuh-muhs)
Having an exceedingly bad reputation; notorious.} manner.  I do not know the
particulars, but I know very well that Mr. Darcy is not in the
least to blame, that he cannot bear to hear George Wickham
mentioned, and that though my brother thought that he could not
well avoid including him in his invitation to the officers, he
was excessively glad to find that he had taken himself out of the
way.  His coming into the country at all is a most insolent\index{insolent} thing,
indeed, and I wonder how he could presume to do it.  I pity you,
Miss Eliza, for this discovery of your favourite's guilt; but
really, considering his \textbf{descent}\index{descent}\stepcounter{footnote}\footnotetext{
\textbf{descent} \emph{n.}  (dee-SEHNT)
a downward slope.}, one could not expect much better."

"His guilt and his descent\index{descent} appear by your account to be the
same," said Elizabeth angrily; "for I have heard you accuse him
of nothing worse than of being the son of Mr. Darcy's steward\index{steward},
and of \emph{that}, I can assure you, he informed me himself."

"I beg your pardon," replied Miss Bingley, turning away with a
sneer.  "Excuse my interference--it was kindly meant."

"Insolent\index{Insolent} girl!" said Elizabeth to herself.  "You are much
mistaken if you expect to influence me by such a paltry\index{paltry} attack
as this.  I see nothing in it but your own wilful ignorance and
the \textbf{malice}\index{malice}\stepcounter{footnote}\footnotetext{
\textbf{malice} \emph{n.}  (MAL-lihs)
hate or extreme ill will.} of Mr. Darcy."  She then sought her eldest sister, who
has undertaken to make inquiries on the same subject\index{subject} of Bingley.
Jane met her with a smile of such sweet complacency\index{complacency}, a glow of
such happy expression, as sufficiently marked how well she was
satisfied with the occurrences of the evening.  Elizabeth instantly
read her feelings, and at that moment solicitude\index{solicitude} for Wickham,
resentment\index{resentment} against his enemies, and everything else, gave way
before the hope of Jane's being in the fairest way for happiness.

"I want to know," said she, with a countenance\index{countenance} no less smiling
than her sister's, "what you have learnt about Mr. Wickham.
But perhaps you have been too pleasantly engaged to think of
any third person; in which case you may be sure of my pardon."

"No," replied Jane, "I have not forgotten him; but I have nothing
satisfactory to tell you.  Mr. Bingley does not know the whole of
his history, and is quite ignorant of the circumstances which have
principally\index{principally} offended Mr. Darcy; but he will vouch\index{vouch} for the good
conduct, the \textbf{probity}\index{probity}\stepcounter{footnote}\footnotetext{
\textbf{probity} \emph{n.}  (PROH-bih-tee)
complete and confirmed integrity.}, and honour of his friend, and is perfectly
convinced that Mr. Wickham has deserved much less attention
from Mr. Darcy than he has received; and I am sorry to say by
his account as well as his sister's, Mr. Wickham is by no means a
respectable young man.  I am afraid he has been very \textbf{imprudent}\index{imprudent}\stepcounter{footnote}\footnotetext{
\textbf{imprudent} \emph{adj.}  (ihm-PROO-dehnt)
lacking wise self-restraint; indiscreet.},
and has deserved to lose Mr. Darcy's regard."

"Mr. Bingley does not know Mr. Wickham himself?"

"No; he never saw him till the other morning at Meryton."

"This account then is what he has received from Mr. Darcy.
I am satisfied.  But what does he say of the living?"

"He does not exactly recollect the circumstances, though he has
heard them from Mr. Darcy more than once, but he believes that
it was left to him \emph{conditionally} only."

"I have not a doubt of Mr. Bingley's sincerity," said Elizabeth
warmly; "but you must excuse my not being convinced by
assurances only.  Mr. Bingley's defense of his friend was a very
able one, I dare say; but since he is unacquainted with several
parts of the story, and has learnt the rest from that friend
himself, I shall venture to still think of both gentlemen as I
did before."

She then changed the discourse\index{discourse} to one more gratifying to each,
and on which there could be no difference of sentiment\index{sentiment}.
Elizabeth listened with delight to the happy, though modest
hopes which Jane entertained of Mr. Bingley's regard, and said
all in her power to heighten her confidence\index{confidence} in it.  On their being
joined by Mr. Bingley himself, Elizabeth withdrew to Miss
Lucas; to whose inquiry after the pleasantness of her last partner
she had scarcely replied, before Mr. Collins came up to them,
and told her with great \textbf{exultation}\index{exultation}\stepcounter{footnote}\footnotetext{
\textbf{exultation} \emph{n.}  (ehg-zuhl-TAY-shun)
a feeling of extreme joy.} that he had just been so
fortunate as to make a most important discovery.

"I have found out," said he, "by a singular\index{singular} accident, that there
is now in the room a near relation of my patroness\index{patroness}.  I happened
to overhear the gentleman himself mentioning to the young lady who
does the honours of the house the names of his cousin Miss de
Bourgh, and of her mother Lady Catherine.  How wonderfully these
sort of things occur!  Who would have thought of my meeting with,
perhaps, a nephew of Lady Catherine de Bourgh in this assembly!
I am most thankful that the discovery is made in time for me to
pay my respects to him, which I am now going to do, and trust
he will excuse my not having done it before.  My total ignorance
of the connection must plead my apology."

"You are not going to introduce yourself to Mr. Darcy!"

"Indeed I am.  I shall entreat\index{entreat} his pardon for not having done it
earlier.  I believe him to be Lady Catherine's \emph{nephew}.  It will
be in my power to assure him that her ladyship was quite well
yesterday se'nnight."

Elizabeth tried hard to \textbf{dissuade}\index{dissuade}\stepcounter{footnote}\footnotetext{
\textbf{dissuade} \emph{tr.v.}  (sihs-SWAYD)
to turn away from by persuasion.} him from such a scheme,
assuring him that Mr. Darcy would consider his addressing\index{addressing} him
without introduction as an impertinent\index{impertinent} freedom, rather than a
compliment to his aunt; that it was not in the least necessary
there should be any notice on either side; and that if it were,
it must belong to Mr. Darcy, the superior\index{superior} in consequence\index{consequence}, to
begin the acquaintance.  Mr. Collins listened to her with the
determined air of following his own inclination\index{inclination}, and, when she
ceased speaking, replied thus:

"My dear Miss Elizabeth, I have the highest opinion in the world
in your excellent judgement in all matters within the \textbf{scope}\index{scope}\stepcounter{footnote}\footnotetext{
\textbf{scope} \emph{n.}  (SKOHP)
} of
your understanding; but permit me to say, that there must be a
wide difference between the established forms of ceremony amongst
the \textbf{laity}\index{laity}\stepcounter{footnote}\footnotetext{
\textbf{laity} \emph{n.}  (LAY-ih-tee)
 everyone except the clergy.}, and those which regulate the clergy; for, give me
leave to observe that I consider the clerical office as equal in
point of dignity with the highest rank in the kingdom--provided
that a proper humility\index{humility} of behaviour is at the same time
maintained.  You must therefore allow me to follow the dictates
of my \textbf{conscience}\index{conscience}\stepcounter{footnote}\footnotetext{
\textbf{conscience} \emph{n.}  (KAHN-shehns)
a feeling of shame when you do something immoral.} on this occasion, which leads me to perform what
I look on as a point of duty.  Pardon me for neglecting to profit
by your advice, which on every other subject\index{subject} shall be my constant
guide, though in the case before us I consider myself more fitted
by education and \textbf{habitual}\index{habitual}\stepcounter{footnote}\footnotetext{
\textbf{habitual} \emph{adj.}  (huh-BIH-choo-uhl)
made a norm, custom, or habit.} study to decide on what is right than
a young lady like yourself."  And with a low bow he left her to
attack Mr. Darcy, whose reception of his advances she eagerly
watched, and whose astonishment at being so addressed\index{addressed} was very
evident.  Her cousin prefaced his speech with a solemn\index{solemn} bow and
though she could not hear a word of it, she felt as if hearing
it all, and saw in the motion of his lips the words "apology,"
"Hunsford," and "Lady Catherine de Bourgh."  It vexed\index{vexed} her to
see him expose himself to such a man.  Mr. Darcy was eyeing him
with unrestrained wonder, and when at last Mr. Collins allowed
him time to speak, replied with an air of distant civility\index{civility}.  Mr.
Collins, however, was not discouraged from speaking again, and
Mr. Darcy's contempt\index{contempt} seemed abundantly increasing with the length
of his second speech, and at the end of it he only made him a
slight bow, and moved another way.  Mr. Collins then returned
to Elizabeth.

"I have no reason, I assure you," said he, "to be dissatisfied
with my reception.  Mr. Darcy seemed much pleased with the
attention.  He answered me with the utmost civility\index{civility}, and even
paid me the compliment of saying that he was so well convinced
of Lady Catherine's \textbf{discernment}\index{discernment}\stepcounter{footnote}\footnotetext{
\textbf{discernment} \emph{n.}  (dihs-SERN-mehnt)
the trait of judging wisely and objectively.} as to be certain she could never
bestow\index{bestow} a favour unworthily.  It was really a very handsome
thought.  Upon the whole, I am much pleased with him."

As Elizabeth had no longer any interest of her own to pursue,
she turned her attention almost entirely on her sister and Mr.
Bingley; and the train of agreeable reflections which her
observations gave birth to, made her perhaps almost as happy as
Jane.  She saw her in idea settled in that very house, in all the
felicity\index{felicity} which a marriage of true affection could bestow\index{bestow}; and she
felt capable, under such circumstances, of endeavouring\index{endeavouring} even to
like Bingley's two sisters.  Her mother's thoughts she plainly
saw were bent the same way, and she determined not to venture
near her, lest she might hear too much.  When they sat down to
supper, therefore, she considered it a most unlucky \textbf{perverseness}\index{perverseness}\stepcounter{footnote}\footnotetext{
\textbf{perverse} \emph{adj.} \textbf{-ness}  (per-VERS)
deliberate unruliness, often in a sexual or violent sense.}
which placed them within one of each other; and deeply was she
vexed\index{vexed} to find that her mother was talking to that one person
(Lady Lucas) freely, openly, and of nothing else but her
expectation that Jane would soon be married to Mr. Bingley.  It
was an animating subject\index{subject}, and Mrs. Bennet seemed incapable of
fatigue while enumerating the advantages of the match.  His
being such a charming young man, and so rich, and living but
three miles from them, were the first points of \textbf{self-gratulation}\index{self-gratulation}\stepcounter{footnote}\footnotetext{
\textbf{} \emph{}  ()
};
and then it was such a comfort to think how fond the two sisters
were of Jane, and to be certain that they must desire the
connection as much as she could do.  It was, moreover, such a
promising thing for her younger daughters, as Jane's marrying so
greatly must throw them in the way of other rich men; and lastly,
it was so pleasant at her time of life to be able to \textbf{consign}\index{consign}\stepcounter{footnote}\footnotetext{
\textbf{consign} \emph{v.} \textbf{-ed}  (kuhn-SYN)
give over to another for care or safekeeping.} her
single daughters to the care of their sister, that she might not
be obliged to go into company more than she liked.  It was
necessary to make this circumstance a matter of pleasure,
because on such occasions it is the \textbf{etiquette}\index{etiquette}\stepcounter{footnote}\footnotetext{
\textbf{etiquette} \emph{n.}  (EH-tih-keht)
rules governing socially acceptable behavior.}; but no one was less
likely than Mrs. Bennet to find comfort in staying home at any
period of her life.  She concluded with many good wishes that
Lady Lucas might soon be equally fortunate,  though evidently
and \textbf{triumphantly}\index{triumphantly}\stepcounter{footnote}\footnotetext{
\textbf{triumphant} \emph{adj.} \textbf{-ly}  (try-UHM-fant)
victorious.} believing there was no chance of it.

In vain did Elizabeth endeavour\index{endeavour} to check the rapidity of her
mother's words, or persuade her to describe her felicity\index{felicity} in a
less audible whisper; for, to her inexpressible vexation\index{vexation}, she
could perceive\index{perceive} that the chief of it was overheard by Mr. Darcy,
who sat opposite to them.  Her mother only scolded her for
being nonsensical\index{nonsensical}.

"What is Mr. Darcy to me, pray, that I should be afraid of him?
I am sure we owe him no such particular civility\index{civility} as to be obliged
to say nothing \emph{he} may not like to hear."

"For heaven's sake, madam, speak lower.  What advantage can
it be for you to offend Mr. Darcy?  You will never recommend
yourself to his friend by so doing!"

Nothing that she could say, however, had any influence.  Her
mother would talk of her views in the same intelligible tone.
Elizabeth blushed and blushed again with shame and vexation\index{vexation}.
She could not help frequently glancing her eye at Mr. Darcy,
though every glance convinced her of what she dreaded; for
though he was not always looking at her mother, she was
convinced that his attention was \textbf{invariably}\index{invariably}\stepcounter{footnote}\footnotetext{
\textbf{invariable} \emph{adj.} \textbf{-ly}  (ihn-VAIR-ee-uh-bl)
constant; not changing.} fixed by her.
The expression of his face changed gradually from \textbf{indignant}\index{indignant}\stepcounter{footnote}\footnotetext{
\textbf{indignant} \emph{adj.}  (ihn-DIHG-nuhnt)
angered at something unjust or wrong.}
contempt\index{contempt} to a composed and steady\index{steady} gravity\index{gravity}.

At length, however, Mrs. Bennet had no more to say; and Lady
Lucas, who had been long yawning at the repetition of delights
which she saw no likelihood of sharing, was left to the comforts
of cold ham and chicken.  Elizabeth now began to revive.  But
not long was the interval of tranquillity\index{tranquillity}; for, when supper was
over, singing was talked of, and she had the mortification\index{mortification} of
seeing Mary, after very little entreaty\index{entreaty}, preparing to oblige the
company.  By many significant looks and silent entreaties, did
she endeavour\index{endeavour} to prevent such a proof of complaisance\index{complaisance}, but in
vain; Mary would not understand them; such an opportunity of
exhibiting was delightful to her, and she began her song.
Elizabeth's eyes were fixed on her with most painful sensations,
and she watched her progress through the several stanzas with
an impatience which was very ill rewarded at their close; for
Mary, on receiving, amongst the thanks of the table, the hint of
a hope that she might be prevailed\index{prevailed} on to favour them again, after
the pause of half a minute\index{minute} began another.  Mary's powers were
by no means fitted for such a display; her voice was weak, and
her manner affected.  Elizabeth was in agonies.  She looked at
Jane, to see how she bore it; but Jane was very composedly
talking to Bingley.  She looked at his two sisters, and saw
them making signs of \textbf{derision}\index{derision}\stepcounter{footnote}\footnotetext{
\textbf{derision} \emph{n.}  (dih-RIH-zhun)
exposing someone to laughter or ridicule.} at each other, and at Darcy, who
continued, however, \textbf{imperturbably}\index{imperturbably}\stepcounter{footnote}\footnotetext{
\textbf{imperturbable} \emph{adj.} \textbf{-ly}  (ihm-per-TER-buh-bl)
Cool and unshakable demeanor.} grave\index{grave}.  She looked at her
father to entreat\index{entreat} his interference, lest Mary should be singing all
night.  He took the hint, and when Mary had finished her second
song, said aloud, "That will do extremely well, child.  You have
delighted us long enough.  Let\index{Let} the other young ladies have time
to exhibit."

Mary, though pretending not to hear, was somewhat disconcerted\index{disconcerted};
and Elizabeth, sorry for her, and sorry for her father's speech,
was afraid her anxiety had done no good.  Others of the party
were now applied to.

"If I," said Mr. Collins, "were so fortunate as to be able to sing,
I should have great pleasure, I am sure, in obliging the company
with an air; for I consider music as a very innocent diversion,
and perfectly compatible with the profession of a clergyman.  I do
not mean, however, to assert that we can be justified in devoting
too much of our time to music, for there are certainly other things
to be attended to.  The rector\index{rector} of a parish\index{parish} has much to do.  In the
first place, he must make such an agreement for \textbf{tithes}\index{tithes}\stepcounter{footnote}\footnotetext{
\textbf{tithe} \emph{th}  (TYDH)
 Ten percent of one's income given to chairity or cause, especially to a church.} as a may be
beneficial to himself and not offensive to his patron.  He must
write his own sermons; and the time that remains will not be too
much for his parish\index{parish} duties, and the care and improvement of his
dwelling, which he cannot be excused from making as a comfortable
as possible.  And I do not think it of light importance that he
should have attentive and \textbf{conciliatory}\index{conciliatory}\stepcounter{footnote}\footnotetext{
\textbf{conciliate} \emph{v.} \textbf{-ing}  (kuhn-SIL-lih-ate)
overcoming animosity or hostility; to appease.} manner towards everybody,
especially towards those to whom he owes his preferment.  I
cannot \textbf{acquit}\index{acquit}\stepcounter{footnote}\footnotetext{
\textbf{acquit} \emph{tr.v.}  (uh-KWIT)
 to pronounce not guilty of criminal charges.} him of that duty; nor could I think well of the
man who should omit an occasion of testifying his respect
towards anybody connected with the family."  And with a bow to
Mr. Darcy, he concluded his speech, which had been spoken so
loud as a to be heard by half the room.  Many stared--many smiled;
but no one looked more amused than Mr. Bennet himself, while
his wife seriously commended\index{commended} Mr. Collins for having spoken so
sensibly, and observed in a half-whisper to Lady Lucas, that he
was a remarkably clever, good kind of young man.

To Elizabeth it appeared that, had her family made an agreement
to expose themselves as a much as a they could during the
evening, it would have been impossible for them to play their
parts with more spirit or finer success; and happy did she think
it for Bingley and her sister that some of the exhibition had
escaped his notice, and that his feelings were not of a sort to
be much distressed by the folly\index{folly} which he must have witnessed.
That his two sisters and Mr. Darcy, however, should have such
an opportunity of ridiculing her relations, was bad enough, and
she could not determine whether the silent contempt\index{contempt} of the
gentleman, or the insolent\index{insolent} smiles of the ladies, were more
intolerable\index{intolerable}.

The rest of the evening brought her little amusement.  She was
teased by Mr. Collins, who continued most perseveringly by her
side, and though he could not prevail\index{prevail} on her to dance with him
again, put it out of her power to dance with others.  In vain
did she entreat\index{entreat} him to stand up with somebody else, and offer to
introduce him to any young lady in the room.  He assured her,
that as to dancing, he was perfectly indifferent\index{indifferent} to it; that his
chief object was by delicate attentions to recommend himself to
her and that he should therefore make a point of remaining close
to her the whole evening.  There was no arguing upon such a
project.  She owed her greatest relief to her friend Miss Lucas,
who often joined them, and good-naturedly engaged Mr. Collins's
conversation to herself.

She was at least free from the offense of Mr. Darcy's further
notice; though often standing within a very short distance of her,
quite disengaged, he never came near enough to speak.  She felt
it to be the \textbf{probable}\index{probable}\stepcounter{footnote}\footnotetext{
\textbf{probable} \emph{adj.}  (PRAW-buh-bl)
likely but not certain} consequence\index{consequence} of her \textbf{allusions}\index{allusions}\stepcounter{footnote}\footnotetext{
\textbf{allusion} \emph{n.}  (al-LOO-zhun)
an indirect reference.} to Mr. Wickham,
and rejoiced in it.

The Longbourn party were the last of all the company to depart,
and, by a \textbf{manoeuvre}\index{manoeuvre}\stepcounter{footnote}\footnotetext{
\textbf{manoeuvre} \emph{n.}  (muh-NOO-ver)
} of Mrs. Bennet, had to wait for their
carriage a quarter of an hour after everybody else was gone,
which gave them time to see how heartily\index{heartily} they were wished
away by some of the family.  Mrs. Hurst and her sister scarcely
opened their mouths, except to complain of fatigue, and were
evidently impatient to have the house to themselves.  They
repulsed every attempt of Mrs. Bennet at conversation, and by
so doing threw a \textbf{languor}\index{languor}\stepcounter{footnote}\footnotetext{
\textbf{languor} \emph{n.}  (LAN-gor)
 a feeling of lack of interest or energy.} over the whole party, which was very
little relieved by the long speeches of Mr. Collins, who was
complimenting Mr. Bingley and his sisters on the elegance of
their entertainment, and the hospitality and politeness which had
marked their behaviour to their guests.  Darcy said nothing at all.
Mr. Bennet, in equal silence, was enjoying the scene.  Mr.
Bingley and Jane were standing together, a little detached from
the rest, and talked only to each other.  Elizabeth preserved as
steady\index{steady} a silence as either Mrs. Hurst or Miss Bingley; and even
Lydia was too much fatigued to utter more than the occasional
exclamation of "Lord, how tired I am!" accompanied by a
violent yawn.

When at length they arose to take leave, Mrs. Bennet was most
pressingly civil in her hope of seeing the whole family soon
at Longbourn, and addressed\index{addressed} herself especially to Mr. Bingley,
to assure him how happy he would make them by eating a family
dinner with them at any time, without the ceremony of a formal
invitation.  Bingley was all grateful pleasure, and he readily\index{readily}
engaged for taking the earliest opportunity of waiting on her,
after his return from London, whither he was obliged to go the
next day for a short time.

Mrs. Bennet was perfectly satisfied, and quitted the house under
the delightful persuasion that, allowing for the necessary
preparations of settlements, new carriages, and wedding clothes,
she should undoubtedly see her daughter settled at Netherfield in
the course of three or four months.  Of having another daughter
married to Mr. Collins, she thought with equal certainty, and
with considerable, though not equal, pleasure.  Elizabeth was the
least dear to her of all her children; and though the man and the
match were quite good enough for \emph{her}, the worth of each was
eclipsed by Mr. Bingley and Netherfield.



\chapter{19}


The next day opened a new scene at Longbourn.  Mr. Collins
made his declaration in form.  Having resolved to do it without
loss of time, as his leave of absence extended only to the
following Saturday, and having no feelings of \textbf{diffidence}\index{diffidence}\stepcounter{footnote}\footnotetext{
\textbf{diffidence} \emph{n.}  (DIHF-fih-dense)
a lack of self-confidence.} to make
it distressing to himself even at the moment, he set about it
in a very orderly manner, with all the observances, which
he supposed a regular\index{regular} part of the business.  On finding Mrs.
Bennet, Elizabeth, and one of the younger girls together, soon
after breakfast, he addressed\index{addressed} the mother in these words:

"May I hope, madam, for your interest with your fair daughter
Elizabeth, when I solicit\index{solicit} for the honour of a private\index{private} audience
with her in the course of this morning?"

Before Elizabeth had time for anything but a blush of surprise,
Mrs. Bennet answered instantly, "Oh dear!--yes--certainly.  I
am sure Lizzy will be very happy--I am sure she can have no
objection.  Come, Kitty, I want you upstairs."  And, gathering
her work together, she was hastening away, when Elizabeth
called out:

"Dear madam, do not go.  I beg you will not go.  Mr. Collins
must excuse me.  He can have nothing to say to me that anybody
need not hear.  I am going away myself."

"No, no, nonsense, Lizzy.  I desire you to stay where you are."
And upon Elizabeth's seeming really, with vexed\index{vexed} and embarrassed
looks, about to escape, she added: "Lizzy, I \emph{insist} upon your
staying and hearing Mr. Collins."

Elizabeth would not oppose such an \textbf{injunction}\index{injunction}\stepcounter{footnote}\footnotetext{
\textbf{injunction} \emph{n.}  (ihn-JUHNK-shun)
a formal command or admonition.}--and a moment's
consideration making her also sensible that it would be wisest to
get it over as soon and as quietly as possible, she sat down again
and tried to conceal, by incessant\index{incessant} employment the feelings which
were divided between distress and diversion. Mrs. Bennet and Kitty
walked off, and as soon as they were gone, Mr. Collins began.

"Believe me, my dear Miss Elizabeth, that your modesty, so far
from doing you any disservice, rather adds to your other
perfections.  You would have been less amiable\index{amiable} in my eyes had
there \emph{not} been this little unwillingness; but allow me to assure
you, that I have your respected mother's permission for this
address\index{address}.  You can hardly doubt the \textbf{purport}\index{purport}\stepcounter{footnote}\footnotetext{
\textbf{purport} \emph{tr.v.}  (per-PORT)
 the evident, but often false, meaning of a communication.} of my discourse\index{discourse},
however your natural delicacy\index{delicacy} may lead you to \textbf{dissemble}\index{dissemble}\stepcounter{footnote}\footnotetext{
\textbf{dissemble} \emph{v.}  (dihs-SEHM-bl)
to hide under a false appearance.}; my
attentions have been too marked to be mistaken.  Almost as soon
as I entered the house, I singled you out as the companion of
my future life.  But before I am run away with by my feelings on
this subject\index{subject}, perhaps it would be advisable for me to state my
reasons for marrying--and, moreover, for coming into Hertfordshire
with the design of selecting a wife, as I certainly did."

The idea of Mr. Collins, with all his solemn\index{solemn} composure\index{composure}, being
run away with by his feelings, made Elizabeth so near laughing,
that she could not use the short pause he allowed in any attempt
to stop him further, and he continued:

"My reasons for marrying are, first, that I think it a right
thing for every clergyman in easy circumstances (like myself) to
set the example of matrimony\index{matrimony} in his parish\index{parish}; secondly, that I am
convinced that it will add very greatly to my happiness; and
thirdly--which perhaps I ought to have mentioned earlier, that
it is the particular advice and recommendation of the very noble
lady whom I have the honour of calling patroness\index{patroness}.  Twice has
she condescended\index{condescended} to give me her opinion (unasked too!) on this
subject\index{subject}; and it was but the very Saturday night before I left
Hunsford--between our pools at quadrille\index{quadrille}, while Mrs. Jenkinson
was arranging Miss de Bourgh's footstool, that she said, 'Mr.
Collins, you must marry.  A clergyman like you must marry.
Choose properly, choose a gentlewoman for \emph{my} sake; and for
your \emph{own}, let\index{let} her be an active, useful sort of person, not brought
up high, but able to make a small income go a good way.  This is
my advice.  Find such a woman as soon as you can, bring her to
Hunsford, and I will visit her.'  Allow me, by the way, to
observe, my fair cousin, that I do not reckon the notice and
kindness of Lady Catherine de Bourgh as among the least of the
advantages in my power to offer.  You will find her manners
beyond anything I can describe; and your wit and vivacity\index{vivacity}, I
think, must be acceptable to her, especially when tempered with
the silence and respect which her rank will inevitably excite.
Thus much for my general intention in favour of matrimony\index{matrimony};
it remains to be told why my views were directed towards
Longbourn instead of my own neighbourhood, where I can
assure you there are many amiable\index{amiable} young women.  But the fact
is, that being, as I am, to inherit this estate after the death of
your honoured father (who, however, may live many years
longer), I could not satisfy myself without resolving to choose a
wife from among his daughters, that the loss to them might be as
little as possible, when the \textbf{melancholy}\index{melancholy}\stepcounter{footnote}\footnotetext{
\textbf{melancholy} \emph{n.}  (MEH-luhn-kaw-lee)
Sadness or depression.} event takes place--which,
however, as I have already said, may not be for several years.
This has been my motive, my fair cousin, and I flatter myself it
will not sink me in your esteem.  And now nothing remains but
for me but to assure you in the most animated language of the
violence of my affection.  To fortune I am perfectly indifferent\index{indifferent},
and shall make no demand of that nature on your father, since I
am well aware that it could not be complied with; and that one
thousand pounds in the four per cents, which will not be yours
till after your mother's decease, is all that you may ever be
entitled to.  On that head, therefore, I shall be uniformly silent;
and you may assure yourself that no ungenerous reproach\index{reproach} shall
ever pass my lips when we are married."

It was absolutely necessary to interrupt him now.

"You are too hasty, sir," she cried.  "You forget that I have
made no answer.  Let\index{Let} me do it without further loss of time.
Accept my thanks for the compliment you are paying me.  I am
very sensible of the honour of your proposals, but it is
impossible for me to do otherwise than to decline them."

"I am not now to learn," replied Mr. Collins, with a formal wave
of the hand, "that it is usual with young ladies to reject the
addresses of the man whom they secretly mean to accept, when
he first applies for their favour; and that sometimes the refusal
is repeated a second, or even a third time.  I am therefore by no
means discouraged by what you have just said, and shall hope to
lead you to the altar \textbf{ere}\index{ere}\stepcounter{footnote}\footnotetext{
\textbf{ere} \emph{prep.}  (AIR)
} long."

"Upon my word, sir," cried Elizabeth, "your hope is a rather
extraordinary one after my declaration.  I do assure you that I
am not one of those young ladies (if such young ladies there are)
who are so daring as to risk their happiness on the chance of
being asked a second time.  I am perfectly serious in my refusal.
You could not make \emph{me} happy, and I am convinced that I am
the last woman in the world who could make you so.  Nay, were
your friend Lady Catherine to know me, I am persuaded she
would find me in every respect ill qualified for the situation."

"Were it certain that Lady Catherine would think so," said Mr.
Collins very gravely\index{gravely}--"but I cannot imagine that her ladyship
would at all disapprove of you.  And you may be certain when I
have the honour of seeing her again, I shall speak in the very
highest terms of your modesty, \textbf{economy}\index{economy}\stepcounter{footnote}\footnotetext{
\textbf{economy} \emph{n.}  (ee-KAHN-noh-mee)
the efficient use of resources.}, and other amiable\index{amiable}
qualification."

"Indeed, Mr. Collins, all praise of me will be unnecessary.
You must give me leave to judge for myself, and pay me the
compliment of believing what I say.  I wish you very happy and
very rich, and by refusing your hand, do all in my power to
prevent your being otherwise.  In making me the offer, you must
have satisfied the delicacy\index{delicacy} of your feelings with regard to my
family, and may take possession of Longbourn estate whenever
it falls, without any self-reproach\index{reproach}.  This matter may be
considered, therefore, as finally settled."  And rising as she
thus spoke, she would have quitted the room, had Mr. Collins
not thus addressed\index{addressed} her:

"When I do myself the honour of speaking to you next on the
subject\index{subject}, I shall hope to receive a more favourable answer than
you have now given me; though I am far from accusing you of
cruelty at present, because I know it to be the established
custom of your sex to reject a man on the first application\index{application}, and
perhaps you have even now said as much to encourage my suit
as would be consistent with the true delicacy\index{delicacy} of the female
character."

"Really, Mr. Collins," cried Elizabeth with some warmth, "you
puzzle me exceedingly.  If what I have \textbf{hitherto}\index{hitherto}\stepcounter{footnote}\footnotetext{
\textbf{hitherto} \emph{th}  (HIH-dher-too)
before; until this time.} said can appear
to you in the form of encouragement, I know not how to express
my refusal in such a way as to convince you of its being one."

"You must give me leave to flatter myself, my dear cousin, that
your refusal of my addresses is merely words of course.  My
reasons for believing it are briefly these: It does not appear
to me that my hand is unworthy your acceptance, or that the
establishment I can offer would be any other than highly
desirable.  My situation in life, my connections with the family
of de Bourgh, and my relationship to your own, are circumstances
highly in my favour; and you should take it into further
consideration, that in spite of your \textbf{manifold}\index{manifold}\stepcounter{footnote}\footnotetext{
\textbf{manifold} \emph{adj.}  (MAN-ih-fohld)
Many and varied.} attractions, it is
by no means certain that another offer of marriage may ever be
made you.  Your portion is unhappily so small that it will in
all likelihood undo the effects of your loveliness and amiable\index{amiable}
qualifications.  As I must therefore conclude that you are not
serious in your rejection of me, I shall choose to attribute it
to your wish of increasing my love by suspense, according to the
usual practice of elegant females."

"I do assure you, sir, that I have no pretensions\index{pretensions} whatever to that
kind of elegance which consists in tormenting a respectable man.
I would rather be paid the compliment of being believed sincere.
I thank you again and again for the honour you have done me in
your proposals, but to accept them is absolutely impossible.  My
feelings in every respect forbid it.  Can I speak plainer?  Do not
consider me now as an elegant female, intending to plague\index{plague} you,
but as a rational\index{rational} creature, speaking the truth from her heart."

"You are uniformly charming!" cried he, with an air of awkward
gallantry\index{gallantry}; "and I am persuaded that when \textbf{sanctioned}\index{sanctioned}\stepcounter{footnote}\footnotetext{
\textbf{sanction} \emph{n.} \textbf{-ed}  (SANK-shun)
given authoritative approval.} by the
express authority of both your excellent parents, my proposals
will not fail of being acceptable."

To such \textbf{perseverance}\index{perseverance}\stepcounter{footnote}\footnotetext{
\textbf{perseverance} \emph{n.}  (per-suh-VEE-ruhnse)
steadfast commitment to an action or toward a goal.} in wilful self-deception Elizabeth would
make no reply, and immediately and in silence withdrew;
determined, if he persisted in considering her repeated
refusals as flattering encouragement, to apply to her father,
whose negative might be uttered in such a manner as to be
decisive, and whose behavior at least could not be mistaken
for the affectation\index{affectation} and \textbf{coquetry}\index{coquetry}\stepcounter{footnote}\footnotetext{
\textbf{coquetry} \emph{n.}  (KOH-keht-ree)
playful behavior intended to arouse sexual interest.} of an elegant female.



\chapter{20}


Mr. Collins was not left long to the silent contemplation\index{contemplation} of his
successful love; for Mrs. Bennet, having \textbf{dawdled}\index{dawdled}\stepcounter{footnote}\footnotetext{
\textbf{dawdle} \emph{v.} \textbf{-ed}  (DAWD-uhl)
 To take more time than necessary.} about in the
\textbf{vestibule}\index{vestibule}\stepcounter{footnote}\footnotetext{
\textbf{vestibule} \emph{n.}  (VES-tih-BYOOL)
a large entrance or reception room or area.} to watch for the end of the conference, no sooner saw
Elizabeth open the door and with quick step pass her towards
the staircase, than she entered the breakfast-room, and
congratulated both him and herself in warm terms on the happy
prospect\index{prospect} or their nearer connection.  Mr. Collins received and
returned these \textbf{felicitations}\index{felicitations}\stepcounter{footnote}\footnotetext{
\textbf{felicitation} \emph{n.}  (fuh-lih-sih-TAY-SHUN)
Congratulations.} with equal pleasure, and then
proceeded to relate the particulars of their interview, with the
result of which he trusted he had every reason to be satisfied,
since the refusal which his cousin had \textbf{steadfastly}\index{steadfastly}\stepcounter{footnote}\footnotetext{
\textbf{steadfast} \emph{adj.} \textbf{-ly}  (STEHD-fast)
with resolute determination.} given him
would naturally flow from her bashful modesty and the genuine
delicacy\index{delicacy} of her character.

This information, however, startled Mrs. Bennet; she would
have been glad to be equally satisfied that her daughter had
meant to encourage him by protesting against his proposals,
but she dared not believe it, and could not help saying so.

"But, depend upon it, Mr. Collins," she added, "that Lizzy shall
be brought to reason.  I will speak to her about it directly.
She is a very \textbf{headstrong}\index{headstrong}\stepcounter{footnote}\footnotetext{
\textbf{headstrong} \emph{adj.}  (HEHD-strawng)
habitually disposed to disobedience and opposition.}, foolish girl, and does not know her
own interest but I will \emph{make} her know it."

"Pardon me for interrupting you, madam," cried Mr. Collins;
"but if she is really headstrong\index{headstrong} and foolish, I know not whether
she would altogether be a very desirable wife to a man in my
situation, who naturally looks for happiness in the marriage
state.  If therefore she actually persists in rejecting my suit,
perhaps it were better not to force her into accepting me,
because if liable to such defects of temper, she could not
contribute much to my felicity\index{felicity}."

"Sir, you quite misunderstand me," said Mrs. Bennet, alarmed.
"Lizzy is only headstrong\index{headstrong} in such matters as these.  In everything
else she is as good-natured a girl as ever lived.  I will go
directly to Mr. Bennet, and we shall very soon settle it with her,
I am sure."

She would not give him time to reply, but hurrying instantly to
her husband, called out as she entered the library, "Oh! Mr.
Bennet, you are wanted immediately; we are all in an uproar.
You must come and make Lizzy marry Mr. Collins, for she vows
she will not have him, and if you do not make haste he will
change his mind and not have \emph{her}."

Mr. Bennet raised his eyes from his book as she entered, and
fixed them on her face with a calm unconcern which was not in
the least altered by her communication.

"I have not the pleasure of understanding you," said he, when
she had finished her speech.  "Of what are you talking?"

"Of Mr. Collins and Lizzy.  Lizzy declares she will not have
Mr. Collins, and Mr. Collins begins to say that he will not
have Lizzy."

"And what am I to do on the occasion?  It seems an hopeless
business."

"Speak to Lizzy about it yourself.  Tell her that you insist upon
her marrying him."

"Let\index{Let} her be called down.  She shall hear my opinion."

Mrs. Bennet rang the bell, and Miss Elizabeth was summoned\index{summoned} to
the library.

"Come here, child," cried her father as she appeared.  "I have
sent for you on an affair of importance.  I understand that Mr.
Collins has made you an offer of marriage.  Is it true?"  Elizabeth
replied that it was.  "Very well--and this offer of marriage you
have refused?"

"I have, sir."

"Very well.  We now come to the point.  Your mother insists
upon your accepting it.  Is it not so, Mrs. Bennet?"

"Yes, or I will never see her again."

"An unhappy alternative is before you, Elizabeth.  From this day
you must be a stranger to one of your parents.  Your mother will
never see you again if you do \emph{not} marry Mr. Collins, and I will
never see you again if you \emph{do}."

Elizabeth could not but smile at such a conclusion of such a
beginning, but Mrs. Bennet, who had persuaded herself that her
husband regarded the affair as she wished, was excessively
disappointed.

"What do you mean, Mr. Bennet, in talking this way?  You
promised me to \emph{insist} upon her marrying him."

"My dear," replied her husband, "I have two small favours to
request.  First, that you will allow me the free use of my
understanding on the present occasion; and secondly, of my
room.  I shall be glad to have the library to myself as soon as
may be."

Not yet, however, in spite of her disappointment in her husband,
did Mrs. Bennet give up the point.  She talked to Elizabeth again
and again; \textbf{coaxed}\index{coaxed}\stepcounter{footnote}\footnotetext{
\textbf{coax} \emph{v.} \textbf{-ed}  (KOAX)
 to persuade, cajole, or convince by flattery.} and threatened her by turns.  She \textbf{endeavoured}\index{endeavoured}\stepcounter{footnote}\footnotetext{
\textbf{} \emph{}  ()
}\index{endeavoured}
to secure Jane in her interest; but Jane, with all possible
mildness, declined interfering; and Elizabeth, sometimes with
real earnestness, and sometimes with playful \textbf{gaiety}\index{gaiety}\stepcounter{footnote}\footnotetext{
\textbf{gaiety} \emph{n.}  (GAY-eh-tee)
a merry, festive feeling.}, replied to
her attacks.  Though her manner varied, however, her determination
never did.

Mr. Collins, meanwhile, was meditating in solitude\index{solitude} on what had
passed.  He thought too well of himself to comprehend\index{comprehend} on what
motives his cousin could refuse him; and though his pride was
hurt, he suffered in no other way.  His regard for her was quite
imaginary; and the possibility of her deserving her mother's
reproach\index{reproach} prevented his feeling any regret.

While the family were in this confusion, Charlotte Lucas came to
spend the day with them.  She was met in the vestibule\index{vestibule} by Lydia,
who, flying to her, cried in a half whisper, "I am glad you are
come, for there is such fun here!  What do you think has
happened this morning?  Mr. Collins has made an offer to Lizzy,
and she will not have him."

Charlotte hardly had time to answer, before they were joined by
Kitty, who came to tell the same news; and no sooner had they
entered the breakfast-room, where Mrs. Bennet was alone, than
she likewise began on the subject\index{subject}, calling on Miss Lucas for her
compassion, and entreating\index{entreating} her to persuade her friend Lizzy to
comply\index{comply} with the wishes of all her family.  "Pray do, my dear
Miss Lucas," she added in a melancholy\index{melancholy} tone, "for nobody is on
my side, nobody takes part with me.  I am cruelly used, nobody
feels for my poor nerves."

Charlotte's reply was spared\index{spared} by the entrance of Jane and
Elizabeth.

"Aye, there she comes," continued Mrs. Bennet, "looking as
unconcerned as may be, and caring no more for us than if we
were at York, provided she can have her own way.  But I tell
you, Miss Lizzy--if you take it into your head to go on refusing
every offer of marriage in this way, you will never get a husband
at all--and I am sure I do not know who is to maintain you
when your father is dead.  I shall not be able to keep you--and
so I warn you.  I have done with you from this very day.  I told
you in the library, you know, that I should never speak to you
again, and you will find me as good as my word.  I have no
pleasure in talking to undutiful children.  Not that I have much
pleasure, indeed, in talking to anybody.  People who suffer as
I do from nervous complaints can have no great inclination\index{inclination} for
talking.  Nobody can tell what I suffer!  But it is always so.
Those who do not complain are never pitied."

Her daughters listened in silence to this effusion\index{effusion}, sensible
that any attempt to reason with her or soothe her would only
increase the irritation.  She talked on, therefore, without
interruption from any of them, till they were joined by Mr.
Collins, who entered the room with an air more stately\index{stately} than
usual, and on perceiving whom, she said to the girls, "Now, I do
insist upon it, that you, all of you, hold your tongues, and
let\index{let} me and Mr. Collins have a little conversation together."

Elizabeth passed quietly out of the room, Jane and Kitty
followed, but Lydia stood her ground, determined to hear all she
could; and Charlotte, detained first by the civility\index{civility} of Mr. Collins,
whose inquiries after herself and all her family were very minute\index{minute},
and then by a little curiosity, satisfied herself with walking to
the window and pretending not to hear.  In a \textbf{doleful}\index{doleful}\stepcounter{footnote}\footnotetext{
\textbf{doleful} \emph{adj.}  (DOAL-fuhl)
filled with or evoking sadness.} voice Mrs.
Bennet began the projected conversation: "Oh! Mr. Collins!"

"My dear madam," replied he, "let\index{let} us be for ever silent on this
point.  Far be it from me," he presently continued, in a voice that
marked his displeasure, "to resent the behaviour of your daughter.
\textbf{Resignation}\index{resignation}\stepcounter{footnote}\footnotetext{
\textbf{resignation} \emph{n.}  (reh-zihg-NAY-shun)
acceptance of despair; surrender.} to inevitable evils is the evil duty of us all; the
peculiar\index{peculiar} duty of a young man who has been so fortunate as I
have been in early preferment; and I trust I am \textbf{resigned}\index{resigned}\stepcounter{footnote}\footnotetext{
\textbf{re-sign} \emph{tr.v.} \textbf{-ed}  (rih-ZYN)
}.
Perhaps not the less so from feeling a doubt of my positive
happiness had my fair cousin honoured me with her hand; for I
have often observed that resignation\index{resignation} is never so perfect as when
the blessing denied begins to lose somewhat of its value in our
estimation.  You will not, I hope, consider me as showing any
disrespect to your family, my dear madam, by thus withdrawing
my pretensions\index{pretensions} to your daughter's favour, without having paid
yourself and Mr. Bennet the compliment of requesting you to
\textbf{interpose}\index{interpose}\stepcounter{footnote}\footnotetext{
\textbf{interpose} \emph{v.}  (IHN-ter-pohz)
to insert between other parts.} your authority in my behalf.  My conduct may, I fear,
be objectionable in having accepted my dismission from your
daughter's lips instead of your own.  But we are all liable to
error.  I have certainly meant well through the whole affair.  My
object has been to secure an amiable\index{amiable} companion for myself, with
due consideration for the advantage of all your family, and if
my \emph{manner} has been at all reprehensible\index{reprehensible}, I here beg leave to
apologise."



\chapter{21}


The discussion of Mr. Collins's offer was now nearly at an end,
and Elizabeth had only to suffer from the uncomfortable feelings
necessarily attending it, and occasionally from some \textbf{peevish}\index{peevish}\stepcounter{footnote}\footnotetext{
\textbf{peevish} \emph{adj.}  (PEE-vihsh)
 easily irritated or annoyed.}
allusions\index{allusions} of her mother.  As for the gentleman himself, \emph{his}
feelings were chiefly expressed, not by embarrassment or
\textbf{dejection}\index{dejection}\stepcounter{footnote}\footnotetext{
\textbf{dejection} \emph{n.}  (dih-JEHK-shun)
a state of melancholy depression; feeling down.}, or by trying to avoid her, but by stiffness of manner
and resentful silence.  He scarcely ever spoke to her, and the
\textbf{assiduous}\index{assiduous}\stepcounter{footnote}\footnotetext{
\textbf{assiduous} \emph{adj.} \textbf{-ly}  (uh-SIHD-yoo-uhs)
marked by care and persistent effort.} attentions which he had been so sensible of himself
were transferred for the rest of the day to Miss Lucas, whose
civility\index{civility} in listening to him was a seasonable relief to them all,
and especially to her friend.

The morrow produced no \textbf{abatement}\index{abatement}\stepcounter{footnote}\footnotetext{
\textbf{abatement} \emph{n.}  (uh-BATE-muhnt)
an interruption in the intensity or amount of something.} of Mrs. Bennet's ill-humour
or ill health.  Mr. Collins was also in the same state of angry
pride.  Elizabeth had hoped that his resentment\index{resentment} might shorten his
visit, but his plan did not appear in the least affected by it.
He was always to have gone on Saturday, and to Saturday he meant
to stay.

After breakfast, the girls walked to Meryton to inquire if Mr.
Wickham were returned, and to \textbf{lament}\index{lament}\stepcounter{footnote}\footnotetext{
\textbf{lament} \emph{v.} \textbf{-ing}  (luh-MEHNT)
a cry of sorrow and grief.} over his absence from
the Netherfield ball.  He joined them on their entering the town,
and attended them to their aunt's where his regret and vexation\index{vexation},
and the concern of everybody, was well talked over.  To
Elizabeth, however, he voluntarily acknowledged that the
necessity of his absence \emph{had} been \textbf{self-imposed}\index{self-imposed}\stepcounter{footnote}\footnotetext{
\textbf{self-imposed} \emph{adj.}  (self-ihm-POHZD)
voluntarily assumed or endured.}\index{imposed}.

"I found," said he, "as the time drew near that I had better not
meet Mr. Darcy; that to be in the same room, the same party
with him for so many hours together, might be more than I could
bear, and that scenes might arise unpleasant to more than
myself."

She highly approved his forbearance\index{forbearance}, and they had leisure for a
full discussion of it, and for all the commendation\index{commendation} which they
civilly \textbf{bestowed}\index{bestowed}\stepcounter{footnote}\footnotetext{
\textbf{bestow} \emph{tr.v.} \textbf{-ed}  (bih-STOU)
 to give something formally or officially.} on each other, as Wickham and another officer
walked back with them to Longbourn, and during the walk he
particularly attended to her.  His accompanying them was a
double advantage; she felt all the compliment it offered to
herself, and it was most acceptable as an occasion of introducing
him to her father and mother.

Soon after their return, a letter was delivered to Miss Bennet;
it came from Netherfield.  The envelope contained a sheet of
elegant, little, hot-pressed paper, well covered with a lady's fair,
flowing hand; and Elizabeth saw her sister's countenance\index{countenance} change
as she read it, and saw her dwelling intently on some particular
passages.  Jane recollected herself soon, and putting the letter
away, tried to join with her usual cheerfulness in the general
conversation; but Elizabeth felt an anxiety on the subject\index{subject} which
drew off her attention even from Wickham; and no sooner had
he and he companion taken leave, than a glance from Jane
invited her to follow her upstairs.  When they had gained their
own room, Jane, taking out the letter, said:

"This is from Caroline Bingley; what it contains has surprised me
a good deal.  The whole party have left Netherfield by this time,
and are on their way to town--and without any intention of coming
back again.  You shall hear what she says."

She then read the first sentence aloud, which comprised the
information of their having just resolved to follow their brother
to town directly, and of their meaning to dine in Grosvenor
Street, where Mr. Hurst had a house.  The next was in these
words: "I do not pretend to regret anything I shall leave in
Hertfordshire, except your society, my dearest friend; but we
will hope, at some future period, to enjoy many returns of that
delightful \textbf{intercourse}\index{intercourse}\stepcounter{footnote}\footnotetext{
\textbf{intercourse} \emph{n.}  (IHN-ter-kors)
communication between individuals.} we have known, and in the meanwhile may
lessen the pain of separation by a very frequent and most
unreserved correspondence.  I depend on you for that."  To
these highflown expressions Elizabeth listened with all the
insensibility of distrust; and though the suddenness of their
removal surprised her, she saw nothing in it really to lament\index{lament};
it was not to be supposed that their absence from Netherfield
would prevent Mr. Bingley's being there; and as to the loss of
their society, she was persuaded that Jane must cease to regard
it, in the enjoyment of his.

"It is unlucky," said she, after a short pause, "that you should
not be able to see your friends before they leave the country.
But may we not hope that the period of future happiness to
which Miss Bingley looks forward\index{forward} may arrive earlier than she is
aware, and that the delightful intercourse\index{intercourse} you have known as
friends will be renewed with yet greater satisfaction as sisters?
Mr. Bingley will not be detained in London by them."

"Caroline decidedly says that none of the party will return into
Hertfordshire this winter.  I will read it to you:"

"When my brother left us yesterday, he imagined that the
business which took him to London might be concluded in three
or four days; but as we are certain it cannot be so, and at the
same time convinced that when Charles gets to town he will be
in no hurry to leave it again, we have determined on following
him thither, that he may not be obliged to spend his vacant hours
in a comfortless hotel.  Many of my acquaintances are already
there for the winter; I wish that I could hear that you, my dearest
friend, had any intention of making one of the crowd--but of
that I despair.  I sincerely hope your Christmas in Hertfordshire
may \textbf{abound}\index{abound}\stepcounter{footnote}\footnotetext{
\textbf{abound} \emph{intr.v.}  (uh-BOWND)
 to exist in large quantities.} in the gaieties which that season generally brings,
and that your \textbf{beaux}\index{beaux}\stepcounter{footnote}\footnotetext{
\textbf{beau} \emph{n.} \textbf{-x}  (BOH)
 A boyfriend.} will be so numerous as to prevent your
feeling the loss of the three of whom we shall deprive you."

"It is evident by this," added Jane, "that he comes back no more
this winter."

"It is only evident that Miss Bingley does not mean that he
\emph{should}."

"Why will you think so?  It must be his own doing.  He is his
own master.  But you do not know \emph{all}.  I \emph{will} read you the
passage which particularly hurts me.  I will have no reserves
from \emph{you}."

"Mr. Darcy is impatient to see his sister; and, to confess the
truth, \emph{we} are scarcely less eager to meet her again.  I really do
not think Georgiana Darcy has her equal for beauty, elegance,
and accomplishments; and the affection she inspires in Louisa
and myself is heightened into something still more interesting,
from the hope we dare entertain of her being hereafter our
sister.  I do not know whether I ever before mentioned to you
my feelings on this subject\index{subject}; but I will not leave the country
without confiding them, and I trust you will not esteem them
unreasonable.  My brother admires her greatly already; he will
have frequent opportunity now of seeing her on the most
intimate footing; her relations all wish the connection as much as
his own; and a sister's partiality\index{partiality} is not misleading me, I think,
when I call Charles most capable of engaging any woman's
heart.  With all these circumstances to favour an attachment, and
nothing to prevent it, am I wrong, my dearest Jane, in indulging
the hope of an event which will secure the happiness of so many?"

"What do you think of \emph{this} sentence, my dear Lizzy?" said
Jane as she finished it.  "Is it not clear enough?  Does it not
expressly declare that Caroline neither expects nor wishes me to
be her sister; that she is perfectly convinced of her brother's
indifference\index{indifference}; and that if she suspects the nature of my feelings
for him, she means (most kindly!) to put me on my guard?  Can
there be any other opinion on the subject\index{subject}?"

"Yes, there can; for mine is totally different.  Will you hear it?"

"Most willingly."

"You shall have it in a few words.  Miss Bingley sees that her
brother is in love with you, and wants him to marry Miss Darcy.
She follows him to town in hope of keeping him there, and tries
to persuade you that he does not care about you."

Jane shook her head.

"Indeed, Jane, you ought to believe me.  No one who has ever
seen you together can doubt his affection.  Miss Bingley, I am
sure, cannot.  She is not such a simpleton\index{simpleton}.  Could she have seen
half as much love in Mr. Darcy for herself, she would have
ordered her wedding clothes.  But the case is this: We are not
rich enough or grand enough for them; and she is the more
anxious to get Miss Darcy for her brother, from the notion that
when there has been \emph{one} \textbf{intermarriage}\index{intermarriage}\stepcounter{footnote}\footnotetext{
\textbf{intermarry} \emph{intr.v.}  (ihn-ter-MAIR-ree)
To marry a member of another group.}, she may have less
trouble in achieving a second; in which there is certainly some
\textbf{ingenuity}\index{ingenuity}\stepcounter{footnote}\footnotetext{
\textbf{ingenuity} \emph{n.}  (ihn-jeh-NYOO-ih-tee)
cleverness; the power of creative imagination.}, and I dare say it would succeed, if Miss de Bourgh
were out of the way.  But, my dearest Jane, you cannot seriously
imagine that because Miss Bingley tells you her brother greatly
admires Miss Darcy, he is in the smallest degree less sensible
of \emph{your} merit than when he took leave of you on Tuesday, or
that it will be in her power to persuade him that, instead of
being in love with you, he is very much in love with her friend."

"If we thought alike of Miss Bingley," replied Jane, "your
representation of all this might make me quite easy.  But I know
the foundation is unjust.  Caroline is incapable of wilfully
deceiving anyone; and all that I can hope in this case is that
she is deceiving herself."

"That is right.  You could not have started a more happy idea,
since you will not take comfort in mine.  Believe her to be
deceived, by all means.  You have now done your duty by her,
and must fret\index{fret} no longer."

"But, my dear sister, can I be happy, even supposing the best, in
accepting a man whose sisters and friends are all wishing him to
marry elsewhere?"

"You must decide for yourself," said Elizabeth; "and if, upon
mature deliberation\index{deliberation}, you find that the misery of \textbf{disobliging}\index{disobliging}\stepcounter{footnote}\footnotetext{
\textbf{disoblige} \emph{tr.v.} \textbf{-ing}  (dihs-uh-BLYJ)
to be intentionally unaccommodating.} his
two sisters is more than equivalent to the happiness of being his
wife, I advise you by all means to refuse him."

"How can you talk so?" said Jane, faintly smiling.  "You must
know that though I should be exceedingly grieved\index{grieved} at their
\textbf{disapprobation}\index{disapprobation}\stepcounter{footnote}\footnotetext{
\textbf{disapprobation} \emph{n.}  (dihs-ap-proh-BAY-shun)
an expression of strong disapproval.}, I could not hesitate."

"I did not think you would; and that being the case, I cannot
consider your situation with much compassion."

"But if he returns no more this winter, my choice will never be
required.  A thousand things may arise in six months!"

The idea of his returning no more Elizabeth treated with the
utmost contempt\index{contempt}.  It appeared to her merely the suggestion of
Caroline's interested wishes, and she could not for a moment
suppose that those wishes, however openly or artfully spoken,
could influence a young man so totally independent of everyone.

She represented to her sister as forcibly as possible what she
felt on the subject\index{subject}, and had soon the pleasure of seeing its
happy effect.  Jane's temper was not \textbf{desponding}\index{desponding}\stepcounter{footnote}\footnotetext{
\textbf{despond} \emph{intr.v.} \textbf{-ing}  (dih-SPAHND)
to become discouraged, gloomy, or feel let down.}, and she was
gradually led to hope, though the diffidence\index{diffidence} of affection
sometimes overcame the hope, that Bingley would return to
Netherfield and answer every wish of her heart.

They agreed that Mrs. Bennet should only hear of the departure
of the family, without being alarmed on the score of the
gentleman's conduct; but even this partial\index{partial} communication gave
her a great deal of concern, and she \textbf{bewailed}\index{bewailed}\stepcounter{footnote}\footnotetext{
\textbf{bewail} \emph{tr.v.} \textbf{-ed}  (bee-WAYL)
to lament or cry intensely over.} it as exceedingly
unlucky that the ladies should happen to go away just as they
were all getting so intimate together.  After lamenting\index{lamenting} it,
however, at some length, she had the consolation\index{consolation} that Mr.
Bingley would be soon down again and soon dining at Longbourn,
and the conclusion of all was the comfortable declaration,
that though he had been invited only to a family dinner, she
would take care to have two full courses.



\chapter{22}


The Bennets were engaged to dine with the Lucases and again
during the chief of the day was Miss Lucas so kind as to listen
to Mr. Collins.  Elizabeth took an opportunity of thanking her.
"It keeps him in good humour," said she, "and I am more obliged
to you than I can express."  Charlotte assured her friend of
her satisfaction in being useful, and that it amply repaid her
for the little sacrifice of her time.  This was very amiable\index{amiable},
but Charlotte's kindness extended farther than Elizabeth had
any conception of; its object was nothing else than to secure
her from any return of Mr. Collins's addresses, by engaging
them towards herself.  Such was Miss Lucas's scheme; and
appearances were so favourable, that when they parted at night,
she would have felt almost secure of success if he had not been
to leave Hertfordshire so very soon.  But here she did injustice\index{injustice}
to the fire and independence of his character, for it led
him to escape out of Longbourn House the next morning with
admirable slyness, and hasten to Lucas Lodge to throw himself
at her feet.  He was anxious to avoid the notice of his cousins,
from a conviction\index{conviction} that if they saw him depart, they could not
fail to conjecture\index{conjecture} his design, and he was not willing to have
the attempt known till its success might be known likewise; for
though feeling almost secure, and with reason, for Charlotte had
been tolerably encouraging, he was comparatively\index{comparatively} \textbf{diffident}\index{diffident}\stepcounter{footnote}\footnotetext{
\textbf{diffident} \emph{adj.}  (DIHF-fih-dehnt)
lacking self-confidence.} since
the adventure of Wednesday.  His reception, however, was of
the most flattering kind.  Miss Lucas perceived\index{perceived} him from an
upper window as he walked towards the house, and instantly set
out to meet him accidentally in the lane.  But little had she
dared to hope that so much love and \textbf{eloquence}\index{eloquence}\stepcounter{footnote}\footnotetext{
\textbf{eloquence} \emph{n.}  (EH-loh-kwehnse)
powerful and effective language.} awaited her there.

In as short a time as Mr. Collins's long speeches would allow,
everything was settled between them to the satisfaction of both;
and as they entered the house he earnestly\index{earnestly} entreated\index{entreated} her to name
the day that was to make him the happiest of men; and though
such a solicitation must be waived for the present, the lady felt
no inclination\index{inclination} to trifle with his happiness.  The stupidity with
which he was favoured by nature must guard his courtship from
any charm that could make a woman wish for its continuance;
and Miss Lucas, who accepted him solely\index{solely} from the pure and
disinterested\index{disinterested} desire of an establishment, cared not how soon
that establishment were gained.

Sir William and Lady Lucas were speedily applied to for their
consent; and it was bestowed\index{bestowed} with a most joyful alacrity\index{alacrity}.  Mr.
Collins's present circumstances made it a most eligible match for
their daughter, to whom they could give little fortune; and his
prospects\index{prospects} of future wealth were exceedingly fair.  Lady Lucas
began directly to calculate, with more interest than the
matter had ever excited before, how many years longer Mr.
Bennet was likely to live; and Sir William gave it as his decided
opinion, that whenever Mr. Collins should be in possession of the
Longbourn estate, it would be highly expedient that both he and
his wife should make their appearance at St. James's.  The whole
family, in short, were properly overjoyed on the occasion.
The younger girls formed hopes of \emph{coming} \emph{out} a year or two
sooner than they might otherwise have done; and the boys were
relieved from their \textbf{apprehension}\index{apprehension}\stepcounter{footnote}\footnotetext{
\textbf{apprehension} \emph{n.}  (ap-rih-HEHN-shun)
fearful expectation or anticipation.} of Charlotte's dying an old
maid.  Charlotte herself was tolerably composed.  She had
gained her point, and had time to consider of it.  Her reflections
were in general satisfactory.  Mr. Collins, to be sure, was
neither sensible nor agreeable; his society was \textbf{irksome}\index{irksome}\stepcounter{footnote}\footnotetext{
\textbf{irksome} \emph{adj.}  (ERK-suhm)
Something that is annoying, boring, bothersome or tedious.}, and his
attachment to her must be imaginary.  But still he would be her
husband.  Without thinking highly either of men or matrimony\index{matrimony},
marriage had always been her object; it was the only \textbf{provision}\index{provision}\stepcounter{footnote}\footnotetext{
\textbf{provision} \emph{n.}  (proh-VIH-zhun)
a stipulated condition.}
for well-educated young women of small fortune, and however
uncertain of giving happiness, must be their pleasantest
\textbf{preservative}\index{preservative}\stepcounter{footnote}\footnotetext{
\textbf{preservative} \emph{adj.}  (prih-ZER-vuh-tihv)
} from want.  This preservative\index{preservative} she had now
obtained; and at the age of twenty-seven, without having ever
been handsome, she felt all the good luck of it.  The least
agreeable circumstance in the business was the surprise it must
occasion to Elizabeth Bennet, whose friendship she valued
beyond that of any other person.  Elizabeth would wonder, and
probably would blame her; and though her resolution\index{resolution} was not to
be shaken, her feelings must be hurt by such a disapprobation\index{disapprobation}.
She resolved to give her the information herself, and therefore
charged Mr. Collins, when he returned to Longbourn to dinner,
to drop no hint of what had passed before any of the family.  A
promise of secrecy was of course very dutifully given, but it
could not be kept without difficulty; for the curiosity excited
by his long absence burst forth in such very direct questions on
his return as required some ingenuity\index{ingenuity} to evade, and he was at the
same time exercising great self-denial, for he was longing to
publish his prosperous love.

As he was to begin his journey too early on the morrow to see
any of the family, the ceremony of leave-taking was performed
when the ladies moved for the night; and Mrs. Bennet, with
great politeness and cordiality, said how happy they should be
to see him at Longbourn again, whenever his engagements might
allow him to visit them.

"My dear madam," he replied, "this invitation is particularly
gratifying, because it is what I have been hoping to receive; and
you may be very certain that I shall \textbf{avail}\index{avail}\stepcounter{footnote}\footnotetext{
\textbf{avail} \emph{v.}  (uh-VAYL)
 to be useful to; to help.} myself of it as soon
as possible."

They were all astonished; and Mr. Bennet, who could by no
means wish for so speedy a return, immediately said:

"But is there not danger of Lady Catherine's disapprobation\index{disapprobation}
here, my good sir?  You had better neglect your relations than
run the risk of offending your patroness\index{patroness}."

"My dear sir," replied Mr. Collins," I am particularly obliged
to you for this friendly caution, and you may depend upon my not
taking so material a step without her ladyship's \textbf{concurrence}\index{concurrence}\stepcounter{footnote}\footnotetext{
\textbf{concurrence} \emph{n.}  (kuhn-KER-rense)
}."

"You cannot be too much upon your guard.  Risk anything
rather than her displeasure; and if you find it likely to be raised
by your coming to us again, which I should think exceedingly
probable\index{probable}, stay quietly at home, and be satisfied that \emph{we} shall
take no offence."

"Believe me, my dear sir, my gratitude is warmly excited by
such affectionate attention; and depend upon it, you will speedily
receive from me a letter of thanks for this, and for every other
mark of your regard during my stay in Hertfordshire.  As for my
fair cousins, though my absence may not be long enough to
render\index{render} it necessary, I shall now take the liberty of wishing them
health and happiness, not excepting my cousin Elizabeth."

With proper civilities the ladies then withdrew; all of them
equally surprised that he \textbf{meditated}\index{meditated}\stepcounter{footnote}\footnotetext{
\textbf{} \emph{}  ()
} a quick return.  Mrs. Bennet
wished to understand by it that he thought of paying his
addresses to one of her younger girls, and Mary might have been
prevailed\index{prevailed} on to accept him.  She rated his abilities much higher
than any of the others; there was a solidity\index{solidity} in his reflections
which often struck her, and though by no means so clever as
herself, she thought that if encouraged to read and improve
himself by such an example as hers, he might become a very
agreeable companion.  But on the following morning, every
hope of this kind was done away.  Miss Lucas called soon after
breakfast, and in a private\index{private} conference with Elizabeth related the
event of the day before.

The possibility of Mr. Collins's fancying herself in love with her
friend had once occurred to Elizabeth within the last day or two;
but that Charlotte could encourage him seemed almost as far
from possibility as she could encourage him herself, and her
astonishment was consequently so great as to overcome at first
the bounds of decorum\index{decorum}, and she could not help crying out:

"Engaged to Mr. Collins!  My dear Charlotte--impossible!"

The steady\index{steady} countenance\index{countenance} which Miss Lucas had commanded in
telling her story, gave way to a momentary confusion here on
receiving so direct a reproach\index{reproach}; though, as it was no more than
she expected, she soon regained her composure\index{composure}, and calmly
replied:

"Why should you be surprised, my dear Eliza?  Do you think it
incredible that Mr. Collins should be able to procure\index{procure} any
woman's good opinion, because he was not so happy as to
succeed with you?"

But Elizabeth had now recollected herself, and making a strong
effort for it, was able to assure with tolerable\index{tolerable} firmness that the
prospect\index{prospect} of their relationship was highly grateful to her, and
that she wished her all imaginable happiness.

"I see what you are feeling," replied Charlotte.  "You must be
surprised, very much surprised--so lately as Mr. Collins was
wishing to marry you.  But when you have had time to think it
over, I hope you will be satisfied with what I have done.  I am
not romantic, you know; I never was.  I ask only a comfortable
home; and considering Mr. Collins's character, connection, and
situation in life, I am convinced that my chance of happiness
with him is as fair as most people can boast on entering the
marriage state."

Elizabeth quietly answered "Undoubtedly;" and after an
awkward pause, they returned to the rest of the family.
Charlotte did not stay much longer, and Elizabeth was then left
to reflect on what she had heard.  It was a long time before she
became at all \textbf{reconciled}\index{reconciled}\stepcounter{footnote}\footnotetext{
\textbf{reconcile} \emph{v.} \textbf{-ed}  (REH-kuhn-syl)
come to terms; to reestablish a relationship.} to the idea of so unsuitable a match.
The strangeness of Mr. Collins's making two offers of marriage
within three days was nothing in comparison of his being now
accepted.  She had always felt that Charlotte's opinion of
matrimony\index{matrimony} was not exactly like her own, but she had not
supposed it to be possible that, when called into action, she
would have sacrificed every better feeling to worldly advantage.
Charlotte the wife of Mr. Collins was a most humiliating picture!
And to the pang of a friend disgracing herself and sunk in her
esteem, was added the distressing conviction\index{conviction} that it was
impossible for that friend to be tolerably happy in the lot
she had chosen.



\chapter{23}


Elizabeth was sitting with her mother and sisters, reflecting on
what she had heard, and doubting whether she was authorised to
mention it, when Sir William Lucas himself appeared, sent by his
daughter, to announce her engagement to the family.  With many
compliments to them, and much self-gratulation on the prospect\index{prospect}
of a connection between the houses, he unfolded the matter--to
an audience not merely wondering, but \textbf{incredulous}\index{incredulous}\stepcounter{footnote}\footnotetext{
\textbf{incredulous} \emph{adj.}  (ihn-KREH-dyoo-luhs)
not willing to believe; skeptical.}; for Mrs.
Bennet, with more perseverance\index{perseverance} than politeness, protested he
must be entirely mistaken; and Lydia, always unguarded and
often uncivil\index{uncivil}, \textbf{boisterously}\index{boisterously}\stepcounter{footnote}\footnotetext{
\textbf{boisterous} \emph{adj.} \textbf{-ly}  (BOYS-truhs)
} exclaimed:

"Good Lord! Sir William, how can you tell such a story?  Do not
you know that Mr. Collins wants to marry Lizzy?"

Nothing less than the complaisance\index{complaisance} of a courtier could have
borne\index{borne} without anger such treatment; but Sir William's good
breeding carried him through it all; and though he begged leave
to be positive as to the truth of his information, he listened
to all their impertinence\index{impertinence} with the most \textbf{forbearing}\index{forbearing}\stepcounter{footnote}\footnotetext{
\textbf{forbear} \emph{v.}  (for-BAIR)
to restrain oneself or to refrain from.} courtesy.

Elizabeth, feeling it incumbent\index{incumbent} on her to relieve him from so
unpleasant a situation, now put herself forward\index{forward} to confirm his
account, by mentioning her prior knowledge of it from Charlotte
herself; and endeavoured\index{endeavoured} to put a stop to the exclamations of her
mother and sisters by the earnestness of her congratulations to
Sir William, in which she was readily\index{readily} joined by Jane, and by
making a variety of remarks on the happiness that might be
expected from the match, the excellent character of Mr. Collins,
and the convenient distance of Hunsford from London.

Mrs. Bennet was in fact too much overpowered to say a great
deal while Sir William remained; but no sooner had he left them
than her feelings found a rapid\index{rapid} vent.  In the first place, she
persisted in disbelieving the whole of the matter; secondly, she
was very sure that Mr. Collins had been taken in; thirdly, she
trusted that they would never be happy together; and fourthly,
that the match might be broken off.  Two \textbf{inferences}\index{inferences}\stepcounter{footnote}\footnotetext{
\textbf{inference} \emph{n.}  (IHN-fer-ense)
a logical conclusion drawn from factual information.}, however,
were plainly \textbf{deduced}\index{deduced}\stepcounter{footnote}\footnotetext{
\textbf{deduce} \emph{tr.v.} \textbf{-ed}  (dih-DYOOS)
to reach a conclusion logically.} from the whole: one, that Elizabeth was
the real cause of the mischief; and the other that she herself had
been \textbf{barbarously}\index{barbarously}\stepcounter{footnote}\footnotetext{
\textbf{barbarous} \emph{adj.} \textbf{-ly}  (BAR-buh-uhs)
disposed to inflicting pain or suffering.} misused by them all; and on these two points
she principally\index{principally} dwelt during the rest of the day.  Nothing could
console\index{console} and nothing could \textbf{appease}\index{appease}\stepcounter{footnote}\footnotetext{
\textbf{appease} \emph{tr.v.} \textbf{-ed}  (uh-PEEZ)
} her.  Nor did that day wear
out her resentment\index{resentment}.  A week elapsed before she could see
Elizabeth without scolding her, a month passed away before she
could speak to Sir William or Lady Lucas without being rude,
and many months were gone before she could at all forgive their
daughter.

Mr. Bennet's emotions were much more \textbf{tranquil}\index{tranquil}\stepcounter{footnote}\footnotetext{
\textbf{tranquil} \emph{adj.} \textbf{-ly}  (TRAN-kwihl)
without emotional agitation; calm.} on the occasion,
and such as he did experience he pronounced to be of a most
agreeable sort; for it gratified him, he said, to discover that
Charlotte Lucas, whom he had been used to think tolerably
sensible, was as foolish as his wife, and more foolish than his
daughter!

Jane confessed herself a little surprised at the match; but she
said less of her astonishment than of her earnest\index{earnest} desire for their
happiness; nor could Elizabeth persuade her to consider it as
\textbf{improbable}\index{improbable}\stepcounter{footnote}\footnotetext{
\textbf{improbable} \emph{adj.}  (ihm-PRAH-buh-bl)
not likely to happen or be true.}.  Kitty and Lydia were far from envying Miss Lucas,
for Mr. Collins was only a clergyman; and it affected them in no
other way than as a piece of news to spread at Meryton.

Lady Lucas could not be insensible of triumph\index{triumph} on being able to
\textbf{retort}\index{retort}\stepcounter{footnote}\footnotetext{
\textbf{retort} \emph{n.}  (rih-TORT)
to reply to a question or remark (especially a witty or critical reply).} on Mrs. Bennet the comfort of having a daughter well
married; and she called at Longbourn rather oftener than usual
to say how happy she was, though Mrs. Bennet's sour looks and
ill-natured remarks might have been enough to drive happiness
away.

Between Elizabeth and Charlotte there was a restraint which
kept them mutually silent on the subject\index{subject}; and Elizabeth felt
persuaded that no real confidence\index{confidence} could ever subsist\index{subsist} between
them again.  Her disappointment in Charlotte made her turn with
fonder regard to her sister, of whose \textbf{rectitude}\index{rectitude}\stepcounter{footnote}\footnotetext{
\textbf{rectitude} \emph{n.}  (REHK-tih-tood)
uprightness as a consequence of being honorable and honest; righteousness.} and delicacy\index{delicacy}
she was sure her opinion could never be shaken, and for whose
happiness she grew daily more anxious, as Bingley had now been
gone a week and nothing more was heard of his return.

Jane had sent Caroline an early answer to her letter, and was
counting the days till she might reasonably hope to hear again.
The promised letter of thanks from Mr. Collins arrived on
Tuesday, addressed\index{addressed} to their father, and written with all the
solemnity\index{solemnity} of gratitude which a twelvemonth's abode\index{abode} in the
family might have prompted.  After discharging his conscience\index{conscience}
on that head, he proceeded to inform them, with many \textbf{rapturous}\index{rapturous}\stepcounter{footnote}\footnotetext{
\textbf{rapturous} \emph{adj.} \textbf{-ly}  (RAP-cher-uhs)
3}
expressions, of his happiness in having obtained the affection of
their amiable\index{amiable} neighbour, Miss Lucas, and then explained that it
was merely with the view of enjoying her society that he had
been so ready to close with their kind wish of seeing him again
at Longbourn, whither he hoped to be able to return on Monday
fortnight\index{fortnight}; for Lady Catherine, he added, so heartily\index{heartily} approved
his marriage, that she wished it to take place as soon as possible,
which he trusted would be an unanswerable argument with his
amiable\index{amiable} Charlotte to name an early day for making him the
happiest of men.

Mr. Collins's return into Hertfordshire was no longer a matter
of pleasure to Mrs. Bennet.  On the contrary, she was as much
disposed\index{disposed} to complain of it as her husband.  It was very strange
that he should come to Longbourn instead of to Lucas Lodge; it
was also very inconvenient and exceedingly troublesome.  She
hated having visitors in the house while her health was so
indifferent\index{indifferent}, and lovers were of all people the most disagreeable.
Such were the gentle murmurs of Mrs. Bennet, and they gave way
only to the greater distress of Mr. Bingley's continued absence.

Neither Jane nor Elizabeth were comfortable on this subject\index{subject}.
Day after day passed away without bringing any other \textbf{tidings}\index{tidings}\stepcounter{footnote}\footnotetext{
\textbf{tiding} \emph{n.}  (TY-dihng)
 new information about specific and timely events; news.} of
him than the report which shortly prevailed\index{prevailed} in Meryton of his
coming no more to Netherfield the whole winter; a report which
highly \textbf{incensed}\index{incensed}\stepcounter{footnote}\footnotetext{
\textbf{incense} \emph{tr.v.} \textbf{-ed}  (ihn-SEHNS)
to cause anger or a strong feeling of injustice.} Mrs. Bennet, and which she never failed to
contradict as a most scandalous\index{scandalous} falsehood.

Even Elizabeth began to fear--not that Bingley was indifferent\index{indifferent}--but
that his sisters would be successful in keeping him away.
Unwilling as she was to admit an idea so destructive of Jane's
happiness, and so dishonorable to the stability of her lover, she
could not prevent its frequently occurring.  The united efforts of
his two unfeeling sisters and of his overpowering friend, assisted
by the attractions of Miss Darcy and the amusements of London
might be too much, she feared, for the strength of his attachment.

As for Jane, \emph{her} anxiety under this suspense was, of course,
more painful than Elizabeth's, but whatever she felt she was
desirous of concealing, and between herself and Elizabeth,
therefore, the subject\index{subject} was never \textbf{alluded}\index{alluded}\stepcounter{footnote}\footnotetext{
\textbf{allude} \emph{intr.v.} \textbf{-ed}  (al-LOOD)
} to.  But as no such
delicacy\index{delicacy} restrained her mother, an hour seldom passed in which
she did not talk of Bingley, express her impatience for his arrival,
or even require Jane to confess that if he did not come back she
would think herself very ill used.  It needed all Jane's steady\index{steady}
mildness to bear these attacks with tolerable\index{tolerable} tranquillity\index{tranquillity}.

Mr. Collins returned most \textbf{punctually}\index{punctually}\stepcounter{footnote}\footnotetext{
\textbf{punctual} \emph{adj.} \textbf{-ity}  (PUHNK-choo-uhl)
on time; prompt.} on Monday fortnight\index{fortnight}, but
his reception at Longbourn was not quite so gracious as it had
been on his first introduction.  He was too happy, however, to
need much attention; and luckily for the others, the business
of love-making relieved them from a great deal of his company.
The chief of every day was spent by him at Lucas Lodge, and he
sometimes returned to Longbourn only in time to make an
apology for his absence before the family went to bed.

Mrs. Bennet was really in a most pitiable state.  The very
mention of anything concerning the match threw her into an
agony of ill-humour, and wherever she went she was sure of
hearing it talked of.  The sight of Miss Lucas was odious\index{odious} to
her.  As her \textbf{successor}\index{successor}\stepcounter{footnote}\footnotetext{
\textbf{successor} \emph{n.}  (suhk-SEHS-sawr)
a person who follows another person in office or position.} in that house, she regarded her with
jealous \textbf{abhorrence}\index{abhorrence}\stepcounter{footnote}\footnotetext{
\textbf{abhorrence} \emph{n.}  (ab-HAWR-rense)
hate coupled with disgust.}.  Whenever Charlotte came to see them,
she concluded her to be anticipating the hour of possession;
and whenever she spoke in a low voice to Mr. Collins, was
convinced that they were talking of the Longbourn estate, and
resolving to turn herself and her daughters out of the house,
as soon as Mr. Bennet were dead.  She complained bitterly of
all this to her husband.

"Indeed, Mr. Bennet," said she, "it is very hard to think that
Charlotte Lucas should ever be mistress of this house, that I
should be forced to make way for \emph{her}, and live to see her take
her place in it!"

"My dear, do not give way to such gloomy thoughts.  Let\index{Let} us
hope for better things.  Let\index{Let} us flatter ourselves that I may be
the survivor."

This was not very consoling to Mrs. Bennet, and therefore, instead
of making any answer, she went on as before.

"I cannot bear to think that they should have all this estate.
If it was not for the entail\index{entail}, I should not mind it."

"What should not you mind?"

"I should not mind anything at all."

"Let\index{Let} us be thankful that you are preserved from a state of such
insensibility."

"I never can be thankful, Mr. Bennet, for anything about the
entail\index{entail}.  How anyone could have the conscience\index{conscience} to entail\index{entail} away an
estate from one's own daughters, I cannot understand; and all
for the sake of Mr. Collins too!  Why should \emph{he} have it more
than anybody else?"

"I leave it to yourself to determine," said Mr. Bennet.



\chapter{24}


Miss Bingley's letter arrived, and put an end to doubt.  The very
first sentence conveyed the assurance of their being all settled
in London for the winter, and concluded with her brother's regret
at not having had time to pay his respects to his friends in
Hertfordshire before he left the country.

Hope was over, entirely over; and when Jane could attend to the
rest of the letter, she found little, except the professed affection
of the writer, that could give her any comfort.  Miss Darcy's
praise occupied the chief of it.  Her many attractions were again
dwelt on, and Caroline boasted joyfully of their increasing
intimacy, and ventured to predict the accomplishment of the
wishes which had been unfolded in her former letter.  She wrote
also with great pleasure of her brother's being an inmate of Mr.
Darcy's house, and mentioned with raptures\index{raptures} some plans of the
latter with regard to new furniture.

Elizabeth, to whom Jane very soon communicated the chief of
all this, heard it in silent indignation\index{indignation}.  Her heart was divided
between concern for her sister, and resentment\index{resentment} against all others.
To Caroline's assertion\index{assertion} of her brother's being partial\index{partial} to Miss
Darcy she paid no credit.  That he was really fond of Jane, she
doubted no more than she had ever done; and much as she had
always been disposed\index{disposed} to like him, she could not think without
anger, hardly without contempt\index{contempt}, on that easiness of temper, that
want of proper resolution\index{resolution}, which now made him the slave of his
designing friends, and led him to sacrifice of his own happiness
to the caprice\index{caprice} of their inclination\index{inclination}.  Had his own happiness,
however, been the only sacrifice, he might have been allowed to
sport with it in whatever manner he thought best, but her sister's
was involved in it, as she thought he must be sensible himself.
It was a subject\index{subject}, in short, on which reflection would be long
indulged, and must be \textbf{unavailing}\index{unavailing}\stepcounter{footnote}\footnotetext{
\textbf{unavailing} \emph{adj.}  (uhn-uh-VAY-ling)
producing no result or effect.}.  She could think of nothing
else; and yet whether Bingley's regard had really died away, or
were \textbf{suppressed}\index{suppressed}\stepcounter{footnote}\footnotetext{
\textbf{suppress} \emph{tr.v.} \textbf{-ed}  (suhp-PRESS)
to put down or subdue by force or authority.} by his friends' interference; whether he had
been aware of Jane's attachment, or whether it had escaped his
observation; whatever were the case, though her opinion of him
must be materially affected by the difference, her sister's
situation remained the same, her peace equally wounded.

A day or two passed before Jane had courage to speak of her
feelings to Elizabeth; but at last, on Mrs. Bennet's leaving them
together, after a longer irritation than usual about Netherfield
and its master, she could not help saying:

"Oh, that my dear mother had more command over herself!  She
can have no idea of the pain she gives me by her continual
reflections on him.  But I will not \textbf{repine}\index{repine}\stepcounter{footnote}\footnotetext{
\textbf{repine} \emph{intr.v.} \textbf{-ing}  (rih-PYN)
to wish discontentedly; to dejectedly complain.}.  It cannot last long.
He will be forgot, and we shall all be as we were before."

Elizabeth looked at her sister with incredulous\index{incredulous} solicitude\index{solicitude}, but
said nothing.

"You doubt me," cried Jane, slightly colouring; "indeed, you
have no reason.  He may live in my memory as the most amiable\index{amiable}
man of my acquaintance, but that is all.  I have nothing either
to hope or fear, and nothing to reproach\index{reproach} him with.  Thank God! I
have not \emph{that} pain.  A little time, therefore--I shall certainly
try to get the better."

With a stronger voice she soon added, "I have this comfort
immediately, that it has not been more than an error of fancy on
my side, and that it has done no harm to anyone but myself."

"My dear Jane!" exclaimed Elizabeth, "you are too good.  Your
sweetness and disinterestedness are really angelic; I do not know
what to say to you.  I feel as if I had never done you justice, or
loved you as you deserve."

Miss Bennet eagerly disclaimed all extraordinary merit, and
threw back the praise on her sister's warm affection.

"Nay," said Elizabeth, "this is not fair.  \emph{you} wish to think all
the world respectable, and are hurt if I speak ill of anybody.  I
only want to think \emph{you} perfect, and you set yourself against it.
Do not be afraid of my running into any excess, of my \textbf{encroaching}\index{encroaching}\stepcounter{footnote}\footnotetext{
\textbf{encroach} \emph{intr.v.} \textbf{-ing}  (ehn-KROACH)
to gradually intrude or take over something without right or permision.}
on your privilege of universal good-will.  You need not.  There
are few people whom I really love, and still fewer of whom I think
well.  The more I see of the world, the more am I dissatisfied
with it; and every day confirms my belief of the inconsistency\index{inconsistency} of
all human characters, and of the little dependence that can be
placed on the appearance of merit or sense.  I have met with two
instances lately, one I will not mention; the other is Charlotte's
marriage.  It is unaccountable!  In every view it is unaccountable!"

"My dear Lizzy, do not give way to such feelings as these.  They
will ruin your happiness.  You do not make allowance enough
for difference of situation and temper.  Consider Mr. Collins's
respectability, and Charlotte's steady\index{steady}, \textbf{prudent}\index{prudent}\stepcounter{footnote}\footnotetext{
\textbf{prudent} \emph{adj.} \textbf{-ly}  (PROO-duhnt)
 exercising good sense, tact, and judgement.} character.
Remember that she is one of a large family; that as to fortune,
it is a most eligible match; and be ready to believe, for
everybody's sake, that she may feel something like regard and
esteem for our cousin."

"To oblige you, I would try to believe almost anything, but no
one else could be benefited by such a belief as this; for were I
persuaded that Charlotte had any regard for him, I should only
think worse of her understanding than I now do of her heart.
My dear Jane, Mr. Collins is a conceited\index{conceited}, pompous\index{pompous}, narrow-minded,
silly man; you know he is, as well as I do; and you must feel,
as well as I do, that the woman who married him cannot have a
proper way of thinking.  You shall not defend her, though it is
Charlotte Lucas.  You shall not, for the sake of one individual,
change the meaning of principle and \textbf{integrity}\index{integrity}\stepcounter{footnote}\footnotetext{
\textbf{integrity} \emph{n.}  (ihn-TEH-grih-tee)
moral soundness.}, nor endeavour\index{endeavour} to
persuade yourself or me, that selfishness is \textbf{prudence}\index{prudence}\stepcounter{footnote}\footnotetext{
\textbf{prudence} \emph{n.}  (PROO-dehnse)
discretion; knowing how to avoid embarrassment or distress.}, and
insensibility of danger security for happiness."

"I must think your language too strong in speaking of both,"
replied Jane; "and I hope you will be convinced of it by seeing
them happy together.  But enough of this.  You alluded\index{alluded} to
something else.  You mentioned \emph{two} instances.  I cannot
misunderstand you, but I entreat\index{entreat} you, dear Lizzy, not to pain
me by thinking \emph{that} \emph{person} to blame, and saying your opinion
of him is sunk.  We must not be so ready to fancy ourselves
intentionally injured.  We must not expect a lively young man to
be always so guarded and \textbf{circumspect}\index{circumspect}\stepcounter{footnote}\footnotetext{
\textbf{circumspect} \emph{adj.}  (SER-kuhm-spehkt)
heedful of potential consequences; prudent.}.  It is very often nothing
but our own vanity\index{vanity} that deceives us.  Women fancy admiration
means more than it does."

"And men take care that they should."

"If it is designedly done, they cannot be justified; but I have no
idea of there being so much design in the world as some persons
imagine."

"I am far from attributing any part of Mr. Bingley's conduct to
design," said Elizabeth; "but without scheming to do wrong, or
to make others unhappy, there may be error, and there may be
misery.  Thoughtlessness, want of attention to other people's
feelings, and want of resolution\index{resolution}, will do the business."

"And do you impute\index{impute} it to either of those?"

"Yes; to the last.  But if I go on, I shall displease you by saying
what I think of persons you esteem.  Stop me whilst you can."

"You persist, then, in supposing his sisters influence him?"

"Yes, in \textbf{conjunction}\index{conjunction}\stepcounter{footnote}\footnotetext{
\textbf{conjunction} \emph{n.}  (kuhn-JUHNK-shun)
something that joins or connects.} with his friend."

"I cannot believe it.  Why should they try to influence him?
They can only wish his happiness; and if he is attached to me,
no other woman can secure it."

"Your first position is false.  They may wish many things besides
his happiness; they may wish his increase of wealth and
consequence\index{consequence}; they may wish him to marry a girl who has all the
importance of money, great connections, and pride."

"Beyond a doubt, they \emph{do} wish him to choose Miss Darcy,"
replied Jane; "but this may be from better feelings than you are
supposing.  They have known her much longer than they have
known me; no wonder if they love her better.  But, whatever
may be their own wishes, it is very unlikely they should have
opposed their brother's.  What sister would think herself at
liberty to do it, unless there were something very objectionable?
If they believed him attached to me, they would not try to part
us; if he were so, they could not succeed.  By supposing such
an affection, you make everybody acting unnaturally and wrong,
and me most unhappy.  Do not distress me by the idea.  I am not
ashamed of having been mistaken--or, at least, it is light, it
is nothing in comparison of what I should feel in thinking ill
of him or his sisters.  Let\index{Let} me take it in the best light, in
the light in which it may be understood."

Elizabeth could not oppose such a wish; and from this time
Mr. Bingley's name was scarcely ever mentioned between them.

Mrs. Bennet still continued to wonder and repine\index{repine} at his returning
no more, and though a day seldom passed in which Elizabeth did
not account for it clearly, there was little chance of her ever
considering it with less \textbf{perplexity}\index{perplexity}\stepcounter{footnote}\footnotetext{
\textbf{perplexity} \emph{n.}  (per-PLEHX-sih-tee)
the state of being puzzled or confused.}.  Her daughter endeavoured\index{endeavoured}
to convince her of what she did not believe herself, that his
attentions to Jane had been merely the effect of a common and
transient liking, which ceased when he saw her no more; but
though the probability of the statement was admitted at the time,
she had the same story to repeat every day.  Mrs. Bennet's best
comfort was that Mr. Bingley must be down again in the
summer.

Mr. Bennet treated the matter differently.  "So, Lizzy," said he
one day, "your sister is crossed in love, I find.  I congratulate
her.  Next to being married, a girl likes to be crossed a little in
love now and then.  It is something to think of, and it gives her a
sort of distinction among her companions.  When is your turn to
come?  You will hardly bear to be long outdone by Jane.  Now is
your time.  Here are officers enough in Meryton to disappoint all
the young ladies in the country.  Let\index{Let} Wickham be \emph{your} man.  He
is a pleasant fellow, and would \textbf{jilt}\index{jilt}\stepcounter{footnote}\footnotetext{
\textbf{jilt} \emph{tr.v.}  (JIHLT)
 To deceive or drop (a lover) suddenly or callously, especially on the wedding day.} you creditably."

"Thank you, sir, but a less agreeable man would satisfy me.  We
must not all expect Jane's good fortune."

"True," said Mr. Bennet, "but it is a comfort to think that
whatever of that kind may befall you, you have an affectionate
mother who will make the most of it."

Mr. Wickham's society was of material service in dispelling the
gloom which the late perverse\index{perverse} occurrences had thrown on many
of the Longbourn family.  They saw him often, and to his other
recommendations was now added that of general unreserve.
The whole of what Elizabeth had already heard, his claims on
Mr. Darcy, and all that he had suffered from him, was now
openly acknowledged and publicly \textbf{canvassed}\index{canvassed}\stepcounter{footnote}\footnotetext{
\textbf{canvass} \emph{n.} \textbf{-ed}  (KAN-vass)
}; and everybody
was pleased to know how much they had always disliked Mr.
Darcy before they had known anything of the matter.

Miss Bennet was the only creature who could suppose there might
be any \textbf{extenuating}\index{extenuating}\stepcounter{footnote}\footnotetext{
\textbf{extenuate} \emph{tr.v.} \textbf{-ing}  (ehx-TEHN-yoo-ayt)
to lessen or to try to lessen the seriousness or extent of.} circumstances in the case, unknown to the
society of Hertfordshire; her mild and steady\index{steady} candour always
pleaded for allowances, and urged the possibility of mistakes--but
by everybody else Mr. Darcy was condemned as the worst of men.



\chapter{25}


After a week spent in professions of love and schemes of felicity\index{felicity},
Mr. Collins was called from his amiable\index{amiable} Charlotte by the arrival
of Saturday.  The pain of separation, however, might be
\textbf{alleviated}\index{alleviated}\stepcounter{footnote}\footnotetext{
\textbf{alleviate} \emph{tr.v.} \textbf{-ed}  (uh-LEE-vee-ATE)
to provide physical relief, as from pain; make easier.} on his side, by preparations for the reception of his
bride; as he had reason to hope, that shortly after his return into
Hertfordshire, the day would be fixed that was to make him the
happiest of men.  He took leave of his relations at Longbourn
with as much solemnity\index{solemnity} as before; wished his fair cousins health
and happiness again, and promised their father another letter of
thanks.

On the following Monday, Mrs. Bennet had the pleasure of
receiving her brother and his wife, who came as usual to spend
the Christmas at Longbourn.  Mr. Gardiner was a sensible,
gentlemanlike man, greatly superior\index{superior} to his sister, as well by
nature as education.  The Netherfield ladies would have had
difficulty in believing that a man who lived by trade, and within
view of his own warehouses, could have been so well-bred and
agreeable.  Mrs. Gardiner, who was several years younger than
Mrs. Bennet and Mrs. Phillips, was an amiable\index{amiable}, intelligent,
elegant woman, and a great favourite with all her Longbourn
nieces.  Between the two eldest and herself especially, there
subsisted\index{subsisted} a particular regard.  They had frequently been staying
with her in town.

The first part of Mrs. Gardiner's business on her arrival was to
distribute her presents and describe the newest fashions\index{fashions}.  When
this was done she had a less active part to play.  It became her
turn to listen.  Mrs. Bennet had many \textbf{grievances}\index{grievances}\stepcounter{footnote}\footnotetext{
\textbf{grievance} \emph{n.}  (GREE-vense)
a situation or curcumstance viewed as reason for complaint.} to relate, and
much to complain of.  They had all been very ill-used since she
last saw her sister.  Two of her girls had been upon the point of
marriage, and after all there was nothing in it.

"I do not blame Jane," she continued, "for Jane would have got
Mr. Bingley if she could.  But Lizzy!  Oh, sister!  It is very hard
to think that she might have been Mr. Collins's wife by this time,
had it not been for her own perverseness.  He made her an offer
in this very room, and she refused him.  The consequence\index{consequence} of it is,
that Lady Lucas will have a daughter married before I have, and
that the Longbourn estate is just as much entailed\index{entailed} as ever.  The
Lucases are very artful people indeed, sister.  They are all for
what they can get.  I am sorry to say it of them, but so it is.
It makes me very nervous and poorly, to be thwarted so in my own
family, and to have neighbours who think of themselves before
anybody else.  However, your coming just at this time is the
greatest of comforts, and I am very glad to hear what you tell us,
of long sleeves."

Mrs. Gardiner, to whom the chief of this news had been given
before, in the course of Jane and Elizabeth's correspondence
with her, made her sister a slight answer, and, in compassion to
her nieces, turned the conversation.

When alone with Elizabeth afterwards, she spoke more on the
subject\index{subject}.  "It seems likely to have been a desirable match for
Jane," said she.  "I am sorry it went off.  But these things
happen so often!  A young man, such as you describe Mr. Bingley,
so easily falls in love with a pretty girl for a few weeks, and
when accident separates them, so easily forgets her, that these
sort of inconsistencies are very frequent."

"An excellent consolation\index{consolation} in its way," said Elizabeth, "but it will
not do for \emph{us}.  We do not suffer by \emph{accident}.  It does not often
happen that the interference of friends will persuade a young
man of independent fortune to think no more of a girl whom he
was violently in love with only a few days before."

"But that expression of 'violently in love' is so \textbf{hackneyed}\index{hackneyed}\stepcounter{footnote}\footnotetext{
\textbf{hackneyed} \emph{adj.}  (HAK-need)
Overfamiliar through overuse; trite.},
so doubtful, so indefinite, that it gives me very little idea.
It is as often applied to feelings which arise from a half-hour's
acquaintance, as to a real, strong attachment.  Pray, how
\emph{violent} \emph{was} Mr. Bingley's love?"

"I never saw a more promising inclination\index{inclination}; he was growing quite
inattentive to other people, and wholly\index{wholly} engrossed\index{engrossed} by her.  Every
time they met, it was more decided and remarkable.  At his own
ball he offended two or three young ladies, by not asking them to
dance; and I spoke to him twice myself, without receiving an answer.
Could there be finer symptoms?  Is not general \textbf{incivility}\index{incivility}\stepcounter{footnote}\footnotetext{
\textbf{incivility} \emph{n.}  (ihn-sih-VIH-lih-tee)
deliberate discourtesy.} the very
essence of love?"

"Oh, yes!--of that kind of love which I suppose him to have felt.
Poor Jane!  I am sorry for her, because, with her disposition\index{disposition}, she
may not get over it immediately.  It had better have happened to
\emph{you}, Lizzy; you would have laughed yourself out of it sooner.
But do you think she would be prevailed\index{prevailed} upon to go back with
us?  Change of scene might be of service--and perhaps a little
relief from home may be as useful as anything."

Elizabeth was exceedingly pleased with this proposal, and felt
persuaded of her sister's ready \textbf{acquiescence}\index{acquiescence}\stepcounter{footnote}\footnotetext{
\textbf{acquiescence} \emph{n.}  (ak-wee-ESS-sense)
passive acceptance or agreement.}.

"I hope," added Mrs. Gardiner, "that no consideration with
regard to this young man will influence her.  We live in so
different a part of town, all our connections are so different, and,
as you well know, we go out so little, that it is very improbable\index{improbable}
that they should meet at all, unless he really comes to see her."

"And \emph{that} is quite impossible; for he is now in the custody of
his friend, and Mr. Darcy would no more suffer him to call on
Jane in such a part of London!  My dear aunt, how could you
think of it?  Mr. Darcy may perhaps have \emph{heard} of such a
place as Gracechurch Street, but he would hardly think a
month's \textbf{ablution}\index{ablution}\stepcounter{footnote}\footnotetext{
\textbf{ablution} \emph{n.}  (ab-BLOO-shun)
washing or imersing one's body in water, especially as a religious act.} enough to cleanse him from its \textbf{impurities}\index{impurities}\stepcounter{footnote}\footnotetext{
\textbf{impurity} \emph{n.}  (ihm-PYOO-rih-tee)
Contamination, pollution, or a foreign substance.}, were
he once to enter it; and depend upon it, Mr. Bingley never stirs
without him."

"So much the better.  I hope they will not meet at all.  But does
not Jane correspond with his sister?  \emph{she} will not be able to
help calling."

"She will drop the acquaintance entirely."

But in spite of the certainty in which Elizabeth affected to place
this point, as well as the still more interesting one of Bingley's
being withheld from seeing Jane, she felt a solicitude\index{solicitude} on the
subject\index{subject} which convinced her, on examination, that she did not
consider it entirely hopeless.  It was possible, and sometimes she
thought it probable\index{probable}, that his affection might be reanimated, and
the influence of his friends successfully \textbf{combated}\index{combated}\stepcounter{footnote}\footnotetext{
\textbf{combat} \emph{v.} \textbf{-ed}  (KAHM-bat)
 to fight in battle.} by the more
natural influence of Jane's attractions.

Miss Bennet accepted her aunt's invitation with pleasure; and
the Bingleys were no otherwise in her thoughts at the same time,
than as she hoped by Caroline's not living in the same house
with her brother, she might occasionally spend a morning with
her, without any danger of seeing him.

The Gardiners stayed a week at Longbourn; and what with the
Phillipses, the Lucases, and the officers, there was not a day
without its engagement.  Mrs. Bennet had so carefully provided
for the entertainment of her brother and sister, that they did
not once sit down to a family dinner.  When the engagement was
for home, some of the officers always made part of it--of
which officers Mr. Wickham was sure to be one; and on these
occasion, Mrs. Gardiner, rendered\index{rendered} suspicious by Elizabeth's
warm commendation\index{commendation}, narrowly observed them both.  Without
supposing them, from what she saw, to be very seriously in love,
their preference of each other was plain enough to make her a
little uneasy; and she resolved to speak to Elizabeth on the
subject\index{subject} before she left Hertfordshire, and represent to her the
imprudence\index{imprudence} of encouraging such an attachment.

To Mrs. Gardiner, Wickham had one means of affording pleasure,
unconnected with his general powers.  About ten or a dozen years
ago, before her marriage, she had spent a considerable time in
that very part of Derbyshire to which he belonged.  They had,
therefore, many acquaintances in common; and though Wickham had
been little there since the death of Darcy's father, it was yet
in his power to give her fresher intelligence of her former
friends than she had been in the way of procuring.

Mrs. Gardiner had seen Pemberley, and known the late Mr. Darcy
by character perfectly well.  Here consequently was an
inexhaustible subject\index{subject} of discourse\index{discourse}.  In comparing her recollection\index{recollection}
of Pemberley with the minute\index{minute} description which Wickham could
give, and in bestowing\index{bestowing} her tribute of praise on the character of
its late possessor, she was delighting both him and herself.  On
being made acquainted with the present Mr. Darcy's treatment
of him, she tried to remember some of that gentleman's \textbf{reputed}\index{reputed}\stepcounter{footnote}\footnotetext{
\textbf{repute} \emph{tr.v.} \textbf{-ed}  (rih-PYOOT)
to ascribe a reputation to.}
disposition\index{disposition} when quite a lad which might agree with it, and was
confident at last that she recollected having heard Mr.
Fitzwilliam Darcy formerly spoken of as a very proud, ill-natured
boy.



\chapter{26}


Mrs. Gardiner's caution to Elizabeth was punctually\index{punctually} and kindly
given on the first favourable opportunity of speaking to her
alone; after honestly telling her what she thought, she thus went
on:

"You are too sensible a girl, Lizzy, to fall in love merely because
you are warned against it; and, therefore, I am not afraid of
speaking openly.  Seriously, I would have you be on your guard.
Do not involve yourself or endeavour\index{endeavour} to involve him in an
affection which the want of fortune would make so very
imprudent\index{imprudent}.  I have nothing to say against \emph{him}; he is a most
interesting young man; and if he had the fortune he ought to
have, I should think you could not do better.  But as it is, you
must not let\index{let} your fancy run away with you.  You have sense, and
we all expect you to use it.  Your father would depend on
\emph{your} resolution\index{resolution} and good conduct, I am sure.  You must not
disappoint your father."

"My dear aunt, this is being serious indeed."

"Yes, and I hope to engage you to be serious likewise."

"Well, then, you need not be under any alarm.  I will take care
of myself, and of Mr. Wickham too.  He shall not be in love with
me, if I can prevent it."

"Elizabeth, you are not serious now."

"I beg your pardon, I will try again.  At present I am not in
love with Mr. Wickham; no, I certainly am not.  But he is, beyond
all comparison, the most agreeable man I ever saw--and if he
becomes really attached to me--I believe it will be better that
he should not.  I see the imprudence\index{imprudence} of it.  Oh! \emph{that} abominable\index{abominable}
Mr. Darcy!  My father's opinion of me does me the greatest
honour, and I should be miserable to forfeit it.  My father,
however, is partial\index{partial} to Mr. Wickham.  In short, my dear aunt,
I should be very sorry to be the means of making any of you
unhappy; but since we see every day that where there is
affection, young people are seldom withheld by immediate want
of fortune from entering into engagements with each other, how
can I promise to be wiser than so many of my fellow-creatures
if I am tempted, or how am I even to know that it would be
wisdom  to resist?  All that I can promise you, therefore, is not
to be in a hurry.  I will not be in a hurry to believe myself his
first object.  When I am in company with him, I will not be
wishing.  In short, I will do my best."

"Perhaps it will be as well if you discourage his coming here so
very often.  At least, you should not \emph{remind} you mother of
inviting him."

"As I did the other day," said Elizabeth with a conscious smile:
"very true, it will be wise in me to refrain from \emph{that}.  But do
not imagine that he is always here so often.  It is on your
account that he has been so frequently invited this week.  You
know my mother's ideas as to the necessity of constant company
for her friends.  But really, and upon my honour, I will try to do
what I think to be the wisest; and now I hope you are satisfied."

Her aunt assured her that she was, and Elizabeth having thanked
her for the kindness of her hints, they parted; a wonderful
instance of advice being given on such a point, without being
resented.

Mr. Collins returned into Hertfordshire soon after it had been
quitted by the Gardiners and Jane; but as he took up his abode\index{abode}
with the Lucases, his arrival was no great inconvenience to Mrs.
Bennet.  His marriage was now fast approaching, and she was at
length so far resigned as to think it inevitable, and even
repeatedly to say, in an ill-natured tone, that she "\emph{wished} they
might be happy."  Thursday was to be the wedding day, and on
Wednesday Miss Lucas paid her farewell visit; and when she
rose to take leave, Elizabeth, ashamed of her mother's
ungracious and reluctant good wishes, and sincerely affected
herself, accompanied her out of the room.  As they went
downstairs together, Charlotte said:

"I shall depend on hearing from you very often, Eliza."

"\emph{that} you certainly shall."

"And I have another favour to ask you.  Will you come and see
me?"

"We shall often meet, I hope, in Hertfordshire."

"I am not likely to leave Kent for some time.  Promise me,
therefore, to come to Hunsford."

Elizabeth could not refuse, though she foresaw little pleasure
in the visit.

"My father and Maria are coming to me in March," added
Charlotte, "and I hope you will consent to be of the party.
Indeed, Eliza, you will be as welcome as either of them."

The wedding took place; the bride and bridegroom set off for
Kent from the church door, and everybody had as much to say,
or to hear, on the subject\index{subject} as usual.  Elizabeth soon heard from
her friend; and their correspondence was as regular\index{regular} and frequent
as it had ever been; that it should be equally unreserved was
impossible.  Elizabeth could never address\index{address} her without feeling
that all the comfort of intimacy was over, and though determined
not to slacken as a correspondent, it was for the sake of what
had been, rather than what was.  Charlotte's first letters were
received with a good deal of eagerness; there could not but be
curiosity to know how she would speak of her new home, how
she would like Lady Catherine, and how happy she would dare
pronounce herself to be; though, when the letters were read,
Elizabeth felt that Charlotte expressed herself on every point
exactly as she might have foreseen.  She wrote cheerfully,
seemed surrounded with comforts, and mentioned nothing which
she could not praise.  The house, furniture, neighbourhood, and
roads, were all to her taste, and Lady Catherine's behaviour
was most friendly and obliging.  It was Mr. Collins's picture
of Hunsford and Rosings \textbf{rationally}\index{rationally}\stepcounter{footnote}\footnotetext{
\textbf{rational} \emph{adj.} \textbf{-ly}  (RA-shun-uhl)
} softened; and Elizabeth
perceived\index{perceived} that she must wait for her own visit there to know the
rest.

Jane had already written a few lines to her sister to announce
their safe arrival in London; and when she wrote again, Elizabeth
hoped it would be in her power to say something of the
Bingleys.

Her impatience for this second letter was as well rewarded as
impatience generally is.  Jane had been a week in town without
either seeing or hearing from Caroline.  She accounted for it,
however, by supposing that her last letter to her friend from
Longbourn had by some accident been lost.

"My aunt," she continued, "is going to-morrow into that part of
the town, and I shall take the opportunity of calling in Grosvenor
Street."

She wrote again when the visit was paid, and she had seen Miss
Bingley.  "I did not think Caroline in spirits," were her words,
"but she was very glad to see me, and \textbf{reproached}\index{reproached}\stepcounter{footnote}\footnotetext{
\textbf{reproach} \emph{tr.v.} \textbf{-ed}  (rih-PROACH)
to express criticism, disapproval, or to rebuke.} me for giving
her no notice of my coming to London.  I was right, therefore,
my last letter had never reached her.  I inquired after their
brother, of course.  He was well, but so much engaged with Mr.
Darcy that they scarcely ever saw him.  I found that Miss Darcy
was expected to dinner.  I wish I could see her.  My visit was
not long, as Caroline and Mrs. Hurst were going out.  I dare say
I shall see them soon here."

Elizabeth shook her head over this letter.  It convinced her that
accident only could discover to Mr. Bingley her sister's being in
town.

Four weeks passed away, and Jane saw nothing of him.  She
endeavoured\index{endeavoured} to persuade herself that she did not regret it; but
she could no longer be blind to Miss Bingley's inattention.  After
waiting at home every morning for a fortnight\index{fortnight}, and inventing
every evening a fresh excuse for her, the visitor did at last
appear; but the shortness of her stay, and yet more, the alteration
of her manner would allow Jane to deceive herself no longer.
The letter which she wrote on this occasion to her sister will
prove what she felt.

"My dearest Lizzy will, I am sure, be incapable of triumphing\index{triumphing} in
her better judgement, at my expense, when I confess myself to
have been entirely deceived in Miss Bingley's regard for me.
But, my dear sister, though the event has proved you right, do
not think me \textbf{obstinate}\index{obstinate}\stepcounter{footnote}\footnotetext{
\textbf{obstinate} \emph{adj.}  (AHB-stih-nuht)
stubborn.} if I still assert that, considering what
her behaviour was, my confidence\index{confidence} was as natural as your suspicion.
I do not at all comprehend\index{comprehend} her reason for wishing to be intimate
with me; but if the same circumstances were to happen again, I
am sure I should be deceived again.  Caroline did not return my
visit till yesterday; and not a note, not a line, did I receive
in the meantime.  When she did come, it was very evident that she
had no pleasure in it; she made a slight, formal apology, for not
calling before, said not a word of wishing to see me again, and
was in every respect so altered a creature, that when she went
away I was perfectly resolved to continue the acquaintance no
longer.  I pity, though I cannot help blaming her.  She was very
wrong in singling me out as she did; I can safely say that every
advance to intimacy began on her side.  But I pity her, because
she must feel that she has been acting wrong, and because I am
very sure that anxiety for her brother is the cause of it.  I need
not explain myself farther; and though \emph{we} know this anxiety to
be quite needless, yet if she feels it, it will easily account
for her behaviour to me; and so deservedly dear as he is to his
sister, whatever anxiety she must feel on his behalf is natural
and amiable\index{amiable}.  I cannot but wonder, however, at her having any
such fears now, because, if he had at all cared about me, we
must have met, long ago.  He knows of my being in town, I am
certain, from something she said herself; and yet it would seem,
by her manner of talking, as if she wanted to persuade herself
that he is really partial\index{partial} to Miss Darcy.  I cannot understand it.
If I were not afraid of judging harshly, I should be almost
tempted to say that there is a strong appearance of \textbf{duplicity}\index{duplicity}\stepcounter{footnote}\footnotetext{
\textbf{duplicity} \emph{n.}  (doo-PLIH-sih-tee)
a fraudulent or deceptive representation.} in
all this.  But I will endeavour\index{endeavour} to banish every painful thought,
and think only of what will make me happy--your affection, and
the invariable\index{invariable} kindness of my dear uncle and aunt.  Let\index{Let} me hear
from you very soon.  Miss Bingley said something of his never
returning to Netherfield again, of giving up the house, but not
with any certainty.  We had better not mention it.  I am extremely
glad that you have such pleasant accounts from our friends at
Hunsford.  Pray go to see them, with Sir William and Maria.  I am
sure you will be very comfortable there.--Yours, etc."

This letter gave Elizabeth some pain; but her spirits returned as
she considered that Jane would no longer be \textbf{duped}\index{duped}\stepcounter{footnote}\footnotetext{
\textbf{dupe} \emph{n.} \textbf{-ed}  (DOOP)
to trick or deceive.}, by the sister
at least.  All expectation from the brother was now absolutely
over.  She would not even wish for a renewal of his attentions.
His character sunk on every review of it; and as a punishment for
him, as well as a possible advantage to Jane, she seriously hoped
he might really soon marry Mr. Darcy's sister, as by Wickham's
account, she would make him abundantly regret what he had
thrown away.

Mrs. Gardiner about this time reminded Elizabeth of her promise
concerning that gentleman, and required information; and
Elizabeth had such to send as might rather give \textbf{contentment}\index{contentment}\stepcounter{footnote}\footnotetext{
\textbf{contentment} \emph{n.}  (kuhn-TEHNT-muhnt)
satisfaction; happiness with one's situation in life.} to
her aunt than to herself.  His apparent partiality\index{partiality} had subsided,
his attentions were over, he was the admirer of some one else.
Elizabeth was watchful enough to see it all, but she could see it
and write of it without material pain.  Her heart had been but
slightly touched, and her vanity\index{vanity} was satisfied with believing that
\emph{she} would have been his only choice, had fortune permitted it.
The sudden acquisition of ten thousand pounds was the most
remarkable charm of the young lady to whom he was now rendering\index{rendering}
himself agreeable; but Elizabeth, less clear-sighted perhaps in
this case than in Charlotte's, did not quarrel with him for his
wish of independence.  Nothing, on the contrary, could be more
natural; and while able to suppose that it cost him a few struggle
to relinquish her, she was ready to allow it a wise and desirable
measure for both, and could very sincerely wish him happy.

All this was acknowledged to Mrs. Gardiner; and after relating
the circumstances, she thus went on: "I am now convinced, my
dear aunt, that I have never been much in love; for had I really
experienced that pure and elevating passion, I should at present
detest his very name, and wish him all manner of evil.  But my
feelings are not only cordial\index{cordial} towards \emph{him}; they are even
impartial\index{impartial} towards Miss King.  I cannot find out that I hate her at
all, or that I am in the least unwilling to think her a very good
sort of girl.  There can be no love in all this.  My watchfulness
has been \textbf{effectual}\index{effectual}\stepcounter{footnote}\footnotetext{
\textbf{effectual} \emph{adj.} \textbf{-ly}  (ih-FEHK-choo-uhl)
}; and though I certainly should be a more
interesting object to all my acquaintances were I distractedly\index{distractedly}
in love with him, I cannot say that I regret my comparative\index{comparative}
insignificance.  Importance may sometimes be purchased too
dearly.  Kitty and Lydia take his defection much more to heart
than I do.  They are young in the ways of the world, and not
yet open to the mortifying\index{mortifying} conviction\index{conviction} that handsome young men
must have something to live on as well as the plain."



\chapter{27}


With no greater events than these in the Longbourn family, and
otherwise diversified by little beyond the walks to Meryton,
sometimes dirty and sometimes cold, did January and February
pass away.  March was to take Elizabeth to Hunsford.  She had
not at first thought very seriously of going thither; but Charlotte,
she soon found, was depending on the plan and she gradually
learned to consider it herself with greater pleasure as well as
greater certainty.  Absence had increased her desire of seeing
Charlotte again, and weakened her disgust of Mr. Collins.  There
was novelty\index{novelty} in the scheme, and as, with such a mother and such
uncompanionable sisters, home could not be faultless, a little
change was not unwelcome for its own sake.  The journey
would moreover give her a peep at Jane; and, in short, as the
time drew near, she would have been very sorry for any delay.
Everything, however, went on smoothly, and was finally settled
according to Charlotte's first sketch.  She was to accompany Sir
William and his second daughter.  The improvement of spending
a night in London was added in time, and the plan became
perfect as plan could be.

The only pain was in leaving her father, who would certainly
miss her, and who, when it came to the point, so little liked her
going, that he told her to write to him, and almost promised to
answer her letter.

The farewell between herself and Mr. Wickham was perfectly
friendly; on his side even more.  His present pursuit could not
make him forget that Elizabeth had been the first to excite and to
deserve his attention, the first to listen and to pity, the first
to be admired; and in his manner of bidding her \textbf{adieu}\index{adieu}\stepcounter{footnote}\footnotetext{
\textbf{adieu} \emph{interj.}  (ad-YOOH)
}, wishing
her every enjoyment, reminding her of what she was to expect in
Lady Catherine de Bourgh, and trusting their opinion of her--their
opinion of everybody--would always coincide, there was a solicitude\index{solicitude},
an interest which she felt must ever attach her to him with a most
sincere regard; and she parted from him convinced that, whether
married or single, he must always be her model of the amiable\index{amiable} and
pleasing.

Her fellow-travellers the next day were not of a kind to make her
think him less agreeable.  Sir William Lucas, and his daughter
Maria, a good-humoured girl, but as empty-headed as himself,
had nothing to say that could be worth hearing, and were
listened to with about as much delight as the rattle of the chaise\index{chaise}.
Elizabeth loved \textbf{absurdities}\index{absurdities}\stepcounter{footnote}\footnotetext{
\textbf{} \emph{}  ()
}, but she had known Sir William's too
long.  He could tell her nothing new of the wonders of his
presentation and knighthood; and his civilities were worn out,
like his information.

It was a journey of only twenty-four miles, and they began it so
early as to be in Gracechurch Street by noon.  As they drove
to Mr. Gardiner's door, Jane was at a drawing-room window
watching their arrival; when they entered the passage she was
there to welcome them, and Elizabeth, looking earnestly\index{earnestly} in her
face, was pleased to see it healthful and lovely as ever.  On the
stairs were a troop of little boys and girls, whose eagerness for
their cousin's appearance would not allow them to wait in the
drawing-room, and whose shyness, as they had not seen her for
a twelvemonth, prevented their coming lower.  All was joy and
kindness.  The day passed most pleasantly away; the morning in
bustle and shopping, and the evening at one of the theatres.

Elizabeth then contrived\index{contrived} to sit by her aunt.  Their first object was
her sister; and she was more grieved\index{grieved} than astonished to hear, in
reply to her minute\index{minute} inquiries, that though Jane always struggled
to support her spirits, there were periods of dejection\index{dejection}.  It was
reasonable, however, to hope that they would not continue long.
Mrs. Gardiner gave her the particulars also of Miss Bingley's
visit in Gracechurch Street, and repeated conversations
occurring at different times between Jane and herself, which
proved that the former had, from her heart, given up the
acquaintance.

Mrs. Gardiner then \textbf{rallied}\index{rallied}\stepcounter{footnote}\footnotetext{
\textbf{rally} \emph{v.} \textbf{-ed}  (RAL-lee)
 to assemble or join together for a common purpose.} her niece on  Wickham's desertion,
and complimented her on bearing it so well.

"But my dear Elizabeth," she added, "what sort of girl is Miss
King?  I should be sorry to think our friend \textbf{mercenary}\index{mercenary}\stepcounter{footnote}\footnotetext{
\textbf{mercenary} \emph{adj.}  (MER-sehn-nair-ree)
}."

"Pray, my dear aunt, what is the difference in matrimonial
affairs, between the mercenary\index{mercenary} and the prudent\index{prudent} motive?  Where
does discretion\index{discretion} end, and \textbf{avarice}\index{avarice}\stepcounter{footnote}\footnotetext{
\textbf{avarice} \emph{n.}  (AV-uh-rihss)
insatiable desire for wealth; greed.} begin?  Last Christmas you
were afraid of his marrying me, because it would be imprudent\index{imprudent};
and now, because he is trying to get a girl with only ten
thousand pounds, you want to find out that he is mercenary\index{mercenary}."

"If you will only tell me what sort of girl Miss King is, I shall
know what to think."

"She is a very good kind of girl, I believe.  I know no harm of
her."

"But he paid her not the smallest attention till her grandfather's
death made her mistress of this fortune."

"No--what should he?  If it were not allowable for him to gain
\emph{my} affections because I had no money, what occasion could
there be for making love to a girl whom he did not care about,
and who was equally poor?"

"But there seems an indelicacy in directing his attentions
towards her so soon after this event."

"A man in distressed circumstances has not time for all those
elegant \textbf{decorums}\index{decorums}\stepcounter{footnote}\footnotetext{
\textbf{} \emph{}  ()
}\index{decorums} which other people may observe.  If \emph{she} does
not object to it, why should \emph{we}?"

"\emph{her} not objecting does not justify \emph{him}.  It only shows her
being deficient in something herself--sense or feeling."

"Well," cried Elizabeth, "have it as you choose.  \emph{he} shall be
mercenary\index{mercenary}, and \emph{she} shall be foolish."

"No, Lizzy, that is what I do \emph{not} choose.  I should be sorry,
you know, to think ill of a young man who has lived so long in
Derbyshire."

"Oh! if that is all, I have a very poor opinion of young men
who live in Derbyshire; and their intimate friends who live
in Hertfordshire are not much better.  I am sick of them all.
Thank Heaven!  I am going to-morrow where I shall find a man
who has not one agreeable quality, who has neither manner nor
sense to recommend him.  Stupid men are the only ones worth
knowing, after all."

"Take care, Lizzy; that speech \textbf{savours}\index{savours}\stepcounter{footnote}\footnotetext{
\textbf{savour} \emph{n.}  (SAY-ver)
 to taste with appreciation or pleasure.} strongly of disappointment."

Before they were separated by the conclusion of the play, she
had the unexpected happiness of an invitation to accompany her
uncle and aunt in a tour of pleasure which they proposed taking
in the summer.

"We have not determined how far it shall carry us," said Mrs.
Gardiner, "but, perhaps, to the Lakes."

No scheme could have been more agreeable to Elizabeth, and
her acceptance of the invitation was most ready and grateful.
"Oh, my dear, dear aunt," she rapturously\index{rapturously} cried, "what delight!
what felicity\index{felicity}!  You give me fresh life and vigour.  Adieu\index{Adieu} to
disappointment and \textbf{spleen}\index{spleen}\stepcounter{footnote}\footnotetext{
\textbf{spleen} \emph{n.}  (SPLEEN)
 a large dark-red organ on the left side of the body.}.  What are young men to rocks and
mountains?  Oh! what hours of transport we shall spend!  And
when we \emph{do} return, it shall not be like other travellers,
without being able to give one accurate idea of anything.  We
\emph{will} know where we have gone--we \emph{will} recollect what we have
seen.  Lakes, mountains, and rivers shall not be jumbled together
in our imaginations; nor when we attempt to describe any
particular scene, will we begin quarreling about its relative
situation.  Let\index{Let} \emph{our} first effusions\index{effusions} be less insupportable\index{insupportable} than
those of the \textbf{generality}\index{generality}\stepcounter{footnote}\footnotetext{
\textbf{generality} \emph{n.}  (jeh-ner-RAL-ih-tee)
} of travellers."



\chapter{28}


Every object in the next day's journey was new and interesting
to Elizabeth; and her spirits were in a state of enjoyment; for
she had seen her sister looking so well as to banish all fear for
her health, and the prospect\index{prospect} of her northern tour was a constant
source of delight.

When they left the high road for the lane to Hunsford, every eye
was in search of the Parsonage, and every turning expected to
bring it in view.  The \textbf{palings}\index{palings}\stepcounter{footnote}\footnotetext{
\textbf{paling} \emph{n.}  (PAY-ling)
} of Rosings Park was their boundary
on one side.  Elizabeth smiled at the recollection\index{recollection} of all that she
had heard of its inhabitants.

At length the Parsonage was \textbf{discernible}\index{discernible}\stepcounter{footnote}\footnotetext{
\textbf{discernible} \emph{adj.}  (dihs-SER-nih-bl)
perceptible by the senses or intellect.}.  The garden sloping to
the road, the house standing in it, the green \textbf{pales}\index{pales}\stepcounter{footnote}\footnotetext{
\textbf{pale} \emph{n.}  (PAYL)
}, and the laurel
\textbf{hedge}\index{hedge}\stepcounter{footnote}\footnotetext{
\textbf{hedge} \emph{n.}  (HEHDGE)
}, everything declared they were arriving.  Mr. Collins and
Charlotte appeared at the door, and the carriage stopped at the
small gate which led by a short gravel walk to the house, amidst
the nods and smiles of the whole party.  In a moment they were
all out of the chaise\index{chaise}, rejoicing at the sight of each other.
Mrs. Collins welcomed her friend with the liveliest pleasure,
and Elizabeth was more and more satisfied with coming when she
found herself so affectionately received.  She saw instantly that
her cousin's manners were not altered by his marriage; his formal
civility\index{civility} was just what it had been, and he detained her some
minutes\index{minutes} at the gate to hear and satisfy his inquiries after
all her family.  They were then, with no other delay than his
pointing out the neatness of the entrance, taken into the house;
and as soon as they were in the parlour, he welcomed them a
second time, with \textbf{ostentatious}\index{ostentatious}\stepcounter{footnote}\footnotetext{
\textbf{ostentatious} \emph{adj.}  (ahs-tehn-TAY-shuhs)
intended to attract notice and impress others; tawdry or vulgar.} formality to his humble abode\index{abode},
and punctually\index{punctually} repeated all his wife's offers of refreshment.

Elizabeth was prepared to see him in his glory; and she could
not help in fancying that in displaying the good proportion of
the room, its aspect\index{aspect} and its furniture, he addressed\index{addressed} himself
particularly to her, as if wishing to make her feel what she
had lost in refusing him.  But though everything seemed neat
and comfortable, she was not able to gratify him by any sigh of
repentance, and rather looked with wonder at her friend that
she could have so cheerful an air with such a companion.  When
Mr. Collins said anything of which his wife might reasonably be
ashamed, which certainly was not unseldom, she involuntarily
turned her eye on Charlotte.  Once or twice she could discern
a faint blush; but in general Charlotte wisely did not hear.
After sitting long enough to admire every article of furniture in
the room, from the sideboard to the fender, to give an account
of their journey, and of all that had happened in London, Mr.
Collins invited them to take a stroll in the garden, which was
large and well laid out, and to the \textbf{cultivation}\index{cultivation}\stepcounter{footnote}\footnotetext{
\textbf{cultivation} \emph{n.}  (kuhl-tih-VAY-shun)
socialization through training and education.} of which he
attended himself.  To work in this garden was one of his most
respectable pleasures; and Elizabeth admired the command of
countenance\index{countenance} with which Charlotte talked of the healthfulness of
the exercise, and owned she encouraged it as much as possible.
Here, leading the way through every walk and cross walk, and
scarcely allowing them an interval to utter the praises he asked
for, every view was pointed out with a minuteness which left
beauty entirely behind.  He could number the fields in every
direction, and could tell how many tress there were in the most
distant clump.  But of all the views which his garden, or which
the country or kingdom could boast, none were to be compared
with the prospect\index{prospect} of Rosings, afforded by an opening in the trees
that bordered the park nearly opposite the front of his house.  It
was a handsome modern building, well situated on rising ground.

From his garden, Mr. Collins would have led them round his two
meadows; but the ladies, not having shoes to encounter the
remains of a white frost, turned back; and while Sir William
accompanied him, Charlotte took her sister and friend over the
house, extremely well pleased, probably, to have the opportunity
of showing it without her husband's help.  It was rather small,
but well built and convenient; and everything was fitted up and
arranged with a neatness and consistency of which Elizabeth
gave Charlotte all the credit.  When Mr. Collins could be
forgotten, there was really an air of great comfort throughout,
and by Charlotte's evident enjoyment of it, Elizabeth supposed
he must be often forgotten.

She had already learnt that Lady Catherine was still in the
country.  It was spoken of again while they were at dinner,
when Mr. Collins joining in, observed:

"Yes, Miss Elizabeth, you will have the honour of seeing Lady
Catherine de Bourgh on the ensuing Sunday at church, and I
need not say you will be delighted with her.  She is all affability
and condescension\index{condescension}, and I doubt not but you will be honoured
with some portion of her notice when service is over.  I have
scarcely any hesitation in saying she will include you and my
sister Maria in every invitation with which she honours us during
your stay here.  Her behaviour to my dear Charlotte is charming.
We dine at Rosings twice every week, and are never allowed to
walk home.  Her ladyship's carriage is regularly\index{regularly} ordered for us.
I \emph{should} say, one of her ladyship's carriages, for she has
several."

"Lady Catherine is a very respectable, sensible woman indeed,"
added Charlotte, "and a most attentive neighbour."

"Very true, my dear, that is exactly what I say.  She is the sort
of woman whom one cannot regard with too much deference\index{deference}."

The evening was  spent chiefly in talking over Hertfordshire
news, and telling again what had already been written; and when
it closed, Elizabeth, in the solitude\index{solitude} of her chamber, had to
meditate\index{meditate} upon Charlotte's degree of contentment\index{contentment}, to understand
her address\index{address} in guiding, and composure\index{composure} in bearing with, her
husband, and to acknowledge that it was all done very well.  She
had also to anticipate how her visit would pass, the quiet \textbf{tenor}\index{tenor}\stepcounter{footnote}\footnotetext{
\textbf{tenor} \emph{n.}  (TEN-uhr)
}
of their usual employments, the \textbf{vexatious}\index{vexatious}\stepcounter{footnote}\footnotetext{
\textbf{vexatious} \emph{adj.}  (vex-SAY-shus)
causing irritation or annoyance.} interruptions of Mr.
Collins, and the gaieties of their intercourse\index{intercourse} with Rosings.
A lively imagination soon settled it all.

About the middle of the next day, as she was in her room getting
ready for a walk, a sudden noise below seemed to speak the
whole house in confusion; and, after listening a moment, she
heard somebody running upstairs in a violent hurry, and calling
loudly after her.  She opened the door and met Maria in the
landing place, who, breathless with agitation, cried out--

"Oh, my dear Eliza! pray make haste and come into the dining-room,
for there is such a sight to be seen!  I will not tell you what
it is.  Make haste, and come down this moment."

Elizabeth asked questions in vain; Maria would tell her nothing
more, and down they ran into the dining-room, which fronted
the lane, in quest of this wonder; It was two ladies stopping in
a low phaeton\index{phaeton} at the garden gate.

"And is this all?" cried Elizabeth.  "I expected at least that the
pigs were got into the garden, and here is nothing but Lady
Catherine and her daughter."

"La! my dear," said Maria, quite shocked at the mistake, "it is
not Lady Catherine.  The old lady is Mrs. Jenkinson, who lives
with them; the other is Miss de Bourgh.  Only look at her.  She
is quite a little creature.  Who would have thought that she could
be so thin and small?"

"She is abominably rude to keep Charlotte out of doors in all this
wind.  Why does she not come in?"

"Oh, Charlotte says she hardly ever does.  It is the greatest of
favours when Miss de Bourgh comes in."

"I like her appearance," said Elizabeth, struck with other ideas.
"She looks sickly and cross.  Yes, she will do for him very well.
She will make him a very proper wife."

Mr. Collins and Charlotte were both standing at the gate in
conversation with the ladies; and Sir William, to Elizabeth's
high diversion, was stationed in the doorway, in earnest\index{earnest}
contemplation\index{contemplation} of the greatness before him, and constantly
bowing whenever Miss de Bourgh looked that way.

At length there was nothing more to be said; the ladies drove on,
and the others returned into the house.  Mr. Collins no sooner
saw the two girls than he began to congratulate them on their
good fortune, which Charlotte explained by letting them know
that the whole party was asked to dine at Rosings the next day.



\chapter{29}


Mr. Collins's triumph\index{triumph}, in consequence\index{consequence} of this invitation, was
complete.  The power of displaying the grandeur\index{grandeur} of his patroness\index{patroness}
to his wondering visitors, and of letting them see her civility\index{civility}
towards himself and his wife, was exactly what he had wished
for; and that an opportunity of doing it should be given so soon,
was such an instance of Lady Catherine's condescension\index{condescension}, as he
knew not how to admire enough.

"I confess," said he, "that I should not have been at all surprised
by her ladyship's asking us on Sunday to drink tea and spend the
evening at Rosings.  I rather expected, from my knowledge of
her affability, that it would happen.  But who could have
foreseen such an attention as this?  Who could have imagined
that we should receive an invitation to dine there (an invitation,
moreover, including the whole party) so immediately after your
arrival!"

"I am the less surprised at what has happened," replied Sir
William, "from that knowledge of what the manners of the great
really are, which my situation in life has allowed me to acquire.
About the court, such instances of elegant breeding are not
uncommon."

Scarcely anything was talked of the whole day or next morning
but their visit to Rosings.  Mr. Collins was carefully instructing
them in what they were to expect, that the sight of such rooms,
so many servants, and so splendid a dinner, might not wholly\index{wholly}
overpower them.

When the ladies were separating for the toilette, he said to
Elizabeth--

"Do not make yourself uneasy, my dear cousin, about your
apparel.  Lady Catherine is far from requiring that elegance of
dress in us which becomes herself and her daughter.  I would
advise you merely to put on whatever of your clothes is superior\index{superior}
to the rest--there is no occasion for anything more.  Lady
Catherine will not think the worse of you for being simply
dressed.  She likes to have the distinction of rank preserved."

While they were dressing, he came two or three times to their
different doors, to recommend their being quick, as Lady
Catherine very much objected to be kept waiting for her dinner.
Such \textbf{formidable}\index{formidable}\stepcounter{footnote}\footnotetext{
\textbf{formidable} \emph{adj.}  (FORM-ih-duh-bl)
extremely impressive in strength or excellence; inspiring fear.} accounts of her ladyship, and her manner of
living, quite frightened Maria Lucas who had been little used to
company, and she looked forward\index{forward} to her introduction at Rosings
with as much apprehension\index{apprehension} as her father had done to his
presentation at St. James's.

As the weather was fine, they had a pleasant walk of about half
a mile across the park.  Every park has its beauty and its
prospects\index{prospects}; and Elizabeth saw much to be pleased with, though
she could not be in such raptures\index{raptures} as Mr. Collins expected the
scene to inspire, and was but slightly affected by his enumeration
of the windows in front of the house, and his relation of what the
glazing altogether had originally cost Sir Lewis de Bourgh.

When they ascended the steps to the hall, Maria's alarm was
every moment increasing, and even Sir William did not look
perfectly calm.  Elizabeth's courage did not fail her.  She had
heard nothing of Lady Catherine that spoke her awful from any
extraordinary talents or miraculous virtue\index{virtue}, and the mere
stateliness of money or rank she thought she could witness
without \textbf{trepidation}\index{trepidation}\stepcounter{footnote}\footnotetext{
\textbf{trepidation} \emph{n.}  (treh-pih-DAY-shun)
a feeling of alarm or dread.}.

From the entrance-hall, of which Mr. Collins pointed out, with a
rapturous\index{rapturous} air, the fine proportion and the finished ornaments,
they followed the servants through an \textbf{ante-chamber}\index{ante-chamber}\stepcounter{footnote}\footnotetext{
\textbf{antechamber} \emph{n.}  (AN-tih-CHAYM-ber)
an entrance or reception room or area.}, to the room
where Lady Catherine, her daughter, and Mrs. Jenkinson were
sitting.  Her ladyship, with great condescension\index{condescension}, arose to receive
them; and as Mrs. Collins had settled it with her husband that the
office of introduction should be hers, it was performed in a
proper manner, without any of those apologies and thanks which
he would have thought necessary.

In spite of having been at St. James's Sir William was so
completely awed\index{awed} by the grandeur\index{grandeur} surrounding him, that he had
but just courage enough to make a very low bow, and take his
seat without saying a word; and his daughter, frightened almost
out of her senses, sat on the edge of her chair, not knowing
which way to look.  Elizabeth found herself quite equal to the
scene, and could observe the three ladies before her composedly.
Lady Catherine was a tall, large woman, with strongly-marked
features, which might once have been handsome.  Her air was
not conciliating, nor was her manner of receiving them such as
to make her visitors forget their inferior\index{inferior} rank.  She was not
rendered\index{rendered} formidable\index{formidable} by silence; but whatever she said was
spoken in so authoritative\index{authoritative} a tone, as marked her self-importance,
and brought Mr. Wickham immediately to Elizabeth's mind; and
from the observation of the day altogether, she believed Lady
Catherine to be exactly what he represented.

When, after examining the mother, in whose countenance\index{countenance} and
\textbf{deportment}\index{deportment}\stepcounter{footnote}\footnotetext{
\textbf{deportment} \emph{n.}  (dee-PORT-muhnt)
behavior or personal conduct.} she soon found some resemblance of Mr. Darcy, she
turned her eyes on the daughter, she could almost have joined
in Maria's astonishment at her being so thin and so small.  There
was neither in figure nor face any likeness between the ladies.
Miss de Bourgh was pale\index{pale} and sickly; her features, though not
plain, were insignificant; and she spoke very little, except in
a low voice, to Mrs. Jenkinson, in whose appearance there was
nothing remarkable, and who was entirely engaged in listening to
what she said, and placing a screen in the proper direction before
her eyes.

After sitting a few minutes\index{minutes}, they were all sent to one of the
windows to admire the view, Mr. Collins attending them to
point out its beauties, and Lady Catherine kindly informing
them that it was much better worth looking at in the summer.

The dinner was exceedingly handsome, and there were all the
servants and all the articles of plate which Mr. Collins had
promised; and, as he had likewise foretold, he took his seat at
the bottom of the table, by her ladyship's desire, and looked as
if he felt that life could furnish nothing greater.  He carved,
and ate, and praised with delighted alacrity\index{alacrity}; and every dish was
commended\index{commended}, first by him and then by Sir William, who was now
enough recovered to echo whatever his son-in-law said, in a
manner which Elizabeth wondered Lady Catherine could bear.
But Lady Catherine seemed gratified by their excessive admiration,
and gave most gracious smiles, especially when any dish on the
table proved a novelty\index{novelty} to them.  The party did not supply much
conversation.  Elizabeth was ready to speak whenever there was
an opening, but she was seated between Charlotte and Miss de
Bourgh--the former of whom was engaged in listening to Lady
Catherine, and the latter said not a word to her all dinner-time.
Mrs. Jenkinson was chiefly employed in watching how little Miss
de Bourgh ate, pressing her to try some other dish, and fearing
she was \textbf{indisposed}\index{indisposed}\stepcounter{footnote}\footnotetext{
\textbf{indispose} \emph{tr.v.} \textbf{-ed}  (ihn-dihs-POAZ)
(usually followed by `to') strongly opposed; disinclined.}.  Maria thought speaking out of the question,
and the gentlemen did nothing but eat and admire.

When the ladies returned to the drawing-room, there was little
to be done but to hear Lady Catherine talk, which she did
without any intermission\index{intermission} till coffee came in, delivering her
opinion on every subject\index{subject} in so decisive a manner, as proved
that she was not used to have her judgement controverted.  She
inquired into Charlotte's domestic\index{domestic} concerns familiarly and
\textbf{minutely}\index{minutely}\stepcounter{footnote}\footnotetext{
\textbf{minutely} \emph{adj.}  (my-NYOOT)
}\index{minutely\index{minutely}}, gave her a great deal of advice as to the management
of them all; told her how everything ought to be regulated in so
small a family as hers, and instructed her as to the care of her
cows and her poultry.  Elizabeth found that nothing was beneath
this great lady's attention, which could furnish her with an
occasion of dictating to others.  In the intervals of her discourse\index{discourse}
with Mrs. Collins, she addressed\index{addressed} a variety of questions to Maria
and Elizabeth, but especially to the latter, of whose connections
she knew the least, and who she observed to Mrs. Collins was a
very genteel\index{genteel}, pretty kind of girl.  She asked her, at different
times, how many sisters she had, whether they were older or
younger than herself, whether any of them were likely to be
married, whether they were handsome, where they had been
educated, what carriage her father kept, and what had been her
mother's maiden name?  Elizabeth felt all the impertinence\index{impertinence} of her
questions but answered them very composedly.  Lady Catherine
then observed,

"Your father's estate is entailed\index{entailed} on Mr. Collins, I think.  For
your sake," turning to Charlotte, "I am glad of it; but otherwise I
see no occasion for entailing\index{entailing} estates from the female line.  It was
not thought necessary in Sir Lewis de Bourgh's family.  Do you
play and sing, Miss Bennet?"

"A little."

"Oh! then--some time or other we shall be happy to hear you.
Our instrument is a capital\index{capital} one, probably superior\index{superior} to----You
shall try it some day.  Do your sisters play and sing?"

"One of them does."

"Why did not you all learn?  You ought all to have learned.  The
Miss Webbs all play, and their father has not so good an income
as yours.  Do you draw?"

"No, not at all."

"What, none of you?"

"Not one."

"That is very strange.  But I suppose you had no opportunity.
Your mother should have taken you to town every spring for the
benefit of masters."

"My mother would have had no objection, but my father hates
London."

"Has your governess left you?"

"We never had any governess."

"No governess!  How was that possible?  Five daughters brought
up at home without a governess!  I never heard of such a thing.
Your mother must have been quite a slave to your education."

Elizabeth could hardly help smiling as she assured her that had
not been the case.

"Then, who taught you?  who attended to you?  Without a
governess, you must have been neglected."

"Compared with some families, I believe we were; but such of us
as wished to learn never wanted the means.  We were always
encouraged to read, and had all the masters that were necessary.
Those who chose to be idle, certainly might."

"Aye, no doubt; but that is what a governess will prevent, and
if I had known your mother, I should have advised her most
strenuously to engage one.  I always say that nothing is to be
done in education without steady\index{steady} and regular\index{regular} instruction, and
nobody but a governess can give it.  It is wonderful how many
families I have been the means of supplying in that way.  I am
always glad to get a young person well placed out.  Four nieces
of Mrs. Jenkinson are most delightfully situated through my
means; and it was but the other day that I recommended another
young person, who was merely accidentally mentioned to me,
and the family are quite delighted with her.  Mrs. Collins, did I
tell you of Lady Metcalf's calling yesterday to thank me?  She
finds Miss Pope a treasure.  'Lady Catherine,' said she, 'you
have given me a treasure.'  Are any of your younger sisters out,
Miss Bennet?"

"Yes, ma'am, all."

"All!  What, all five out at once?  Very odd!  And you only
the second.  The younger ones out before the elder ones are
married!  Your younger sisters must be very young?"

"Yes, my youngest is not sixteen.  Perhaps \emph{she} is full young to
be much in company.  But really, ma'am, I think it would be
very hard upon younger sisters, that they should not have their
share of society and amusement, because the elder may not have
the means or inclination\index{inclination} to marry early.  The last-born has as
good a right to the pleasures of youth at the first.  And to be
kept back on \emph{such} a motive!  I think it would not be very likely
to promote sisterly affection or delicacy\index{delicacy} of mind."

"Upon my word," said her ladyship, "you give your opinion very
decidedly for so young a person.  Pray, what is your age?"

"With three younger sisters grown up," replied Elizabeth,
smiling, "your ladyship can hardly expect me to own it."

Lady Catherine seemed quite astonished at not receiving a direct
answer; and Elizabeth suspected herself to be the first creature
who had ever dared to trifle with so much dignified impertinence\index{impertinence}.

"You cannot be more than twenty, I am sure, therefore you need
not conceal your age."

"I am not one-and-twenty."

When the gentlemen had joined them, and tea was over, the
card-tables were placed.  Lady Catherine, Sir William, and Mr.
and Mrs. Collins sat down to quadrille\index{quadrille}; and as Miss de Bourgh
chose to play at \textbf{cassino}\index{cassino}\stepcounter{footnote}\footnotetext{
\textbf{casino} \emph{n.}  (kuh-SEE-noh)
a public room for gambling.}, the two girls had the honour of
assisting Mrs. Jenkinson to make up her party.  Their table was
\textbf{superlatively}\index{superlatively}\stepcounter{footnote}\footnotetext{
\textbf{superlative} \emph{adj.} \textbf{-ly}  (soo-PER-luh-TIHV)
the best; superior.} stupid.  Scarcely a syllable was uttered that did
not relate to the game, except when Mrs. Jenkinson expressed her
fears of Miss de Bourgh's being too hot or too cold, or having
too much or too little light.  A great deal more passed at the
other table.  Lady Catherine was generally speaking--stating
the mistakes of the three others, or relating some anecdote\index{anecdote} of
herself.  Mr. Collins was employed in agreeing to everything
her ladyship said, thanking her for every fish he won, and
apologising if he thought he won too many.  Sir William did not
say much.  He was storing his memory with anecdotes\index{anecdotes} and noble
names.

When Lady Catherine and her daughter had played as long as
they chose, the tables were broken up, the carriage was offered
to Mrs. Collins, gratefully accepted and immediately ordered.
The party then gathered round the fire to hear Lady Catherine
determine what weather they were to have on the morrow.  From
these instructions they were summoned\index{summoned} by the arrival of the
coach; and with many speeches of thankfulness on Mr. Collins's
side and as many bows on Sir William's they departed.  As soon
as they had driven from the door, Elizabeth was called on by her
cousin to give her opinion of all that she had seen at Rosings,
which, for Charlotte's sake, she made more favourable than it
really was.  But her commendation\index{commendation}, though costing her some
trouble, could by no means satisfy Mr. Collins, and he was very
soon obliged to take her ladyship's praise into his own hands.



\chapter{30}


Sir William stayed only a week at Hunsford, but his visit was
long enough to convince him of his daughter's being most
comfortably settled, and of her possessing such a husband and
such a neighbour as were not often met with.  While Sir William
was with them, Mr. Collins devoted his morning to driving him
out in his \textbf{gig}\index{gig}\stepcounter{footnote}\footnotetext{
\textbf{gig} \emph{n.}  (GIG)
}, and showing him the country; but when he went
away, the whole family returned to their usual employments, and
Elizabeth was thankful to find that they did not see more of
her cousin by the alteration, for the chief of the time between
breakfast and dinner was now passed by him either at work in
the garden or in reading and writing, and looking out of the
window in his own book-room, which fronted the road.  The
room in which the ladies sat was backwards.  Elizabeth had at
first rather wondered that Charlotte should not prefer the
dining-parlour for common use; it was a better sized room, and
had a more pleasant aspect\index{aspect}; but she soon saw that her friend
had an excellent reason for what she did, for Mr. Collins would
undoubtedly have been much less in his own apartment\index{apartment}, had they
sat in one equally lively; and she gave Charlotte credit for
the arrangement.

From the drawing-room they could distinguish nothing in the
lane, and were indebted to Mr. Collins for the knowledge of
what carriages went along, and how often especially Miss de
Bourgh drove by in her phaeton\index{phaeton}, which he never failed coming
to inform them of, though it happened almost every day.  She
not unfrequently stopped at the Parsonage, and had a few
minutes\index{minutes}' conversation with Charlotte, but was scarcely ever
prevailed\index{prevailed} upon to get out.

Very few days passed in which Mr. Collins did not walk to
Rosings, and not many in which his wife did not think it
necessary to go likewise; and till Elizabeth recollected that
there might be other family livings to be disposed\index{disposed} of, she could
not understand the sacrifice of so many hours.  Now and then
they were honoured with a call from her ladyship, and nothing
escaped her observation that was passing in the room during
these visits.  She examined into their employments, looked at
their work, and advised them to do it differently; found fault
with the arrangement of the furniture; or detected the housemaid
in negligence; and if she accepted any refreshment, seemed to do
it only for the sake of finding out that Mrs. Collins's joints of
meat were too large for her family.

Elizabeth soon perceived\index{perceived}, that though this great lady was not in
commission of the peace of the county, she was a most active
\textbf{magistrate}\index{magistrate}\stepcounter{footnote}\footnotetext{
\textbf{magistrate} \emph{n.}  (MAJ-jihs-trayt)
a public official authorized to decide questions bought before a court of justice.} in her own parish\index{parish}, the \textbf{minutest}\index{minutest}\stepcounter{footnote}\footnotetext{
\textbf{} \emph{}  ()
} concerns of which
were carried to her by Mr. Collins; and whenever any of the
cottagers were disposed\index{disposed} to be quarrelsome, discontented, or
too poor, she \textbf{sallied}\index{sallied}\stepcounter{footnote}\footnotetext{
\textbf{sally} \emph{intr.v.} \textbf{-ed}  (sal-yoo-TAY-shun)
} forth into the village to settle their
differences, silence their complaints, and scold them into
harmony and plenty.

The entertainment of dining at Rosings was repeated about twice
a week; and, allowing for the loss of Sir William, and there being
only one card-table in the evening, every such entertainment was
the counterpart of the first.  Their other engagements were few,
as the style of living in the neighbourhood in general was beyond
Mr. Collins's reach.  This, however, was no evil to Elizabeth,
and upon the whole she spent her time comfortably enough;
there were half-hours of pleasant conversation with Charlotte,
and the weather was so fine for the time of year that she had
often great enjoyment out of doors.  Her favourite walk, and
where she frequently went while the others were calling on Lady
Catherine, was along the open grove which edged that side of
the park, where there was a nice sheltered path, which no one
seemed to value but herself, and where she felt beyond the reach
of Lady Catherine's curiosity.

In this quiet way, the first fortnight\index{fortnight} of her visit soon passed
away.  Easter was approaching, and the week preceding it was
to bring an addition to the family at Rosings, which in so small
a circle must be important.  Elizabeth had heard soon after her
arrival that Mr. Darcy was expected there in the course of a few
weeks, and though there were not many of her acquaintances whom
she did not prefer, his coming would furnish one comparatively\index{comparatively}
new to look at in their Rosings parties, and she might be amused
in seeing how hopeless Miss Bingley's designs on him were, by
his behaviour to his cousin, for whom he was evidently destined
by Lady Catherine, who talked of his coming with the greatest
satisfaction, spoke of him in terms of the highest admiration,
and seemed almost angry to find that he had already been
frequently seen by Miss Lucas and herself.

His arrival was soon known at the Parsonage; for Mr. Collins
was walking the whole morning within view of the lodges
opening into Hunsford Lane, in order to have the earliest
assurance of it, and after making his bow as the carriage turned
into the Park, hurried home with the great intelligence.  On the
following morning he hastened to Rosings to pay his respects.
There were two nephews of Lady Catherine to require them, for
Mr. Darcy had brought with him a Colonel Fitzwilliam, the
younger son of his uncle Lord ----, and, to the great surprise
of all the party, when Mr. Collins returned, the gentleman
accompanied him.  Charlotte had seen them from her husband's
room, crossing the road, and immediately running into the other,
told the girls what an honour they might expect, adding:

"I may thank you, Eliza, for this piece of civility\index{civility}.  Mr. Darcy
would never have come so soon to wait upon me."

Elizabeth had scarcely time to disclaim all right to the
compliment, before their approach was announced by the
door-bell, and shortly afterwards the three gentlemen entered
the room.  Colonel Fitzwilliam, who led the way, was about
thirty, not handsome, but in person and address\index{address} most truly the
gentleman.  Mr. Darcy looked just as he had been used to look
in Hertfordshire--paid his compliments, with his usual reserve,
to Mrs. Collins, and whatever might be his feelings toward her
friend, met her with every appearance of composure\index{composure}.  Elizabeth
merely curtseyed to him without saying a word.

Colonel Fitzwilliam entered into conversation directly with
the readiness and ease of a well-bred man, and talked very
pleasantly; but his cousin, after having addressed\index{addressed} a slight
observation on the house and garden to Mrs. Collins, sat for
some time without speaking to anybody.  At length, however,
his civility\index{civility} was so far awakened as to inquire of Elizabeth after
the health of her family.  She answered him in the usual way,
and after a moment's pause, added:

"My eldest sister has been in town these three months.  Have
you never happened to see her there?"

She was perfectly sensible that he never had; but she wished
to see whether he would betray any consciousness of what had
passed between the Bingleys and Jane, and she thought he
looked a little confused as he answered that he had never been
so fortunate as to meet Miss Bennet.  The subject\index{subject} was pursued
no farther, and the gentlemen soon afterwards went away.



\chapter{31}


Colonel Fitzwilliam's manners were very much admired at the
Parsonage, and the ladies all felt that he must add considerably
to the pleasures of their engagements at Rosings.  It was some
days, however, before they received any invitation thither--for
while there were visitors in the house, they could not be
necessary; and it was not till Easter-day, almost a week after the
gentlemen's arrival, that they were honoured by such an
attention, and then they were merely asked on leaving church to
come there in the evening.  For the last week they had seen very
little of Lady Catherine or her daughter.  Colonel Fitzwilliam had
called at the Parsonage more than once during the time, but Mr.
Darcy they had seen only at church.

The invitation was accepted of course, and at a proper hour they
joined the party in Lady Catherine's drawing-room.  Her ladyship
received them civilly, but it was plain that their company was by
no means so acceptable as when she could get nobody else; and
she was, in fact, almost engrossed\index{engrossed} by her nephews, speaking to
them, especially to Darcy, much more than to any other person
in the room.

Colonel Fitzwilliam seemed really glad to see them; anything
was a welcome relief to him at Rosings; and Mrs. Collins's
pretty friend had moreover caught his fancy very much.  He now
seated himself by her, and talked so agreeably of Kent and
Hertfordshire, of travelling and staying at home, of new books
and music, that Elizabeth had never been half so well entertained
in that room before; and they conversed with so much spirit and
flow, as to draw the attention of Lady Catherine herself, as well
as of Mr. Darcy.  \emph{his} eyes had been soon and repeatedly turned
towards them with a look of curiosity; and that her ladyship,
after a while, shared the feeling, was more openly acknowledged,
for she did not scruple\index{scruple} to call out:

"What is that you are saying, Fitzwilliam?  What is it you are
talking of?  What are you telling Miss Bennet?  Let\index{Let} me hear
what it is."

"We are speaking of music, madam," said he, when no longer
able to avoid a reply.

"Of music!  Then pray speak aloud.  It is of all subjects\index{subjects} my
delight.  I must have my share in the conversation if you are
speaking of music.  There are few people in England, I suppose,
who have more true enjoyment of music than myself, or a better
natural taste.  If I had ever learnt, I should have been a great
proficient.  And so would Anne, if her health had allowed her to
apply.  I am confident that she would have performed delightfully.
How does Georgiana get on, Darcy?"

Mr. Darcy spoke with affectionate praise of his sister's
proficiency.

"I am very glad to hear such a good account of her," said Lady
Catherine; "and pray tell her from me, that she cannot expect to
excel if she does not practice a good deal."

"I assure you, madam," he replied, "that she does not need such
advice.  She practises very constantly."

"So much the better.  It cannot be done too much; and when I
next write to her, I shall charge her not to neglect it on any
account.  I often tell young ladies that no excellence in music
is to be acquired without constant practice.  I have told Miss
Bennet several times, that she will never play really well unless
she practises more; and though Mrs. Collins has no instrument,
she is very welcome, as I have often told her, to come to Rosings
every day, and play on the pianoforte\index{pianoforte} in Mrs. Jenkinson's room.
She would be in nobody's way, you know, in that part of the house."

Mr. Darcy looked a little ashamed of his aunt's ill-breeding, and
made no answer.

When coffee was over, Colonel Fitzwilliam reminded Elizabeth
of having promised to play to him; and she sat down directly to
the instrument.  He drew a chair near her.  Lady Catherine
listened to half a song, and then talked, as before, to her other
nephew; till the latter walked away from her, and making with
his usual deliberation\index{deliberation} towards the pianoforte\index{pianoforte} stationed himself
so as to command a full view of the fair performer's countenance\index{countenance}.
Elizabeth saw what he was doing, and at the first convenient
pause, turned to him with an arch\index{arch} smile, and said:

"You mean to frighten me, Mr. Darcy, by coming in all this state
to hear me?  I will not be alarmed though your sister \emph{does} play
so well.  There is a stubbornness about me that never can bear to
be frightened at the will of others.  My courage always rises at
every attempt to intimidate me."

"I shall not say you are mistaken," he replied, "because you
could not really believe me to entertain any design of alarming
you; and I have had the pleasure of your acquaintance long
enough to know that you find great enjoyment in occasionally
professing opinions which in fact are not your own."

Elizabeth laughed heartily\index{heartily} at this picture of herself, and said to
Colonel Fitzwilliam, "Your cousin will give you a very pretty
notion of me, and teach you not to believe a word I say.  I am
particularly unlucky in meeting with a person so able to expose
my real character, in a part of the world where I had hoped to
pass myself off with some degree of credit.  Indeed, Mr. Darcy,
it is very ungenerous in you to mention all that you knew to my
disadvantage in Hertfordshire--and, give me leave to say, very
\textbf{impolitic}\index{impolitic}\stepcounter{footnote}\footnotetext{
\textbf{impolitic} \emph{adj.}  (ihm-PAH-lih-tihk)
 Socially unwise; indiscreet; inexpedient; undiplomatic.} too--for it is provoking me to \textbf{retaliate}\index{retaliate}\stepcounter{footnote}\footnotetext{
\textbf{retaliate} \emph{v.}  (rih-TAL-lih-ayt)
strike back or take revenge; to respond 'tit-for-tat'.}, and such
things may come out as will shock your relations to hear."

"I am not afraid of you," said he, smilingly.

"Pray let\index{let} me hear what you have to accuse him of," cried
Colonel Fitzwilliam.  "I should like to know how he behaves
among strangers."

"You shall hear then--but prepare yourself for something very
dreadful.  The first time of my ever seeing him in Hertfordshire,
you must know, was at a ball--and at this ball, what do you
think he did?  He danced only four dances, though gentlemen
were scarce; and, to my certain knowledge, more than one
young lady was sitting down in want of a partner.  Mr. Darcy,
you cannot deny the fact."

"I had not at that time the honour of knowing any lady in the
assembly beyond my own party."

"True; and nobody can ever be introduced in a ball-room.  Well,
Colonel Fitzwilliam, what do I play next?  My fingers wait your
orders."

"Perhaps," said Darcy, "I should have judged better, had I
sought an introduction; but I am ill-qualified to recommend
myself to strangers."

"Shall we ask your cousin the reason of this?" said Elizabeth,
still addressing\index{addressing} Colonel Fitzwilliam.  "Shall we ask him why a
man of sense and education, and who has lived in the world, is
ill qualified to recommend himself to strangers?"

"I can answer your question," said Fitzwilliam, "without
applying to him.  It is because he will not give himself the
trouble."

"I certainly have not the talent which some people possess," said
Darcy, "of conversing easily with those I have never seen before.
I cannot catch their tone of conversation, or appear interested
in their concerns, as I often see done."

"My fingers," said Elizabeth, "do not move over this instrument
in the masterly manner which I see so many women's do.  They
have not the same force or rapidity, and do not produce the
same expression.  But then I have always supposed it to be my
own fault--because I will not take the trouble of practising.
It is not that I do not believe \emph{my} fingers as capable as any
other woman's of superior\index{superior} execution\index{execution}."

Darcy smiled and said, "You are perfectly right.  You have
employed your time much better.  No one admitted to the
privilege of hearing you can think anything wanting.  We neither
of us perform to strangers."

Here they were interrupted by Lady Catherine, who called out to
know what they were talking of.  Elizabeth immediately began
playing again.  Lady Catherine approached, and, after listening
for a few minutes\index{minutes}, said to Darcy:

"Miss Bennet would not play at all \textbf{amiss}\index{amiss}\stepcounter{footnote}\footnotetext{
\textbf{amiss} \emph{adj.}  (uh-MIHS)
 not functioning properly; out of order.} if she practised more,
and could have the advantage of a London master.  She has a
very good notion of fingering, though her taste is not equal to
Anne's.  Anne would have been a delightful performer, had her
health allowed her to learn."

Elizabeth looked at Darcy to see how cordially\index{cordially} he assented\index{assented} to
his cousin's praise; but neither at that moment nor at any other
could she discern any symptom of love; and from the whole of
his behaviour to Miss de Bourgh she derived this comfort for
Miss Bingley, that he might have been just as likely to marry
\emph{her}, had she been his relation.

Lady Catherine continued her remarks on Elizabeth's performance,
mixing with them many instructions on execution\index{execution} and taste.
Elizabeth received them with all the forbearance\index{forbearance} of civility\index{civility},
and, at the request of the gentlemen, remained at the instrument
till her ladyship's carriage was ready to take them all home.



\chapter{32}


Elizabeth was sitting by herself the next morning, and writing to
Jane while Mrs. Collins and Maria were gone on business into
the village, when she was startled by a ring at the door, the
certain signal of a visitor.  As she had heard no carriage, she
thought it not unlikely to be Lady Catherine, and under that
apprehension\index{apprehension} was putting away her half-finished letter that she
might escape all impertinent\index{impertinent} questions, when the door opened,
and, to her very great surprise, Mr. Darcy, and Mr. Darcy only,
entered the room.

He seemed astonished too on finding her alone, and apologised
for his intrusion by letting her know that he had understood all
the ladies were to be within.

They then sat down, and when her inquiries after Rosings were
made, seemed in danger of sinking into total silence.  It was
absolutely necessary, therefore, to think of something, and in
this emergence recollecting \emph{when} she had seen him last in
Hertfordshire, and feeling curious to know what he would say
on the subject\index{subject} of their hasty departure, she observed:

"How very suddenly you all quitted Netherfield last November,
Mr. Darcy!  It must have been a most agreeable surprise to Mr.
Bingley to see you all after him so soon; for, if I recollect right,
he went but the day before.  He and his sisters were well, I hope,
when you left London?"

"Perfectly so, I thank you."

She found that she was to receive no other answer, and, after a
short pause added:

"I think I have understood that Mr. Bingley has not much idea of
ever returning to Netherfield again?"

"I have never heard him say so; but it is probable\index{probable} that he may
spend very little of his time there in the future.  He has many
friends, and is at a time of life when friends and engagements are
continually increasing."

"If he means to be but little at Netherfield, it would be better for
the neighbourhood that he should give up the place entirely, for
then we might possibly get a settled family there.  But, perhaps,
Mr. Bingley did not take the house so much for the convenience
of the neighbourhood as for his own, and we must expect him to
keep it or quit it on the same principle."

"I should not be surprised," said Darcy, "if he were to give it up
as soon as any eligible purchase offers."

Elizabeth made no answer.  She was afraid of talking longer of
his friend; and, having nothing else to say, was now determined
to leave the trouble of finding a subject\index{subject} to him.

He took the hint, and soon began with, "This seems a very
comfortable house.  Lady Catherine, I believe, did a great deal to
it when Mr. Collins first came to Hunsford."

"I believe she did--and I am sure she could not have bestowed\index{bestowed}
her kindness on a more grateful object."

"Mr. Collins appears to be very fortunate in his choice of a wife."

"Yes, indeed, his friends may well rejoice in his having met with
one of the very few sensible women who would have accepted
him, or have made him happy if they had.  My friend has an
excellent understanding--though I am not certain that I consider
her marrying Mr. Collins as the wisest thing she ever did.  She
seems perfectly happy, however, and in a \textbf{prudential}\index{prudential}\stepcounter{footnote}\footnotetext{
\textbf{prudential} \emph{adj.}  (proo-DEHN-chuhl)
good judgment, or common sense, and the ability to avoid risk or embarassment.} light it is
certainly a very good match for her."

"It must be very agreeable for her to be settled within so easy a
distance of her own family and friends."

"An easy distance, do you call it?  It is nearly fifty miles."

"And what is fifty miles of good road?  Little more than half a
day's journey.  Yes, I call it a \emph{very} easy distance."

"I should never have considered the distance as one of the
\emph{advantages} of the match," cried Elizabeth.  "I should never
have said Mrs. Collins was settled \emph{near} her family."

"It is a proof of your own attachment to Hertfordshire.
Anything beyond the very neighbourhood of Longbourn, I
suppose, would appear far."

As he spoke there was a sort of smile which Elizabeth fancied
she understood; he must be supposing her to be thinking of Jane
and Netherfield, and she blushed as she answered:

"I do not mean to say that a woman may not be settled too near
her family.  The far and the near must be relative, and depend on
many varying circumstances.  Where there is fortune to make the
expenses of travelling unimportant, distance becomes no evil.
But that is not the case \emph{here}.  Mr. and Mrs. Collins have a
comfortable income, but not such a one as will allow of frequent
journeys--and I am persuaded my friend would not call herself
\emph{near} her family under less than \emph{half} the present distance."

Mr. Darcy drew his chair a little towards her, and said, "\emph{you}
cannot have a right to such very strong local attachment.
\emph{you} cannot have been always at Longbourn."

Elizabeth looked surprised.  The gentleman experienced some
change of feeling; he drew back his chair, took a newspaper
from the table, and glancing over it, said, in a colder voice:

"Are you pleased with Kent?"

A short dialogue on the subject\index{subject} of the country ensued, on either
side calm and concise--and soon put an end to by the entrance
of Charlotte and her sister, just returned from her walk.  The
tete-a-tete surprised them.  Mr. Darcy related the mistake which
had occasioned his intruding on Miss Bennet, and after sitting a
few minutes\index{minutes} longer without saying much to anybody, went away.

"What can be the meaning of this?" said Charlotte, as soon as he
was gone.  "My dear, Eliza, he must be in love with you, or he
would never have called us in this familiar way."

But when Elizabeth told of his silence; it did not seem very
likely, even to Charlotte's wishes, to be the case; and after
various conjectures\index{conjectures}, they could at last only suppose his visit to
proceed from the difficulty of finding anything to do, which was
the more probable\index{probable} from the time of year.  All field sports were
over.  Within doors there was Lady Catherine, books, and a
billiard-table, but gentlemen cannot always be within doors; and
in the nearness of the Parsonage, or the pleasantness of the walk
to it, or of the people who lived in it, the two cousins found a
temptation from this period of walking thither almost every day.
They called at various times of the morning, sometimes
separately, sometimes together, and now and then accompanied
by their aunt.  It was plain to them all that Colonel Fitzwilliam
came because he had pleasure in their society, a persuasion
which of course recommended him still more; and Elizabeth was
reminded by her own satisfaction in being with him, as well as by
his evident admiration of her, of her former favourite George
Wickham; and though, in comparing them, she saw there was
less captivating softness in Colonel Fitzwilliam's manners, she
believed he might have the best informed mind.

But why Mr. Darcy came so often to the Parsonage, it was more
difficult to understand.  It could not be for society, as he
frequently sat there ten minutes\index{minutes} together without opening his
lips; and when he did speak, it seemed the effect of necessity
rather than of choice--a sacrifice to propriety\index{propriety}, not a pleasure
to himself.  He seldom appeared really animated.  Mrs. Collins
knew not what to make of him.  Colonel Fitzwilliam's occasionally
laughing at his stupidity, proved that he was generally different,
which her own knowledge of him could not have told her; and as
she would liked to have believed this change the effect of love,
and the object of that love her friend Eliza, she set herself
seriously to work to find it out.  She watched him whenever they
were at Rosings, and whenever he came to Hunsford; but without
much success.  He certainly looked at her friend a great deal,
but the expression of that look was disputable.  It was an
earnest\index{earnest}, steadfast\index{steadfast} gaze, but she often doubted whether there
were much admiration in it, and sometimes it seemed nothing but
absence of mind.

She had once or twice suggested to Elizabeth the possibility of
his being partial\index{partial} to her, but Elizabeth always laughed at the idea;
and Mrs. Collins did not think it right to press the subject\index{subject}, from
the danger of raising expectations which might only end in
disappointment; for in her opinion it admitted not of a doubt,
that all her friend's dislike would vanish, if she could suppose
him to be in her power.


In her kind schemes for Elizabeth, she sometimes planned her
marrying Colonel Fitzwilliam.  He was beyond comparison the
most pleasant man; he certainly admired her, and his situation in
life was most eligible; but, to counterbalance these advantages,
Mr. Darcy had considerable patronage\index{patronage} in the church, and his
cousin could have none at all.



\chapter{33}


More than once did Elizabeth, in her ramble\index{ramble} within the park,
unexpectedly meet Mr. Darcy.  She felt all the perverseness of
the \textbf{mischance}\index{mischance}\stepcounter{footnote}\footnotetext{
\textbf{mischance} \emph{n.}  (mihs-CHANS)
a mishap; bad luck.} that should bring him where no one else was brought,
and, to prevent its ever happening again, took care to inform him
at first that it was a favourite haunt of hers.  How it could occur
a second time, therefore, was very odd!  Yet it did, and even a
third.  It seemed like wilful ill-nature, or a voluntary \textbf{penance}\index{penance}\stepcounter{footnote}\footnotetext{
\textbf{penance} \emph{n.}  (PEH-nuhnse)
 remorse for past misconduct, often accompanied by self-inflicted suffering meant to atone.},
for on these occasions it was not merely a few formal inquiries
and an awkward pause and then away, but he actually thought it
necessary to turn back and walk with her.  He never said a great
deal, nor did she give herself the trouble of talking or of
listening much; but it struck her in the course of their third
\textbf{rencontre}\index{rencontre}\stepcounter{footnote}\footnotetext{
\textbf{} \emph{}  ()
} that he was asking some odd unconnected questions--about
her pleasure in being at Hunsford, her love of \textbf{solitary}\index{solitary}\stepcounter{footnote}\footnotetext{
\textbf{solitary} \emph{adj.}  (SAW-lih-the-ree)
single and isolated from others; remote or secluded.} walks, and
her opinion of Mr. and Mrs. Collins's happiness; and that in
speaking of Rosings and her not perfectly understanding the house,
he seemed to expect that whenever she came into Kent again she
would be staying \emph{there} too.  His words seemed to imply it.  Could
he have Colonel Fitzwilliam in his thoughts?  She supposed, if he
meant anything, he must mean and allusion\index{allusion} to what might arise in
that quarter.  It distressed her a little, and she was quite glad
to find herself at the gate in the pales\index{pales} opposite the Parsonage.

She was engaged one day as she walked, in perusing Jane's last
letter, and dwelling on some passages which proved that Jane
had not written in spirits, when, instead of being again surprised
by Mr. Darcy, she saw on looking up that Colonel Fitzwilliam
was meeting her.  Putting away the letter immediately and
forcing a smile, she said:

"I did not know before that you ever walked this way."

"I have been making the tour of the park," he replied, "as I
generally do every year, and intend to close it with a call at the
Parsonage.  Are you going much farther?"

"No, I should have turned in a moment."

And accordingly she did turn, and they walked towards the
Parsonage together.

"Do you certainly leave Kent on Saturday?" said she.

"Yes--if Darcy does not put it off again.  But I am at his
disposal.  He arranges the business just as he pleases."

"And if not able to please himself in the arrangement, he has
at least pleasure in the great power of choice.  I do not know
anybody who seems more to enjoy the power of doing what he
likes than Mr. Darcy."

"He likes to have his own way very well," replied Colonel
Fitzwilliam.  "But so we all do.  It is only that he has better
means of having it than many others, because he is rich, and
many others are poor.  I speak feelingly.  A younger son, you
know, must be \textbf{inured}\index{inured}\stepcounter{footnote}\footnotetext{
\textbf{enure} \emph{v.} \textbf{-ed}  (ehn-YOOR)
 to get used to something unpleasant; same as 'inure'.} to self-denial and dependence."

"In my opinion, the younger son of an earl can know very
little of either.  Now seriously, what have you ever known of
self-denial and dependence?  When have you been prevented by
want of money from going wherever you chose, or procuring
anything you had a fancy for?"

"These are home questions--and perhaps I cannot say that I
have experienced many hardships of that nature.  But in matters
of greater weight, I may suffer from want of money.  Younger
sons cannot marry where they like."

"Unless where they like women of fortune, which I think they
very often do."

"Our habits of expense make us too dependent, and there are too
many in my rank of life who can afford to marry without some
attention to money."

"Is this," thought Elizabeth, "meant for me?" and she coloured
at the idea; but, recovering herself, said in a lively tone, "And
pray, what is the usual price of an earl's younger son?  Unless
the elder brother is very sickly, I suppose you would not ask
above fifty thousand pounds."

He answered her in the same style, and the subject\index{subject} dropped.  To
interrupt a silence which might make him fancy her affected with
what had passed, she soon afterwards said:

"I imagine your cousin brought you down with him chiefly for
the sake of  having someone at his disposal.  I wonder he does
not marry, to secure a lasting convenience of that kind.  But,
perhaps, his sister does as well for the present, and, as she is
under his sole care, he may do what he likes with her."

"No," said Colonel Fitzwilliam, "that is an advantage which he
must divide with me.  I am joined with him in the guardianship
of Miss Darcy."

"Are you indeed?  And pray what sort of guardians do you
make?  Does your charge give you much trouble?  Young ladies
of her age are sometimes a little difficult to manage, and if she
has the true Darcy spirit, she may like to have her own way."

As she spoke she observed him looking at her earnestly\index{earnestly}; and
the manner in which he immediately asked her why she supposed
Miss Darcy likely to give them any uneasiness, convinced her
that she had somehow or other got pretty near the truth.  She
directly replied:

"You need not be frightened.  I never heard any harm of her; and
I dare say she is one of the most \textbf{tractable}\index{tractable}\stepcounter{footnote}\footnotetext{
\textbf{tractable} \emph{adj.}  (TRAK-tuh-bl)
Easily managed or controlled.} creatures in the world.
She is a very great favourite with some ladies of my acquaintance,
Mrs. Hurst and Miss Bingley.  I think I have heard you say that
you know them."

"I know them a little.  Their brother is a pleasant gentlemanlike
man--he is a great friend of Darcy's."

"Oh! yes," said Elizabeth drily; "Mr. Darcy is uncommonly kind
to Mr. Bingley, and takes a \textbf{prodigious}\index{prodigious}\stepcounter{footnote}\footnotetext{
\textbf{prodigious} \emph{adj.} \textbf{-ly}  (proh-DIH-juhs)
exceptional in size or extent; huge.} deal of care of him."

"Care of him!  Yes, I really believe Darcy \emph{does} take care of
him in those points where he most wants care.  From something
that he told me in our journey hither, I have reason to think
Bingley very much indebted to him.  But I ought to beg his
pardon, for I have no right to suppose that Bingley was the
person meant.  It was all conjecture\index{conjecture}."

"What is it you mean?"

"It is a circumstance which Darcy could not wish to be generally
known, because if it were to get round to the lady's family, it
would be an unpleasant thing."

"You may depend upon my not mentioning it."

"And remember that I have not much reason for supposing it
to be Bingley.  What he told me was merely this: that he
congratulated himself on having lately saved a friend from
the inconveniences of a most imprudent\index{imprudent} marriage, but without
mentioning names or any other particulars, and I only suspected
it to be Bingley from believing him the kind of young man to get
into a scrape of that sort, and from knowing them to have been
together the whole of last summer."

"Did Mr. Darcy give you reasons for this interference?"

"I understood that there were some very strong objections
against the lady."

"And what arts did he use to separate them?"

"He did not talk to me of his own arts," said Fitzwilliam, smiling.
"He only told me what I have now told you."

Elizabeth made no answer, and walked on, her heart swelling
with indignation\index{indignation}.  After watching her a little, Fitzwilliam asked
her why she was so thoughtful.

"I am thinking of what you have been telling me," said she.
"Your cousin's conduct does not suit my feelings.  Why was he
to be the judge?"

"You are rather disposed\index{disposed} to call his interference \textbf{officious}\index{officious}\stepcounter{footnote}\footnotetext{
\textbf{officious} \emph{adj.} \textbf{-ness}  (uh-FIH-shus)
a strong inclination to advance oneself or one's ideas when unwanted.}?"

"I do not see what right Mr. Darcy had to decide on the
propriety\index{propriety} of his friend's inclination\index{inclination}, or why, upon his own
judgement alone, he was to determine and direct in what manner
his friend was to be happy.  But," she continued, recollecting
herself, "as we know none of the particulars, it is not fair to
condemn him.  It is not to be supposed that there was much
affection in the case."

"That is not an unnatural surmise\index{surmise}," said Fitzwilliam, "but it is a
lessening of the honour of my cousin's triumph\index{triumph} very sadly."

This was spoken jestingly; but it appeared to her so just a picture
of Mr. Darcy, that she would not trust herself with an answer,
and therefore, abruptly changing the conversation talked on
indifferent\index{indifferent} matters until they reached the Parsonage.  There, shut
into her own room, as soon as their visitor left them, she could
think without interruption of all that she had heard.  It was not
to be supposed that any other people could be meant than those
with whom she was connected.  There could not exist in the
world \emph{two} men over whom Mr. Darcy could have such boundless
influence.  That he had been concerned in the measures taken to
separate Bingley and Jane she had never doubted; but she had
always attributed to Miss Bingley the principal\index{principal} design and
arrangement of them.  If his own vanity\index{vanity}, however, did not mislead
him, \emph{he} was the cause, his pride and caprice\index{caprice} were the cause, of
all that Jane had suffered, and still continued to suffer.  He
had ruined for a while every hope of happiness for the most
affectionate, generous heart in the world; and no one could say
how lasting an evil he might have inflicted.

"There were some very strong objections against the lady,"
were Colonel Fitzwilliam's words; and those strong objections
probably were, her having one uncle who was a country attorney,
and another who was in business in London.

"To Jane herself," she exclaimed, "there could be no possibility
of objection; all loveliness and goodness as she is!--her
understanding excellent, her mind improved, and her manners
captivating.  Neither could anything be urged against my father,
who, though with some peculiarities, has abilities Mr. Darcy
himself need not disdain, and respectability which he will
probably never each."  When she thought of her mother, her
confidence\index{confidence} gave way a little; but she would not allow that any
objections \emph{there} had material weight with Mr. Darcy, whose
pride, she was convinced, would receive a deeper wound from
the want of importance in his friend's connections, than from
their want of sense; and she was quite decided, at last, that he
had been partly governed by this worst kind of pride, and partly
by the wish of retaining Mr. Bingley for his sister.

The agitation and tears which the subject\index{subject} occasioned, brought on
a headache; and it grew so much worse towards the evening,
that, added to her unwillingness to see Mr. Darcy, it determined
her not to attend her cousins to Rosings, where they were
engaged to drink tea.  Mrs. Collins, seeing that she was really
unwell, did not press her to go and as much as possible
prevented her husband from pressing her; but Mr. Collins could
not conceal his apprehension\index{apprehension} of Lady Catherine's being rather
displeased by her staying at home.



\chapter{34}


When they were gone, Elizabeth, as if intending to \textbf{exasperate}\index{exasperate}\stepcounter{footnote}\footnotetext{
\textbf{exasperate} \emph{tr.v.}  (ehx-ZASS-per-ayt)
to make worse; to make very impatient or annoyed.}
herself as much as possible against Mr. Darcy, chose for her
employment the examination of all the letters which Jane had
written to her since her being in Kent.  They contained no actual
complaint, nor was there any revival of past occurrences, or any
communication of present suffering.  But in all, and in almost
every line of each, there was a want of that cheerfulness which
had been used to characterise her style, and which, proceeding
from the \textbf{serenity}\index{serenity}\stepcounter{footnote}\footnotetext{
\textbf{serenity} \emph{n.}  (suh-REH-nih-tee)
the absence of stress or anxiety; peaceful calm; tranquility.} of a mind at ease with itself and kindly
disposed\index{disposed} towards everyone, had been scarcely ever clouded.
Elizabeth noticed every sentence conveying the idea of uneasiness,
with an attention which it had hardly received on the first
\textbf{perusal}\index{perusal}\stepcounter{footnote}\footnotetext{
\textbf{peruse} \emph{tr.v.}  (per-ROOZ)
 to read or examine carefully, as if to remember.}.  Mr. Darcy's shameful boast of what misery he had been
able to inflict, gave her a \textbf{keener}\index{keener}\stepcounter{footnote}\footnotetext{
\textbf{keen} \emph{n.} \textbf{-er}  (KEEN)
} sense of her sister's
sufferings.  It was some consolation\index{consolation} to think that his visit
to Rosings was to end on the day after the next--and, a still
greater, that in less than a fortnight\index{fortnight} she should herself be with
Jane again, and enabled to contribute to the recovery of her
spirits, by all that affection could do.

She could not think of Darcy's leaving Kent without remembering
that his cousin was to go with him; but Colonel Fitzwilliam had
made it clear that he had no intentions at all, and agreeable
as he was, she did not mean to be unhappy about him.

While settling this point, she was suddenly roused by the sound
of the door-bell, and her spirits were a little fluttered by the
idea of its being Colonel Fitzwilliam himself, who had once
before called late in the evening, and might now come to inquire
particularly after her.  But this idea was soon banished, and
her spirits were very differently affected, when, to her utter
amazement, she saw Mr. Darcy walk into the room.  In an
hurried manner he immediately began an inquiry after her health,
imputing his visit to a wish of hearing that she were better.
She answered him with cold civility\index{civility}.  He sat down for a few
moments, and then getting up, walked about the room.  Elizabeth
was surprised, but said not a word.  After a silence of
several minutes\index{minutes}, he came towards her in an agitated manner,
and thus began:

"In vain I have struggled.  It will not do.  My feelings will not
be \textbf{repressed}\index{repressed}\stepcounter{footnote}\footnotetext{
\textbf{repress} \emph{v.} \textbf{-ed}  (rih-PREHSS)
to put down by force or intimidation.}.  You must allow me to tell you how \textbf{ardently}\index{ardently}\stepcounter{footnote}\footnotetext{
\textbf{ardent} \emph{adj.} \textbf{-ly}  (AR-dehnt)
 with strong emotion and intense enthusiasm.} I admire
and love you."

Elizabeth's astonishment was beyond expression.  She stared,
coloured, doubted, and was silent.  This he considered sufficient
encouragement; and the avowal\index{avowal} of all that he felt, and had long
felt for her, immediately followed.  He spoke well; but there
were feelings besides those of the heart to be detailed; and he
was not more eloquent\index{eloquent} on the subject\index{subject} of tenderness than of pride.
His sense of her inferiority--of its being a \textbf{degradation}\index{degradation}\stepcounter{footnote}\footnotetext{
\textbf{degradation} \emph{n.}  (deh-greh-DAY-shun)
changing to a lower, or less respected, state or status.}--of the
family obstacles which had always opposed to inclination\index{inclination}, were
dwelt on with a warmth which seemed due to the consequence\index{consequence} he
was wounding, but was very unlikely to recommend his suit.

In spite of her deeply-rooted dislike, she could not be insensible
to the compliment of such a man's affection, and though her
intentions did not vary for an instant, she was at first sorry for
the pain he was to receive; till, roused to resentment\index{resentment} by his
subsequent language, she lost all compassion in anger.  She
tried, however, to compose herself to answer him with patience,
when he should have done.  He concluded with representing to
her the strength of that attachment which, in spite of all his
endeavours\index{endeavours}, he had found impossible to conquer; and with
expressing his hope that it would now be rewarded by her
acceptance of his hand.  As he said this, she could easily
see that he had no doubt of a favourable answer.  He \emph{spoke} of
apprehension\index{apprehension} and anxiety, but his countenance\index{countenance} expressed real
security.  Such a circumstance could only exasperate\index{exasperate} farther,
and, when he ceased, the colour rose into her cheeks, and she
said:

"In such cases as this, it is, I believe, the established mode\index{mode}
to express a sense of obligation for the sentiments\index{sentiments} \textbf{avowed}\index{avowed}\stepcounter{footnote}\footnotetext{
\textbf{avow} \emph{tr.v.} \textbf{-ed}  (uh-VOW)
 to boldly and openly declare.},
however unequally they may be returned.  It is natural that
obligation should be felt, and if I could \emph{feel} gratitude, I would
now thank you.  But I cannot--I have never desired your good
opinion, and you have certainly bestowed\index{bestowed} it most unwillingly.  I
am sorry to have occasioned pain to anyone.  It has been most
unconsciously\index{unconsciously} done, however, and I hope will be of short
duration.  The feelings which, you tell me, have long prevented
the acknowledgment of your regard, can have little difficulty in
overcoming it after this explanation."

Mr. Darcy, who was leaning against the mantelpiece with his
eyes fixed on her face, seemed to catch her words with no less
resentment\index{resentment} than surprise.  His complexion became pale\index{pale} with
anger, and the disturbance of his mind was visible in every
feature.  He was struggling for the appearance of composure\index{composure},
and would not open his lips till he believed himself to have
attained it.  The pause was to Elizabeth's feelings dreadful.
At length, with a voice of forced calmness, he said:

"And this is all the reply which I am to have the honour of
expecting!  I might, perhaps, wish to be informed why, with so
little \emph{endeavour\index{endeavour}} at civility\index{civility}, I am thus rejected.  But it is of
small importance."

"I might as well inquire," replied she, "why with so evident a
desire of offending and insulting me, you chose to tell me that
you liked me against your will, against your reason, and even
against your character?  Was not this some excuse for incivility\index{incivility},
if I \emph{was} uncivil\index{uncivil}?  But I have other \textbf{provocations}\index{provocations}\stepcounter{footnote}\footnotetext{
\textbf{provocation} \emph{n.}  (praw-voh-KAY-shun)
unfriendly behavior that aggravates or irritates someone; an incitement.}.  You know I
have.  Had not my feelings decided against you--had they been
indifferent\index{indifferent}, or had they even been favourable, do you think that
any consideration would tempt me to accept the man who has
been the means of ruining, perhaps for ever, the happiness of a
most beloved sister?"

As she pronounced these words, Mr. Darcy changed colour; but
the emotion was short, and he listened without attempting
to interrupt her while she continued:

"I have every reason in the world to think ill of you.  No motive
can excuse the unjust and ungenerous part you acted \emph{there}.
You dare not, you cannot deny, that you have been the principal\index{principal},
if not the only means of dividing them from each other--of
exposing one to the censure\index{censure} of the world for caprice\index{caprice} and
instability, and the other to its derision\index{derision} for disappointed hopes,
and involving them both in misery of the \textbf{acutest}\index{acutest}\stepcounter{footnote}\footnotetext{
\textbf{} \emph{}  ()
} kind."

She paused, and saw with no slight indignation\index{indignation} that he was
listening with an air which proved him wholly\index{wholly} unmoved by any
feeling of remorse.  He even looked at her with a smile of
affected \textbf{incredulity}\index{incredulity}\stepcounter{footnote}\footnotetext{
\textbf{incredulity} \emph{n.}  (ihn-kreh-DYOO-lih-tee)
doubt about the truth of something; disbelief.}.

"Can you deny that you have done it?" she repeated.

With assumed tranquillity\index{tranquillity} he then replied: "I have no wish of
denying that I did everything in my power to separate my friend
from your sister, or that I rejoice in my success.  Towards \emph{him}
I have been kinder than towards myself."

Elizabeth disdained the appearance of noticing this civil
reflection, but its meaning did not escape, nor was it likely to
conciliate\index{conciliate} her.

"But it is not merely this affair," she continued, "on which my
dislike is founded.  Long before it had taken place my opinion
of you was decided.  Your character was unfolded in the \textbf{recital}\index{recital}\stepcounter{footnote}\footnotetext{
\textbf{recital} \emph{n.}  (rih-SYT-uhl)
 a performance of music or dance especially by soloists.}
which I received many months ago from Mr. Wickham.  On this
subject\index{subject}, what can you have to say?  In what imaginary act
of friendship can you here defend yourself?  or under what
misrepresentation can you here impose\index{impose} upon others?"

"You take an eager interest in that gentleman's concerns," said
Darcy, in a less tranquil\index{tranquil} tone, and with a heightened colour.

"Who that knows what his misfortunes have been, can help
feeling an interest in him?"

"His misfortunes!" repeated Darcy contemptuously; "yes, his
misfortunes have been great indeed."

"And of your \textbf{infliction}\index{infliction}\stepcounter{footnote}\footnotetext{
\textbf{infliction} \emph{n.}  (ihn-FLIHK-shun)
the act of imposing something unpleasant.}," cried Elizabeth with energy.  "You
have reduced him to his present state of poverty--comparative\index{comparative}
poverty.  You have withheld the advantages which you must
know to have been designed for him.  You have deprived the
best years of his life of that independence which was no less his
due than his desert.  You have done all this!  and yet you can
treat the mention of his misfortune with contempt\index{contempt} and ridicule."

"And this," cried Darcy, as he walked with quick steps across
the room, "is your opinion of me!  This is the estimation in
which you hold me!  I thank you for explaining it so fully.  My
faults, according to this calculation, are heavy indeed!  But
perhaps," added he, stopping in his walk, and turning towards
her, "these offenses might have been overlooked, had not your
pride been hurt by my honest confession of the scruples\index{scruples} that had
long prevented my forming any serious design.  These bitter
accusations might have been suppressed\index{suppressed}, had I, with greater
policy, concealed my struggles, and flattered you into the belief
of my being \textbf{impelled}\index{impelled}\stepcounter{footnote}\footnotetext{
\textbf{impel} \emph{tr.v.} \textbf{-ed}  (ihm-PEHL)
 to urge or force to action through moral pressure.} by unqualified, \textbf{unalloyed}\index{unalloyed}\stepcounter{footnote}\footnotetext{
\textbf{unalloyed} \emph{adj.}  (uhn-al-LOYD)
Not mixed with other metals.} inclination\index{inclination}; by
reason, by reflection, by everything.  But disguise of every sort
is my abhorrence\index{abhorrence}.  Nor am I ashamed of the feelings I related.
They were natural and just.  Could you expect me to rejoice in
the inferiority of your connections?--to congratulate myself on
the hope of relations, whose condition in life is so decidedly
beneath my own?"

Elizabeth felt herself growing more angry every moment; yet she
tried to the utmost to speak with composure\index{composure} when she said:

"You are mistaken, Mr. Darcy, if you suppose that the mode\index{mode} of
your declaration affected me in any other way, than as it spared\index{spared}
the concern which I might have felt in refusing you, had you
behaved in a more gentlemanlike manner."

She saw him start at this, but he said nothing, and she continued:

"You could not have made the offer of your hand in any possible
way that would have tempted me to accept it."

Again his astonishment was obvious; and he looked at her with
an expression of mingled incredulity\index{incredulity} and mortification\index{mortification}.  She went
on:

"From the very beginning--from the first moment, I may almost
say--of my acquaintance with you, your manners, impressing me
with the fullest belief of your arrogance, your conceit\index{conceit}, and your
selfish disdain of the feelings of others, were such as to form the
groundwork of disapprobation\index{disapprobation} on which succeeding events have
built so immovable a dislike; and I had not known you a month
before I felt that you were the last man in the world whom I
could ever be prevailed\index{prevailed} on to marry."

"You have said quite enough, madam.  I perfectly comprehend\index{comprehend}
your feelings, and have now only to be ashamed of what my own
have been.  Forgive me for having taken up so much of your
time, and accept my best wishes for your health and happiness."

And with these words he hastily left the room, and Elizabeth
heard him the next moment open the front door and quit the
house.

The tumult\index{tumult} of her mind, was now painfully great.  She knew not
how to support herself, and from actual weakness sat down and
cried for half-an-hour.  Her astonishment, as she reflected on
what had passed, was increased by every review of it.  That she
should receive an offer of marriage from Mr. Darcy!  That he
should have been in love with her for so many months!  So much
in love as to wish to marry her in spite of all the objections
which had made him prevent his friend's marrying her sister,
and which must appear at least with equal force in his own
case--was almost incredible!  It was gratifying to have inspired
unconsciously\index{unconsciously} so strong an affection.  But his pride, his
abominable\index{abominable} pride--his shameless avowal\index{avowal} of what he had done with
respect to Jane--his unpardonable assurance in acknowledging,
though he could not justify it, and the unfeeling manner in
which he had mentioned Mr. Wickham, his cruelty towards whom
he had not attempted to deny, soon overcame the pity which the
consideration of his attachment had for a moment excited.  She
continued in very agitated reflections till the sound of Lady
Catherine's carriage made her feel how unequal she was to
encounter Charlotte's observation, and hurried her away to
her room.



\chapter{35}


Elizabeth awoke the next morning to the same thoughts and
meditations which had at length closed her eyes.  She could
not yet recover from the surprise of what had happened; it was
impossible to think of anything else; and, totally indisposed\index{indisposed}
for employment, she resolved, soon after breakfast, to indulge
herself in air and exercise.  She was proceeding directly to her
favourite walk, when the recollection\index{recollection} of Mr. Darcy's sometimes
coming there stopped her, and instead of entering the park, she
turned up the lane, which led farther from the turnpike-road.
The park paling\index{paling} was still the boundary on one side, and she
soon passed one of the gates into the ground.

After walking two or three times along that part of the lane, she
was tempted, by the pleasantness of the morning, to stop at the
gates and look into the park.  The five weeks which she had now
passed in Kent had made a great difference in the country, and
every day was adding to the \textbf{verdure}\index{verdure}\stepcounter{footnote}\footnotetext{
\textbf{verdure} \emph{n.}  (VER-djer)
 lush green plant growth and foliage.} of the early trees.  She was
on the point of continuing her walk, when she caught a glimpse
of a gentleman within the sort of grove which edged the park;
he was moving that way; and, fearful of its being Mr. Darcy,
she was directly retreating.  But the person who advanced was
now near enough to see her, and stepping forward\index{forward} with eagerness,
pronounced her name.  She had turned away; but on hearing
herself called, though in a voice which proved it to be Mr.
Darcy, she moved again towards the gate.  He had by that time
reached it also, and, holding out a letter, which she instinctively
took, said, with a look of haughty\index{haughty} composure\index{composure}, "I have been
walking in the grove some time in the hope of meeting you.  Will
you do me the honour of reading that letter?"  And then, with a
slight bow, turned again into the \textbf{plantation}\index{plantation}\stepcounter{footnote}\footnotetext{
\textbf{plantation} \emph{n.}  (plan-TAY-shun)
an estate where cash crops are grown on a large scale.}, and was soon out of
sight.

With no expectation of pleasure, but with the strongest curiosity,
Elizabeth opened the letter, and, to her still increasing wonder,
perceived\index{perceived} an envelope containing two sheets of letter-paper,
written quite through, in a very close hand.  The envelope itself
was likewise full.  Pursuing her way along the lane, she then
began it.  It was dated from Rosings, at eight o'clock in the
morning, and was as follows:--

"Be not alarmed, madam, on receiving this letter, by the
apprehension\index{apprehension} of its containing any repetition of those sentiments\index{sentiments}
or renewal of those offers which were last night so disgusting to
you.  I write without any intention of paining you, or humbling
myself, by dwelling on wishes which, for the happiness of both,
cannot be too soon forgotten; and the effort which the formation
and the perusal of this letter must occasion, should have been
spared\index{spared}, had not my character required it to be written and read.
You must, therefore, pardon the freedom with which I demand
your attention; your feelings, I know, will bestow\index{bestow} it unwillingly,
but I demand it of your justice.

"Two offenses of a very different nature, and by no means of
equal magnitude, you last night laid to my charge.  The first
mentioned was, that, regardless of the sentiments\index{sentiments} of either, I had
detached Mr. Bingley from your sister, and the other, that I had,
in \textbf{defiance}\index{defiance}\stepcounter{footnote}\footnotetext{
\textbf{defiance} \emph{n.}  (dih-FY-ans)
brazen resistance to authority.} of various claims, in defiance\index{defiance} of honour and
humanity, ruined the immediate prosperity and blasted the
prospects\index{prospects} of Mr. Wickham.  Wilfully and \textbf{wantonly}\index{wantonly}\stepcounter{footnote}\footnotetext{
\textbf{wanton} \emph{adj.} \textbf{-ly}  (WAHN-tuhn)
 in a lewd, immoral, or promiscuous manner.} to have
thrown off the companion of my youth, the acknowledged
favourite of my father, a young man who had scarcely any other
dependence than on our patronage\index{patronage}, and who had been brought
up to expect its exertion\index{exertion}, would be a \textbf{depravity}\index{depravity}\stepcounter{footnote}\footnotetext{
\textbf{depravity} \emph{n.}  (dih-PRA-vih-tee)
moral perversion; a degenerate act or practice.}, to which the
separation of two young persons, whose affection could be the
growth of only a few weeks, could bear no comparison.  But
from the severity of that blame which was last night so liberally
bestowed\index{bestowed}, respecting each circumstance, I shall hope to be in the
future secured, when the following account of my actions and
their motives has been read.  If, in the explanation of them,
which is due to myself, I am under the necessity of relating
feelings which may be offensive to yours, I can only say that I
am sorry.  The necessity must be obeyed, and further apology
would be absurd.

"I had not been long in Hertfordshire, before I saw, in common
with others, that Bingley preferred your elder sister to any other
young woman in the country.  But it was not till the evening of
the dance at Netherfield that I had any apprehension\index{apprehension} of his
feeling a serious attachment.  I had often seen him in love before.
At that ball, while I had the honour of dancing with you, I was
first made acquainted, by Sir William Lucas's accidental
information, that Bingley's attentions to your sister had given
rise to a general expectation of their marriage.  He spoke of it
as a certain event, of which the time alone could be undecided.
From that moment I observed my friend's behaviour attentively;
and I could then perceive\index{perceive} that his partiality\index{partiality} for Miss Bennet
was beyond what I had ever witnessed in him.  Your sister I
also watched.  Her look and manners were open, cheerful, and
engaging as ever, but without any symptom of peculiar\index{peculiar} regard,
and I remained convinced from the evening's \textbf{scrutiny}\index{scrutiny}\stepcounter{footnote}\footnotetext{
\textbf{scrutiny} \emph{n.}  (SKROO-tih-nee)
A careful examination (as for mistakes).}, that
though she received his attentions with pleasure, she did not
invite them by any participation of sentiment\index{sentiment}.  If \emph{you} have not
been mistaken here, \emph{I} must have been in error.  Your superior\index{superior}
knowledge of your sister must make the latter probable\index{probable}.  If it be
so, if I have been misled by such error to inflict pain on her,
your resentment\index{resentment} has not been unreasonable.  But I shall not scruple\index{scruple}
to assert, that the serenity\index{serenity} of your sister's countenance\index{countenance} and air
was such as might have given the most acute\index{acute} observer a conviction\index{conviction}
that, however amiable\index{amiable} her temper, her heart was not likely to be
easily touched.  That I was desirous of believing her indifferent\index{indifferent}
is certain--but I will venture to say that my investigation and
decisions are not usually influenced by my hopes or fears.  I did
not believe her to be indifferent\index{indifferent} because I wished it; I believed
it on impartial\index{impartial} conviction\index{conviction}, as truly as I wished it in reason.
My objections to the marriage were not merely those which I last
night acknowledged to have the utmost force of passion to put
aside, in my own case; the want of connection could not be so
great an evil to my friend as to me.  But there were other causes
of \textbf{repugnance}\index{repugnance}\stepcounter{footnote}\footnotetext{
\textbf{repugnance} \emph{n.}  (rih-PUHG-nuhns)
intense aversion or dislike.}; causes which, though still existing, and existing
to an equal degree in both instances, I had myself endeavoured\index{endeavoured}
to forget, because they were not immediately before me.  These
causes must be stated, though briefly.  The situation of
your mother's family, though objectionable, was nothing in
comparison to that total want of propriety\index{propriety} so frequently, so
almost uniformly betrayed by herself, by your three younger
sisters, and occasionally even by your father.  Pardon me.  It
pains me to offend you.  But amidst your concern for the defects
of  your nearest relations, and your displeasure at this
representation of them, let\index{let} it give you consolation\index{consolation} to consider
that, to have conducted yourselves so as to avoid any share of
the like censure\index{censure}, is praise no less generally bestowed\index{bestowed} on you and
your elder sister, than it is honourable to the sense and
disposition\index{disposition} of both.  I will only say farther that from what passed
that evening, my opinion of all parties was confirmed, and every
inducement\index{inducement} heightened which could have led me before, to
preserve my friend from what I esteemed a most unhappy
connection.  He left Netherfield for London, on the day
following, as you, I am certain, remember, with the design of
soon returning.

"The part which I acted is now to be explained.  His sisters'
uneasiness had been equally excited with my own; our coincidence
of feeling was soon discovered, and, alike sensible that no time
was to be lost in detaching their brother, we shortly resolved
on joining him directly in London.  We accordingly went--and
there I readily\index{readily} engaged in the office of pointing out to my
friend the certain evils of such a choice.  I described, and
enforced them earnestly\index{earnestly}.  But, however this \textbf{remonstrance}\index{remonstrance}\stepcounter{footnote}\footnotetext{
\textbf{remonstrance} \emph{n.}  (rih-MAHN-struhns)
an expression of opposition to a course of action.} might
have staggered or delayed his determination, I do not suppose
that it would ultimately have prevented the marriage, had it not
been seconded by the assurance that I hesitated not in giving, of
your sister's indifference\index{indifference}.  He had before believed her to return
his affection with sincere, if not with equal regard.  But Bingley
has great natural modesty, with a stronger dependence on my
judgement than on his own.  To convince him, therefore, that he
had deceived himself, was no very difficult point.  To persuade
him against returning into Hertfordshire, when that conviction\index{conviction}
had been given, was scarcely the work of a moment.  I cannot
blame myself for having done thus much.  There is but one part
of my conduct in the whole affair on which I do not reflect with
satisfaction; it is that I condescended\index{condescended} to adopt the measures of
art so far as to conceal from him your sister's being in town.  I
knew it myself, as it was known to Miss Bingley; but her brother
is even yet ignorant of it.  That they might have met without ill
consequence\index{consequence} is perhaps probable\index{probable}; but his regard did not appear
to me enough extinguished for him to see her without some danger.
Perhaps this concealment, this disguise was beneath me; it is
done, however, and it was done for the best.  On this subject\index{subject}
I have nothing more to say, no other apology to offer.  If I
have wounded your sister's feelings, it was unknowingly done and
though the motives which governed me may to you very naturally
appear insufficient, I have not yet learnt to condemn them.

"With respect to that other, more weighty accusation, of having
injured Mr. Wickham, I can only \textbf{refute}\index{refute}\stepcounter{footnote}\footnotetext{
\textbf{refute} \emph{tr.v.}  (rih-FYOOT)
to prove to be false or incorrect.} it by laying before you
the whole of his connection with my family.  Of what he has
\emph{particularly} accused me I am ignorant; but of the truth of
what I shall relate, I can summon\index{summon} more than one witness of
undoubted veracity\index{veracity}.

"Mr. Wickham is the son of a very respectable man, who had for
many years the management of all the Pemberley estates, and
whose good conduct in the discharge of his trust naturally
inclined my father to be of service to him; and on George
Wickham, who was his godson\index{godson}, his kindness was therefore
liberally bestowed\index{bestowed}.  My father supported him at school, and
afterwards at Cambridge--most important assistance, as his own
father, always poor from the extravagance of his wife, would
have been unable to give him a gentleman's education.  My
father was not only fond of this young man's society, whose
manner were always engaging; he had also the highest opinion of
him, and hoping the church would be his profession, intended to
provide for him in it.  As for myself, it is many, many years since
I first began to think of him in a very different manner.  The
vicious \textbf{propensities}\index{propensities}\stepcounter{footnote}\footnotetext{
\textbf{propensity} \emph{n.}  (proh-PEHN-sih-tee)
an inclination or tendency.}--the want of principle, which he was careful
to guard from the knowledge of his best friend, could not escape
the observation of a young man of nearly the same age with
himself, and who had opportunities of seeing him in unguarded
moments, which Mr. Darcy could not have.  Here again shall
give you pain--to what degree you only can tell.  But whatever
may be the sentiments\index{sentiments} which Mr. Wickham has created, a
suspicion of their nature shall not prevent me from unfolding
his real character--it adds even another motive.

"My excellent father died about five years ago; and his attachment
to Mr. Wickham was to the last so steady\index{steady}, that in his will he
particularly recommended it to me, to promote his advancement in
the best manner that his profession might allow--and if he took
orders, desired that a valuable family living might be his as soon
as it became vacant.  There was also a \textbf{legacy}\index{legacy}\stepcounter{footnote}\footnotetext{
\textbf{legacy} \emph{n.}  (LEH-guh-see)
a gift of personal property by will or inheritance.} of one thousand
pounds.  His own father did not long survive mine, and within half
a year from these events, Mr. Wickham wrote to inform me that,
having finally resolved against taking orders, he hoped I should
not think it unreasonable for him to expect some more immediate
\textbf{pecuniary}\index{pecuniary}\stepcounter{footnote}\footnotetext{
\textbf{pecuniary} \emph{adj.}  (puh-KYOO-nuh-ree)
relating to or involving money.} advantage, in \textbf{lieu}\index{lieu}\stepcounter{footnote}\footnotetext{
\textbf{lieu} \emph{n.}  (LYOO)
 in place of a position or function normally served by another.} of the preferment, by which he could
not be benefited.  He had some intention, he added, of studying
law, and I must be aware that the interest of one thousand pounds
would be a very insufficient support therein.  I rather wished,
than believed him to be sincere; but, at any rate, was perfectly
ready to accede\index{accede} to his proposal.  I knew that Mr. Wickham
ought not to be a clergyman; the business was therefore soon
settled--he resigned all claim to assistance in the church, were
it possible that he could ever be in a situation to receive it,
and accepted in return three thousand pounds.  All connection
between us seemed now dissolved.  I thought too ill of him to
invite him to Pemberley, or admit his society in town.  In town
I believe he chiefly lived, but his studying the law was a mere
\textbf{pretence}\index{pretence}\stepcounter{footnote}\footnotetext{
\textbf{pretence} \emph{n.}  (see)
a false or unsupportable quality; same as pretense.}, and being now free from all restraint, his life was a
life of idleness and \textbf{dissipation}\index{dissipation}\stepcounter{footnote}\footnotetext{
\textbf{dissipation} \emph{n.}  (dihs-sih-PAY-shun)
}.  For about three years I heard
little of him; but on the decease of the incumbent\index{incumbent} of the living
which had been designed for him, he applied to me again by letter
for the presentation.  His circumstances, he assured me, and I
had no difficulty in believing it, were exceedingly bad.  He had
found the law a most unprofitable study, and was now absolutely
resolved on being ordained, if I would present him to the living
in question--of which he trusted there could be little doubt, as
he was well assured that I had no other person to provide for,
and I could not have forgotten my revered father's intentions.
You will hardly blame me for refusing to comply\index{comply} with this entreaty\index{entreaty},
or for resisting every repetition to it.  His resentment\index{resentment} was in
proportion to the distress of his circumstances--and he was
doubtless as violent in his abuse of me to others as in his
reproaches to myself.  After this period every appearance of
acquaintance was dropped.  How he lived I know not.  But last
summer he was again most painfully \textbf{obtruded}\index{obtruded}\stepcounter{footnote}\footnotetext{
\textbf{obtrude} \emph{v.} \textbf{-ed}  (ahb-TROOD)
to forcibly impose (oneself or one's opinion) on others.} on my notice.

"I must now mention a circumstance which I would wish to
forget myself, and which no obligation less than the present
should induce\index{induce} me to unfold to any human being.  Having said
thus much, I feel no doubt of your secrecy.  My sister, who is
more than ten years my junior, was left to the guardianship of
my mother's nephew, Colonel Fitzwilliam, and myself.  About a
year ago, she was taken from school, and an establishment
formed for her in London; and last summer she went with the
lady who presided over it, to Ramsgate; and thither also went
Mr. Wickham, undoubtedly by design; for there proved to have
been a prior acquaintance between him and Mrs. Younge, in
whose character we were most unhappily deceived; and by
her \textbf{connivance}\index{connivance}\stepcounter{footnote}\footnotetext{
\textbf{connivance} \emph{n.}  (kuhn-NYV-vuhns)
agreement on a secret plot.} and aid, he so far recommended himself to
Georgiana, whose affectionate heart retained a strong impression
of his kindness to her as a child, that she was persuaded to
believe herself in love, and to consent to an \textbf{elopement}\index{elopement}\stepcounter{footnote}\footnotetext{
\textbf{elope} \emph{intr.v.} \textbf{-ment}  (ee-LOAP)
 the act of running away with a lover (usually to get married).}.  She was
then but fifteen, which must be her excuse; and after stating her
imprudence\index{imprudence}, I am happy to add, that I owed the knowledge of it
to herself.  I joined them unexpectedly a day or two before the
intended elopement, and then Georgiana, unable to support the
idea of grieving and offending a brother whom she almost
looked up to as a father, acknowledged the whole to me.  You
may imagine what I felt and how I acted.  Regard for my sister's
credit and feelings prevented any public exposure; but I wrote
to Mr. Wickham, who left the place immediately, and Mrs. Younge
was of course removed from her charge.  Mr. Wickham's chief
object was unquestionably my sister's fortune, which is thirty
thousand pounds; but I cannot help supposing that the hope of
revenging himself on me was a strong inducement\index{inducement}.  His revenge
would have been complete indeed.

"This, madam, is a faithful narrative of every event in which
we have been concerned together; and if you do not absolutely
reject it as false, you will, I hope, acquit\index{acquit} me henceforth
of cruelty towards Mr. Wickham.  I know not in what manner,
under what form of falsehood he had imposed\index{imposed} on you; but his
success is not perhaps to be wondered at.  Ignorant as you
previously were of everything concerning either, detection
could not be in your power, and suspicion certainly not in
your inclination\index{inclination}.

"You may possibly wonder why all this was not told you last
night; but I was not then master enough of myself to know what
could or ought to be revealed.  For the truth of everything here
related, I can appeal more particularly to the testimony of
Colonel Fitzwilliam, who, from our near relationship and
constant intimacy, and, still more, as one of the executors of
my father's will, has been unavoidably acquainted with every
particular of these transactions.  If your abhorrence\index{abhorrence} of \emph{me}
should make \emph{my} assertions\index{assertions} valueless, you cannot be prevented
by the same cause from confiding in my cousin; and that there
may be the possibility of consulting him, I shall endeavour\index{endeavour} to
find some opportunity of putting this letter in your hands in
the course of the morning.  I will only add, God bless you.

"\emph{fitzwilliam} \emph{darcy}"



\chapter{36}


If Elizabeth, when Mr. Darcy gave her the letter, did not expect
it to contain a renewal of his offers, she had formed no
expectation at all of its contents.  But such as they were, it
may well be supposed how eagerly she went through them, and what
a \textbf{contrariety}\index{contrariety}\stepcounter{footnote}\footnotetext{
\textbf{contrariety} \emph{n.}  (kuhn-trair-RY-eh-tee)
} of emotion they excited.  Her feelings as she
read were scarcely to be defined.  With amazement did she first
understand that he believed any apology to be in his power; and
steadfastly\index{steadfastly} was she persuaded, that he could have no explanation
to give, which a just sense of shame would not conceal.  With a
strong prejudice\index{prejudice} against everything he might say, she began his
account of what had happened at Netherfield.  She read with an
eagerness which hardly left her power of comprehension, and
from impatience of knowing what the next sentence might bring,
was incapable of attending to the sense of the one before her
eyes.  His belief of her sister's insensibility she instantly
resolved to be false; and his account of the real, the worst
objections to the match, made her too angry to have any wish of
doing him justice.  He expressed no regret for what he had done
which satisfied her; his style was not \textbf{penitent}\index{penitent}\stepcounter{footnote}\footnotetext{
\textbf{penitent} \emph{adj.}  (PEHN-ih-tuhnt)
feeling or expressing remorse for one's misdeeds.}, but haughty\index{haughty}.
It was all pride and \textbf{insolence}\index{insolence}\stepcounter{footnote}\footnotetext{
\textbf{insolence} \emph{n.}  (IHN-soh-lense)
the trait of being rude and impertinent; inclined to take liberties.}.

But when this subject\index{subject} was succeeded by his account of Mr.
Wickham--when she read with somewhat clearer attention a
relation of events which, if true, must overthrow every cherished
opinion of his worth, and which bore so alarming an affinity\index{affinity} to
his own history of himself--her feelings were yet more acutely\index{acutely}
painful and more difficult of definition.  Astonishment,
apprehension\index{apprehension}, and even horror, oppressed her.  She wished to
discredit it entirely, repeatedly exclaiming, "This must be false!
This cannot be!  This must be the grossest falsehood!"--and
when she had gone through the whole letter, though scarcely
knowing anything of the last page or two, put it hastily away,
protesting that she would not regard it, that she would never
look in it again.

In this \textbf{perturbed}\index{perturbed}\stepcounter{footnote}\footnotetext{
\textbf{perturb} \emph{tr.v.} \textbf{-ed}  (per-TERB)
to agitate; to disturb deeply; to make uneasy.} state of mind, with thoughts that could rest on
nothing, she walked on; but it would not do; in half a minute\index{minute} the
letter was unfolded again, and collecting herself as well as she
could, she again began the mortifying\index{mortifying} perusal of all that related
to Wickham, and commanded herself so far as to examine the
meaning of every sentence.  The account of his connection with
the Pemberley family was exactly what he had related himself;
and the kindness of the late Mr. Darcy, though she had not
before known its extent, agreed equally well with his own
words.  So far each recital\index{recital} confirmed the other; but when she
came to the will, the difference was great.  What Wickham had
said of the living was fresh in her memory, and as she recalled
his very words, it was impossible not to feel that there was gross
duplicity\index{duplicity} on one side or the other; and, for a few moments, she
flattered herself that her wishes did not err.  But when she
read and re-read with the closest attention, the particulars
immediately following of Wickham's resigning all pretensions\index{pretensions} to
the living, of his receiving in lieu\index{lieu} so considerable a sum as
three thousand pounds, again was she forced to hesitate.  She
put down the letter, weighed every circumstance with what she
meant to be impartiality--deliberated on the probability of each
statement--but with little success.  On both sides it was only
assertion\index{assertion}.  Again she read on; but every line proved more clearly
that the affair, which she had believed it impossible that any
contrivance\index{contrivance} could so represent as to render\index{render} Mr. Darcy's conduct
in it less than infamous\index{infamous}, was capable of a turn which must make
him entirely blameless throughout the whole.

The extravagance and general \textbf{profligacy}\index{profligacy}\stepcounter{footnote}\footnotetext{
\textbf{profligate} \emph{adj.}  (PRAW-flih-guht)
} which he scrupled\index{scrupled} not
to lay at Mr. Wickham's charge, exceedingly shocked her; the
more so, as she could bring no proof of its injustice\index{injustice}.  She had
never heard of him before his entrance into the ----shire Militia\index{Militia},
in which he had engaged at the persuasion of the young man
who, on meeting him accidentally in town, had there renewed a
slight acquaintance.  Of his former way of life nothing had been
known in Hertfordshire but what he told himself.  As to his real
character, had information been in her power, she had never felt
a wish of inquiring.  His countenance\index{countenance}, voice, and manner had
established him at once in the possession of every virtue\index{virtue}.  She
tried to recollect some instance of goodness, some distinguished
trait of integrity\index{integrity} or benevolence\index{benevolence}, that might rescue him from the
attacks of Mr. Darcy; or at least, by the predominance of virtue\index{virtue},
atone\index{atone} for those casual errors under which she would endeavour\index{endeavour}
to class what Mr. Darcy had described as the idleness and vice\index{vice}
of many years' continuance.  But no such recollection\index{recollection} befriended
her.  She could see him instantly before her, in every charm of
air and address\index{address}; but she could remember no more substantial
good than the general approbation\index{approbation} of the neighbourhood, and
the regard which his social powers had gained him in the mess.
After pausing on this point a considerable while, she once more
continued to read.  But, alas! the story which followed, of his
designs on Miss Darcy, received some confirmation from what
had passed between Colonel Fitzwilliam and herself only the
morning before; and at last she was referred for the truth of
every particular to Colonel Fitzwilliam himself--from whom she
had previously received the information of his near concern in
all his cousin's affairs, and whose character she had no reason
to question.  At one time she had almost resolved on applying
to him, but the idea was checked by the awkwardness of the
application\index{application}, and at length wholly\index{wholly} banished by the conviction\index{conviction}
that Mr. Darcy would never have hazarded such a proposal, if
he had not been well assured of his cousin's corroboration.

She perfectly remembered everything that had passed in
conversation between Wickham and herself, in their first evening
at Mr. Phillips's.  Many of his expressions were still fresh in
her memory.  She was \emph{now} struck with the \textbf{impropriety}\index{impropriety}\stepcounter{footnote}\footnotetext{
\textbf{impropriety} \emph{n.}  (ihm-proh-PRY-eh-tee)
} of such
communications to a stranger, and wondered it had escaped her
before.  She saw the indelicacy of putting himself forward\index{forward} as
he had done, and the inconsistency\index{inconsistency} of his professions with his
conduct.  She remembered that he had boasted of having no fear
of seeing Mr. Darcy--that Mr. Darcy might leave the country,
but that \emph{he} should stand his ground; yet he had avoided the
Netherfield ball the very next week.  She remembered also that,
till the Netherfield family had quitted the country, he had told
his story to no one but herself; but that after their removal it
had been everywhere discussed; that he had then no reserves, no
scruples\index{scruples} in sinking Mr. Darcy's character, though he had assured
her that respect for the father would always prevent his exposing
the son.

How differently did everything now appear in which he was
concerned!  His attentions to Miss King were now the consequence\index{consequence}
of views solely\index{solely} and hatefully mercenary\index{mercenary}; and the mediocrity of
her fortune proved no longer the moderation of his wishes, but
his eagerness to grasp at anything.  His behaviour to herself
could now have had no tolerable\index{tolerable} motive; he had either been
deceived with regard to her fortune, or had been gratifying his
vanity\index{vanity} by encouraging the preference which she believed she had
most incautiously shown.  Every lingering struggle in his favour
grew fainter and fainter; and in farther justification of Mr.
Darcy, she could not but allow Mr. Bingley, when questioned
by Jane, had long ago asserted his blamelessness in the affair;
that proud and repulsive as were his manners, she had never, in
the whole course of their acquaintance--an acquaintance which
had latterly brought them much together, and given her a sort of
intimacy with his ways--seen anything that betrayed him to be
unprincipled or unjust--anything that spoke him of \textbf{irreligious}\index{irreligious}\stepcounter{footnote}\footnotetext{
\textbf{irreligious} \emph{adj.}  (ihr-ree-LIH-juhs)
hostile or indifferent to religion.}
or \textbf{immoral}\index{immoral}\stepcounter{footnote}\footnotetext{
\textbf{immoral} \emph{adj.}  (ihm-MAW-ruhl)
not adhering to ethical or moral principles; impure; licentious; wicked.} habits; that among his own connections he was
esteemed and valued--that even Wickham had allowed him
merit as a brother, and that she had often heard him speak so
affectionately of his sister as to prove him capable of \emph{some}
amiable\index{amiable} feeling; that had his actions been what Mr. Wickham
represented them, so gross a violation of everything right could
hardly have been concealed from the world; and that friendship
between a person capable of it, and such an amiable\index{amiable} man as Mr.
Bingley, was incomprehensible.

She grew absolutely ashamed of herself.  Of neither Darcy nor
Wickham could she think without feeling she had been blind,
partial\index{partial}, prejudiced\index{prejudiced}, absurd.

"How despicably I have acted!" she cried; "I, who have prided
myself on my discernment\index{discernment}!  I, who have valued myself on my
abilities!  who have often disdained the generous candour of my
sister, and gratified my vanity\index{vanity} in useless or blameable mistrust!
How humiliating is this discovery!  Yet, how just a humiliation!
Had I been in love, I could not have been more wretchedly\index{wretchedly} blind!
But vanity\index{vanity}, not love, has been my folly\index{folly}.  Pleased with the
preference of one, and offended by the neglect of the other,
on the very beginning of our acquaintance, I have courted
prepossession\index{prepossession} and ignorance, and driven reason away, where
either were concerned.  Till this moment I never knew myself."

From herself to Jane--from Jane to Bingley, her thoughts were
in a line which soon brought to her recollection\index{recollection} that Mr. Darcy's
explanation \emph{there} had appeared very insufficient, and she read
it again.  Widely different was the effect of a second perusal.
How could she deny that credit to his assertions\index{assertions} in one instance,
which she had been obliged to give in the other?  He declared
himself to be totally unsuspicious of her sister's attachment;
and she could not help remembering what Charlotte's opinion
had always been.  Neither could she deny the justice of his
description of Jane.  She felt that Jane's feelings, though \textbf{fervent}\index{fervent}\stepcounter{footnote}\footnotetext{
\textbf{fervent} \emph{adj.} \textbf{-ly}  (FER-vehnt)
with passionate zeal or fervor.},
were little displayed, and that there was a constant complacency\index{complacency}
in her air and manner not often united with great sensibility.

When she came to that part of the letter in which her family were
mentioned in terms of such mortifying\index{mortifying}, yet merited reproach\index{reproach}, her
sense of shame was severe\index{severe}.  The justice of the charge struck her
too forcibly for denial, and the circumstances to which he
particularly alluded\index{alluded} as having passed at the Netherfield ball,
and as confirming all his first disapprobation\index{disapprobation}, could not have
made a stronger impression on his mind than on hers.

The compliment to herself and her sister was not unfelt.  It
soothed, but it could not console\index{console} her for the contempt\index{contempt} which had
thus been self-attracted by the rest of her family; and as she
considered that Jane's disappointment had in fact been the work
of her nearest relations, and reflected how materially the credit
of both must be hurt by such impropriety\index{impropriety} of conduct, she felt
depressed beyond anything she had ever known before.

After wandering along the lane for two hours, giving way to
every variety of thought--re-considering events, determining
probabilities, and reconciling herself, as well as she could, to
a change so sudden and so important, fatigue, and a recollection\index{recollection}
of her long absence, made her at length return home; and she
entered the house with the wish of appearing cheerful as usual,
and the resolution\index{resolution} of repressing\index{repressing} such reflections as must make
her unfit for conversation.

She was immediately told that the two gentlemen from Rosings
had each called during her absence; Mr. Darcy, only for a few
minutes\index{minutes}, to take leave--but that Colonel Fitzwilliam had been
sitting with them at least an hour, hoping for her return, and
almost resolving to walk after her till she could be found.
Elizabeth could but just \emph{affect} concern in missing him; she
really rejoiced at it.  Colonel Fitzwilliam was no longer an
object; she could think only of her letter.



\chapter{37}


The two gentlemen left Rosings the next morning, and Mr.
Collins having been in waiting near the lodges, to make them
his parting \textbf{obeisance}\index{obeisance}\stepcounter{footnote}\footnotetext{
\textbf{obeisance} \emph{n.}  (oh-BEE-sense)
a body movement, such as a bow or curtsy, intended to show respect or submission.}, was able to bring home the pleasing
intelligence, of their appearing in very good health, and in as
tolerable\index{tolerable} spirits as could be expected, after the melancholy\index{melancholy} scene
so lately gone through at Rosings.  To Rosings he then hastened,
to console\index{console} Lady Catherine and her daughter; and on his return
brought back, with great satisfaction, a message from her
ladyship, \textbf{importing}\index{importing}\stepcounter{footnote}\footnotetext{
\textbf{import} \emph{v.} \textbf{-ing}  (IHM-port)
comercially bringing in goods from a foreign country, usually for resale.} that she felt herself so dull as to make
her very desirous of having them all to dine with her.

Elizabeth could not see Lady Catherine without recollecting that,
had she chosen it, she might by this time have been presented to
her as her future niece; nor could she think, without a smile, of
what her ladyship's indignation\index{indignation} would have been.  "What would
she have said? how would she have behaved?" were questions with
which she amused herself.

Their first subject\index{subject} was the \textbf{diminution}\index{diminution}\stepcounter{footnote}\footnotetext{
\textbf{diminution} \emph{n.}  (dihm-yoo-NIH-shun)
a decrease or reduction; an act of making smaller.} of the Rosings party.
"I assure you, I feel it exceedingly," said Lady Catherine; "I
believe no one feels the loss of friends so much as I do.  But
I am particularly attached to these young men, and know them to
be so much attached to me!  They were excessively sorry to go!
But so they always are.  The dear Colonel rallied his spirits
tolerably till just at last; but Darcy seemed to feel it most
acutely\index{acutely}, more, I think, than last year.  His attachment to
Rosings certainly increases."

Mr. Collins had a compliment, and an allusion\index{allusion} to throw in here,
which were kindly smiled on by the mother and daughter.

Lady Catherine observed, after dinner, that Miss Bennet seemed
out of spirits, and immediately accounting for it by herself,
by supposing that she did not like to go home again so soon,
she added:

"But if that is the case, you must write to your mother and beg
that you may stay a little longer.  Mrs. Collins will be very glad
of your company, I am sure."

"I am much obliged to your ladyship for your kind invitation,"
replied Elizabeth, "but it is not in my power to accept it.
I must be in town next Saturday."

"Why, at that rate, you will have been here only six weeks.  I
expected you to stay two months.  I told Mrs. Collins so before
you came.  There can be no occasion for your going so soon.
Mrs. Bennet could certainly spare you for another fortnight\index{fortnight}."

"But my father cannot.  He wrote last week to hurry my return."

"Oh! your father of course may spare you, if your mother can.
Daughters are never of so much consequence\index{consequence} to a father.  And if
you will stay another \emph{month} complete, it will be in my power
to take one of you as far as London, for I am going there early
in June, for a week; and as Dawson does not object to the
barouche-box, there will be very good room for one of you--and
indeed, if the weather should happen to be cool, I should not
object to taking you both, as you are neither of you large."

"You are all kindness, madam; but I believe we must \textbf{abide}\index{abide}\stepcounter{footnote}\footnotetext{
\textbf{abide} \emph{v.} \textbf{-ing}  (uh-BIDE)
} by
our original plan."

Lady Catherine seemed resigned.  "Mrs. Collins, you must send
a servant with them.  You know I always speak my mind, and I
cannot bear the idea of two young women travelling \textbf{post}\index{post}\stepcounter{footnote}\footnotetext{
\textbf{post} \emph{n.}  (POHST)
} by
themselves.  It is highly improper.  You must contrive\index{contrive} to send
somebody.  I have the greatest dislike in the world to that sort
of thing.  Young women should always be properly guarded and
attended, according to their situation in life.  When my niece
Georgiana went to Ramsgate last summer, I made a point of her
having two men-servants go with her.  Miss Darcy, the daughter
of Mr. Darcy, of Pemberley, and Lady Anne, could not have
appeared with propriety\index{propriety} in a different manner.  I am excessively
attentive to all those things.  You must send John with the young
ladies, Mrs. Collins.  I am glad it occurred to me to mention it;
for it would really be discreditable to \emph{you} to let\index{let} them go
alone."

"My uncle is to send a servant for us."

"Oh!  Your uncle!  He keeps a man-servant, does he?  I am very
glad you have somebody who thinks of these things.  Where
shall you change horses?  Oh!  Bromley, of course.  If you
mention my name at the Bell, you will be attended to."

Lady Catherine had many other questions to ask respecting their
journey, and as she did not answer them all herself, attention was
necessary, which Elizabeth believed to be lucky for her; or, with
a mind so occupied, she might have forgotten where she was.
Reflection must be reserved for solitary\index{solitary} hours; whenever she
was alone, she gave way to it as the greatest relief; and not a
day went by without a solitary\index{solitary} walk, in which she might indulge
in all the delight of unpleasant recollections\index{recollections}.

Mr. Darcy's letter she was in a fair way of soon knowing by
heart.  She studied every sentence; and her feelings towards its
writer were at times widely different.  When she remembered the
style of his address\index{address}, she was still full of indignation\index{indignation}; but when
she considered how unjustly she had condemned and upbraided\index{upbraided} him,
her anger was turned against herself; and his disappointed
feelings became the object of compassion.  His attachment
excited gratitude, his general character respect; but she could
not approve him; nor could she for a moment repent her refusal,
or feel the slightest inclination\index{inclination} ever to see him again.  In
her own past behaviour, there was a constant source of vexation\index{vexation}
and regret; and in the unhappy defects of her family, a subject\index{subject}
of yet heavier \textbf{chagrin}\index{chagrin}\stepcounter{footnote}\footnotetext{
\textbf{chagrin} \emph{n.}  (shuh-GRIN)
strong feelings of embarrassment or hurt pride.}.  They were hopeless of remedy.  Her father,
contented with laughing at them, would never exert himself to
restrain the wild giddiness of his youngest daughters; and her
mother, with manners so far from right herself, was entirely
insensible of the evil.  Elizabeth had frequently united with Jane
in an endeavour\index{endeavour} to check the imprudence\index{imprudence} of Catherine and Lydia;
but while they were supported by their mother's indulgence, what
chance could there be of improvement?  Catherine, weak-spirited,
irritable, and completely under Lydia's guidance, had been always
affronted\index{affronted} by their advice; and Lydia, self-willed and careless,
would scarcely give them a hearing.  They were ignorant, idle, and
vain.  While there was an officer in Meryton, they would flirt
with him; and while Meryton was within a walk of Longbourn, they
would be going there forever.

Anxiety on Jane's behalf was another \textbf{prevailing}\index{prevailing}\stepcounter{footnote}\footnotetext{
\textbf{prevailing} \emph{adj.}  (prih-VAYL-ling)
most frequent or common.}\index{prevailing\index{prevailing}} concern; and
Mr. Darcy's explanation, by restoring Bingley to all her former
good opinion, heightened the sense of what Jane had lost.  His
affection was proved to have been sincere, and his conduct
cleared of all blame, unless any could attach to the implicitness
of his confidence\index{confidence} in his friend.  How grievous\index{grievous} then was the
thought that, of a situation so desirable in every respect, so
\textbf{replete}\index{replete}\stepcounter{footnote}\footnotetext{
\textbf{replete} \emph{adj.}  (rih-PLEET)
} with advantage, so promising for happiness, Jane had
been deprived, by the folly\index{folly} and \textbf{indecorum}\index{indecorum}\stepcounter{footnote}\footnotetext{
\textbf{indecorum} \emph{n.}  (ihn-duh-KOR-uhm)
an act of impropriety or undue intimacy.} of her own family!

When to these recollections\index{recollections} was added the development of Wickham's
character, it may be easily believed that the happy spirits which
had seldom been depressed before, were now so much affected as to
make it almost impossible for her to appear tolerably cheerful.

Their engagements at Rosings were as frequent during the last
week of her stay as they had been at first.  The very last evening
was spent there; and her ladyship again inquired minutely\index{minutely}\index{minutely\index{minutely}} into
the particulars of their journey, gave them directions as to the
best method of packing, and was so urgent on the necessity of
placing gowns in the only right way, that Maria thought herself
obliged, on her return, to undo all the work of the morning, and
pack her trunk afresh.

When they parted, Lady Catherine, with great condescension\index{condescension},
wished them a good journey, and invited them to come to
Hunsford again next year; and Miss de Bourgh exerted herself
so far as to curtsey and hold out her hand to both.



\chapter{38}


On Saturday morning Elizabeth and Mr. Collins met for breakfast
a few minutes\index{minutes} before the others appeared; and he took the
opportunity of paying the parting civilities which he deemed
indispensably necessary.

"I know not, Miss Elizabeth," said he, "whether Mrs. Collins has
yet expressed her sense of your kindness in coming to us; but I
am very certain you will not leave the house without receiving
her thanks for it.  The favor of your company has been much
felt, I assure you.  We know how little there is to tempt anyone
to our humble abode\index{abode}.  Our plain manner of living, our small
rooms and few domestics\index{domestics}, and the little we see of the world,
must make Hunsford extremely dull to a young lady like
yourself; but I hope you will believe us grateful for the
condescension\index{condescension}, and that we have done everything in our power
to prevent your spending your time unpleasantly."

Elizabeth was eager with her thanks and assurances of happiness.
She had spent six weeks with great enjoyment; and the pleasure
of being with Charlotte, and the kind attentions she had received,
must make \emph{her} feel the obliged.  Mr. Collins was gratified, and
with a more smiling solemnity\index{solemnity} replied:

"It gives me great pleasure to hear that you have passed your
time not disagreeably.  We have certainly done our best; and
most fortunately having it in our power to introduce you to very
superior\index{superior} society, and, from our connection with Rosings, the
frequent means of varying the humble home scene, I think we
may flatter ourselves that your Hunsford visit cannot have been
entirely irksome\index{irksome}.  Our situation with regard to Lady Catherine's
family is indeed the sort of extraordinary advantage and blessing
which few can boast.  You see on what a footing we are.  You
see how continually we are engaged there.  In truth I must
acknowledge that, with all the disadvantages of this humble
parsonage, I should not think anyone abiding in it an object of
compassion, while they are sharers of our intimacy at Rosings."

Words were insufficient for the elevation of his feelings; and
he was obliged to walk about the room, while Elizabeth tried to
unite civility\index{civility} and truth in a few short sentences.

"You may, in fact, carry a very favourable report of us into
Hertfordshire, my dear cousin.  I flatter myself at least that you
will be able to do so.  Lady Catherine's great attentions to Mrs.
Collins you have been a daily witness of; and altogether I trust
it does not appear that your friend has drawn an unfortunate--but
on this point it will be as well to be silent.  Only let\index{let} me
assure you, my dear Miss Elizabeth, that I can from my heart most
cordially\index{cordially} wish you equal felicity\index{felicity} in marriage.  My dear Charlotte
and I have but one mind and one way of thinking.  There is in
everything a most remarkable resemblance of character and ideas
between us.  We seem to have been designed for each other."

Elizabeth could safely say that it was a great happiness where
that was the case, and with equal sincerity could add, that she
firmly believed and rejoiced in his domestic\index{domestic} comforts.  She was
not sorry, however, to have the recital\index{recital} of them interrupted by
the lady from whom they sprang.  Poor Charlotte!  it was
melancholy\index{melancholy} to leave her to such society!  But she had chosen it
with her eyes open; and though evidently regretting that her
visitors were to go, she did not seem to ask for compassion.
Her home and her housekeeping, her parish\index{parish} and her poultry, and
all their dependent concerns, had not yet lost their charms.

At length the chaise\index{chaise} arrived, the trunks were fastened on, the
parcels placed within, and it was pronounced to be ready.  After
an affectionate parting between the friends, Elizabeth was
attended to the carriage by Mr. Collins, and as they walked
down the garden he was commissioning her with his best
respects to all her family, not forgetting his thanks for the
kindness he had received at Longbourn in the winter, and his
compliments to Mr. and Mrs. Gardiner, though unknown.  He
then handed her in, Maria followed, and the door was on the
point of being closed, when he suddenly reminded them, with
some \textbf{consternation}\index{consternation}\stepcounter{footnote}\footnotetext{
\textbf{consternation} \emph{n.}  (kahn-ster-NAY-shun)
a combination of fear, amazement, and dismay in face of danger.}, that they had hitherto\index{hitherto} forgotten to leave
any message for the ladies at Rosings.

"But," he added, "you will of course wish to have your humble
respects delivered to them, with your grateful thanks for their
kindness to you while you have been here."

Elizabeth made no objection; the door was then allowed to be
shut, and the carriage drove off.

"Good gracious!" cried Maria, after a few minutes\index{minutes}' silence, "it
seems but a day or two since we first came! and yet how many
things have happened!"

"A great many indeed," said her companion with a sigh.

"We have dined nine times at Rosings, besides drinking tea there
twice!  How much I shall have to tell!"

Elizabeth added privately\index{privately}, "And how much I shall have to conceal!"

Their journey was performed without much conversation, or any
alarm; and within four hours of their leaving Hunsford they
reached Mr. Gardiner's house, where they were to remain a few
days.

Jane looked well, and Elizabeth had little opportunity of
studying her spirits, amidst the various engagements which the
kindness of her aunt had reserved for them.  But Jane was to go
home with her, and at Longbourn there would be leisure enough
for observation.

It was not without an effort, meanwhile, that she could wait
even for Longbourn, before she told her sister of Mr. Darcy's
proposals.  To know that she had the power of revealing what
would so exceedingly astonish Jane, and must, at the same time,
so highly gratify whatever of her own vanity\index{vanity} she had not yet
been able to reason away, was such a temptation to openness
as nothing could have conquered but the state of indecision
in which she remained as to the extent of what she should
communicate; and her fear, if she once entered on the subject\index{subject},
of being hurried into repeating something of Bingley which
might only grieve\index{grieve} her sister further.



\chapter{39}


It was the second week in May, in which the three young ladies
set out together from Gracechurch Street for the town of ----,
in Hertfordshire; and, as they drew near the appointed inn where
Mr. Bennet's carriage was to meet them, they quickly perceived\index{perceived},
in token of the coachman's punctuality, both Kitty and Lydia
looking out of a dining-room upstairs.  These two girls had been
above an hour in the place, happily employed in visiting an
opposite milliner\index{milliner}, watching the \textbf{sentinel}\index{sentinel}\stepcounter{footnote}\footnotetext{
\textbf{sentinel} \emph{n.}  (SEHN-tih-nehl)
a guard or sentry.} on guard, and dressing a
salad and cucumber.

After welcoming their sisters, they triumphantly\index{triumphantly} displayed a table
set out with such cold meat as an inn larder usually affords,
exclaiming, "Is not this nice?  Is not this an agreeable surprise?"

"And we mean to treat you all," added Lydia, "but you must lend
us the money, for we have just spent ours at the shop out there."
Then, showing her purchases--"Look here, I have bought this bonnet.
I do not think it is very pretty; but I thought I might as well
buy it as not.  I shall pull it to pieces as soon as I get home,
and see if I can make it up any better."

And when her sisters abused it as ugly, she added, with perfect
unconcern, "Oh! but there were two or three much uglier in the
shop; and when I have bought some prettier-coloured satin to
trim it with fresh, I think it will be very tolerable\index{tolerable}.  Besides,
it will not much signify what one wears this summer, after the
----shire have left Meryton, and they are going in a fortnight\index{fortnight}."

"Are they indeed!" cried Elizabeth, with the greatest satisfaction.

"They are going to be encamped near Brighton; and I do so
want papa to take us all there for the summer!  It would be such
a delicious scheme; and I dare say would hardly cost anything at
all.  Mamma would like to go too of all things!  Only think what
a miserable summer else we shall have!"

"Yes," thought Elizabeth, "\emph{that} would be a delightful scheme
indeed, and completely do for us at once.  Good Heaven!
Brighton, and a whole campful of soldiers, to us, who have been
overset already by one poor regiment\index{regiment} of militia\index{militia}, and the monthly
balls of Meryton!"

"Now I have got some news for you," said Lydia, as they sat down
at table.  "What do you think?  It is excellent news--capital\index{capital}
news--and about a certain person we all like!"

Jane and Elizabeth looked at each other, and the waiter was told
he need not stay.  Lydia laughed, and said:

"Aye, that is just like your formality and discretion\index{discretion}.  You
thought the waiter must not hear, as if he cared!  I dare say he
often hears worse things said than I am going to say.  But he is
an ugly fellow!  I am glad he is gone.  I never saw such a long
chin in my life.  Well, but now for my news; it is about dear
Wickham; too good for the waiter, is it not?  There is no danger
of Wickham's marrying Mary King.  There's for you!  She is gone
down to her uncle at Liverpool: gone to stay.  Wickham is safe."

"And Mary King is safe!" added Elizabeth; "safe from a
connection imprudent\index{imprudent} as to fortune."

"She is a great fool for going away, if she liked him."

"But I hope there is no strong attachment on either side,"
said Jane.

"I am sure there is not on \emph{his}.  I will answer for it, he never
cared three straws about her--who could about such a nasty
little freckled thing?"

Elizabeth was shocked to think that, however incapable of
such coarseness of \emph{expression} herself, the coarseness of the
\emph{sentiment\index{sentiment}} was little other than her own breast had \textbf{harboured}\index{harboured}\stepcounter{footnote}\footnotetext{
\textbf{} \emph{}  ()
}
and fancied liberal!

As soon as all had ate, and the elder ones paid, the carriage was
ordered; and after some contrivance\index{contrivance}, the whole party, with all
their boxes, work-bags, and parcels, and the unwelcome addition
of Kitty's and Lydia's purchases, were seated in it.

"How nicely we are all crammed in," cried Lydia.  "I am glad I
bought my bonnet, if it is only for the fun of having another
bandbox!  Well, now let\index{let} us be quite comfortable and snug, and
talk and laugh all the way home.  And in the first place, let\index{let}
us hear what has happened to you all since you went away.  Have
you seen any pleasant men?  Have you had any flirting?  I was
in great hopes that one of you would have got a husband before
you came back.  Jane will be quite an old maid soon, I declare.
She is almost three-and-twenty!  Lord, how ashamed I should be of
not being married before three-and-twenty!  My aunt Phillips wants
you so to get husbands, you can't think.  She says Lizzy had
better have taken Mr. Collins; but \emph{I} do not think there would
have been any fun in it.  Lord! how I should like to be married
before any of you; and then I would \textbf{chaperon}\index{chaperon}\stepcounter{footnote}\footnotetext{
\textbf{chaperon} \emph{n.}  (sha--per-OAN)
someone who supervises a young woman or gatherings of young people.} you about to all
the balls.  Dear me! we had such a good piece of fun the other
day at Colonel Forster's.  Kitty and me were to spend the day
there, and Mrs. Forster promised to have a little dance in the
evening; (by the bye, Mrs. Forster and me are \emph{such} friends!) and
so she asked the two Harringtons to come, but Harriet was ill,
and so Pen was forced to come by herself; and then, what do you
think we did?  We dressed up Chamberlayne in woman's clothes on
purpose to pass for a lady, only think what fun!  Not a soul
knew of it, but Colonel and Mrs. Forster, and Kitty and me,
except my aunt, for we were forced to borrow one of her gowns;
and you cannot imagine how well he looked!  When Denny, and
Wickham, and Pratt, and two or three more of the men came in,
they did not know him in the least.  Lord! how I laughed! and
so did Mrs. Forster.  I thought I should have died.  And \emph{that}
made the men suspect something, and then they soon found out
what was the matter."

With such kinds of histories of their parties and good jokes, did
Lydia, assisted by Kitty's hints and additions, endeavour\index{endeavour} to
amuse her companions all the way to Longbourn.  Elizabeth
listened as little as she could, but there was no escaping the
frequent mention of Wickham's name.

Their reception at home was most kind.  Mrs. Bennet rejoiced to
see Jane in undiminished beauty; and more than once during
dinner did Mr. Bennet say voluntarily to Elizabeth:

"I am glad you are come back, Lizzy."

Their party in the dining-room was large, for almost all the
Lucases came to meet Maria and hear the news; and various
were the subjects\index{subjects} that occupied them: Lady Lucas was inquiring
of Maria, after the welfare and poultry of her eldest daughter;
Mrs. Bennet was doubly engaged, on one hand collecting an
account of the present fashions\index{fashions} from Jane, who sat some way
below her, and, on the other, retailing\index{retailing} them all to the younger
Lucases; and Lydia, in a voice rather louder than any other
person's, was enumerating the various pleasures of the morning
to anybody who would hear her.

"Oh! Mary," said she, "I wish you had gone with us, for we had
such fun!  As we went along, Kitty and I drew up the blinds,
and pretended there was nobody in the coach; and I should have
gone so all the way, if Kitty had not been sick; and when we got
to the George, I do think we behaved very handsomely, for we
treated the other three with the nicest cold luncheon in the
world, and if you would have gone, we would have treated you
too.  And then when we came away it was such fun!  I thought
we never should have got into the coach.  I was ready to die
of laughter.  And then we were so merry all the way home! we
talked and laughed so loud, that anybody might have heard us
ten miles off!"

To this Mary very gravely\index{gravely} replied, "Far be it from me, my dear
sister, to \textbf{depreciate}\index{depreciate}\stepcounter{footnote}\footnotetext{
\textbf{depreciate} \emph{v.}  (dih-PREE-shee-ayt)
to lower the price or value of something; to belittle.} such pleasures!  They would doubtless be
\textbf{congenial}\index{congenial}\stepcounter{footnote}\footnotetext{
\textbf{congenial} \emph{adj.}  (kuhn-JEEN-yuhl)
suitable to your needs, taste, or nature; warm and friendly.} with the generality\index{generality} of female minds.  But I confess
they would have no charms for \emph{me}--I should infinitely prefer a
book."

But of this answer Lydia heard not a word.  She seldom listened
to anybody for more than half a minute\index{minute}, and never attended to
Mary at all.

In the afternoon Lydia was urgent with the rest of the girls
to walk to Meryton, and to see how everybody went on; but
Elizabeth steadily opposed the scheme.  It should not be said
that the Miss Bennets could not be at home half a day before
they were in pursuit of the officers.  There was another reason
too for her opposition.  She dreaded seeing Mr. Wickham again,
and was resolved to avoid it as long as possible.  The comfort
to \emph{her} of the regiment\index{regiment}'s approaching removal was indeed beyond
expression.  In a fortnight\index{fortnight} they were to go--and once gone, she
hoped there could be nothing more to plague\index{plague} her on his account.

She had not been many hours at home before she found that the
Brighton scheme, of which Lydia had given them a hint at the inn,
was under frequent discussion between her parents.  Elizabeth
saw directly that her father had not the smallest intention of
yielding; but his answers were at the same time so vague and
\textbf{equivocal}\index{equivocal}\stepcounter{footnote}\footnotetext{
\textbf{equivocal} \emph{adj.}  (ee-KWIH-vuh-kuhl)
open to two or more interpretations; (often) intended to mislead.}, that her mother, though often \textbf{disheartened}\index{disheartened}\stepcounter{footnote}\footnotetext{
\textbf{dishearten} \emph{tr.v.} \textbf{-ed}  (dihs-HAR-tehn)
to reduce in hope or enthusiasm.}, had never
yet despaired of succeeding at last.



\chapter{40}


Elizabeth's impatience to acquaint Jane with what had happened
could no longer be overcome; and at length, resolving to
suppress\index{suppress} every particular in which her sister was concerned,
and preparing her to be surprised, she related to her the next
morning the chief of the scene between Mr. Darcy and herself.

Miss Bennet's astonishment was soon lessened by the strong
sisterly partiality\index{partiality} which made any admiration of Elizabeth appear
perfectly natural; and all surprise was shortly lost in other
feelings.  She was sorry that Mr. Darcy should have delivered his
sentiments\index{sentiments} in a manner so little suited to recommend them; but
still more was she grieved\index{grieved} for the unhappiness which her sister's
refusal must have given him.

"His being so sure of succeeding was wrong," said she, "and
certainly ought not to have appeared; but consider how much it
must increase his disappointment!"

"Indeed," replied Elizabeth, "I am heartily\index{heartily} sorry for him; but he
has other feelings, which will probably soon drive away his
regard for me.  You do not blame me, however, for refusing him?"

"Blame you!  Oh, no."

"But you blame me for having spoken so warmly of Wickham?"

"No--I do not know that you were wrong in saying what you
did."

"But you \emph{will} know it, when I tell you what happened the very
next day."

She then spoke of the letter, repeating the whole of its contents
as far as they concerned George Wickham.  What a stroke was
this for poor Jane! who would willingly have gone through the
world without believing that so much wickedness existed in the
whole race of mankind, as was here collected in one individual.
Nor was Darcy's \textbf{vindication}\index{vindication}\stepcounter{footnote}\footnotetext{
\textbf{vindication} \emph{n.}  (vihn-dih-KAY-shun)
}, though grateful to her feelings,
capable of consoling her for such discovery.  Most earnestly\index{earnestly} did
she labour to prove the probability of error, and seek to clear the
one without involving the other.

"This will not do," said Elizabeth; "you never will be able to
make both of them good for anything.  Take your choice, but
you must be satisfied with only one.  There is but such a quantity
of merit between them; just enough to make one good sort of
man; and of late it has been shifting about pretty much.  For my
part, I am inclined to believe it all Darcy's; but you shall do
as you choose."

It was some time, however, before a smile could be extorted\index{extorted}
from Jane.

"I do not know when I have been more shocked," said she.
"Wickham so very bad!  It is almost past belief.  And poor Mr.
Darcy!  Dear Lizzy, only consider what he must have suffered.
Such a disappointment! and with the knowledge of your ill
opinion, too! and having to relate such a thing of his sister!
It is really too distressing.  I am sure you must feel it so."

"Oh! no, my regret and compassion are all done away by seeing
you so full of both.  I know you will do him such ample\index{ample} justice,
that I am growing every moment more unconcerned and indifferent\index{indifferent}.
Your \textbf{profusion}\index{profusion}\stepcounter{footnote}\footnotetext{
\textbf{profusion} \emph{n.}  (proh-FYOO-zhun)
extremely abundant.} makes me saving; and if you lament\index{lament} over him much
longer, my heart will be as light as a feather."

"Poor Wickham! there is such an expression of goodness in his
countenance\index{countenance}! such an openness and gentleness in his manner!"

"There certainly was some great mismanagement in the education
of those two young men.  One has got all the goodness, and the
other all the appearance of it."

"I never thought Mr. Darcy so deficient in the \emph{appearance}
of it as you used to do."

"And yet I meant to be uncommonly clever in taking so decided
a dislike to him, without any reason.  It is such a spur to one's
genius, such an opening for wit, to have a dislike of that kind.
One may be continually abusive without saying anything just; but
one cannot always be laughing at a man without now and then
stumbling on something witty."

"Lizzy, when you first read that letter, I am sure you could not
treat the matter as you do now."

"Indeed, I could not.  I was uncomfortable enough, I may say
unhappy.  And with no one to speak to about what I felt, no
Jane to comfort me and say that I had not been so very weak and
vain and nonsensical\index{nonsensical} as I knew I had!  Oh! how I wanted you!"

"How unfortunate that you should have used such very strong
expressions in speaking of Wickham to Mr. Darcy, for now they
\emph{do} appear wholly\index{wholly} undeserved."

"Certainly.  But the misfortune of speaking with bitterness
is a most natural consequence\index{consequence} of the prejudices\index{prejudices} I had been
encouraging.  There is one point on which I want your advice.
I want to be told whether I ought, or ought not, to make our
acquaintances in general understand Wickham's character."

Miss Bennet paused a little, and then replied, "Surely there can
be no occasion for exposing him so dreadfully.  What is your
opinion?"

"That it ought not to be attempted.  Mr. Darcy has not
authorised me to make his communication public.  On the
contrary, every particular relative to his sister was meant to
be kept as much as possible to myself; and if I endeavour\index{endeavour} to
undeceive people as to the rest of his conduct, who will believe
me?  The general prejudice\index{prejudice} against Mr. Darcy is so violent, that
it would be the death of half the good people in Meryton to
attempt to place him in an amiable\index{amiable} light.  I am not equal to it.
Wickham will soon be gone; and therefore it will not signify to
anyone here what he really is.  Some time hence it will be all
found out, and then we may laugh at their stupidity in not
knowing it before.  At present I will say nothing about it."

"You are quite right.  To have his errors made public might ruin
him for ever.  He is now, perhaps, sorry for what he has done,
and anxious to re-establish a character.  We must not make him
desperate."

The tumult\index{tumult} of Elizabeth's mind was \textbf{allayed}\index{allayed}\stepcounter{footnote}\footnotetext{
\textbf{allay} \emph{tr.v.} \textbf{-ed}  (uh-LAY)
To relieve or reduce in intensity.} by this conversation.
She had got rid of two of the secrets which had weighed on her
for a fortnight\index{fortnight}, and was certain of a willing listener in Jane,
whenever she might wish to talk again of either.  But there was
still something lurking behind, of which prudence\index{prudence} forbade the
disclosure.  She dared not relate the other half of Mr. Darcy's
letter, nor explain to her sister how sincerely she had been
valued by her friend.  Here was knowledge in which no one
could \textbf{partake}\index{partake}\stepcounter{footnote}\footnotetext{
\textbf{partake} \emph{v.}  (par-TAYK)
 to take or receive a share of; to participate.}; and she was sensible that nothing less than a
perfect understanding between the parties could justify her in
throwing off this last \textbf{encumbrance}\index{encumbrance}\stepcounter{footnote}\footnotetext{
\textbf{encumbrance} \emph{n.}  (ehn-KUHM-brehnse)
a burden or obstacle.} of mystery.  "And then," said
she, "if that very improbable\index{improbable} event should ever take place, I
shall merely be able to tell what Bingley may tell in a much more
agreeable manner himself.  The liberty of communication cannot
be mine till it has lost all its value!"

She was now, on being settled at home, at leisure to observe the
real state of her sister's spirits.  Jane was not happy.  She still
cherished a very tender affection for Bingley.  Having never even
fancied herself in love before, her regard had all the warmth of
first attachment, and, from her age and disposition\index{disposition}, greater
steadiness than most first attachments often boast; and so
fervently\index{fervently} did she value his \textbf{remembrance}\index{remembrance}\stepcounter{footnote}\footnotetext{
\textbf{remembrance} \emph{n.}  (rih-MEHM-bruhns)
}, and prefer him to every
other man, that all her good sense, and all her attention to the
feelings of her friends, were \textbf{requisite}\index{requisite}\stepcounter{footnote}\footnotetext{
\textbf{requisite} \emph{adj.}  (REHK-wih-ziht)
indispensable, required, or essential.} to check the indulgence of
those regrets which must have been injurious to her own health
and their tranquillity\index{tranquillity}.

"Well, Lizzy," said Mrs. Bennet one day, "what is your opinion
\emph{now} of this sad business of Jane's?  For my part, I am
determined never to speak of it again to anybody.  I told my
sister Phillips so the other day.  But I cannot find out that Jane
saw anything of him in London.  Well, he is a very undeserving
young man--and I do not suppose there's the least chance in the
world of her ever getting him now.  There is no talk of his
coming to Netherfield again in the summer; and I have inquired
of everybody, too, who is likely to know."

"I do not believe he will ever live at Netherfield any more."

"Oh well! it is just as he chooses.  Nobody wants him to come.
Though I shall always say he used my daughter extremely ill; and
if I was her, I would not have put up with it.  Well, my comfort
is, I am sure Jane will die of a broken heart; and then he will
be sorry for what he has done."

But as Elizabeth could not receive comfort from any such
expectation, she made no answer.

"Well, Lizzy," continued her mother, soon afterwards, "and so
the Collinses live very comfortable, do they?  Well, well, I only
hope it will last.  And what sort of table do they keep?  Charlotte
is an excellent manager, I dare say.  If she is half as sharp as
her mother, she is saving enough.  There is nothing extravagant in
\emph{their} housekeeping, I dare say."

"No, nothing at all."

"A great deal of good management, depend upon it.  Yes, yes.
\emph{they} will take care not to outrun their income.  \emph{they} will
never be distressed for money.  Well, much good may it do
them!  And so, I suppose, they often talk of having Longbourn
when your father is dead.  They look upon it as quite their own,
I dare say, whenever that happens."

"It was a subject\index{subject} which they could not mention before me."

"No; it would have been strange if they had; but I make no
doubt they often talk of it between themselves.  Well, if they
can be easy with an estate that is not lawfully their own, so
much the better.  I should be ashamed of having one that was
only entailed\index{entailed} on me."



\chapter{41}


The first week of their return was soon gone.  The second began.
It was the last of the regiment\index{regiment}'s stay in Meryton, and all the
young ladies in the neighbourhood were \textbf{drooping}\index{drooping}\stepcounter{footnote}\footnotetext{
\textbf{droop} \emph{v.} \textbf{-ing}  (DROOP)
 hanging down (as from exhaustion or weakness).} \textbf{apace}\index{apace}\stepcounter{footnote}\footnotetext{
\textbf{apace} \emph{adv.}  (uh-PACE)
 quick, or with rapid movement.}.  The
dejection\index{dejection} was almost universal.  The elder Miss Bennets alone
were still able to eat, drink, and sleep, and pursue the usual
course of their employments.  Very frequently were they
reproached\index{reproached} for this insensibility by Kitty and Lydia, whose
own misery was extreme, and who could not comprehend\index{comprehend} such
hard-heartedness in any of the family.

"Good Heaven! what is to become of us?  What are we to do?"
would they often exclaiming the bitterness of woe.  "How can
you be smiling so, Lizzy?"

Their affectionate mother shared all their grief; she remembered
what she had herself endured on a similar occasion, five-and-twenty
years ago.

"I am sure," said she, "I cried for two days together when
Colonel Miller's regiment\index{regiment} went away.  I thought I should have
broken my heart."

"I am sure I shall break \emph{mine}," said Lydia.

"If one could but go to Brighton!" observed Mrs. Bennet.

"Oh, yes!--if one could but go to Brighton!  But papa is so
disagreeable."

"A little sea-bathing would set me up forever."

"And my aunt Phillips is sure it would do \emph{me} a great deal of
good," added Kitty.

Such were the kind of \textbf{lamentations}\index{lamentations}\stepcounter{footnote}\footnotetext{
\textbf{lamentation} \emph{n.}  (lam-ehn-TAY-shun)
a cry of sorrow and grief.} resounding perpetually\index{perpetually}
through Longbourn House.  Elizabeth tried to be \textbf{diverted}\index{diverted}\stepcounter{footnote}\footnotetext{
\textbf{divert} \emph{v.} \textbf{-ed}  (dih-VERT)
 to turn aside; to turn away from.} by
them; but all sense of pleasure was lost in shame.  She felt anew
the justice of Mr. Darcy's objections; and never had she been so
much disposed\index{disposed} to pardon his interference in the views of his
friend.

But the gloom of Lydia's prospect\index{prospect} was shortly cleared away; for
she received an invitation from Mrs. Forster, the wife of the
colonel of the regiment\index{regiment}, to accompany her to Brighton.  This
invaluable\index{invaluable} friend was a very young woman, and very lately
married.  A resemblance in good humour and good spirits had
recommended her and Lydia to each other, and out of their
\emph{three} months' acquaintance they had been intimate \emph{two}.

The rapture\index{rapture} of Lydia on this occasion, her adoration of Mrs.
Forster, the delight of Mrs. Bennet, and the mortification\index{mortification} of
Kitty, are scarcely to be described.  Wholly\index{Wholly} inattentive to her
sister's feelings, Lydia flew about the house in restless ecstasy\index{ecstasy},
calling for everyone's congratulations, and laughing and talking
with more violence than ever; whilst the luckless Kitty continued
in the parlour repined\index{repined} at her fate in terms as unreasonable as her
accent was peevish\index{peevish}.

"I cannot see why Mrs. Forster should not ask \emph{me} as well as
Lydia," said she, "Though I am \emph{not} her particular friend.  I
have just as much right to be asked as she has, and more too,
for I am two years older."

In vain did Elizabeth attempt to make her reasonable, and Jane
to make her resigned.  As for Elizabeth herself, this invitation
was so far from exciting in her the same feelings as in her mother
and Lydia, that she considered it as the death warrant of all
possibility of common sense for the latter; and detestable as such
a step must make her were it known, she could not help secretly
advising her father not to let\index{let} her go.  She represented to him all
the improprieties of Lydia's general behaviour, the little
advantage she could derive from the friendship of such a woman
as Mrs. Forster, and the probability of her being yet more
imprudent\index{imprudent} with such a companion at Brighton, where the
temptations must be greater than at home.  He heard her
attentively, and then said:

"Lydia will never be easy until she has exposed herself in some
public place or other, and we can never expect her to do it with
so little expense or inconvenience to her family as under the
present circumstances."

"If you were aware," said Elizabeth, "of the very great
disadvantage to us all which must arise from the public notice
of Lydia's unguarded and imprudent\index{imprudent} manner--nay, which has
already arisen from it, I am sure you would judge differently in
the affair."

"Already arisen?" repeated Mr. Bennet.  "What, has she
frightened away some of your lovers?  Poor little Lizzy!  But do
not be cast down.  Such \textbf{squeamish}\index{squeamish}\stepcounter{footnote}\footnotetext{
\textbf{squeamish} \emph{adj.}  (SKWEE-mish)
easily sickened or disgusted.} youths as cannot bear to be
connected with a little absurdity are not worth a regret.  Come,
let\index{let} me see the list of pitiful fellows who have been kept \textbf{aloof}\index{aloof}\stepcounter{footnote}\footnotetext{
\textbf{aloof} \emph{adj.}  (uh-LOOF)
 emotionally reserved or remote.}
by Lydia's folly\index{folly}."

"Indeed you are mistaken.  I have no such injuries to resent.
It is not of particular, but of general evils, which I am now
complaining.  Our importance, our respectability in the world must
be affected by the wild \textbf{volatility}\index{volatility}\stepcounter{footnote}\footnotetext{
\textbf{volatile} \emph{adj.}  (VAW-luh-TILE)
being easily excited.}, the assurance and disdain of
all restraint which mark Lydia's character.  Excuse me, for I must
speak plainly.  If you, my dear father, will not take the trouble
of checking her \textbf{exuberant}\index{exuberant}\stepcounter{footnote}\footnotetext{
\textbf{exuberant} \emph{adj.}  (ehg-ZYOO-ber-ehnT)
feeling unrestrained joy.} spirits, and of teaching her that her
present pursuits are not to be the business of her life, she will
soon be beyond the reach of amendment\index{amendment}.  Her character will be
fixed, and she will, at sixteen, be the most determined flirt that
ever made herself or her family ridiculous; a flirt, too, in the
worst and meanest degree of flirtation; without any attraction
beyond youth and a tolerable\index{tolerable} person; and, from the ignorance and
emptiness of her mind, wholly\index{wholly} unable to \textbf{ward}\index{ward}\stepcounter{footnote}\footnotetext{
\textbf{Ward} \emph{Robert Elsmere}  (WAWRD)
} off any portion of
that universal contempt\index{contempt} which her rage for admiration will excite.
In this danger Kitty also is \textbf{comprehended}\index{comprehended}\stepcounter{footnote}\footnotetext{
\textbf{comprehend} \emph{tr.v.} \textbf{-ed}  (kahm-prih-HEHND)
to fully understand or grasp.}.  She will follow wherever
Lydia leads.  Vain, ignorant, idle, and absolutely uncontrolled!
Oh! my dear father, can you suppose it possible that they will not
be censured\index{censured} and despised wherever they are known, and that their
sisters will not be often involved in the disgrace?"

Mr. Bennet saw that her whole heart was in the subject\index{subject}, and
affectionately taking her hand said in reply:

"Do not make yourself uneasy, my love.  Wherever you and Jane
are known you must be respected and valued; and you will not
appear to less advantage for having a couple of--or I may say,
three--very silly sisters.  We shall have no peace at Longbourn
if Lydia does not go to Brighton.  Let\index{Let} her go, then.  Colonel
Forster is a sensible man, and will keep her out of any real
mischief; and she is luckily too poor to be an object of prey
to anybody.  At Brighton she will be of less importance even as
a common flirt than she has been here.  The officers will find
women better worth their notice.  Let\index{Let} us hope, therefore, that
her being there may teach her her own insignificance.  At any
rate, she cannot grow many degrees worse, without authorising
us to lock her up for the rest of her life."

With this answer Elizabeth was forced to be content; but her
own opinion continued the same, and she left him disappointed
and sorry.  It was not in her nature, however, to increase her
vexations\index{vexations} by dwelling on them.  She was confident of having
performed her duty, and to fret\index{fret} over unavoidable evils, or
\textbf{augment}\index{augment}\stepcounter{footnote}\footnotetext{
\textbf{augment} \emph{v.}  (awg-MEHNT)
to enlarge, increase, or intensify.} them by anxiety, was no part of her disposition\index{disposition}.

Had Lydia and her mother known the substance of her conference
with her father, their indignation\index{indignation} would hardly have found
expression in their united \textbf{volubility}\index{volubility}\stepcounter{footnote}\footnotetext{
\textbf{voluble} \emph{adj.}  (VAWL-yoo-bl)
 fluent, ready, and easy in speach.}.  In Lydia's imagination,
a visit to Brighton comprised every possibility of earthly
happiness.  She saw, with the creative eye of fancy, the streets
of that gay bathing-place covered with officers.  She saw
herself the object of attention, to tens and to scores of them
at present unknown.  She saw all the glories of the camp--its
tents stretched forth in beauteous uniformity of lines, crowded
with the young and the gay, and dazzling with scarlet; and, to
complete the view, she saw herself seated beneath a tent, tenderly
flirting with at least six officers at once.

Had she known her sister sought to tear her from such prospects\index{prospects}
and such realities as these, what would have been her sensations?
They could have been understood only by her mother, who might
have felt nearly the same.  Lydia's going to Brighton was all
that consoled\index{consoled} her for her melancholy\index{melancholy} conviction\index{conviction} of her husband's
never intending to go there himself.

But they were entirely ignorant of what had passed; and their
raptures\index{raptures} continued, with little intermission\index{intermission}, to the very day of
Lydia's leaving home.

Elizabeth was now to see Mr. Wickham for the last time.
Having been frequently in company with him since her return,
agitation was pretty well over; the agitations of formal partiality\index{partiality}
entirely so.  She had even learnt to detect, in the very gentleness
which had first delighted her, an affectation\index{affectation} and a sameness to
disgust and weary.  In his present behaviour to herself,
moreover, she had a fresh source of displeasure, for the
inclination\index{inclination} he soon testified of renewing those intentions which
had marked the early part of their acquaintance could only serve,
after what had since passed, to provoke\index{provoke} her.  She lost all concern
for him in finding herself thus selected as the object of such idle
and \textbf{frivolous}\index{frivolous}\stepcounter{footnote}\footnotetext{
\textbf{frivolous} \emph{adj.}  (FRIH-vuh-luhs)
not serious in content, attitude, or behavior; trivial.} gallantry\index{gallantry}; and while she steadily repressed\index{repressed} it, could
not but feel the reproof\index{reproof} contained in his believing, that however
long, and for whatever cause, his attentions had been withdrawn,
her vanity\index{vanity} would be gratified, and her preference secured at any
time by their renewal.

On the very last day of the regiment\index{regiment}'s remaining at Meryton, he
dined, with other of the officers, at Longbourn; and so little
was Elizabeth disposed\index{disposed} to part from him in good humour, that on
his making some inquiry as to the manner in which her time had
passed at Hunsford, she mentioned Colonel Fitzwilliam's and
Mr. Darcy's having both spent three weeks at Rosings, and
asked him, if he was acquainted with the former.

He looked surprised, displeased, alarmed; but with a moment's
recollection\index{recollection} and a returning smile, replied, that he had formerly
seen him often; and, after observing that he was a very
gentlemanlike man, asked her how she had liked him.  Her
answer was warmly in his favour.  With an air of indifference\index{indifference} he
soon afterwards added:

"How long did you say he was at Rosings?"

"Nearly three weeks."

"And you saw him frequently?"

"Yes, almost every day."

"His manners are very different from his cousin's."

"Yes, very different.  But I think Mr. Darcy improves upon
acquaintance."

"Indeed!" cried Mr. Wickham with a look which did not escape
her.  "And pray, may I ask?--"  But checking himself, he added,
in a gayer tone, "Is it in address\index{address} that he improves?  Has he
deigned\index{deigned} to add aught of civility\index{civility} to his ordinary style?--for I
dare not hope," he continued in a lower and more serious tone,
"that he is improved in essentials."

"Oh, no!" said Elizabeth.  "In essentials, I believe, he is very
much what he ever was."

While she spoke, Wickham looked as if scarcely knowing
whether to rejoice over her words, or to distrust their meaning.
There was a something in her countenance\index{countenance} which made him listen
with an \textbf{apprehensive}\index{apprehensive}\stepcounter{footnote}\footnotetext{
\textbf{apprehensive} \emph{adj.}  (ap-ree-HEHN-sihv)
worried, anxious, or fearful.} and anxious attention, while she added:

"When I said that he improved on acquaintance, I did not mean
that his mind or his manners were in a state of improvement, but
that, from knowing him better, his disposition\index{disposition} was better
understood."

Wickham's alarm now appeared in a heightened complexion and
agitated look; for a few minuted\index{minuted} he was silent, till, shaking
off his embarrassment, he turned to her again, and said in the
gentlest of accents:

"You, who so well know my feeling towards Mr. Darcy, will
readily\index{readily} comprehend\index{comprehend} how sincerely I must rejoice that he is
wise enough to assume even the \emph{appearance} of what is right.
His pride, in that direction, may be of service, if not to himself,
to many others, for it must only \textbf{deter}\index{deter}\stepcounter{footnote}\footnotetext{
\textbf{deter} \emph{v.}  (dih-TER)
 to persuade someone not to do something.} him from such foul
misconduct as I have suffered by.  I only fear that the sort of
cautiousness to which you, I imagine, have been alluding, is
merely adopted on his visits to his aunt, of whose good opinion
and judgement he stands much in awe\index{awe}.  His fear of her has
always operated, I know, when they were together; and a good
deal is to be imputed\index{imputed} to his wish of forwarding\index{forwarding} the match with
Miss de Bourgh, which I am certain he has very much at heart."

Elizabeth could not repress\index{repress} a smile at this, but she answered only
by a slight inclination\index{inclination} of the head.  She saw that he wanted to
engage her on the old subject\index{subject} of his grievances\index{grievances}, and she was in
no humour to indulge him.  The rest of the evening passed with
the \emph{appearance}, on his side, of usual cheerfulness, but with
no further attempt to distinguish Elizabeth; and they parted at
last with mutual civility\index{civility}, and possibly a mutual desire of never
meeting again.

When the party broke up, Lydia returned with Mrs. Forster to
Meryton, from whence they were to set out early the next
morning.  The separation between her and her family was rather
noisy than pathetic.  Kitty was the only one who shed tears; but
she did weep from vexation\index{vexation} and envy.  Mrs. Bennet was diffuse\index{diffuse}
in her good wishes for the felicity\index{felicity} of her daughter, and
impressive in her injunctions\index{injunctions} that she should not miss the
opportunity of enjoying herself as much as possible--advice
which there was every reason to believe would be well attended
to; and in the \textbf{clamorous}\index{clamorous}\stepcounter{footnote}\footnotetext{
\textbf{clamorous} \emph{adj.}  (KLAM-mer-uhs)
a sustained din as if from a crowd; offensively loud.} happiness of Lydia herself in bidding
farewell, the more gentle adieus\index{adieus} of her sisters were uttered
without being heard.



\chapter{42}


Had Elizabeth's opinion been all drawn from her own family,
she could not have formed a very pleasing opinion of \textbf{conjugal}\index{conjugal}\stepcounter{footnote}\footnotetext{
\textbf{conjugal} \emph{adj.}  (KAHN-joo-guhl)
relating to marriage.}
felicity\index{felicity} or domestic\index{domestic} comfort.  Her father, captivated by youth
and beauty, and that appearance of good humour which youth
and beauty generally give, had married a woman whose weak
understanding and \textbf{illiberal}\index{illiberal}\stepcounter{footnote}\footnotetext{
\textbf{illiberal} \emph{adj.}  (ihl-LIH-ber-uhl)
narrow-minded or bigoted about one's opinions.} mind had very early in their marriage
put an end to all real affection for her.  Respect, esteem, and
confidence\index{confidence} had vanished for ever; and all his views of domestic\index{domestic}
happiness were overthrown.  But Mr. Bennet was not of a
disposition\index{disposition} to seek comfort for the disappointment which his
own imprudence\index{imprudence} had brought on, in any of those pleasures which
too often console\index{console} the unfortunate for their folly\index{folly} of their vice\index{vice}.
He was fond of the country and of books; and from these tastes had
arisen his principal\index{principal} enjoyments.  To his wife he was very little
otherwise indebted, than as her ignorance and folly\index{folly} had
contributed to his amusement.  This is not the sort of happiness
which a man would in general wish to owe to his wife; but
where other powers of entertainment are wanting, the true
\textbf{philosopher}\index{philosopher}\stepcounter{footnote}\footnotetext{
\textbf{philosopher} \emph{n.}  (fih-LAH-suh-fer)
} will derive benefit from such as are given.

Elizabeth, however, had never been blind to the impropriety\index{impropriety} of
her father's behaviour as a husband.  She had always seen it with
pain; but respecting his abilities, and grateful for his affectionate
treatment of herself, she endeavoured\index{endeavoured} to forget what she could
not overlook, and to banish from her thoughts that continual
breach\index{breach} of conjugal\index{conjugal} obligation and decorum\index{decorum} which, in exposing
his wife to the contempt\index{contempt} of her own children, was so highly
reprehensible\index{reprehensible}.  But she had never felt so strongly as now the
disadvantages which must attend the children of so unsuitable a
marriage, nor ever been so fully aware of the evils arising from
so ill-judged a direction of talents; talents, which, rightly used,
might at least have preserved the respectability of his daughters,
even if incapable of enlarging the mind of his wife.

When Elizabeth had rejoiced over Wickham's departure she
found little other cause for satisfaction in the loss of the
regiment\index{regiment}.  Their parties abroad were less varied than before, and
at home she had a mother and sister whose constant \textbf{repinings}\index{repinings}\stepcounter{footnote}\footnotetext{
\textbf{} \emph{}  ()
} at
the dullness of everything around them threw a real gloom over
their domestic\index{domestic} circle; and, though Kitty might in time regain her
natural degree of sense, since the disturbers of her brain were
removed, her other sister, from whose disposition\index{disposition} greater evil
might be apprehended\index{apprehended}, was likely to be hardened in all her
folly\index{folly} and assurance by a situation of such double danger as a
watering-place and a camp.  Upon the whole, therefore, she
found, what has been sometimes been found before, that an
event to which she had been looking with impatient desire did
not, in taking place, bring all the satisfaction she had promised
herself.  It was consequently necessary to name some other
period for the \textbf{commencement}\index{commencement}\stepcounter{footnote}\footnotetext{
\textbf{commencement} \emph{n.}  (kuh-MENS-mehnt)
} of actual felicity\index{felicity}--to have some
other point on which her wishes and hopes might be fixed, and
by again enjoying the pleasure of anticipation, console\index{console} herself for
the present, and prepare for another disappointment.  Her tour
to the Lakes was now the object of her happiest thoughts; it was
her best consolation\index{consolation} for all the uncomfortable hours which the
discontentedness of her mother and Kitty made inevitable; and
could she have included Jane in the scheme, every part of it
would have been perfect.

"But it is fortunate," thought she, "that I have something to wish
for.  Were the whole arrangement complete, my disappointment
would be certain.  But here, by carrying with me one ceaseless
source of regret in my sister's absence, I may reasonably hope to
have all my expectations of pleasure realised.  A scheme of
which every part promises delight can never be successful; and
general disappointment is only warded\index{warded} off by the defence of
some little peculiar\index{peculiar} vexation\index{vexation}."

When Lydia went away she promised to write very often and
very minutely\index{minutely}\index{minutely\index{minutely}} to her mother and Kitty; but her letters were
always long expected, and always very short.  Those to her
mother contained little else than that they were just returned
from the library, where such and such officers had attended
them, and where she had seen such beautiful ornaments as made
her quite wild; that she had a new gown, or a new parasol, which
she would have described more fully, but was obliged to leave
off in a violent hurry, as Mrs. Forster called her, and they were
going off to the camp; and from her correspondence with her
sister,  there was still less to be learnt--for her letters to
Kitty, though rather longer, were much too full of lines under the
words to be made public.

After the first fortnight\index{fortnight} or three weeks of her absence, health,
good humour, and cheerfulness began to reappear at Longbourn.
Everything wore a happier aspect\index{aspect}.  The families who had been in
town for the winter came back again, and summer finery and
summer engagements arose.  Mrs. Bennet was restored to her
usual \textbf{querulous}\index{querulous}\stepcounter{footnote}\footnotetext{
\textbf{querulous} \emph{adj.}  (KWEH-ruh-luhs)
constantly complaining.} serenity\index{serenity}; and, by the middle of June, Kitty was
so much recovered as to be able to enter Meryton without tears;
an event of such happy promise as to make Elizabeth hope that
by the following Christmas she might be so tolerably reasonable
as not to mention an officer above once a day, unless, by some
cruel and malicious\index{malicious} arrangement at the War Office, another
regiment\index{regiment} should be quartered in Meryton.

The time fixed for the beginning of their northern tour was now
fast approaching, and a fortnight\index{fortnight} only was wanting of it, when
a letter arrived from Mrs. Gardiner, which at once delayed its
commencement\index{commencement} and \textbf{curtailed}\index{curtailed}\stepcounter{footnote}\footnotetext{
\textbf{curtail} \emph{tr.v.} \textbf{-ed}  (ker-TAYL)
 to cut short.} its extent.  Mr. Gardiner would be
prevented by business from setting out till a fortnight\index{fortnight} later in
July, and must be in London again within a month, and as that
left too short a period for them to go so far, and see so much
as they had proposed, or at least to see it with the leisure and
comfort they had built on, they were obliged to give up the
Lakes, and substitute a more contracted tour, and, according
to the present plan, were to go no farther northwards than
Derbyshire.  In that county there was enough to be seen to
occupy the chief of their three weeks; and to Mrs. Gardiner it
had a peculiarly\index{peculiarly} strong attraction.  The town where she had
formerly passed some years of her life, and where they were now
to spend a few days, was probably as great an object of her
curiosity as all the celebrated beauties of Matlock, Chatsworth,
Dovedale, or the Peak.

Elizabeth was excessively disappointed; she had set her heart on
seeing the Lakes, and still thought there might have been time
enough.  But it was her business to be satisfied--and certainly
her temper to be happy; and all was soon right again.

With the mention of Derbyshire there were many ideas connected.
It was impossible for her to see the word without thinking of
Pemberley and its owner.  "But surely," said she, "I may enter
his county without \textbf{impunity}\index{impunity}\stepcounter{footnote}\footnotetext{
\textbf{impunity} \emph{n.}  (ihm-PYOO-nih-tee)
exemption from punishment or loss.}, and rob it of a few \textbf{petrified}\index{petrified}\stepcounter{footnote}\footnotetext{
\textbf{petrify} \emph{v.} \textbf{-ed}  (PEHT-rih-fy)
} \textbf{spars}\index{spars}\stepcounter{footnote}\footnotetext{
\textbf{spar} \emph{n.}  (SPAHR)
}
without his perceiving me."

The period of expectation was now doubled.  Four weeks were to
pass away before her uncle and aunt's arrival.  But they did pass
away, and Mr. and Mrs. Gardiner, with their four children, did
at length appear at Longbourn.  The children, two girls of six
and eight years old, and two younger boys, were to be left under
the particular care of their cousin Jane, who was the general
favourite, and whose steady\index{steady} sense and sweetness of temper exactly
adapted her for attending to them in every way--teaching them,
playing with them, and loving them.

The Gardiners stayed only one night at Longbourn, and set off
the next morning with Elizabeth in pursuit of novelty\index{novelty} and
amusement.  One enjoyment was certain--that of suitableness
of companions; a suitableness which comprehended\index{comprehended} health and
temper to bear inconveniences--cheerfulness to enhance every
pleasure--and affection and intelligence, which might supply
it among themselves if there were disappointments abroad.

It is not the object of this work to give a description of
Derbyshire, nor of any of the remarkable places through which
their route thither lay; Oxford, Blenheim, Warwick, Kenilworth,
Birmingham, etc. are sufficiently known.  A small part of
Derbyshire is all the present concern.  To the little town of
Lambton, the scene of Mrs. Gardiner's former residence, and
where she had lately learned some acquaintance still remained,
they bent their steps, after having seen all the principal\index{principal} wonders
of the country; and within five miles of Lambton, Elizabeth
found from her aunt that Pemberley was situated.  It was not in
their direct road, nor more than a mile or two out of it.  In
talking over their route the evening before, Mrs. Gardiner
expressed an inclination\index{inclination} to see the place again.  Mr. Gardiner
declared his willingness, and Elizabeth was applied to for her
approbation\index{approbation}.

"My love, should not you like to see a place of which you have
heard so much?" said her aunt; "a place, too, with which so
many of your acquaintances are connected.  Wickham passed all
his youth there, you know."

Elizabeth was distressed.  She felt that she had no business at
Pemberley, and was obliged to assume a disinclination\index{disinclination} for seeing
it.  She must own that she was tired of seeing great houses; after
going over so many, she really had no pleasure in fine carpets or
satin curtains.

Mrs. Gardiner abused her stupidity.  "If it were merely a fine
house richly furnished," said she, "I should not care about it
myself; but the grounds are delightful.  They have some of the
finest woods in the country."

Elizabeth said no more--but her mind could not \textbf{acquiesce}\index{acquiesce}\stepcounter{footnote}\footnotetext{
\textbf{acquiesce} \emph{intr.v.}  (ak-wee-ESS)
to agree or express agreement; to consent.}.
The possibility of meeting Mr. Darcy, while viewing the place,
instantly occurred.  It would be dreadful!  She blushed at the
very idea, and thought it would be better to speak openly to
her aunt than to run such a risk.  But against this there were
objections; and she finally resolved that it could be the last
resource, if her private\index{private} inquiries to the absence of the family
were unfavourably answered.

Accordingly, when she retired at night, she asked the chambermaid
whether Pemberley were not a very fine place? what was the name
of its proprietor\index{proprietor}? and, with no little alarm, whether the family
were down for the summer?  A most welcome negative followed the
last question--and her alarms now being removed, she was at
leisure to feel a great deal of curiosity to see the house herself;
and when the subject\index{subject} was revived the next morning, and she was
again applied to, could readily\index{readily} answer, and with a proper air of
indifference\index{indifference}, that she had not really any dislike to the scheme.
To Pemberley, therefore, they were to go.



\chapter{43}


Elizabeth, as they drove along, watched for the first appearance
of Pemberley Woods with some \textbf{perturbation}\index{perturbation}\stepcounter{footnote}\footnotetext{
\textbf{perturbation} \emph{n.}  (per-ter-BAY-shun)
an unhappy and worried mental state.}; and when at
length they turned in at the lodge, her spirits were in a high
flutter.

The park was very large, and contained great variety of ground.
They entered it in one of its lowest points, and drove for some
time through a beautiful wood stretching over a wide extent.

Elizabeth's mind was too full for conversation, but she saw and
admired every remarkable spot and point of view.  They
gradually ascended for half-a-mile, and then found themselves
at the top of a considerable \textbf{eminence}\index{eminence}\stepcounter{footnote}\footnotetext{
\textbf{eminence} \emph{n.}  (EHM-mih-nehnse)
high status importance owing to marked superiority.}, where the wood ceased,
and the eye was instantly caught by Pemberley House, situated
on the opposite side of a valley, into which the road with some
abruptness wound.  It was a large, handsome stone building,
standing well on rising ground, and backed by a ridge of high
woody hills; and in front, a stream of some natural importance
was swelled into greater, but without any artificial appearance.
Its banks were neither formal nor falsely adorned.  Elizabeth
was delighted.  She had never seen a place for which nature
had done more, or where natural beauty had been so little
counteracted by an awkward taste.  They were all of them warm
in their admiration; and at that moment she felt that to be
mistress of Pemberley might be something!

They descended the hill, crossed the bridge, and drove to the
door; and, while examining the nearer aspect\index{aspect} of the house, all
her apprehension\index{apprehension} of meeting its owner returned.  She dreaded
lest the chambermaid had been mistaken.  On applying to see
the place, they were admitted into the hall; and Elizabeth, as
they waited for the housekeeper, had leisure to wonder at her
being where she was.

The housekeeper came; a respectable-looking elderly woman,
much less fine, and more civil, than she had any notion of
finding her.  They followed her into the dining-parlour.
It was a large, well proportioned room, handsomely fitted up.
Elizabeth, after slightly surveying it, went to a window to enjoy
its prospect\index{prospect}.  The hill, crowned with wood, which they had
descended, receiving increased abruptness from the distance,
was a beautiful object.  Every disposition\index{disposition} of the ground was
good; and she looked on the whole scene, the river, the trees
scattered on its banks and the winding of the valley, as far as
she could trace it, with delight.  As they passed into other
rooms these objects were taking different positions; but from
every window there were beauties to be seen.  The rooms were
lofty and handsome, and their furniture suitable to the fortune
of its proprietor\index{proprietor}; but Elizabeth saw, with admiration of his taste,
that it was neither \textbf{gaudy}\index{gaudy}\stepcounter{footnote}\footnotetext{
\textbf{gaudy} \emph{adj.}  (GAW-dee)
 tastelessly showy.} nor uselessly fine; with less of
\textbf{splendour}\index{splendour}\stepcounter{footnote}\footnotetext{
\textbf{splendour} \emph{n.}  (SPLEHN-der)
magnificent, splendid, or grand; same as splendor.}, and more real elegance, than the furniture of Rosings.

"And of this place," thought she, "I might have been mistress!
With these rooms I might now have been familiarly acquainted!
Instead of viewing them as a stranger, I might have rejoiced
in them as my own, and welcomed to them as visitors my uncle
and aunt.  But no,"--recollecting herself--"that could never
be; my uncle and aunt would have been lost to me; I should not
have been allowed to invite them."

This was a lucky recollection\index{recollection}--it saved her from something
very like regret.

She longed to inquire of the housekeeper whether her master
was really absent, but had not the courage for it.  At length
however, the question was asked by her uncle; and she turned
away with alarm, while Mrs. Reynolds replied that he was,
adding, "But we expect him to-morrow, with a large party of
friends."  How rejoiced was Elizabeth that their own journey
had not by any circumstance been delayed a day!

Her aunt now called her to look at a picture.  She approached
and saw the likeness of Mr. Wickham, suspended, amongst
several other miniatures, over the mantelpiece.  Her aunt asked
her, smilingly, how she liked it.  The housekeeper came forward\index{forward},
and told them it was a picture of a young gentleman, the son of
her late master's steward\index{steward}, who had been brought up by him at
his own expense.  "He is now gone into the army," she added;
"but I am afraid he has turned out very wild."

Mrs. Gardiner looked at her niece with a smile, but Elizabeth
could not return it.

"And that," said Mrs. Reynolds, pointing to another of the
miniatures, "is my master--and very like him.  It was drawn at
the same time as the other--about eight years ago."

"I have heard much of your master's fine person," said Mrs.
Gardiner, looking at the picture; "it is a handsome face.
But, Lizzy, you can tell us whether it is like or not."

Mrs. Reynolds respect for Elizabeth seemed to increase on this
intimation of her knowing her master.

"Does that young lady know Mr. Darcy?"

Elizabeth coloured, and said: "A little."

"And do not you think him a very handsome gentleman, ma'am?"

"Yes, very handsome."

"I am sure I know none so handsome; but in the gallery
upstairs you will see a finer, larger picture of him than this.
This room was my late master's favourite room, and these
miniatures are just as they used to be then.  He was very fond
of them."

This accounted to Elizabeth for Mr. Wickham's being among them.

Mrs. Reynolds then directed their attention to one of Miss Darcy,
drawn when she was only eight years old.

"And is Miss Darcy as handsome as her brother?" said Mrs. Gardiner.

"Oh! yes--the handsomest young lady that ever was seen; and
so accomplished!--She plays and sings all day long.  In the next
room is a new instrument just come down for her--a present
from my master; she comes here to-morrow with him."

Mr. Gardiner, whose manners were very easy and pleasant,
encouraged her communicativeness by his questions and remarks;
Mrs. Reynolds, either by pride or attachment, had evidently
great pleasure in talking of her master and his sister.

"Is your master much at Pemberley in the course of the year?"

"Not so much as I could wish, sir; but I dare say he may spend
half his time here; and Miss Darcy is always down for the
summer months."

"Except," thought Elizabeth, "when she goes to Ramsgate."

"If your master would marry, you might see more of him."

"Yes, sir; but I do not know when \emph{that} will be.  I do not
know who is good enough for him."

Mr. and Mrs. Gardiner smiled.  Elizabeth could not help saying,
"It is very much to his credit, I am sure, that you should think
so."

"I say no more than the truth, and everybody will say that
knows him," replied the other.  Elizabeth thought this was
going pretty far; and she listened with increasing astonishment
as the housekeeper added, "I have never known a cross word
from him in my life, and I have known him ever since he was
four years old."

This was praise, of all others most extraordinary, most opposite
to her ideas.  That he was not a good-tempered man had been
her firmest opinion.  Her keenest attention was awakened; she
longed to hear more, and was grateful to her uncle for saying:

"There are very few people of whom so much can be said.  You
are lucky in having such a master."

"Yes, sir, I know I am.  If I were to go through the world, I
could not meet with a better.  But I have always observed, that
they who are good-natured when children, are good-natured
when they grow up; and he was always the sweetest-tempered,
most generous-hearted boy in the world."

Elizabeth almost stared at her.  "Can this be Mr. Darcy?"
thought she.

"His father was an excellent man," said Mrs. Gardiner.

"Yes, ma'am, that he was indeed; and his son will be just like
him--just as affable\index{affable} to the poor."

Elizabeth listened, wondered, doubted, and was impatient for
more.  Mrs. Reynolds could interest her on no other point.  She
related the subjects\index{subjects} of the pictures, the dimensions of the rooms,
and the price of the furniture, in vain, Mr. Gardiner, highly
amused by the kind of family prejudice\index{prejudice} to which he attributed
her excessive commendation\index{commendation} of her master, soon led again to
the subject\index{subject}; and she dwelt with energy on his many merits as
they proceeded together up the great staircase.

"He is the best landlord, and the best master," said she, "that
ever lived; not like the wild young men nowadays, who think of
nothing but themselves.  There is not one of his tenants or
servants but will give him a good name.  Some people call him
proud; but I am sure I never saw anything of it.  To my fancy, it
is only because he does not rattle away like other young men."

"In what an amiable\index{amiable} light does this place him!" thought
Elizabeth.

"This fine account of him," whispered her aunt as they walked,
"is not quite consistent with his behaviour to our poor friend."

"Perhaps we might be deceived."

"That is not very likely; our authority was too good."

On reaching the spacious lobby above they were shown into a
very pretty sitting-room, lately fitted up with greater elegance
and lightness than the apartments\index{apartments} below; and were informed that
it was but just done to give pleasure to Miss Darcy, who had
taken a liking to the room when last at Pemberley.

"He is certainly a good brother," said Elizabeth, as she walked
towards one of the windows.

Mrs. Reynolds anticipated Miss Darcy's delight, when she
should enter the room.  "And this is always the way with him,"
she added.  "Whatever can give his sister any pleasure is sure
to be done in a moment.  There is nothing he would not do for
her."

The picture-gallery, and two or three of the principal\index{principal} bedrooms,
were all that remained to be shown.  In the former were many
good paintings; but Elizabeth knew nothing of the art; and from
such as had been already visible below, she had willingly turned
to look at some drawings of Miss Darcy's, in crayons, whose
subjects\index{subjects} were usually more interesting, and also more intelligible.

In the gallery there were many family portraits, but they could
have little to fix the attention of a stranger.  Elizabeth walked
in quest of the only face whose features would be known to her.
At last it arrested her--and she beheld a striking resemblance
to Mr. Darcy, with such a smile over the face as she remembered
to have sometimes seen when he looked at her.  She stood
several minutes\index{minutes} before the picture, in earnest\index{earnest} contemplation\index{contemplation},
and returned to it again before they quitted the gallery.  Mrs.
Reynolds informed them that it had been taken in his father's
lifetime.

There was certainly at this moment, in Elizabeth's mind, a more
gentle sensation towards the original than she had ever felt at
the height of their acquaintance.  The commendation\index{commendation} bestowed\index{bestowed}
on him by Mrs. Reynolds was of no trifling\index{trifling} nature.  What praise
is more valuable than the praise of an intelligent servant?  As a
brother, a landlord, a master, she considered how many people's
happiness were in his guardianship!--how much of pleasure or
pain was it in his power to bestow\index{bestow}!--how much of good or evil
must be done by him!  Every idea that had been brought forward\index{forward}
by the housekeeper was favourable to his character, and as she
stood before the canvas on which he was represented, and fixed
his eyes upon herself, she thought of his regard with a deeper
sentiment\index{sentiment} of gratitude than it had ever raised before; she
remembered its warmth, and softened its impropriety\index{impropriety} of
expression.

When all of the house that was open to general inspection had
been seen, they returned downstairs, and, taking leave of the
housekeeper, were consigned\index{consigned} over to the gardener, who met
them at the hall-door.

As they walked across the hall towards the river, Elizabeth
turned back to look again; her uncle and aunt stopped also, and
while the former was conjecturing as to the date of the building,
the owner of it himself suddenly came forward\index{forward} from the road,
which led behind it to the stables.

They were within twenty yards of each other, and so abrupt was
his appearance, that it was impossible to avoid his sight.  Their
eyes instantly met, and the cheeks of both were overspread with
the deepest blush.  He absolutely started, and for a moment
seemed immovable from surprise; but shortly recovering himself,
advanced towards the party, and spoke to Elizabeth, if not in
terms of perfect composure\index{composure}, at least of perfect civility\index{civility}.

She had instinctively turned away; but stopping on his approach,
received his compliments with an embarrassment impossible to
be overcome.  Had his first appearance, or his resemblance to
the picture they had just been examining, been insufficient
to assure the other two that they now saw Mr. Darcy, the
gardener's expression of surprise, on beholding his master,
must immediately have told it.  They stood a little aloof\index{aloof} while
he was talking to their niece, who, astonished and confused,
scarcely dared lift her eyes to his face, and knew not what
answer she returned to his civil inquiries after her family.
Amazed at the alteration of his manner since they last parted,
every sentence that he uttered was increasing her embarrassment;
and every idea of the impropriety\index{impropriety} of her being found there
recurring to her mind, the few minutes\index{minutes} in which they continued
were some of the most uncomfortable in her life.  Nor did he
seem much more at ease; when he spoke, his accent had none of
its usual \textbf{sedateness}\index{sedateness}\stepcounter{footnote}\footnotetext{
\textbf{sedate} \emph{tr.v.} \textbf{-ness}  (suh-DAYT)
 characterized by dignity, propriety, and calm.}; and he repeated his inquiries as to the
time of her having left Longbourn, and of her having stayed in
Derbyshire, so often, and in so hurried a way, as plainly spoke
the distraction of his thoughts.

At length every idea seemed to fail him; and, after standing a
few moments without saying a word, he suddenly recollected
himself, and took leave.

The others then joined her, and expressed admiration of his
figure; but Elizabeth heard not a word, and wholly\index{wholly} engrossed\index{engrossed}
by her own feelings, followed them in silence.  She was
overpowered by shame and vexation\index{vexation}.  Her coming there was
the most unfortunate, the most ill-judged thing in the world!
How strange it must appear to him!  In what a disgraceful light
might it not strike so vain a man!  It might seem as if she
had purposely thrown herself in his way again!  Oh! why did she
come?  Or, why did he thus come a day before he was expected?
Had they been only ten minutes\index{minutes} sooner, they should have been
beyond the reach of his \textbf{discrimination}\index{discrimination}\stepcounter{footnote}\footnotetext{
\textbf{discrimination} \emph{n.}  (dihs-krih-mih-NAY-shun)
}; for it was plain that
he was that moment arrived--that moment alighted from his
horse or his carriage.  She blushed again and again over the
perverseness of the meeting.  And his behaviour, so strikingly
altered--what could it mean?  That he should even speak to her
was amazing!--but to speak with such civility\index{civility}, to inquire after
her family!  Never in her life had she seen his manners so little
dignified, never had he spoken with such gentleness as on this
unexpected meeting.  What a contrast did it offer to his last
address\index{address} in Rosings Park, when he put his letter into her hand!
She  knew not what to think, or how to account for it.

They had now entered a beautiful walk by the side of the water,
and every step was bringing forward\index{forward} a nobler fall of ground, or
a finer reach of the woods to which they were approaching; but
it was some time before Elizabeth was sensible of any of it;
and, though she answered mechanically to the repeated appeals
of her uncle and aunt, and seemed to direct her eyes to such
objects as they pointed out, she distinguished no part of
the scene.  Her thoughts were all fixed on that one spot of
Pemberley House, whichever it might be, where Mr. Darcy then
was.  She longed to know what at the moment was passing in
his mind--in what manner he thought of her, and whether, in
defiance\index{defiance} of everything, she was still dear to him.  Perhaps he
had been civil only because he felt himself at ease; yet there
had been \emph{that} in his voice which was not like ease.  Whether he
had felt more of pain or of pleasure in seeing her she could
not tell, but he certainly had not seen her with composure\index{composure}.

At length, however, the remarks of her companions on her
absence of mind aroused her, and she felt the necessity of
appearing more like herself.

They entered the woods, and bidding adieu\index{adieu} to the river for a
while, ascended some of the higher grounds; when, in spots where
the opening of the trees gave the eye power to wander, were many
charming views of the valley, the opposite hills, with the long
range of woods overspreading many, and occasionally part of the
stream.  Mr. Gardiner expressed a wish of going round the whole
park, but feared it might be beyond a walk.  With a triumphant\index{triumphant}
smile they were told that it was ten miles round.  It settled the
matter; and they pursued the accustomed \textbf{circuit}\index{circuit}\stepcounter{footnote}\footnotetext{
\textbf{} \emph{}  ()
}; which brought
them again, after some time, in a descent\index{descent} among hanging woods,
to the edge of the water, and one of its narrowest parts.  They
crossed it by a simple bridge, in character with the general air
of the scene; it was a spot less adorned than any they had yet
visited; and the valley, here contracted into a \textbf{glen}\index{glen}\stepcounter{footnote}\footnotetext{
\textbf{glen} \emph{n.}  (GLEHN)
 A narrow, secluded valley (in the mountains).}, allowed
room only for the stream, and a narrow walk amidst the rough
\textbf{coppice}\index{coppice}\stepcounter{footnote}\footnotetext{
\textbf{coppice} \emph{n.}  (KAHP-pihs)
a dense growth of bushes.}-wood which bordered it.  Elizabeth longed to explore its
windings; but when they had crossed the bridge, and perceived\index{perceived}
their distance from the house, Mrs. Gardiner, who was not a
great walker, could go no farther, and thought only of returning
to the carriage as quickly as possible.  Her niece was, therefore,
obliged to submit, and they took their way towards the house on
the opposite side of the river, in the nearest direction; but
their progress was slow, for Mr. Gardiner, though seldom able to
indulge the taste, was very fond of fishing, and was so much
engaged in watching the occasional appearance of some trout in
the water, and talking to the man about them, that he advanced
but little.  Whilst wandering on in this slow manner, they were
again surprised, and Elizabeth's astonishment was quite equal to
what it had been at first, by the sight of Mr. Darcy approaching
them, and at no great distance.  The walk here being here less
sheltered than on the other side, allowed them to see him before
they met.  Elizabeth, however astonished, was at least more
prepared for an interview than before, and resolved to appear
and to speak with calmness, if he really intended to meet them.
For a few moments, indeed, she felt that he would probably strike
into some other path.  The idea lasted while a turning in the
walk concealed him from their view; the turning past, he was
immediately before them.  With a glance, she saw that he had lost
none of his recent civility\index{civility}; and, to imitate his politeness, she
began, as they met, to admire the beauty of the place; but she
had not got beyond the words "delightful," and "charming," when
some unlucky recollections\index{recollections} obtruded\index{obtruded}, and she fancied that praise
of Pemberley from her might be mischievously \textbf{construed}\index{construed}\stepcounter{footnote}\footnotetext{
\textbf{construe} \emph{v.} \textbf{-ing}  (kuhn-STROO)
to make sense of; to interpret.}.  Her
colour changed, and she said no more.

Mrs. Gardiner was standing a little behind; and on her pausing,
he asked her if she would do him the honour of introducing him
to her friends.  This was a stroke of civility\index{civility} for which she
was quite unprepared; and she could hardly suppress\index{suppress} a smile at
his being now seeking the acquaintance of some of those very
people against whom his pride had revolted in his offer to
herself.  "What will be his surprise," thought she, "when he
knows who they are?  He takes them now for people of fashion\index{fashion}."

The introduction, however, was immediately made; and as she
named their relationship to herself, she stole a sly look at
him, to see how he bore it, and was not without the expectation
of his \textbf{decamping}\index{decamping}\stepcounter{footnote}\footnotetext{
\textbf{decamp} \emph{intr.v.} \textbf{-ing}  (dih-KAMP)
To leave or depart suddenly or secretly.} as fast as he could from such disgraceful
companions.  That he was \emph{surprised} by the connection was
evident; he sustained it, however, with \textbf{fortitude}\index{fortitude}\stepcounter{footnote}\footnotetext{
\textbf{fortitude} \emph{n.}  (FOR-tih-tood)
strength of mind that enables one to endure adversity with courage.}, and so far
from going away, turned his back with them, and entered into
conversation with Mr. Gardiner.  Elizabeth could not but be
pleased, could not but triumph\index{triumph}.  It was consoling that he should
know she had some relations for whom there was no need to
blush.  She listened most attentively to all that passed between
them, and gloried in every expression, every sentence of her
uncle, which marked his intelligence, his taste, or his good
manners.

The conversation soon turned upon fishing; and she heard Mr.
Darcy invite him, with the greatest civility\index{civility}, to fish there as often
as he chose while he continued in the neighbourhood, offering
at the same time to supply him with fishing tackle, and pointing
out those parts of the stream where there was usually most
sport.  Mrs. Gardiner, who was walking arm-in-arm with
Elizabeth, gave her a look expressive of wonder.  Elizabeth
said nothing, but it gratified her exceedingly; the compliment
must be all for herself.  Her astonishment, however, was
extreme, and continually was she repeating, "Why is he so
altered?  From what can it proceed?  It cannot be for \emph{me}--it
cannot be for \emph{my} sake that his manners are thus softened.  My
reproofs\index{reproofs} at Hunsford could not work such a change as this.
It is impossible that he should still love me."

After walking some time in this way, the two ladies in front,
the two gentlemen behind, on resuming their places, after
descending to the brink of the river for the better inspection of
some curious water-plant, there chanced to be a little alteration.
It originated in Mrs. Gardiner, who, fatigued by the exercise of
the morning, found Elizabeth's arm inadequate to her support, and
consequently preferred her husband's.  Mr. Darcy took her place
by her niece, and they walked on together.  After a short silence,
the lady first spoke.  She wished him to know that she had
been assured of his absence before she came to the place, and
accordingly began by observing, that his arrival had been very
unexpected--"for your housekeeper," she added, "informed us that
you would certainly not be here till to-morrow; and indeed, before
we left Bakewell, we understood that you were not immediately
expected in the country."  He acknowledged the truth of it all,
and said that business with his steward\index{steward} had occasioned his coming
forward\index{forward} a few hours before the rest of the party with whom he
had been travelling.  "They will join me early to-morrow," he
continued, "and among them are some who will claim an acquaintance
with you--Mr. Bingley and his sisters."

Elizabeth answered only by a slight bow.  Her thoughts were
instantly driven back to the time when Mr. Bingley's name had
been the last mentioned between them; and, if she might judge
by his complexion, \emph{his} mind was not very differently engaged.

"There is also one other person in the party," he continued after
a pause, "who more particularly wishes to be known to you.
Will you allow me, or do I ask too much, to introduce my sister
to your acquaintance during your stay at Lambton?"

The surprise of such an application\index{application} was great indeed; it was too
great for her to know in what manner she \textbf{acceded}\index{acceded}\stepcounter{footnote}\footnotetext{
\textbf{accede} \emph{intr.v.} \textbf{-ing}  (ak-SEED)
} to it.  She
immediately felt that whatever desire Miss Darcy might have of
being acquainted with her must be the work of her brother, and,
without looking farther, it was satisfactory; it was gratifying to
know that his resentment\index{resentment} had not made him think really ill of her.

They now walked on in silence, each of them deep in thought.
Elizabeth was not comfortable; that was impossible; but she was
flattered and pleased.  His wish of introducing his sister to her
was a compliment of the highest kind.  They soon outstripped the
others, and when they had reached the carriage, Mr. and Mrs.
Gardiner were half a quarter of a mile behind.

He then asked her to walk into the house--but she declared
herself not tired, and they stood together on the lawn.  At
such a time much might have been said, and silence was very
awkward.  She wanted to talk, but there seemed to be an
\textbf{embargo}\index{embargo}\stepcounter{footnote}\footnotetext{
\textbf{embargo} \emph{n.}  (ehm-BAR-goh)
 a government order imposing a trade barrier.} on every subject\index{subject}.  At last she recollected that she had
been travelling, and they talked of Matlock and Dove Dale with
great perseverance\index{perseverance}.  Yet time and her aunt moved slowly--and
her patience and her ideas were nearly worn out before the
tete-a-tete was over.  On Mr. and Mrs. Gardiner's coming up
they were all pressed to go into the house and take some
refreshment; but this was declined, and they parted on each
side with utmost politeness.  Mr. Darcy handed the ladies into
the carriage; and when it drove off, Elizabeth saw him walking
slowly towards the house.

The observations of her uncle and aunt now began; and each of
them pronounced him to be infinitely superior\index{superior} to anything they
had expected.  "He is perfectly well behaved, polite, and
unassuming\index{unassuming}," said her uncle.

"There \emph{is} something a little stately\index{stately} in him, to be sure," replied
her aunt, "but it is confined to his air, and is not unbecoming.
I can now say with the housekeeper, that though some people may
call him proud, I have seen nothing of it."

"I was never more surprised than by his behaviour to us.  It was
more than civil; it was really attentive; and there was no
necessity for such attention.  His acquaintance with Elizabeth
was very trifling\index{trifling}."

"To be sure, Lizzy," said her aunt, "he is not so handsome as
Wickham; or, rather, he has not Wickham's countenance\index{countenance}, for
his features are perfectly good.  But how came you to tell me
that he was so disagreeable?"

Elizabeth excused herself as well as she could; said that she had
liked him better when they had met in Kent than before, and that
she had never seen him so pleasant as this morning.

"But perhaps he may be a little \textbf{whimsical}\index{whimsical}\stepcounter{footnote}\footnotetext{
\textbf{whimsical} \emph{adj.}  (HWIM-zih-kl)
determined by impulse or whim rather than necessity or reason.} in his civilities,"
replied her uncle.  "Your great men often are; and therefore I
shall not take him at his word, as he might change his mind
another day, and warn me off his grounds."

Elizabeth felt that they had entirely misunderstood his character,
but said nothing.

"From what we have seen of him," continued Mrs. Gardiner, "I
really should not have thought that he could have behaved in so
cruel a way by anybody as he has done by poor Wickham.  He
has not an ill-natured look.  On the contrary, there is something
pleasing about his mouth when he speaks.  And there is something
of dignity in his countenance\index{countenance} that would not give one an
unfavourable idea of his heart.  But, to be sure, the good lady
who showed us his house did give him a most flaming character!
I could hardly help laughing aloud sometimes.  But he is a
liberal master, I suppose, and \emph{that} in the eye of a servant
comprehends\index{comprehends} every virtue\index{virtue}."

Elizabeth here felt herself called on to say something in
vindication\index{vindication} of his behaviour to Wickham; and therefore gave
them to understand, in as guarded a manner as she could, that
by what she had heard from his relations in Kent, his actions
were capable of a very different construction; and that his
character was by no means so faulty, nor Wickham's so amiable\index{amiable},
as they had been considered in Hertfordshire.  In confirmation
of this, she related the particulars of all the pecuniary\index{pecuniary}
transactions in which they had been connected, without actually
naming her authority, but stating it to be such as such as might
be relied on.

Mrs. Gardiner was surprised and concerned; but as they were
now approaching the scene of her former pleasures, every idea
gave way to the charm of recollection\index{recollection}; and she was too much
engaged in pointing out to her husband all the interesting spots
in its \textbf{environs}\index{environs}\stepcounter{footnote}\footnotetext{
\textbf{environ} \emph{tr.v.}  (ehn-VYR-ern)
 the area in which something exists or lives, especially in a city.} to think of anything else.  Fatigued as she had
been by the morning's walk they had no sooner dined than she
set off again in quest of her former acquaintance, and the
evening was spent in the satisfactions of a intercourse\index{intercourse} renewed
after many years' discontinuance.

The occurrences of the day were too full of interest to leave
Elizabeth much attention for any of these new friends; and she
could do nothing but think, and think with wonder, of Mr.
Darcy's civility\index{civility}, and, above all, of his wishing her to be
acquainted with his sister.



\chapter{44}


Elizabeth had settled it that Mr. Darcy would bring his sister
to visit her the very day after her reaching Pemberley; and was
consequently resolved not to be out of sight of the inn the whole
of that morning.  But her conclusion was false; for on the very
morning after their arrival at Lambton, these visitors came.
They had been walking about the place with some of their new
friends, and were just returning to the inn to dress themselves
for dining with the same family, when the sound of a carriage
drew them to a window, and they saw a gentleman and a lady
in a \textbf{curricle}\index{curricle}\stepcounter{footnote}\footnotetext{
\textbf{curricle} \emph{n.}  (KER-rih-kuhl)
An open, two-wheeled carriage, drawn by two horses side by side.} driving up the street.  Elizabeth immediately
recognizing the \textbf{livery}\index{livery}\stepcounter{footnote}\footnotetext{
\textbf{livery} \emph{n.}  (LIHV-er-ee)
a uniform worn by a menservant or chauffeurs.}, guessed what it meant, and imparted no
small degree of her surprise to her relations by acquainting
them with the honour which she expected.  Her uncle and aunt
were all amazement; and the embarrassment of her manner as
she spoke, joined to the circumstance itself, and many of the
circumstances of the preceding day, opened to them a new idea
on the business.  Nothing had ever suggested it before, but they
felt that there was no other way of accounting for such attentions
from such a quarter than by supposing a partiality\index{partiality} for their
niece.  While these newly-born notions were passing in their heads,
the perturbation\index{perturbation} of Elizabeth's feelings was at every moment
increasing.  She was quite amazed at her own discomposure; but
amongst other causes of disquiet, she dreaded lest the partiality\index{partiality}
of the brother should have said too much in her favour; and, more
than commonly anxious to please, she naturally suspected that
every power of pleasing would fail her.

She retreated from the window, fearful of being seen; and as
she walked up and down the room, endeavouring\index{endeavouring} to compose
herself, saw such looks of inquiring surprise in her uncle and
aunt as made everything worse.

Miss Darcy and her brother appeared, and this formidable\index{formidable}
introduction took place.  With astonishment did Elizabeth see
that her new acquaintance was at least as much embarrassed as
herself.  Since her being at Lambton, she had heard that Miss
Darcy was exceedingly proud; but the observation of a very few
minutes\index{minutes} convinced her that she was only exceedingly shy.  She
found it difficult to obtain even a word from her beyond a
\textbf{monosyllable}\index{monosyllable}\stepcounter{footnote}\footnotetext{
\textbf{monosyllable} \emph{n.}  (MAH-noh-sih-luh-bl)
}.

Miss Darcy was tall, and on a larger scale than Elizabeth;
and, though little more than sixteen, her figure was formed,
and her appearance womanly and graceful.  She was less handsome
than her brother; but there was sense and good humour in her
face, and her manners were perfectly unassuming\index{unassuming} and gentle.
Elizabeth, who had expected to find in her as acute\index{acute} and
unembarrassed an observer as ever Mr. Darcy had been, was
much relieved by discerning such different feelings.

They had not long been together before Mr. Darcy told her that
Bingley was also coming to wait on her; and she had barely time
to express her satisfaction, and prepare for such a visitor, when
Bingley's quick step was heard on the stairs, and in a moment he
entered the room.  All Elizabeth's anger against him had been
long done away; but had she still felt any, it could hardly have
stood its ground against the unaffected\index{unaffected} cordiality with which he
expressed himself on seeing her again.  He inquired in a friendly,
though general way, after her family, and looked and spoke with
the same good-humoured ease that he had ever done.

To Mr. and Mrs. Gardiner he was scarcely a less interesting
personage than to herself.  They had long wished to see him.
The whole party before them, indeed, excited a lively attention.
The suspicions which had just arisen of Mr. Darcy and their niece
directed their observation towards each with an earnest\index{earnest} though
guarded inquiry; and they soon drew from those inquiries the full
conviction\index{conviction} that one of them at least knew what it was to love.
Of the lady's sensations they remained a little in doubt; but that
the gentleman was overflowing with admiration was evident enough.

Elizabeth, on her side, had much to do.  She wanted to
ascertain\index{ascertain} the feelings of each of her visitors; she wanted to
compose her own, and to make herself agreeable to all; and in
the latter object, where she feared most to fail, she was most
sure of success, for those to whom she endeavoured\index{endeavoured} to give
pleasure were prepossessed in her favour.  Bingley was ready,
Georgiana was eager, and Darcy determined, to be pleased.

In seeing Bingley, her thoughts naturally flew to her sister;
and, oh! how ardently\index{ardently} did she long to know whether any of his were
directed in a like manner.  Sometimes she could fancy that he
talked less than on former occasions, and once or twice pleased
herself with the notion that, as he looked at her, he was trying
to trace a resemblance.  But, though this might be imaginary,
she could not be deceived as to his behaviour to Miss Darcy, who
had been set up as a rival to Jane.  No look appeared on either
side that spoke particular regard.  Nothing occurred between
them that could justify the hopes of his sister.  On this point
she was soon satisfied; and two or three little circumstances
occurred ere\index{ere} they parted, which, in her anxious interpretation,
denoted a recollection\index{recollection} of Jane not \textbf{untinctured}\index{untinctured}\stepcounter{footnote}\footnotetext{
\textbf{} \emph{}  ()
} by tenderness,
and a wish of saying more that might lead to the mention of her,
had he dared.  He observed to her, at a moment when the others
were talking together, and in a tone which had something of real
regret, that it "was a very long time since he had had the
pleasure of seeing her;" and, before she could reply, he added,
"It is above eight months.  We have not met since the 26th of
November, when we were all dancing together at Netherfield."

Elizabeth was pleased to find his memory so exact; and he
afterwards took occasion to ask her, when unattended to by
any of the rest, whether \emph{all} her sisters were at Longbourn.
There was not much in the question, nor in the preceding
remark; but there was a look and a manner which gave them
meaning.

It was not often that she could turn her eyes on Mr. Darcy
himself; but, whenever she did catch a glimpse, she saw an
expression of general complaisance\index{complaisance}, and in all that he said
she heard an accent so removed from hauteur\index{hauteur} or disdain of his
companions, as convinced her that the improvement of manners
which she had yesterday witnessed however temporary its
existence might prove, had at least outlived one day.  When she
saw him thus seeking the acquaintance and courting the good
opinion of people with whom any intercourse\index{intercourse} a few months ago
would have been a disgrace--when she saw him thus civil, not
only to herself, but to the very relations whom he had openly
disdained, and recollected their last lively scene in Hunsford
Parsonage--the difference, the change was so great, and struck
so forcibly on her mind, that she could hardly restrain her
astonishment from being visible.  Never, even in the company
of his dear friends at Netherfield, or his dignified relations
at Rosings, had she seen him so desirous to please, so free
from self-consequence\index{consequence} or unbending reserve, as now, when no
importance could result from the success of his endeavours\index{endeavours}, and
when even the acquaintance of those to whom his attentions
were addressed\index{addressed} would draw down the ridicule and censure\index{censure} of
the ladies both of Netherfield as Rosings.

Their visitors stayed with them above half-an-hour; and when
they arose to depart, Mr. Darcy called on his sister to join him
in expressing their wish of seeing Mr. and Mrs. Gardiner, and
Miss Bennet, to dinner at Pemberley, before they left the
country.  Miss Darcy, though with a diffidence\index{diffidence} which marked
her little in the habit of giving invitations, readily\index{readily} obeyed.
Mrs. Gardiner looked at her niece, desirous of knowing how \emph{she},
whom the invitation most concerned, felt disposed\index{disposed} as to its
acceptance, but Elizabeth had turned away her head.  Presuming
however, that this studied avoidance spoke rather a momentary
embarrassment than any dislike of the proposal, and seeing in
her husband, who was fond of society, a perfect willingness to
accept it, she ventured to engage for her attendance, and the
day after the next was fixed on.

Bingley expressed great pleasure in the certainty of seeing
Elizabeth again, having still a great deal to say to her, and many
inquiries to make after all their Hertfordshire friends.  Elizabeth,
construing all this into a wish of hearing her speak of her sister,
was pleased, and on this account, as well as some others, found
herself, when their visitors left them, capable of considering
the last half-hour with some satisfaction, though while it was
passing, the enjoyment of it had been little.  Eager to be alone,
and fearful of inquiries or hints from her uncle and aunt, she
stayed with them only long enough to hear their favourable
opinion of Bingley, and then hurried away to dress.

But she had no reason to fear Mr. and Mrs. Gardiner's curiosity;
it was not their wish to force her communication.  It was evident
that she was much better acquainted with Mr. Darcy than they
had before any idea of; it was evident that he was very much in
love with her.  They saw much to interest, but nothing to justify
inquiry.

Of Mr. Darcy it was now a matter of anxiety to think well; and,
as far as their acquaintance reached, there was no fault to find.
They could not be untouched by his politeness; and had they
drawn his character from their own feelings and his servant's
report, without any reference to any other account, the circle
in Hertfordshire to which he was known would not have recognized
it for Mr. Darcy.  There was now an interest, however, in
believing the housekeeper; and they soon became sensible that
the authority of a servant who had known him since he was four
years old, and whose own manners indicated respectability, was
not to be hastily rejected.  Neither had anything occurred in
the intelligence of their Lambton friends that could materially
lessen its weight.  They had nothing to accuse him of but pride;
pride he probably had, and if not, it would certainly be imputed\index{imputed}
by the inhabitants of a small market-town where the family did
not visit.  It was acknowledged, however, that he was a liberal
man, and did much good among the poor.

With respect to Wickham, the travellers soon found that he was
not held there in much estimation; for though the chief of his
concerns with the son of his patron were imperfectly understood,
it was yet a well-known fact that, on his quitting Derbyshire,
he had left many debts behind him, which Mr. Darcy afterwards
discharged.

As for Elizabeth, her thoughts were at Pemberley this evening
more than the last; and the evening, though as it passed it
seemed long, was not long enough to determine her feelings
towards \emph{one} in that mansion; and she lay awake two whole
hours endeavouring\index{endeavouring} to make them out.  She certainly did not
hate him.  No; hatred had vanished long ago, and she had
almost as long been ashamed of ever feeling a dislike against
him, that could be so called.  The respect created by the
conviction\index{conviction} of his valuable qualities, though at first unwillingly
admitted, had for some time ceased to be \textbf{repugnant}\index{repugnant}\stepcounter{footnote}\footnotetext{
\textbf{repugnant} \emph{adj.}  (rih-PUHG-nuhnt)
offensive to the mind; repulsive.} to her
feeling; and it was now heightened into somewhat of a friendlier
nature, by the testimony so highly in his favour, and bringing
forward\index{forward} his disposition\index{disposition} in so amiable\index{amiable} a light, which yesterday
had produced.  But above all, above respect and esteem, there
was  a motive within her of goodwill which could not be
overlooked.  It was gratitude; gratitude, not merely for having
once loved her, but for loving her still well enough to forgive
all the \textbf{petulance}\index{petulance}\stepcounter{footnote}\footnotetext{
\textbf{petulant} \emph{adj.}  (PEHT-yoo-luhnt)
Overly irritable or peevish; easily annoyed.} and \textbf{acrimony}\index{acrimony}\stepcounter{footnote}\footnotetext{
\textbf{acrimony} \emph{n.}  (AK-rih-moa-nee)
sharp and bitter speech or behavior.} of her manner in rejecting him, and
all the unjust accusations accompanying her rejection.  He who,
she had been persuaded, would avoid her as his greatest enemy,
seemed, on this accidental meeting, most eager to preserve the
acquaintance, and without any indelicate display of regard, or
any peculiarity of manner, where their two selves only were
concerned, was soliciting\index{soliciting} the good opinion of her friends,
and bent on making her known to his sister.  Such a change
in a man of so much pride exciting not only astonishment but
gratitude--for to love, ardent\index{ardent} love, it must be attributed; and
as such its impression on her was of a sort to be encouraged, as
by no means unpleasing, though it could not be exactly defined.
She respected, she esteemed, she was grateful to him, she felt
a real interest in his welfare; and she only wanted to know how
far she wished that welfare to depend upon herself, and how far
it would be for the happiness of both that she should employ the
power, which her fancy told her she still possessed, of bringing
on her the renewal of his addresses.

It had been settled in the evening between the aunt and the
niece, that such a striking civility\index{civility} as Miss Darcy's in coming to
see them on the very day of her arrival at Pemberley, for she
had reached it only to a late breakfast, ought to be imitated,
though it could not be equalled, by some exertion\index{exertion} of politeness
on their side; and, consequently, that it would be highly
expedient to wait on her at Pemberley the following morning.
They were, therefore, to go.  Elizabeth was pleased; though
when she asked herself the reason, she had very little to say in
reply.

Mr. Gardiner left them soon after breakfast.  The fishing scheme
had been renewed the day before, and a positive engagement
made of his meeting some of the gentlemen at Pemberley before
noon.



\chapter{45}


Convinced as Elizabeth now was that Miss Bingley's dislike of
her had originated in jealousy, she could not help feeling how
unwelcome her appearance at Pemberley must be to her, and
was curious to know with how much civility\index{civility} on that lady's side
the acquaintance would now be renewed.

On reaching the house, they were shown through the hall into
the saloon, whose northern aspect\index{aspect} rendered\index{rendered} it delightful for
summer.  Its windows opening to the ground, admitted a most
refreshing view of the high woody hills behind the house,
and of the beautiful oaks and Spanish chestnuts which were
scattered over the intermediate lawn.

In this house they were received by Miss Darcy, who was
sitting there with Mrs. Hurst and Miss Bingley, and the lady
with whom she lived in London.  Georgiana's reception of them
was very civil, but attended with all the embarrassment which,
though proceeding from shyness and the fear of doing wrong,
would easily give to those who felt themselves inferior\index{inferior} the
belief of her being proud and reserved.  Mrs. Gardiner and her
niece, however, did her justice, and pitied her.

By Mrs. Hurst and Miss Bingley they were noticed only by a
curtsey; and, on their being seated, a pause, awkward as such
pauses must always be, succeeded for a few moments.  It was
first broken by Mrs. Annesley, a genteel\index{genteel}, agreeable-looking
woman, whose endeavour\index{endeavour} to introduce some kind of discourse\index{discourse}
proved her to be more truly well-bred than either of the others;
and between her and Mrs. Gardiner, with occasional help from
Elizabeth, the conversation was carried on.  Miss Darcy looked
as if she wished for courage enough to join in it; and sometimes
did venture a short sentence when there was least danger of its
being heard.

Elizabeth soon saw that she was herself closely watched by Miss
Bingley, and that she could not speak a word, especially to Miss
Darcy, without calling her attention.  This observation would not
have prevented her from trying to talk to the latter, had they
not been seated at an inconvenient distance; but she was not sorry
to be spared\index{spared} the necessity of saying much.  Her own thoughts
were employing her.  She expected every moment that some of the
gentlemen would enter the room.  She wished, she feared that
the master of the house might be amongst them; and whether
she wished or feared it most, she could scarcely determine.
After sitting in this manner a quarter of an hour without hearing
Miss Bingley's voice, Elizabeth was roused by receiving from
her a cold inquiry after the health of her family.  She answered
with equal indifference\index{indifference} and \textbf{brevity}\index{brevity}\stepcounter{footnote}\footnotetext{
\textbf{brevity} \emph{n.}  (BREH-vih-tee)
being short or brief in time; short and concise in expression.}, and the others said no more.

The next variation which their visit afforded was produced by
the entrance of servants with cold meat, cake, and a variety of
all the finest fruits in season; but this did not take place till
after many a significant look and smile from Mrs. Annesley to
Miss Darcy had been given, to remind her of her post\index{post}.  There was
now employment for the whole party--for though they could not all
talk, they could all eat; and the beautiful pyramids of grapes,
nectarines, and peaches soon collected them round the table.

While thus engaged, Elizabeth had a fair opportunity of deciding
whether she most feared or wished for the appearance of Mr.
Darcy, by the feelings which prevailed\index{prevailed} on his entering the room;
and then, though but a moment before she had believed her
wishes to predominate, she began to regret that he came.

He had been some time with Mr. Gardiner, who, with two or
three other gentlemen from the house, was engaged by the river,
and had left him only on learning that the ladies of the family
intended a visit to Georgiana that morning.  No sooner did he
appear than Elizabeth wisely resolved to be perfectly easy and
unembarrassed; a resolution\index{resolution} the more necessary to be made, but
perhaps not the more easily kept, because she saw that the
suspicions of the whole party were awakened against them, and
that there was scarcely an eye which did not watch his
behaviour when he first came into the room.  In no countenance\index{countenance}
was attentive curiosity so strongly marked as in Miss Bingley's,
in spite of the smiles which overspread her face whenever she
spoke to one of its objects; for jealousy had not yet made her
desperate, and her attentions to Mr. Darcy were by no means
over.  Miss Darcy, on her brother's entrance, exerted herself
much more to talk, and Elizabeth saw that he was anxious for
his sister and herself to get acquainted, and forwarded\index{forwarded} as much
as possible, every attempt at conversation on either side.  Miss
Bingley saw all this likewise; and, in the imprudence\index{imprudence} of anger,
took the first opportunity of saying, with sneering civility\index{civility}:

"Pray, Miss Eliza, are not the ----shire Militia\index{Militia} removed from
Meryton?  They must be a great loss to \emph{your} family."

In Darcy's presence she dared not mention Wickham's name;
but Elizabeth instantly comprehended\index{comprehended} that he was uppermost in
her thoughts; and the various recollections\index{recollections} connected with him
gave her a moment's distress; but exerting herself vigorously to
\textbf{repel}\index{repel}\stepcounter{footnote}\footnotetext{
\textbf{repel} \emph{v.}  (rih-PEHL)
 to force or drive back; to ward off.} the ill-natured attack, she presently answered the question
in a tolerably detached tone.  While she spoke, an \textbf{involuntary}\index{involuntary}\stepcounter{footnote}\footnotetext{
\textbf{involuntary} \emph{adj.} \textbf{-ly}  (ihn-VAW-luhn-ter-ree)
done against or without one's will.}
glance showed her Darcy, with a heightened complexion,
earnestly\index{earnestly} looking at her, and his sister overcome with confusion,
and unable to lift up her eyes.  Had Miss Bingley known what
pain she was then giving her beloved friend, she undoubtedly
would have refrained from the hint; but she had merely intended
to discompose\index{discompose} Elizabeth by bringing forward\index{forward} the idea of a man
to whom she believed her partial\index{partial}, to make her betray a sensibility
which might injure her in Darcy's opinion, and, perhaps, to
remind the latter of all the follies and absurdities by which
some part of her family were connected with that corps\index{corps}.  Not a
syllable had ever reached her of Miss Darcy's meditated\index{meditated}
elopement.  To no creature had it been revealed, where secrecy
was possible, except to Elizabeth; and from all Bingley's
connections her brother was particularly anxious to conceal it,
from the very wish which Elizabeth had long ago attributed to
him, of their becoming hereafter her own.  He had certainly
formed such a plan, and without meaning that it should effect
his endeavour\index{endeavour} to separate him from Miss Bennet, it is probable\index{probable}
that it might add something to his lively concern for the welfare
of his friend.

Elizabeth's collected behaviour, however, soon quieted his
emotion; and as Miss Bingley, vexed\index{vexed} and disappointed, dared
not approach nearer to Wickham, Georgiana also recovered in
time, though not enough to be able to speak any more.  Her
brother, whose eye she feared to meet, scarcely recollected her
interest in the affair, and the very circumstance which had
been designed to turn his thoughts from Elizabeth seemed to
have fixed them on her more and more cheerfully.

Their visit did not continue long after the question and answer
above mentioned; and while Mr. Darcy was attending them to
their carriage Miss Bingley was venting her feelings in criticisms
on Elizabeth's person, behaviour, and dress.  But Georgiana
would not join her.  Her brother's recommendation was enough
to ensure her favour; his judgement could not err.  And he had
spoken in such terms of Elizabeth as to leave Georgiana without
the power of finding her otherwise than lovely and amiable\index{amiable}.
When Darcy returned to the saloon, Miss Bingley could not help
repeating to him some part of what she had been saying to his
sister.

"How very ill Miss Eliza Bennet looks this morning, Mr. Darcy,"
she cried; "I never in my life saw anyone so much altered as she
is since the winter.  She is grown so brown and coarse!  Louisa
and I were agreeing that we should not have known her again."

However little Mr. Darcy might have liked such an address\index{address}, he
contented himself with coolly replying that he perceived\index{perceived} no
other alteration than her being rather tanned, no miraculous
consequence\index{consequence} of travelling in the summer.

"For my own part," she rejoined, "I must confess that I never
could see any beauty in her.  Her face is too thin; her
complexion has no brilliancy; and her features are not at all
handsome.  Her nose wants character--there is nothing marked
in its lines.  Her teeth are tolerable\index{tolerable}, but not out of the common
way; and as for her eyes, which have sometimes been called so
fine, I could never see anything extraordinary in them.  They
have a sharp, shrewish look, which I do not like at all; and in her
air altogether there is a self-sufficiency without fashion\index{fashion}, which
is intolerable\index{intolerable}."

Persuaded as Miss Bingley was that Darcy admired Elizabeth,
this was not the best method of recommending herself; but
angry people are not always wise; and in seeing him at last look
somewhat \textbf{nettled}\index{nettled}\stepcounter{footnote}\footnotetext{
\textbf{nettle} \emph{n.} \textbf{-ed}  (NET-tl)
to arouse impatience or anger.}, she had all the success she expected.  He was
resolutely\index{resolutely} silent, however, and, from a determination of making
him speak, she continued:

"I remember, when we first knew her in Hertfordshire, how
amazed we all were to find that she was a reputed\index{reputed} beauty; and I
particularly recollect your saying one night, after they had been
dining at Netherfield, '\emph{she} a beauty!--I should as soon call her
mother a wit.'  But afterwards she seemed to improve on you,
and I believe you thought her rather pretty at one time."

"Yes," replied Darcy, who could contain himself no longer,
"but \emph{that} was only when I first saw her, for it is many months
since I have considered her as one of the handsomest women of
my acquaintance."

He then went away, and Miss Bingley was left to all the
satisfaction of having forced him to say what gave no one any
pain but herself.

Mrs. Gardiner and Elizabeth talked of all that had occurred during
their visit, as they returned, except what had particularly
interested them both.  The look and behaviour of everybody they
had seen were discussed, except of the person who had mostly
engaged their attention.  They talked of his sister, his friends,
his house, his fruit--of everything but himself; yet Elizabeth
was longing to know what Mrs. Gardiner thought of him, and Mrs.
Gardiner would have been highly gratified by her niece's beginning
the subject\index{subject}.



\chapter{46}


Elizabeth had been a good deal disappointed in not finding a
letter from Jane on their first arrival at Lambton; and this
disappointment had been renewed on each of the mornings that
had now been spent there; but on the third her repining was
over, and her sister justified, by the receipt of two letters
from her at once, on one of which was marked that it had been
missent elsewhere.  Elizabeth was not surprised at it, as Jane
had written the direction remarkably ill.

They had just been preparing to walk as the letters came in;
and her uncle and aunt, leaving her to enjoy them in quiet, set
off by themselves.  The one missent must first be attended to;
it had been written five days ago.  The beginning contained an
account of all their little parties and engagements, with such
news as the country afforded; but the latter half, which was
dated a day later, and written in evident agitation, gave more
important intelligence.  It was to this effect:

"Since writing the above, dearest Lizzy, something has occurred
of a most unexpected and serious nature; but I am afraid of
alarming you--be assured that we are all well.  What I have to
say relates to poor Lydia.  An express came at twelve last night,
just as we were all gone to bed, from Colonel Forster, to
inform us that she was gone off to Scotland with one of his
officers; to own the truth, with Wickham!  Imagine our surprise.
To Kitty, however, it does not seem so wholly\index{wholly} unexpected.  I
am very, very sorry.  So imprudent\index{imprudent} a match on both sides! But
I am willing to hope the best, and that his character has been
misunderstood.  Thoughtless and indiscreet I can easily believe
him, but this step (and let\index{let} us rejoice over it) marks nothing bad
at heart.  His choice is disinterested\index{disinterested} at least, for he must know
my father can give her nothing.  Our poor mother is sadly
grieved\index{grieved}.  My father bears it better.  How thankful am I that we
never let\index{let} them know what has been said against him; we must
forget it ourselves.  They were off Saturday night about twelve,
as is conjectured\index{conjectured}, but were not missed till yesterday morning at
eight.  The express was sent off directly.  My dear Lizzy, they
must have passed within ten miles of us.  Colonel Forster gives
us reason to expect him here soon.  Lydia left a few lines for
his wife, informing her of their intention.  I must conclude, for
I cannot be long from my poor mother.  I am afraid you will not
be able to make it out, but I hardly know what I have written."

Without allowing herself time for consideration, and scarcely
knowing what she felt, Elizabeth on finishing this letter instantly
seized the other, and opening it with the utmost impatience, read
as follows: it had been written a day later than the conclusion of
the first.

"By this time, my dearest sister, you have received my hurried
letter; I wish this may be more intelligible, but though not
confined for time, my head is so bewildered that I cannot answer
for being coherent.  Dearest Lizzy, I hardly know what I would
write, but I have bad news for you, and it cannot be delayed.
Imprudent\index{Imprudent} as the marriage between Mr. Wickham and our poor
Lydia would be, we are now anxious to be assured it has taken
place, for there is but too much reason to fear they are not gone
to Scotland.  Colonel Forster came yesterday, having left
Brighton the day before, not many hours after the express.
Though Lydia's short letter to Mrs. F. gave them to understand
that they were going to Gretna Green, something was dropped
by Denny expressing his belief that W. never intended to go
there, or to marry Lydia at all, which was repeated to Colonel
F., who, instantly taking the alarm, set off from B. intending to
trace their route.  He did trace them easily to Clapham, but no
further; for on entering that place, they removed into a \textbf{hackney}\index{hackney}\stepcounter{footnote}\footnotetext{
\textbf{hackney} \emph{n.}  (HAK-nee)
}
coach, and dismissed the chaise\index{chaise} that brought them from Epsom.
All that is known after this is, that they were seen to continue
the London road.  I know not what to think.  After making every
possible inquiry on that side London, Colonel F. came on into
Hertfordshire, anxiously renewing them at all the turnpikes, and
at the inns in Barnet and Hatfield, but without any success--no
such people had been seen to pass through.  With the kindest
concern he came on to Longbourn, and broke his apprehensions\index{apprehensions}
to us in a manner most creditable to his heart.  I am sincerely
grieved\index{grieved} for him and Mrs. F., but no one can throw any blame
on them.  Our distress, my dear Lizzy, is very great.  My father
and mother believe the worst, but I cannot think so ill of him.
Many circumstances might make it more eligible for them to be
married privately\index{privately} in town than to pursue their first plan;
and even if \emph{he} could form such a design against a young woman
of Lydia's connections, which is not likely, can I suppose her
so lost to everything?  Impossible!  I grieve\index{grieve} to find, however,
that Colonel F. is not disposed\index{disposed} to depend upon their marriage;
he shook his head when I expressed my hopes, and said he fear
W. was not a man to be trusted.  My poor mother is really ill,
and keeps her room.  Could she exert herself, it would be better;
but this is not to be expected.  And as to my father, I never in
my life saw him so affected.  Poor Kitty has anger for having
concealed their attachment; but as it was a matter of confidence\index{confidence},
one cannot wonder.  I am truly glad, dearest Lizzy, that you
have been spared\index{spared} something of these distressing scenes; but
now, as the first shock is over, shall I own that I long for
your return?  I am not so selfish, however, as to press for it,
if inconvenient.  Adieu\index{Adieu}!  I take up my pen again to do what I
have just told you I would not; but circumstances are such that
I cannot help earnestly\index{earnestly} begging you all to come here as soon as
possible.  I know my dear uncle and aunt so well, that I am not
afraid of requesting it, though I have still something more to
ask of the former.  My father is going to London with Colonel
Forster instantly, to try to discover her.  What he means to do
I am sure I know not; but his excessive distress will not allow
him to pursue any measure in the best and safest way, and
Colonel Forster is obliged to be at Brighton again to-morrow
evening.  In such and \textbf{exigence}\index{exigence}\stepcounter{footnote}\footnotetext{
\textbf{exigence} \emph{n.}  (EHG-zih-jense)
a pressing situation requiring immediate attention.}, my uncle's advice and assistance
would be everything in the world; he will immediately comprehend\index{comprehend}
what I must feel, and I rely upon his goodness."

"Oh! where, where is my uncle?" cried Elizabeth, darting from
her seat as she finished the letter, in eagerness to follow him,
without losing a moment of the time so precious; but as she
reached the door it was opened by a servant, and Mr. Darcy
appeared.  Her pale\index{pale} face and \textbf{impetuous}\index{impetuous}\stepcounter{footnote}\footnotetext{
\textbf{impetuous} \emph{adj.}  (ihm-PEH-choo-uhs)
undue haste; impulsive and suddenly forceful.} manner made him start,
and before he could recover himself to speak, she, in whose
mind every idea was superseded by Lydia's situation, hastily
exclaimed, "I beg your pardon, but I must leave you.  I must
find Mr. Gardiner this moment, on business that cannot be
delayed; I have not an instant to lose."

"Good God! what is the matter?" cried he, with more feeling
than politeness; then recollecting himself, "I will not detain you
a minute\index{minute}; but let\index{let} me, or let\index{let} the servant go after Mr. and Mrs.
Gardiner.  You are not well enough; you cannot go yourself."

Elizabeth hesitated, but her knees trembled under her and she
felt how little would be gained by her attempting to pursue them.
Calling back the servant, therefore, she commissioned him,
though in so breathless an accent as made her almost unintelligible,
to fetch his master and mistress home instantly.

On his quitting the room she sat down, unable to support
herself, and looking so miserably ill, that it was impossible
for Darcy to leave her, or to refrain from saying, in a tone
of gentleness and commiseration, "Let\index{Let} me call your maid.  Is
there nothing you could take to give you present relief?  A
glass of wine; shall I get you one?  You are very ill."

"No, I thank you," she replied, endeavouring\index{endeavouring} to recover herself.
"There is nothing the matter with me.  I am quite well; I am
only distressed by some dreadful news which I have just
received from Longbourn."

She burst into tears as she alluded\index{alluded} to it, and for a few minutes\index{minutes}
could not speak another word.  Darcy, in wretched\index{wretched} suspense,
could only say something indistinctly of his concern, and
observe her in compassionate silence.  At length she spoke
again.  "I have just had a letter from Jane, with such dreadful
news.  It cannot be concealed from anyone.  My younger sister
has left all her friends--has eloped\index{eloped}; has thrown herself into
the power of--of Mr. Wickham.  They are gone off together from
Brighton.  \emph{you} know him too well to doubt the rest.  She has
no money, no connections, nothing that can tempt him to--she
is lost for ever."

Darcy was fixed in astonishment.  "When I consider," she added
in a yet more agitated voice, "that I might have prevented it!
I, who knew what he was.  Had I but explained some part of
it only--some part of what I learnt, to my own family!  Had his
character been known, this could not have happened.  But it is
all--all too late now."

"I am grieved\index{grieved} indeed," cried Darcy; "grieved\index{grieved}--shocked.  But is
it certain--absolutely certain?"

"Oh, yes!  They left Brighton together on Sunday night, and
were traced almost to London, but not beyond; they are
certainly not gone to Scotland."

"And what has been done, what has been attempted, to recover
her?"

"My father is gone to London, and Jane has written to beg my
uncle's immediate assistance; and we shall be off, I hope, in
half-an-hour.  But nothing can be done--I know very well that
nothing can be done.  How is such a man to be worked on?  How
are they even to be discovered?  I have not the smallest hope.
It is every way horrible!"

Darcy shook his head in silent acquiescence\index{acquiescence}.

"When \emph{my} eyes were opened to his real character--Oh! had I
known what I ought, what I dared to do!  But I knew not--I
was afraid of doing too much.  Wretched\index{Wretched}, wretched\index{wretched} mistake!"

Darcy made no answer.  He seemed scarcely to hear her, and
was walking up and down the room in earnest\index{earnest} meditation, his
brow contracted, his air gloomy.  Elizabeth soon observed, and
instantly understood it.  Her power was sinking; everything
\emph{must} sink under such a proof of family weakness, such an
assurance of the deepest disgrace.  She could neither wonder
nor condemn, but the belief of his self-conquest brought nothing
to her consolatory to her bosom, afforded no \textbf{palliation}\index{palliation}\stepcounter{footnote}\footnotetext{
\textbf{palliate} \emph{tr.v.}  (PAL-lih-ayt)
to lessen the seriousness or extent of; to extenuate; to mitigate.} of her
distress.  It was, on the contrary, exactly calculated to make
her understand her own wishes; and never had she so honestly felt
that she could have loved him, as now, when all love must be vain.

But self, though it would intrude, could not engross\index{engross} her.
Lydia--the humiliation, the misery she was bringing on them all,
soon swallowed up every private\index{private} care; and covering her face with
her handkerchief, Elizabeth was soon lost to everything else;
and, after a pause of several minutes\index{minutes}, was only recalled to a
sense of her situation by the voice of her companion, who, in a
manner which, though it spoke compassion, spoke likewise restraint,
said, "I am afraid you have been long desiring my absence, nor
have I anything to plead in excuse of my stay, but real, though
unavailing\index{unavailing} concern.  Would to Heaven that anything could be
either said or done on my part that might offer consolation\index{consolation} to
such distress!  But I will not torment you with vain wishes, which
may seem purposely to ask for your thanks.  This unfortunate
affair will, I fear, prevent my sister's having the pleasure of
seeing you at Pemberley to-day."

"Oh, yes.  Be so kind as to apologise for us to Miss Darcy.  Say
that urgent business calls us home immediately.  Conceal the
unhappy truth as long as it is possible, I know it cannot be long."

He readily\index{readily} assured her of his secrecy; again expressed his sorrow
for her distress, wished it a happier conclusion than there was
at present reason to hope, and leaving his compliments for her
relations, with only one serious, parting look, went away.

As he quitted the room, Elizabeth felt how improbable\index{improbable} it was
that they should ever see each other again on such terms of
cordiality as had marked their several meetings in Derbyshire;
and as she threw a \textbf{retrospective}\index{retrospective}\stepcounter{footnote}\footnotetext{
\textbf{retrospective} \emph{adj.}  (reh-troh-SPEHK-tihv)
Looking back on or thinking about the past.} glance over the whole of their
acquaintance, so full of contradictions and varieties, sighed
at the perverseness of those feelings which would now have
promoted its continuance, and would formerly have rejoiced in
its termination.

If gratitude and esteem are good foundations of affection,
Elizabeth's change of sentiment\index{sentiment} will be neither improbable\index{improbable} nor
faulty.  But if otherwise--if regard springing from such sources
is unreasonable or unnatural, in comparison of what is so often
described as arising on a first interview with its object, and even
before two words have been exchanged, nothing can be said in
her defence, except that she had given somewhat of a trial to the
latter method in her partiality\index{partiality} for Wickham, and that its ill
success might, perhaps, authorise her to seek the other less
interesting mode\index{mode} of attachment.  Be that as it may, she saw him
go with regret; and in this early example of what Lydia's \textbf{infamy}\index{infamy}\stepcounter{footnote}\footnotetext{
\textbf{infamy} \emph{n.}  (IHN-fuh-mee)
Evil fame or reputation.}
must produce, found additional anguish as she reflected on that
wretched\index{wretched} business.  Never, since reading Jane's second letter,
had she entertained a hope of Wickham's meaning to marry her.
No one but Jane, she thought, could flatter herself with such an
expectation.  Surprise was the least of her feelings on this
development.  While the contents of the first letter remained in
her mind, she was all surprise--all astonishment that Wickham
should marry a girl whom it was impossible he could marry
for money; and how Lydia could ever have attached him had
appeared incomprehensible.  But now it was all too natural.  For
such an attachment as this she might have sufficient charms; and
though she did not suppose Lydia to be deliberately engaging in
an elopement without the intention of marriage, she had no
difficulty in believing that neither her virtue\index{virtue} nor her
understanding would preserve her from falling an easy prey.

She had never perceived\index{perceived}, while the regiment\index{regiment} was in Hertfordshire,
that Lydia had any partiality\index{partiality} for him; but she was convinced that
Lydia wanted only encouragement to attach herself to anybody.
Sometimes one officer, sometimes another, had been her favourite,
as their attentions raised them in her opinion.  Her affections
had continually been fluctuating but never without an object.
The mischief of neglect and mistaken indulgence towards such a
girl--oh! how acutely\index{acutely} did she now feel it!

She was wild to be at home--to hear, to see, to be upon the
spot to share with Jane in the cares that must now fall wholly\index{wholly}
upon her, in a family so \textbf{deranged}\index{deranged}\stepcounter{footnote}\footnotetext{
\textbf{derange} \emph{tr.v.} \textbf{-ed}  (dee-RAYNJ)
to make insane.}, a father absent, a mother
incapable of exertion\index{exertion}, and requiring constant attendance; and
though almost persuaded that nothing could be done for Lydia,
her uncle's interference seemed of the utmost importance, and
till he entered the room her impatience was severe\index{severe}.  Mr. and
Mrs. Gardiner had hurried back in alarm, supposing by the
servant's account that their niece was taken suddenly ill; but
satisfying them instantly on that head, she eagerly communicated
the cause of their summons\index{summons}\index{summons\index{summons}}, reading the two letters aloud, and
dwelling on the postscript of the last with trembling energy,
though Lydia had never been a favourite with them, Mr. and
Mrs. Gardiner could not but be deeply \textbf{afflicted}\index{afflicted}\stepcounter{footnote}\footnotetext{
\textbf{afflict} \emph{tr.v.} \textbf{-ed}  (uh-FLIKT)
 to cause pain or suffering.}.  Not Lydia
only, but all were concerned in it; and after the first
exclamations of surprise and horror, Mr. Gardiner promised
every assistance in his power.  Elizabeth, though expecting no
less, thanked him with tears of gratitude; and all three being
\textbf{actuated}\index{actuated}\stepcounter{footnote}\footnotetext{
\textbf{actuate} \emph{tr.v.} \textbf{-ed}  (AK-choo-ate)
to activate or cause to function.} by one spirit, everything relating to their journey was
speedily settled.  They were to be off as soon as possible.  "But
what is to be done about Pemberley?" cried Mrs. Gardiner.
"John told us Mr. Darcy was here when you sent for us; was it
so?"

"Yes; and I told him we should not be able to keep our
engagement.  \emph{that} is all settled."

"What is all settled?" repeated the other, as she ran into her
room to prepare.  "And are they upon such terms as for her to
disclose the real truth?  Oh, that I knew how it was!"

But wishes were vain, or at least could only serve to amuse her
in the hurry and confusion of the following hour.  Had Elizabeth
been at leisure to be idle, she would have remained certain that
all employment was impossible to one so wretched\index{wretched} as herself;
but she had her share of business as well as her aunt, and
amongst the rest there were notes to be written to all their
friends at Lambton, with false excuses for their sudden
departure.  An hour, however, saw the whole completed; and
Mr. Gardiner meanwhile having settled his account at the inn,
nothing remained to be done but to go; and Elizabeth, after all
the misery of the morning, found herself, in a shorter space of
time than she could have supposed, seated in the carriage, and
on the road to Longbourn.



\chapter{47}


"I have been thinking it over again, Elizabeth," said her uncle,
as they drove from the town; "and really, upon serious
consideration, I am much more inclined than I was to judge as
your eldest sister does on the matter.  It appears to me so very
unlikely that any young man should form such a design against
a girl who is by no means unprotected or friendless, and who
was actually staying in his colonel's family, that I am strongly
inclined to hope the best.  Could he expect that her friends
would not step forward\index{forward}?  Could he expect to be noticed again
by the regiment\index{regiment}, after such an affront\index{affront} to Colonel Forster?  His
temptation is not adequate to the risk!"

"Do you really think so?" cried Elizabeth, brightening up for a
moment.

"Upon my word," said Mrs. Gardiner, "I begin to be of your
uncle's opinion.  It is really too great a violation of decency,
honour, and interest, for him to be guilty of.  I cannot think
so very ill of Wickham.  Can you yourself, Lizzy, so wholly\index{wholly} give
him up, as to believe him capable of it?"

"Not, perhaps, of neglecting his own interest; but of every other
neglect I can believe him capable.  If, indeed, it should be so!
But I dare not hope it.  Why should they not go on to Scotland
if that had been the case?"

"In the first place," replied Mr. Gardiner, "there is no absolute
proof that they are not gone to Scotland."

"Oh! but their removing from the chaise\index{chaise} into a hackney\index{hackney} coach is
such a \textbf{presumption}\index{presumption}\stepcounter{footnote}\footnotetext{
\textbf{presumption} \emph{n.}  (prih-ZUHMP-shun)
audacious (even arrogant) behavior that is offensive.}!  And, besides, no traces of them were to be
found on the Barnet road."

"Well, then--supposing them to be in London.  They may be there,
though for the purpose of concealment, for no more exceptional
purpose.  It is not likely that money should be very abundant on
either side; and it might strike them that they could be more
economically, though less \textbf{expeditiously}\index{expeditiously}\stepcounter{footnote}\footnotetext{
\textbf{expeditious} \emph{adj.} \textbf{-ly}  (ehx-pih-DIH-shus)
in an efficient manner.}, married in London
than in Scotland."

"But why all this secrecy?  Why any fear of detection?  Why must
their marriage be private\index{private}?  Oh, no, no--this is not likely.
His most particular friend, you see by Jane's account, was
persuaded of his never intending to marry her.  Wickham will
never marry a woman without some money.  He cannot afford
it.  And what claims has Lydia--what attraction has she beyond
youth, health, and good humour that could make him, for her
sake, forego every chance of benefiting himself by marrying
well?  As to what restraint the apprehensions\index{apprehensions} of disgrace in the
corps\index{corps} might throw on a dishonourable elopement with her, I am
not able to judge; for I know nothing of the effects that such a
step might produce.  But as to your other objection, I am afraid
it will hardly hold good.  Lydia has no brothers to step forward\index{forward};
and he might imagine, from my father's behaviour, from his
\textbf{indolence}\index{indolence}\stepcounter{footnote}\footnotetext{
\textbf{indolence} \emph{n.}  (IHN-doh-lehnse)
laziness as a lifestyle; sloth.} and the little attention he has ever seemed to give
to what was going forward\index{forward} in his family, that \emph{he} would do as
little, and think as little about it, as any father could do,
in such a matter."

"But can you think that Lydia is so lost to everything but love
of him as to consent to live with him on any terms other than
marriage?"

"It does seem, and it is most shocking indeed," replied Elizabeth,
with tears in her eyes, "that a sister's sense of decency and
virtue\index{virtue} in such a point should admit of doubt.  But, really,
I know not what to say.  Perhaps I am not doing her justice.
But she is very young; she has never been taught to think
on serious subjects\index{subjects}; and for the last half-year, nay, for a
twelvemonth--she has been given up to nothing but amusement
and vanity\index{vanity}.  She has been allowed to dispose\index{dispose} of her time in the
most idle and frivolous\index{frivolous} manner, and to adopt any opinions that
came in her way.  Since the ----shire were first quartered in
Meryton, nothing but love, flirtation, and officers have been
in her head.  She has been doing everything in her power by
thinking and talking on the subject\index{subject}, to give greater--what shall
I call it? \textbf{susceptibility}\index{susceptibility}\stepcounter{footnote}\footnotetext{
\textbf{susceptibility} \emph{n.}  (suh-sehp-tih-BIH-lih-tee)
easily affected or influenced.} to her feelings; which are naturally
lively enough.  And we all know that Wickham has every charm of
person and address\index{address} that can captivate a woman."

"But you see that Jane," said her aunt, "does not think so very
ill of Wickham as to believe him capable of the attempt."

"Of whom does Jane ever think ill?  And who is there, whatever
might be their former conduct, that she would think capable of
such an attempt, till it were proved against them?  But Jane
knows, as well as I do, what Wickham really is.  We both know
that he has been profligate\index{profligate} in every sense of the word; that he
has neither integrity\index{integrity} nor honour; that he is as false and
deceitful as he is \textbf{insinuating}\index{insinuating}\stepcounter{footnote}\footnotetext{
\textbf{insinuate} \emph{v.} \textbf{-ing}  (ihn-SIN-yoo-ayt)
}."

"And do you really know all this?" cried Mrs. Gardiner, whose
curiosity as to the mode\index{mode} of her intelligence was all alive.

"I do indeed," replied Elizabeth, colouring.  "I told you, the
other day, of his infamous\index{infamous} behaviour to Mr. Darcy; and you
yourself, when last at Longbourn, heard in what manner he
spoke of the man who had behaved with such forbearance\index{forbearance} and
\textbf{liberality}\index{liberality}\stepcounter{footnote}\footnotetext{
\textbf{liberality} \emph{n.}  (lih-buh-RAL-lih-tee)
} towards him.  And there are other circumstances which
I am not at liberty--which it is not worth while to relate; but
his lies about the whole Pemberley family are endless.  From what
he said of Miss Darcy I was thoroughly prepared to see a proud,
reserved, disagreeable girl.  Yet he knew to the contrary himself.
He must know that she was as amiable\index{amiable} and unpretending as we
have found her."

"But does Lydia know nothing of this?  can she be ignorant of
what you and Jane seem so well to understand?"

"Oh, yes!--that, that is the worst of all.  Till I was in Kent,
and saw so much both of Mr. Darcy and his relation Colonel
Fitzwilliam, I was ignorant of the truth myself.  And when I
returned home, the ----shire was to leave Meryton in a week or
fortnight\index{fortnight}'s time.  As that was the case, neither Jane, to whom
I related the whole, nor I, thought it necessary to make our
knowledge public; for of what use could it apparently be to any
one, that the good opinion which all the neighbourhood had of
him should then be overthrown?  And even when it was settled
that Lydia should go with Mrs. Forster, the necessity of opening
her eyes to his character never occurred to me.  That \emph{she} could
be in any danger from the deception never entered my head.
That such a consequence\index{consequence} as \emph{this} could ensue, you may easily
believe, was far enough from my thoughts."

"When they all removed to Brighton, therefore, you had no
reason, I suppose, to believe them fond of each other?"

"Not the slightest.  I can remember no symptom of affection on
either side; and had anything of the kind been perceptible, you
must be aware that ours is not a family on which it could be
thrown away.  When first he entered the corps\index{corps}, she was ready
enough to admire him; but so we all were.  Every girl in or
near Meryton was out of her senses about him for the first
two months; but he never distinguished \emph{her} by any particular
attention; and, consequently, after a moderate period of
extravagant and wild admiration, her fancy for him gave
way, and others of the regiment\index{regiment}, who treated her with more
distinction, again became her favourites."



It may be easily believed, that however little of novelty\index{novelty} could be
added to their fears, hopes, and conjectures\index{conjectures}, on this interesting
subject\index{subject}, by its repeated discussion, no other could detain them
from it long, during the whole of the journey.  From Elizabeth's
thoughts it was never absent.  Fixed there by the keenest of all
anguish, self-reproach\index{reproach}, she could find no interval of ease or
forgetfulness.

They travelled as expeditiously\index{expeditiously} as possible, and, sleeping one
night on the road, reached Longbourn by dinner time the next
day.  It was a comfort to Elizabeth to consider that Jane could
not have been wearied by long expectations.

The little Gardiners, attracted by the sight of a chaise\index{chaise}, were
standing on the steps of the house as they entered the \textbf{paddock}\index{paddock}\stepcounter{footnote}\footnotetext{
\textbf{paddock} \emph{n.}  (PAD-dahk)
};
and, when the carriage drove up to the door, the joyful surprise
that lighted up their faces, and displayed itself over their whole
bodies, in a variety of capers and frisks, was the first pleasing
earnest\index{earnest} of their welcome.

Elizabeth jumped out; and, after giving each of them a hasty
kiss, hurried into the vestibule\index{vestibule}, where Jane, who came running
down from her mother's apartment\index{apartment}, immediately met her.

Elizabeth, as she affectionately embraced her, whilst tears filled
the eyes of both, lost not a moment in asking whether anything
had been heard of the fugitives.

"Not yet," replied Jane.  "But now that my dear uncle is come,
I hope everything will be well."

"Is my father in town?"

"Yes, he went on Tuesday, as I wrote you word."

"And have you heard from him often?"

"We have heard only twice.  He wrote me a few lines on
Wednesday to say that he had arrived in safety, and to give me
his directions, which I particularly begged him to do.  He merely
added that he should not write again till he had something of
importance to mention."

"And my mother--how is she?  How are you all?"

"My mother is tolerably well, I trust; though her spirits are
greatly shaken.  She is upstairs and will have great satisfaction
in seeing you all.  She does not yet leave her dressing-room.
Mary and Kitty are, thank Heaven, are quite well."

"But you--how are you?" cried Elizabeth.  "You look pale\index{pale}.
How much you must have gone through!"

Her sister, however, assured her of her being perfectly well;
and their conversation, which had been passing while Mr. and
Mrs. Gardiner were engaged with their children, was now put an
end to by the approach of the whole party.  Jane ran to her uncle
and aunt, and welcomed and thanked them both, with alternate
smiles and tears.

When they were all in the drawing-room, the questions which
Elizabeth had already asked were of course repeated by the
others, and they soon found that Jane had no intelligence
to give.  The \textbf{sanguine}\index{sanguine}\stepcounter{footnote}\footnotetext{
\textbf{sanguine} \emph{adj.}  (SAN-gwihn)
} hope of good, however, which the
benevolence\index{benevolence} of her heart suggested had not yet deserted her;
she still expected that it would all end well, and that every
morning would bring some letter, either from Lydia or her
father, to explain their proceedings, and, perhaps, announce
their marriage.

Mrs. Bennet, to whose apartment\index{apartment} they all repaired, after a few
minutes\index{minutes}' conversation together, received them exactly as might
be expected; with tears and lamentations\index{lamentations} of regret, \textbf{invectives}\index{invectives}\stepcounter{footnote}\footnotetext{
\textbf{invective} \emph{n.}  (ihn-VEHK-tihv)
abusive language; words intended to discredit or denounce.}
against the \textbf{villainous}\index{villainous}\stepcounter{footnote}\footnotetext{
\textbf{villainous} \emph{adj.}  (VIHL-leh-nuhs)
extremely wicked.} conduct of Wickham, and complaints of
her own sufferings and ill-usage; blaming everybody but the
person to whose ill-judging indulgence the errors of her
daughter must principally\index{principally} be owing.

"If I had been able," said she, "to carry my point in going to
Brighton, with all my family, \emph{this} would not have happened;
but poor dear Lydia had nobody to take care of her.  Why did
the Forsters ever let\index{let} her go out of their sight?  I am sure there
was some great neglect or other on their side, for she is not the
kind of girl to do such a thing if she had been well looked after.
I always thought they were very unfit to have the charge of her;
but I was \textbf{overruled}\index{overruled}\stepcounter{footnote}\footnotetext{
\textbf{overrule} \emph{tr.v.} \textbf{-ed}  (oh-ver-ROOL)
}, as I always am.  Poor dear child!  And
now here's Mr. Bennet gone away, and I know he will fight
Wickham, wherever he meets him and then he will be killed, and
what is to become of us all?  The Collinses will turn us out
before he is cold in his grave\index{grave}, and if you are not kind to us,
brother, I do not know what we shall do."

They all exclaimed against such terrific ideas; and Mr. Gardiner,
after general assurances of his affection for her and all her
family, told her that he meant to be in London the very next day,
and would assist Mr. Bennet in every endeavour\index{endeavour} for recovering
Lydia.

"Do not give way to useless alarm," added he; "though it is
right to be prepared for the worst, there is no occasion to look
on it as certain.  It is not quite a week since they left Brighton.
In a few days more we may gain some news of them; and till we
know that they are not married, and have no design of marrying,
do not let\index{let} us give the matter over as lost.  As soon as I get to
town I shall go to my brother, and make him come home with
me to Gracechurch Street; and then we may consult together as
to what is to be done."

"Oh! my dear brother," replied Mrs. Bennet, "that is exactly
what I could most wish for.  And now do, when you get to
town, find them out, wherever they may be; and if they are
not married already, \emph{make} them marry.  And as for wedding
clothes, do not let\index{let} them wait for that, but tell Lydia she
shall have as much money as she chooses to buy them, after they
are married.  And, above all, keep Mr. Bennet from fighting.
Tell him what a dreadful state I am in, that I am frighted out
of my wits--and have such tremblings, such flutterings, all
over me--such spasms in my side and pains in my head, and
such beatings at heart, that I can get no rest by night nor by
day.  And tell my dear Lydia not to give any directions about
her clothes till she has seen me, for she does not know which
are the best warehouses.  Oh, brother, how kind you are!  I
know you will contrive\index{contrive} it all."

But Mr. Gardiner, though he assured her again of his earnest\index{earnest}
endeavours\index{endeavours} in the cause, could not avoid recommending moderation
to her, as well in her hopes as her fear; and after talking with
her in this manner till dinner was on the table, they all left
her to vent all her feelings on the housekeeper, who attended
in the absence of her daughters.

Though her brother and sister were persuaded that there was no
real occasion for such a seclusion from the family, they did not
attempt to oppose it, for they knew that she had not prudence\index{prudence}
enough to hold her tongue before the servants, while they
waited at table, and judged it better that \emph{one} only of the
household, and the one whom they could most trust should
comprehend\index{comprehend} all her fears and solicitude\index{solicitude} on the subject\index{subject}.

In the dining-room they were soon joined by Mary and Kitty,
who had been  too busily engaged in their separate apartments\index{apartments}
to make their appearance before.  One came from her books,
and the other from her toilette.  The faces of both, however,
were tolerably calm; and no change was visible in either, except
that the loss of her favourite sister, or the anger which she had
herself \textbf{incurred}\index{incurred}\stepcounter{footnote}\footnotetext{
\textbf{incur} \emph{tr.v.} \textbf{-ed}  (ihn-KER)
 to make oneself subject to (something bad); to sustain (ie. damage).} in this business, had given more of \textbf{fretfulness}\index{fretfulness}\stepcounter{footnote}\footnotetext{
\textbf{fretful} \emph{adj.} \textbf{-ness}  (FREHT-fuhl)
 an irritable or petulant feeling; troubled through nagging anxiety.}
than usual to the accents of Kitty.  As for Mary, she was
mistress enough of herself to whisper to Elizabeth, with a
countenance\index{countenance} of grave\index{grave} reflection, soon after they were seated
at table:

"This is a most unfortunate affair, and will probably be much
talked of.  But we must stem the tide of malice\index{malice}, and pour into
the wounded bosoms of each other the \textbf{balm}\index{balm}\stepcounter{footnote}\footnotetext{
\textbf{balm} \emph{n.}  (BAHLM)
 any substance used for healing and soothing; ointment.} of sisterly consolation\index{consolation}."

Then, perceiving in Elizabeth no inclination\index{inclination} of replying, she
added, "Unhappy as the event must be for Lydia, we may draw
from it this useful lesson: that loss of virtue\index{virtue} in a female is
irretrievable; that one false step involves her in endless ruin;
that her reputation is no less brittle than it is beautiful; and
that she cannot be too much guarded in her behaviour towards the
undeserving of the other sex."

Elizabeth lifted up her eyes in amazement, but was too much
oppressed to make any reply.  Mary, however, continued to
console\index{console} herself with such kind of moral extractions from the
evil before them.

In the afternoon, the two elder Miss Bennets were able to be
for half-an-hour by themselves; and Elizabeth instantly availed\index{availed}
herself of the opportunity of making any inquiries, which Jane
was equally eager to satisfy.  After joining in general
lamentations\index{lamentations} over the dreadful sequel of this event, which
Elizabeth considered as all but certain, and Miss Bennet could
not assert to be wholly\index{wholly} impossible, the former continued the
subject\index{subject}, by saying, "But tell me all and everything about it
which I have not already heard.  Give me further particulars.
What did Colonel Forster say?  Had they no apprehension\index{apprehension} of
anything before the elopement took place?  They must have seen
them together for ever."

"Colonel Forster did own that he had often suspected some
partiality\index{partiality}, especially on Lydia's side, but nothing to give him any
alarm.  I am so grieved\index{grieved} for him!  His behaviour was attentive and
kind to the utmost.  He \emph{was} coming to us, in order to assure us
of his concern, before he had any idea of their not being gone to
Scotland: when that apprehension\index{apprehension} first got abroad, it hastened
his journey."

"And was Denny convinced that Wickham would not marry?  Did
he know of their intending to go off?  Had Colonel Forster
seen Denny himself?"

"Yes; but, when questioned by \emph{him}, Denny denied knowing
anything of their plans, and would not give his real opinion
about it.  He did not repeat his persuasion of their not
marrying--and from \emph{that}, I am inclined to hope, he might
have been misunderstood before."

"And till Colonel Forster came himself, not one of you
entertained a doubt, I suppose, of their being really married?"

"How was it possible that such an idea should enter our brains?
I felt a little uneasy--a little fearful of my sister's happiness
with him in marriage, because I knew that his conduct had not been
always quite right.  My father and mother knew nothing of that;
they only felt how imprudent\index{imprudent} a match it must be.  Kitty then
owned, with a very natural triumph\index{triumph} on knowing more than the
rest of us, that in Lydia's last letter she had prepared her for
such a step.  She had known, it seems, of their being in love with
each other, many weeks."

"But not before they went to Brighton?"

"No, I believe not."

"And did Colonel Forster appear to think well of Wickham
himself?  Does he know his real character?"

"I must confess that he did not speak so well of Wickham as he
formerly did.  He believed him to be imprudent\index{imprudent} and extravagant.
And since this sad affair has taken place, it is said that he
left Meryton greatly in debt; but I hope this may be false."

"Oh, Jane, had we been less secret, had we told what we knew
of him, this could not have happened!"

"Perhaps it would have been better," replied her sister.  "But to
expose the former faults of any person without knowing what
their present feelings were, seemed unjustifiable.  We acted with
the best intentions."

"Could Colonel Forster repeat the particulars of Lydia's note to
his wife?"

"He brought it with him for us to see."

Jane then took it from her pocket-book, and gave it to Elizabeth.
These were the contents:

"\emph{my} \emph{dear} \emph{harriet},

"You will laugh when you know where I am gone, and I cannot
help laughing myself at your surprise to-morrow morning, as
soon as I am missed.  I am going to Gretna Green, and if you
cannot guess with who, I shall think you a simpleton\index{simpleton}, for there
is but one man in the world I love, and he is an angel.  I should
never be happy without him, so think it no harm to be off.  You
need not send them word at Longbourn of my going, if you do
not like it, for it will make the surprise the greater, when I
write to them and sign my name 'Lydia Wickham.'  What a good
joke it will be!  I can hardly write for laughing.  Pray make
my excuses to Pratt for not keeping my engagement, and dancing
with him to-night.  Tell him I hope he will excuse me when he
knows all; and tell him I will dance with him at the next ball
we meet, with great pleasure.  I shall send for my clothes when
I get to Longbourn; but I wish you would tell Sally\index{Sally} to mend a
great slit in my worked muslin gown before they are packed up.
Good-bye.  Give my love to Colonel Forster.  I hope you will
drink to our good journey.

"Your affectionate friend,

"\emph{lydia} \emph{bennet}."

"Oh! thoughtless, thoughtless Lydia!" cried Elizabeth when she
had finished it.  "What a letter is this, to be written at such
a moment!  But at least it shows that \emph{she} was serious on the
subject\index{subject} of their journey.  Whatever he might afterwards
persuade her to, it was not on her side a \emph{scheme} of infamy\index{infamy}.
My poor father! how he must have felt it!"

"I never saw anyone so shocked.  He could not speak a word
for full ten minutes\index{minutes}.  My mother was taken ill immediately,
and the whole house in such confusion!"

"Oh! Jane," cried Elizabeth, "was there a servant belonging to it
who did not know the whole story before the end of the day?"

"I do not know.  I hope there was.  But to be guarded at such a
time is very difficult.  My mother was in hysterics, and though
I endeavoured\index{endeavoured} to give her every assistance in my power, I am
afraid I did not do so much as I might have done!  But the
horror of what might possibly happen almost took from me
my \textbf{faculties}\index{faculties}\stepcounter{footnote}\footnotetext{
\textbf{faculty} \emph{n.}  (FA-kuhl-tee)
An inherent power or ability.}."

"Your attendance upon her has been too much for you.  You do
not look well.  Oh that I had been with you!  you have had
every care and anxiety upon yourself alone."

"Mary and Kitty have been very kind, and would have shared in
every fatigue, I am sure; but I did not think it right for either
of them.  Kitty is slight and delicate; and Mary studies so much,
that her hours of \textbf{repose}\index{repose}\stepcounter{footnote}\footnotetext{
\textbf{repose} \emph{tr.v.}  (rih-POHZ)
} should not be broken in on.  My aunt
Phillips came to Longbourn on Tuesday, after my father went
away; and was so good as to stay till Thursday with me.  She
was of great use and comfort to us all.  And Lady Lucas has
been very kind; she walked here on Wednesday morning to
\textbf{condole}\index{condole}\stepcounter{footnote}\footnotetext{
\textbf{condole} \emph{intr.v.}  (kuhn-DOAL)
 to express sympathy for another's grief.} with us, and offered her services, or any of her
daughters', if they should be of use to us."

"She had better have stayed at home," cried Elizabeth; "perhaps
she \emph{meant} well, but, under such a misfortune as this, one
cannot see too little of one's neighbours.  Assistance is
impossible; \textbf{condolence}\index{condolence}\stepcounter{footnote}\footnotetext{
\textbf{condolence} \emph{n.}  (kuhn-DOH-lense)
an expression of sympathy with another's grief.} insufferable\index{insufferable}.  Let\index{Let} them triumph\index{triumph} over us
at a distance, and be satisfied."

She then proceeded to inquire into the measures which her
father had intended to pursue, while in town, for the recovery
of his daughter.

"He meant I believe," replied Jane, "to go to Epsom, the place
where they last changed horses, see the \textbf{postilions}\index{postilions}\stepcounter{footnote}\footnotetext{
\textbf{} \emph{}  ()
} and try if
anything could be made out from them.  His principal\index{principal} object
must be to discover the number of the hackney\index{hackney} coach which
took them from Clapham.  It had come with a fare from London;
and as he thought that the circumstance of a gentleman and lady's
removing from one carriage into another might be remarked he
meant to make inquiries at Clapham. If he could anyhow discover
at what house the coachman had before set down his fare, he
determined to make inquiries there, and hoped it might not be
impossible to find out the stand and number of the coach.  I do
not know of any other designs that he had formed; but he was in
such a hurry to be gone, and his spirits so greatly discomposed\index{discomposed},
that I had difficulty in finding out even so much as this."



\chapter{48}


The whole party were in hopes of a letter from Mr. Bennet the
next morning, but the post\index{post} came in without bringing a single line
from him.  His family knew him to be, on all common occasions,
a most negligent and \textbf{dilatory}\index{dilatory}\stepcounter{footnote}\footnotetext{
\textbf{dilatory} \emph{adj.}  (DIH-lat-taw-ree)
inclined to waste time and lag behind; intentionally delaying.} correspondent; but at such a time
they had hoped for exertion\index{exertion}.  They were forced to conclude that
he had no pleasing intelligence to send; but even of \emph{that} they
would have been glad to be certain.  Mr. Gardiner had waited
only for the letters before he set off.

When he was gone, they were certain at least of receiving
constant information of what was going on, and their uncle
promised, at parting, to prevail\index{prevail} on Mr. Bennet to return to
Longbourn, as soon as he could, to the great consolation\index{consolation} of his
sister, who considered it as the only security for her husband's
not being killed in a duel.

Mrs. Gardiner and the children were to remain in Hertfordshire
a few days longer, as the former thought her presence might be
serviceable to her nieces.  She shared in their attendance on
Mrs. Bennet, and was a great comfort to them in their hours of
freedom.  Their other aunt also visited them frequently, and
always, as she said, with the design of cheering and heartening
them up--though, as she never came without reporting some
fresh instance of Wickham's extravagance or irregularity, she
seldom went away without leaving them more \textbf{dispirited}\index{dispirited}\stepcounter{footnote}\footnotetext{
\textbf{dispirit} \emph{tr.v.} \textbf{-ed}  (dihs-SPIH-riht)
showing no enthusiasm; forlorn or dejected.} than
she found them.

All Meryton seemed striving to blacken the man who, but three
months before, had been almost an angel of light.  He was
declared to be in debt to every tradesman in the place, and his
intrigues, all honoured with the title of seduction, had been
extended into every tradesman's family.  Everybody declared
that he was the wickedest young man in the world; and everybody
began to find out that they had always distrusted the appearance
of his goodness.  Elizabeth, though she did not credit above
half of what was said, believed enough to make her former
assurance of her sister's ruin more certain; and even Jane,
who believed still less of it, became almost hopeless, more
especially as the time was now come when, if they had gone to
Scotland, which she had never before entirely despaired of,
they must in all probability have gained some news of them.

Mr. Gardiner left Longbourn on Sunday; on Tuesday his wife
received a letter from him; it told them that, on his arrival,
he had immediately found out his brother, and persuaded him to
come to Gracechurch Street; that Mr. Bennet had been to
Epsom and Clapham, before his arrival, but without gaining
any satisfactory information; and that he was now determined
to inquire at all the principal\index{principal} hotels in town, as Mr. Bennet
thought it possible they might have gone to one of them, on
their first coming to London, before they procured\index{procured} lodgings.
Mr. Gardiner himself did not expect any success from this
measure, but as his brother was eager in it, he meant to assist
him in pursuing it.  He added that Mr. Bennet seemed wholly\index{wholly}
disinclined at present to leave London and promised to write
again very soon.  There was also a postscript to this effect:

"I have written to Colonel Forster to desire him to find out,
if possible, from some of the young man's intimates in the
regiment\index{regiment}, whether Wickham has any relations or connections
who would be likely to know in what part of town he has now
concealed himself.  If there were anyone that one could apply
to with a probability of gaining such a clue as that, it might be
of essential consequence\index{consequence}.  At present we have nothing to guide
us.  Colonel Forster will, I dare say, do everything in his power
to satisfy us on this head.  But, on second thoughts, perhaps,
Lizzy could tell us what relations he has now living, better than
any other person."

Elizabeth was at no loss to understand from whence this
deference\index{deference} to her authority proceeded; but it was not in her
power to give any information of so satisfactory a nature as the
compliment deserved.  She had never heard of his having had
any relations, except a father and mother, both of whom had
been dead many years.  It was possible, however, that some of
his companions in the ----shire might be able to give more
information; and though she was not very sanguine\index{sanguine} in expecting
it, the application\index{application} was a something to look forward\index{forward} to.

Every day at Longbourn was now a day of anxiety; but the most
anxious part of each was when the post\index{post} was expected.  The
arrival of letters was the grand object of every morning's
impatience.  Through letters, whatever of good or bad was to
be told would be communicated, and every succeeding day was
expected to bring some news of importance.

But before they heard again from Mr. Gardiner, a letter arrived
for their father, from a different quarter, from Mr. Collins;
which, as Jane had received directions to open all that came for
him in his absence, she accordingly read; and Elizabeth, who
knew what curiosities his letters always were, looked over her,
and read it likewise.  It was as follows:

"\emph{my} \emph{dear} \emph{sir},

"I feel myself called upon, by our relationship, and my situation
in life, to condole\index{condole} with you on the grievous\index{grievous} \textbf{affliction}\index{affliction}\stepcounter{footnote}\footnotetext{
\textbf{affliction} \emph{n.}  (uh-FLIK-shun)
great suffering, distress, or pain.} you are now
suffering under, of which we were yesterday informed by a letter
from Hertfordshire.  Be assured, my dear sir, that Mrs. Collins
and myself sincerely sympathise with you and all your respectable
family, in your present distress, which must be of the bitterest
kind, because proceeding from a cause which no time can remove.
No arguments shall be wanting on my part that can alleviate\index{alleviate} so
severe\index{severe} a misfortune--or that may comfort you, under a circumstance
that must be of all others the most afflicting\index{afflicting} to a parent's mind.
The death of your daughter would have been a blessing in comparison
of this.  And it is the more to be lamented\index{lamented}, because there is
reason to suppose as my dear Charlotte informs me, that this
\textbf{licentiousness}\index{licentiousness}\stepcounter{footnote}\footnotetext{
\textbf{licentious} \emph{adj.} \textbf{-ness}  (ly-SEHN-chuhs)
being lewd and lascivious especially in sexual conduct.} of behaviour in your daughter has proceeded from
a faulty degree of indulgence; though, at the same time, for the
consolation\index{consolation} of yourself and Mrs. Bennet, I am inclined to think
that her own disposition\index{disposition} must be naturally bad, or she could not
be guilty of such an enormity, at so early an age.  Howsoever that
may be, you are grievously\index{grievously} to be pitied; in which opinion I am not
only joined by Mrs. Collins, but likewise by Lady Catherine and
her daughter, to whom I have related the affair.  They agree with
me in \textbf{apprehending}\index{apprehending}\stepcounter{footnote}\footnotetext{
\textbf{apprehend} \emph{v.} \textbf{-ing}  (ap-rih-HEHND)
To take into custody; to arrest.} that this false step in one daughter will be
injurious to the fortunes of all the others; for who, as Lady
Catherine herself condescendingly says, will connect themselves
with such a family?  And this consideration leads me moreover
to reflect, with augmented\index{augmented} satisfaction, on a certain event
of last November; for had it been otherwise, I must have been
involved in all your sorrow and disgrace.  Let\index{Let} me then advise you,
dear sir, to console\index{console} yourself as much as possible, to throw off
your unworthy child from your affection for ever, and leave her
to reap the fruits of her own \textbf{heinous}\index{heinous}\stepcounter{footnote}\footnotetext{
\textbf{heinous} \emph{adj.}  (HAY-nuhs)
 shockingly brutal or cruel.} offense.

"I am, dear sir, etc., etc."

Mr. Gardiner did not write again till he had received an answer
from Colonel Forster; and then he had nothing of a pleasant
nature to send.  It was not known that Wickham had a single
relationship with whom he kept up any connection, and it
was certain that he had no near one living.  His former
acquaintances had been numerous; but since he had been in the
militia\index{militia}, it did not appear that he was on terms of particular
friendship with any of them.  There was no one, therefore, who
could be pointed out as likely to give any news of him.  And
in the wretched\index{wretched} state of his own finances, there was a very
powerful motive for secrecy, in addition to his fear of discovery
by Lydia's relations, for it had just transpired that he had
left gaming debts behind him to a very considerable amount.
Colonel Forster believed that more than a thousand pounds
would be necessary to clear his expenses at Brighton.  He owed
a good deal in town, but his debts of honour were still more
formidable\index{formidable}.  Mr. Gardiner did not attempt to conceal these
particulars from the Longbourn family.  Jane heard them with
horror.  "A gamester!" she cried.  "This is wholly\index{wholly} unexpected.
I had not an idea of it."

Mr. Gardiner added in his letter, that they might expect to see
their father at home on the following day, which was Saturday.
Rendered\index{Rendered} spiritless by the ill-success of all their endeavours\index{endeavours}, he
had yielded to his brother-in-law's entreaty\index{entreaty} that he would return
to his family, and leave it to him to do whatever occasion might
suggest to be advisable for continuing their pursuit.  When Mrs.
Bennet was told of this, she did not express so much satisfaction
as her children expected, considering what her anxiety for his
life had been before.

"What, is he coming home, and without poor Lydia?" she cried.
"Sure he will not leave London before he has found them.  Who
is to fight Wickham, and make him marry her, if he comes away?"

As Mrs. Gardiner began to wish to be at home, it was settled
that she and the children should go to London, at the same time
that Mr. Bennet came from it.  The coach, therefore, took them
the first stage of their journey, and brought its master back
to Longbourn.

Mrs. Gardiner went away in all the perplexity\index{perplexity} about Elizabeth
and her Derbyshire friend that had attended her from that part
of the world.  His name had never been voluntarily mentioned
before them by her niece; and the kind of half-expectation which
Mrs. Gardiner had formed, of their being followed by a letter
from him, had ended in nothing.  Elizabeth had received none
since her return that could come from Pemberley.

The present unhappy state of the family rendered\index{rendered} any other
excuse for the lowness of her spirits unnecessary; nothing,
therefore, could be fairly conjectured\index{conjectured} from \emph{that}, though
Elizabeth, who was by this time tolerably well acquainted with
her own feelings, was perfectly aware that, had she known
nothing of Darcy, she could have borne\index{borne} the dread of Lydia's
infamy\index{infamy} somewhat better.  It would have spared\index{spared} her, she thought,
one sleepless night out of two.

When Mr. Bennet arrived, he had all the appearance of his usual
\textbf{philosophic}\index{philosophic}\stepcounter{footnote}\footnotetext{
\textbf{philosophical} \emph{adj.}  (fih-luh-SAH-fih-kuhl)
meeting trouble with level-headed detachment.} composure\index{composure}.  He said as little as he had ever been in
the habit of saying; made no mention of the business that had
taken him away, and it was some time before his daughters had
courage to speak of it.

It was not till the afternoon, when he had joined them at tea,
that Elizabeth ventured to introduce the subject\index{subject}; and then, on
her briefly expressing her sorrow for what he must have
endured, he replied, "Say nothing of that.  Who should suffer
but myself?  It has been my own doing, and I ought to feel it."

"You must not be too severe\index{severe} upon yourself," replied Elizabeth.

"You may well warn me against such an evil.  Human nature is
so prone to fall into it!  No, Lizzy, let\index{let} me once in my life
feel how much I have been to blame.  I am not afraid of being
overpowered by the impression.  It will pass away soon enough."

"Do you suppose them to be in London?"

"Yes; where else can they be so well concealed?"

"And Lydia used to want to go to London," added Kitty.

"She is happy then," said her father drily; "and her residence
there will probably be of some duration."

Then after a short silence he continued:

"Lizzy, I bear you no ill-will for being justified in your advice
to me last May, which, considering the event, shows some
greatness of mind."

They were interrupted by Miss Bennet, who came to fetch her
mother's tea.

"This is a parade," he cried, "which does one good; it gives such
an elegance to misfortune!  Another day I will do the same; I
will sit in my library, in my nightcap and powdering gown, and
give as much trouble as I can; or, perhaps, I may defer\index{defer} it till
Kitty runs away."

"I am not going to run away, papa," said Kitty fretfully\index{fretfully}.  "If I
should ever go to Brighton, I would behave better than Lydia."

"\emph{you} go to Brighton.  I would not trust you so near it as
Eastbourne for fifty pounds!  No, Kitty, I have at last learnt to
be cautious, and you will feel the effects of it.  No officer is ever
to enter into my house again, nor even to pass through the
village.  Balls will be absolutely prohibited, unless you stand up
with one of your sisters.  And you are never to stir out of doors
till you can prove that you have spent ten minutes\index{minutes} of every day
in a rational\index{rational} manner."

Kitty, who took all these threats in a serious light, began to cry.

"Well, well," said he, "do not make yourself unhappy.  If you
are a good girl for the next ten years, I will take you to a review
at the end of them."



\chapter{49}


Two days after Mr. Bennet's return, as Jane and Elizabeth were
walking together in the shrubbery\index{shrubbery} behind the house, they saw
the housekeeper coming towards them, and, concluding that she
came to call them to their mother, went forward\index{forward} to meet her;
but, instead of the expected summons\index{summons}\index{summons\index{summons}}, when they approached
her, she said to Miss Bennet, "I beg your pardon, madam, for
interrupting you, but I was in hopes you might have got some
good news from town, so I took the liberty of coming to ask."

"What do you mean, Hill?  We have heard nothing from town."

"Dear madam," cried Mrs. Hill, in great astonishment, "don't
you know there is an express come for master from Mr. Gardiner?
He has been here this half-hour, and master has had a letter."

Away ran the girls, too eager to get in to have time for speech.
They ran through the vestibule\index{vestibule} into the breakfast-room; from
thence to the library; their father was in neither; and they
were on the point of seeking him upstairs with their mother,
when they were met by the butler, who said:

"If you are looking for my master, ma'am, he is walking
towards the little \textbf{copse}\index{copse}\stepcounter{footnote}\footnotetext{
\textbf{copse} \emph{n.}  (KAHPS)
 a dense growth of bushes.}."

Upon this information, they instantly passed through the hall
once more, and ran across the lawn after their father, who was
deliberately pursuing his way towards a small wood on one side
of the paddock\index{paddock}.

Jane, who was not so light nor so much in the habit of running
as Elizabeth, soon lagged behind, while her sister, panting for
breath, came up with him, and eagerly cried out:

"Oh, papa, what news--what news?  Have you heard from my
uncle?"

"Yes I have had a letter from him by express."

"Well, and what news does it bring--good or bad?"

"What is there of good to be expected?" said he, taking the
letter from his pocket.  "But perhaps you would like to read it."

Elizabeth impatiently caught it from his hand.  Jane now came up.

"Read it aloud," said their father, "for I hardly know myself what
it is about."

"Gracechurch Street, Monday,
August 2.

"\emph{my} \emph{dear} \emph{brother},

"At last I am able to send you some tidings\index{tidings} of my niece, and
such as, upon the whole, I hope it will give you satisfaction.
Soon after you left me on Saturday, I was fortunate enough to
find out in what part of London they were.  The particulars I
reserve till we meet; it is enough to know they are discovered.
I have seen them both--"

"Then it is as I always hoped," cried Jane; "they are married!"

Elizabeth read on:

"I have seen them both.  They are not married, nor can I find
there was any intention of being so; but if you are willing to
perform the engagements which I have ventured to make on your
side, I hope it will not be long before they are.  All that is
required of you is, to assure to your daughter, by settlement,
her equal share of the five thousand pounds secured among your
children after the decease of yourself and my sister; and,
moreover, to enter into an engagement of allowing her, during
your life, one hundred pounds per annum.  These are conditions
which, considering everything, I had no hesitation in complying\index{complying}
with, as far as I thought myself privileged, for you.  I shall
send this by express, that no time may be lost in bringing me
your answer.  You will easily comprehend\index{comprehend}, from these particulars,
that Mr. Wickham's circumstances are not so hopeless as they
are generally believed to be.  The world has been deceived in
that respect; and I am happy to say there will be some little
money, even when all his debts are discharged, to settle on my
niece, in addition to her own fortune.  If, as I conclude will
be the case, you send me full powers to act in your name
throughout the whole of this business, I will immediately give
directions to Haggerston for preparing a proper settlement.
There will not be the smallest occasion for your coming to town
again; therefore stay quiet at Longbourn, and depend on my
\textbf{diligence}\index{diligence}\stepcounter{footnote}\footnotetext{
\textbf{diligence} \emph{n.}  (DIH-lih-jense)
conscientiousness and persistent in perform a task.} and care.  Send back your answer as fast as you can,
and be careful to write \textbf{explicitly}\index{explicitly}\stepcounter{footnote}\footnotetext{
\textbf{explicit} \emph{adj.} \textbf{-ly}  (ehx-PLIH-siht)
absolutely clear and understood; leaving nothing to implication.}.  We have judged it best that
my niece should be married from this house, of which I hope
you will approve.  She comes to us to-day.  I shall write again
as soon as anything more is determined on.  Yours, etc.,

"\emph{edw}. \emph{gardiner}."

"Is it possible?" cried Elizabeth, when she had finished.  "Can it
be possible that he will marry her?"

"Wickham is not so undeserving, then, as we thought him," said
her sister.  "My dear father, I congratulate you."

"And have you answered the letter?" cried Elizabeth.

"No; but it must be done soon."

Most earnestly\index{earnestly} did she then entreaty\index{entreaty} him to lose no more time
before he wrote.

"Oh! my dear father," she cried, "come back and write immediately.
Consider how important every moment is in such a case."

"Let\index{Let} me write for you," said Jane, "if you dislike the trouble
yourself."

"I dislike it very much," he replied; "but it must be done."

And so saying, he turned back with them, and walked towards
the house.

"And may I ask--" said Elizabeth; "but the terms, I suppose,
must be complied with."

"Complied with!  I am only ashamed of his asking so little."

"And they \emph{must} marry!  Yet he is \emph{such} a man!"

"Yes, yes, they must marry.  There is nothing else to be done.
But there are two things that I want very much to know; one is,
how much money your uncle has laid down to bring it about;
and the other, how am I ever to pay him."

"Money!  My uncle!" cried Jane, "what do you mean, sir?"

"I mean, that no man in his senses would marry Lydia on so
slight a temptation as one hundred a year during my life, and
fifty after I am gone."

"That is very true," said Elizabeth; "though it had not occurred
to me before.  His debts to be discharged, and something still
to remain!  Oh! it must be my uncle's doings!  Generous, good
man, I am afraid he has distressed himself.  A small sum could
not do all this."

"No," said her father; "Wickham's a fool if he takes her with a
farthing less than ten thousand pounds.  I should be sorry to
think so ill of him, in the very beginning of our relationship."

"Ten thousand pounds!  Heaven forbid!  How is half such a
sum to be repaid?"

Mr. Bennet made no answer, and each of them, deep in thought,
continued silent till they reached the house.  Their father then
went on to the library to write, and the girls walked into the
breakfast-room.

"And they are really to be married!" cried Elizabeth, as soon
as they were by themselves.  "How strange this is!  And for
\emph{this} we are to be thankful.  That they should marry, small as
is their chance of happiness, and wretched\index{wretched} as is his character,
we are forced to rejoice.  Oh, Lydia!"

"I comfort myself with thinking," replied Jane, "that he certainly
would not marry Lydia if he had not a real regard for her.
Though our kind uncle has done something towards clearing
him, I cannot believe that ten thousand pounds, or anything like
it, has been advanced.  He has children of his own, and may
have more.  How could he spare half ten thousand pounds?"

"If he were ever able to learn what Wickham's debts have been,"
said Elizabeth, "and how much is settled on his side on our
sister, we shall exactly know what Mr. Gardiner has done for
them, because Wickham has not sixpence of his own.  The
kindness of my uncle and aunt can never be \textbf{requited}\index{requited}\stepcounter{footnote}\footnotetext{
\textbf{requite} \emph{tr.v.} \textbf{-ed}  (rih-KWYT)
 To make repayment or return in kind.}.  Their
taking her home, and affording her their personal protection
and countenance\index{countenance}, is such a sacrifice to her advantage as years
of gratitude cannot enough acknowledge.  By this time she
is actually with them!  If such goodness does not make her
miserable now, she will never deserve to be happy!  What a
meeting for her, when she first sees my aunt!"

"We must endeavour\index{endeavour} to forget all that has passed on either side,"
said Jane: "I hope and trust they will yet be happy.  His
consenting to marry her is a proof, I will believe, that he is
come to a right way of thinking.  Their mutual affection will
steady\index{steady} them; and I flatter myself they will settle so quietly,
and live in so rational\index{rational} a manner, as may in time make their
past imprudence\index{imprudence} forgotten."

"Their conduct has been such," replied Elizabeth, "as neither
you, nor I, nor anybody can ever forget.  It is useless to talk
of it."

It now occurred to the girls that their mother was in all
likelihood perfectly ignorant of what had happened.  They went
to the library, therefore, and asked their father whether he
would not wish them to make it known to her.  He was writing
and, without raising his head, coolly replied:

"Just as you please."

"May we take my uncle's letter to read to her?"

"Take whatever you like, and get away."

Elizabeth took the letter from his writing-table, and they went
upstairs together.  Mary and Kitty were both with Mrs. Bennet:
one communication would, therefore, do for all.  After a slight
preparation for good news, the letter was read aloud.  Mrs.
Bennet could hardly contain herself.  As soon as Jane had read
Mr. Gardiner's hope of Lydia's  being soon married, her joy
burst forth, and every following sentence added to its
\textbf{exuberance}\index{exuberance}\stepcounter{footnote}\footnotetext{
\textbf{exuberance} \emph{n.}  (ehg-ZYOO-ber-ehnse)
joyfully overflowing with enthusiasm.}.  She was now in an irritation as violent from
delight, as she had ever been fidgety from alarm and vexation\index{vexation}.
To know that her daughter would be married was enough.  She
was disturbed by no fear for her felicity\index{felicity}, nor humbled by any
remembrance\index{remembrance} of her misconduct.

"My dear, dear Lydia!" she cried.  "This is delightful indeed!
She will be married!  I shall see her again!  She will be married
at sixteen!  My good, kind brother!  I knew how it would be.  I
knew he would manage everything!  How I long to see her!  and
to see dear Wickham too!  But the clothes, the wedding clothes!
I will write to my sister Gardiner about them directly.  Lizzy,
my dear, run down to your father, and ask him how much he will
give her.  Stay, stay, I will go myself.  Ring the bell, Kitty, for
Hill.  I will put on my things in a moment.  My dear, dear Lydia!
How merry we shall be together when we meet!"

Her eldest daughter endeavoured\index{endeavoured} to give some relief to the
violence of these transports, by leading her thoughts to the
obligations which Mr. Gardiner's behaviour laid them all under.

"For we must attribute this happy conclusion," she added, "in a
great measure to his kindness.  We are persuaded that he has
pledged himself to assist Mr. Wickham with money."

"Well," cried her mother, "it is all very right; who should do it
but her own uncle?  If he had not had a family of his own, I and
my children must have had all his money, you know; and it is the
first time we have ever had anything from him, except a few
presents.  Well!  I am so happy!  In a short time I shall have
a daughter married.  Mrs. Wickham!  How well it sounds!  And
she was only sixteen last June.  My dear Jane, I am in such a
flutter, that I am sure I can't write; so I will dictate, and you
write for me.  We will settle with your father about the money
afterwards; but the things should be ordered immediately."

She was then proceeding to all the particulars of calico,
muslin, and cambric, and would shortly have dictated some very
plentiful orders, had not Jane, though with some difficulty,
persuaded her to wait till her father was at leisure to be
consulted.  One day's delay, she observed, would be of small
importance; and her mother was too happy to be quite so
obstinate\index{obstinate} as usual.  Other schemes, too, came into her head.

"I will go to Meryton," said she, "as soon as I am dressed, and
tell the good, good news to my sister Philips.  And as I come
back, I can call on Lady Lucas and Mrs. Long.  Kitty, run down
and order the carriage.  An airing would do me a great deal of
good, I am sure.  Girls, can I do anything for you in Meryton?
Oh!  Here comes Hill!  My dear Hill, have you heard the good
news?  Miss Lydia is going to be married; and you shall all have
a bowl of punch to make merry at her wedding."

Mrs. Hill began instantly to express her joy.  Elizabeth received
her congratulations amongst the rest, and then, sick of this folly\index{folly},
took refuge in her own room, that she might think with freedom.

Poor Lydia's situation must, at best, be bad enough; but that
it was no worse, she had need to be thankful.  She felt it so;
and though, in looking forward\index{forward}, neither rational\index{rational} happiness nor
worldly prosperity could be justly expected for her sister, in
looking back to what they had feared, only two hours ago, she
felt all the advantages of what they had gained.



\chapter{50}


Mr. Bennet had very often wished before this period of his life
that, instead of spending his whole income, he had laid by an
annual sum for the better provision\index{provision} of his children, and of his
wife, if she survived him.  He now wished it more than ever.
Had he done his duty in that respect, Lydia need not have been
indebted to her uncle for whatever of honour or credit could
now be purchased for her.  The satisfaction of prevailing\index{prevailing}\index{prevailing\index{prevailing}} on
one of the most worthless young men in Great Britain to be her
husband might then have rested in its proper place.

He was seriously concerned that a cause of so little advantage
to anyone should be forwarded\index{forwarded} at the sole expense of his
brother-in-law, and he was determined, if possible, to find out
the extent of his assistance, and to discharge the obligation
as soon as he could.

When first Mr. Bennet had married, economy\index{economy} was held to be
perfectly useless, for, of course, they were to have a son.  The
son was to join in cutting off the entail\index{entail}, as soon as he should
be of age, and the widow and younger children would by that
means be provided for.  Five daughters successively entered the
world, but yet the son was to come; and Mrs. Bennet, for many
years after Lydia's birth, had been certain that he would.  This
event had at last been despaired of, but it was then too late to
be saving.  Mrs. Bennet had no turn for economy\index{economy}, and her
husband's love of independence had alone prevented their
exceeding their income.

Five thousand pounds was settled by marriage articles on Mrs.
Bennet and the children.  But in what proportions it should be
divided amongst the latter depended on the will of the parents.
This was one point, with regard to Lydia, at least, which was
now to be settled, and Mr. Bennet could have no hesitation in
acceding to the proposal before him.  In terms of grateful
acknowledgment for the kindness of his brother, though
expressed most concisely, he then delivered on paper his perfect
approbation\index{approbation} of all that was done, and his willingness to fulfil
the engagements that had been made for him.  He had never before
supposed that, could Wickham be prevailed\index{prevailed} on to marry his
daughter, it would be done with so little inconvenience to
himself as by the present arrangement.  He would scarcely be
ten pounds a year the loser by the hundred that was to be paid
them; for, what with her board and pocket allowance, and the
continual presents in money which passed to her through her
mother's hands, Lydia's expenses had been very little within
that sum.

That it would be done with such trifling\index{trifling} exertion\index{exertion} on his side,
too, was another very welcome surprise; for his wish at present
was to have as little trouble in the business as possible.  When
the first transports of rage which had produced his activity in
seeking her were over, he naturally returned to all his former
indolence\index{indolence}.  His letter was soon dispatched; for, though dilatory\index{dilatory}
in undertaking business, he was quick in its execution\index{execution}.  He
begged to know further particulars of what he was indebted to
his brother, but was too angry with Lydia to send any message
to her.

The good news spread quickly through the house, and with
\textbf{proportionate}\index{proportionate}\stepcounter{footnote}\footnotetext{
\textbf{proportionate} \emph{adj.}  (proh-POR-shun-eht)
agreeing or right in amount, magnitude, or degree.} speed through the neighbourhood.  It was borne\index{borne}
in the latter with decent \textbf{philosophy}\index{philosophy}\stepcounter{footnote}\footnotetext{
\textbf{philosophy} \emph{n.}  (fih-LAH-suh-fee)
}.  To be sure, it would
have been more for the advantage of conversation had Miss Lydia
Bennet come upon the town; or, as the happiest alternative,
been secluded from the world, in some distant farmhouse.
But there was much to be talked of in marrying her; and the
good-natured wishes for her well-doing which had proceeded
before from all the spiteful old ladies in Meryton lost but a
little of their spirit in this change of circumstances, because
with such an husband her misery was considered certain.

It was a fortnight\index{fortnight} since Mrs. Bennet had been downstairs; but on
this happy day she again took her seat at the head of her table,
and in spirits oppressively high.  No sentiment\index{sentiment} of shame gave
a damp to her triumph\index{triumph}.  The marriage of a daughter, which had
been the first object of her wishes since Jane was sixteen, was
now on the point of accomplishment, and her thoughts and her
words ran wholly\index{wholly} on those attendants of elegant \textbf{nuptials}\index{nuptials}\stepcounter{footnote}\footnotetext{
\textbf{nuptial} \emph{adj.}  (NUHP-chih-uhl)
 the social event associated with a wedding ceremony.}, fine
muslins, new carriages, and servants.  She was busily searching
through the neighbourhood for a proper situation for her
daughter, and, without knowing or considering what their
income might be, rejected many as deficient in size and
importance.

"Haye Park might do," said she, "if the Gouldings could quit it--or
the great house at Stoke, if the drawing-room were larger;  but
Ashworth is too far off!  I could not bear to have her ten miles
from me; and as for Pulvis Lodge, the attics are dreadful."

Her husband allowed her to talk on without interruption while
the servants remained.  But when they had withdrawn, he said
to her: "Mrs. Bennet, before you take any or all of these houses
for your son and daughter, let\index{let} us come to a right understanding.
Into \emph{one} house in this neighbourhood they shall never have
admittance.  I will not encourage the \textbf{impudence}\index{impudence}\stepcounter{footnote}\footnotetext{
\textbf{impudence} \emph{n.}  (IHM-pyoo-dense)
offensive boldness; rude and impertinent.} of either,
by receiving them at Longbourn."

A long dispute followed this declaration; but Mr. Bennet was
firm.  It soon led to another; and Mrs. Bennet found, with
amazement and horror, that her husband would not advance a
guinea to buy clothes for his daughter.  He protested that she
should receive from him no mark of affection whatever on the
occasion.  Mrs. Bennet could hardly comprehend\index{comprehend} it.  That his
anger could be carried to such a point of inconceivable
resentment\index{resentment} as to refuse his daughter a privilege without which
her marriage would scarcely seem valid, exceeded all she could
believe possible.  She was more alive to the disgrace which her
want of new clothes must reflect on her daughter's nuptials\index{nuptials}, than
to any sense of shame at her eloping and living with Wickham a
fortnight\index{fortnight} before they took place.

Elizabeth was now most heartily\index{heartily} sorry that she had, from the
distress of the moment, been led to make Mr. Darcy acquainted
with their fears for her sister; for since her marriage would
so shortly give the proper termination to the elopement, they
might hope to conceal its unfavourable beginning from all those
who were not immediately on the spot.

She had no fear of its spreading farther through his means.
There were few people on whose secrecy she would have more
confidently depended; but, at the same time, there was no one
whose knowledge of a sister's frailty would have mortified her
so much--not, however, from any fear of disadvantage from it
individually to herself, for, at any rate, there seemed a gulf
\textbf{impassable}\index{impassable}\stepcounter{footnote}\footnotetext{
\textbf{impassable} \emph{adj.}  (ihm-PAS-suh-bl)
impossible to pass.} between them.  Had Lydia's marriage been concluded
on the most honourable terms, it was not to be supposed that
Mr. Darcy would connect himself with a family where, to every
other objection, would now be added an alliance\index{alliance} and relationship
of the nearest kind with a man whom he so justly scorned.

From such a connection she could not wonder that he would shrink.
The wish of procuring her regard, which she had assured herself
of his feeling in Derbyshire, could not in rational\index{rational} expectation
survive such a blow as this.  She was humbled, she was grieved\index{grieved};
she repented, though she hardly knew of what.  She became jealous
of his esteem, when she could no longer hope to be benefited by it.
She wanted to hear of him, when there seemed the least chance of
gaining intelligence.  She was convinced that she could have been
happy with him, when it was no longer likely they should meet.

What a triumph\index{triumph} for him, as she often thought, could he know
that the proposals which she had proudly \textbf{spurned}\index{spurned}\stepcounter{footnote}\footnotetext{
\textbf{spurn} \emph{v.} \textbf{-ed}  (SPERN)
 to rebuff (by a lover) without warning; to scorn or reject.} only four
months ago, would now have been most gladly and gratefully
received!  He was as generous, she doubted not, as the most
generous of his sex; but while he was mortal, there must be a
triumph\index{triumph}.

She began now to comprehend\index{comprehend} that he was exactly the man
who, in disposition\index{disposition} and talents, would most suit her.  His
understanding and temper, though unlike her own, would have
answered all her wishes.  It was an union that must have been
to the advantage of both; by her ease and liveliness, his mind
might have been softened, his manners improved; and from his
judgement, information, and knowledge of the world, she must
have received benefit of greater importance.

But no such happy marriage could now teach the admiring
multitude what \textbf{connubial}\index{connubial}\stepcounter{footnote}\footnotetext{
\textbf{connubial} \emph{adj.}  (kuhn-NOO-bih-uhl)
related to marriage; in conjugal or familliar manner as if married.} felicity\index{felicity} really was.  A union of a
different tendency, and precluding the possibility of the
other, was soon to be formed in their family.

How Wickham and Lydia were to be supported in tolerable\index{tolerable}
independence, she could not imagine.  But how little of
permanent happiness could belong to a couple who were only
brought together because their passions were stronger than
their virtue\index{virtue}, she could easily conjecture\index{conjecture}.



Mr. Gardiner soon wrote again to his brother.  To Mr. Bennet's
acknowledgments he briefly replied, with assurance of his
eagerness to promote the welfare of any of his family; and
concluded with entreaties that the subject\index{subject} might never be
mentioned to him again.  The principal\index{principal} purport\index{purport} of his letter was
to inform them that Mr. Wickham had resolved on quitting the
militia\index{militia}.

"It was greatly my wish that he should do so," he added, "as
soon as his marriage was fixed on.  And I think you will agree
with me, in considering the removal from that corps\index{corps} as highly
advisable, both on his account and my niece's.  It is Mr. Wickham's
intention to go into the \textbf{regulars}\index{regulars}\stepcounter{footnote}\footnotetext{
\textbf{regular} \emph{adj.}  (REH-gyoo-luer)
}; and among his former friends,
there are still some who are able and willing to assist him in
the army.  He has the promise of an \textbf{ensigncy}\index{ensigncy}\stepcounter{footnote}\footnotetext{
\textbf{} \emph{}  ()
} in General ----'s
regiment\index{regiment}, now quartered in the North.  It is an advantage to have
it so far from this part of the kingdom.  He promises fairly; and
I hope among different people, where they may each have a character
to preserve, they will both be more prudent\index{prudent}.  I have written to
Colonel Forster, to inform him of our present arrangements, and to
request that he will satisfy the various creditors of Mr. Wickham
in and near Brighton, with assurances of speedy payment, for which
I have pledged myself.  And will you give yourself the trouble of
carrying similar assurances to his creditors in Meryton, of whom
I shall \textbf{subjoin}\index{subjoin}\stepcounter{footnote}\footnotetext{
\textbf{subjoin} \emph{tr.v.}  (suhb-JOYN)
To add to the end; to append.} a list according to his information?  He has
given in all his debts; I hope at least he has not deceived us.
Haggerston has our directions, and all will be completed in a week.
They will then join his regiment\index{regiment}, unless they are first invited to
Longbourn; and I understand from Mrs. Gardiner, that my niece is
very desirous of seeing you all before she leaves the South.  She
is well, and begs to be dutifully remembered to you and your
mother.--Yours, etc.,

"E. \emph{gardiner}."

Mr. Bennet and his daughters saw all the advantages of Wickham's
removal from the ----shire as clearly as Mr. Gardiner could do.
But Mrs. Bennet was not so well pleased with it.  Lydia's being
settled in the North, just when she had expected most pleasure and
pride in her company, for she had by no means given up her plan
of their residing in Hertfordshire, was a severe\index{severe} disappointment;
and, besides, it was such a pity that Lydia should be taken from
a regiment\index{regiment} where she was acquainted with everybody, and had so
many favourites.

"She is so fond of Mrs. Forster," said she, "it will be quite
shocking to send her away!  And there are several of the young
men, too, that she likes very much.  The officers may not be so
pleasant in General----'s regiment\index{regiment}."

His daughter's request, for such it might be considered, of being
admitted into her family again before she set off for the North,
received at first an absolute negative.  But Jane and Elizabeth,
who agreed in wishing, for the sake of their sister's feelings and
consequence\index{consequence}, that she should be noticed on her marriage by her
parents, urged him so earnestly\index{earnestly} yet so rationally\index{rationally} and so mildly,
to receive her and her husband at Longbourn, as soon as they
were married, that he was prevailed\index{prevailed} on to think as they thought,
and act as they wished.  And their mother had the satisfaction
of knowing that she would be able to show her married daughter
in the neighbourhood before she was banished to the North.  When
Mr. Bennet wrote again to his brother, therefore, he sent his
permission for them to come; and it was settled, that as soon
as the ceremony was over, they should proceed to Longbourn.
Elizabeth was surprised, however, that Wickham should consent
to such a scheme, and had she consulted only her own inclination\index{inclination},
any meeting with him would have been the last object of her wishes.



\chapter{51}


Their sister's wedding day arrived; and Jane and Elizabeth felt
for her probably more than she felt for herself.  The carriage
was sent to meet them at ----, and they were to return in it
by dinner-time.  Their arrival was dreaded by the elder Miss
Bennets, and Jane more especially, who gave Lydia the feelings
which would have attended herself, had she been the \textbf{culprit}\index{culprit}\stepcounter{footnote}\footnotetext{
\textbf{culprit} \emph{n.}  (KUHL-priht)
someone who perpetrates a crime or wrongdoing.},
and was wretched\index{wretched} in the thought of what her sister must endure.

They came.  The family were assembled in the breakfast room to
receive them.  Smiles decked the face of Mrs. Bennet as the
carriage drove up to the door; her husband looked \textbf{impenetrably}\index{impenetrably}\stepcounter{footnote}\footnotetext{
\textbf{impenetrable} \emph{adj.} \textbf{-ly}  (ihm-PEH-neht-ruh-bl)
}
grave\index{grave}; her daughters, alarmed, anxious, uneasy.

Lydia's voice was heard in the vestibule\index{vestibule}; the door was thrown
open, and she ran into the room.  Her mother stepped forwards\index{forwards},
embraced her, and welcomed her with rapture\index{rapture}; gave her hand,
with an affectionate smile, to Wickham, who followed his lady;
and wished them both joy with an alacrity\index{alacrity} which shewed no doubt
of their happiness.

Their reception from Mr. Bennet, to whom they then turned, was
not quite so cordial\index{cordial}.  His countenance\index{countenance} rather gained in \textbf{austerity}\index{austerity}\stepcounter{footnote}\footnotetext{
\textbf{austerity} \emph{n.}  (aws-STEH-rih-tee)
the trait of great self-denial (especially from worldly pleasures).};
and he scarcely opened his lips.  The easy assurance of the
young couple, indeed, was enough to provoke\index{provoke} him.  Elizabeth was
disgusted, and even Miss Bennet was shocked.  Lydia was Lydia
still; untamed, unabashed, wild, noisy, and fearless.  She turned
from sister to sister, demanding their congratulations; and when
at length they all sat down, looked eagerly round the room, took
notice of some little alteration in it, and observed, with a
laugh, that it was a great while since she had been there.

Wickham was not at all more distressed than herself, but his
manners were always so pleasing, that had his character and his
marriage been exactly what they ought, his smiles and his easy
address\index{address}, while he claimed their relationship, would have
delighted them all.  Elizabeth had not before believed him
quite equal to such assurance; but she sat down, resolving
within herself to draw no limits in future to the impudence\index{impudence} of
an \textbf{impudent}\index{impudent}\stepcounter{footnote}\footnotetext{
\textbf{impudent} \emph{adj.}  (IHM-pyoo-dehnt)
improperly forward or bold; insolent.} man.  She blushed, and Jane blushed; but the
cheeks of the two who caused their confusion suffered no
variation of colour.

There was no want of discourse\index{discourse}.  The bride and her mother could
neither of them talk fast enough; and Wickham, who happened to
sit near Elizabeth, began inquiring after his acquaintance in
that neighbourhood, with a good humoured ease which she felt
very unable to equal in her replies.  They seemed each of them
to have the happiest memories in the world.  Nothing of the
past was recollected with pain; and Lydia led voluntarily to
subjects\index{subjects} which her sisters would not have alluded\index{alluded} to for the
world.

"Only think of its being three months," she cried, "since I
went away; it seems but a fortnight\index{fortnight} I declare; and yet there
have been things enough happened in the time.  Good gracious!
when I went away, I am sure I had no more idea of being married
till I came back again! though I thought it would be very good
fun if I was."

Her father lifted up his eyes.  Jane was distressed.  Elizabeth
looked expressively at Lydia; but she, who never heard nor saw
anything of which she chose to be insensible, gaily continued,
"Oh! mamma, do the people hereabouts know I am married
to-day?  I was afraid they might not; and we overtook William
Goulding in his curricle\index{curricle}, so I was determined he should know
it, and so I let\index{let} down the side-glass next to him, and took off
my glove, and let\index{let} my hand just rest upon the window frame, so
that he might see the ring, and then I bowed and smiled like
anything."

Elizabeth could bear it no longer.  She got up, and ran out of
the room; and returned no more, till she heard them passing
through the hall to the dining parlour.  She then joined them
soon enough to see Lydia, with anxious parade, walk up to her
mother's right hand, and hear her say to her eldest sister,
"Ah! Jane, I take your place now, and you must go lower,
because I am a married woman."

It was not to be supposed that time would give Lydia that
embarrassment from which she had been so wholly\index{wholly} free at first.
Her ease and good spirits increased.  She longed to see Mrs.
Phillips, the Lucases, and all their other neighbours, and to
hear herself called "Mrs. Wickham" by each of them; and in the
mean time, she went after dinner to show her ring, and boast
of being married, to Mrs. Hill and the two housemaids.

"Well, mamma," said she, when they were all returned to the
breakfast room, "and what do you think of my husband?  Is not
he a charming man?  I am sure my sisters must all envy me.  I
only hope they may have half my good luck.  They must all go to
Brighton.  That is the place to get husbands.  What a pity it
is, mamma, we did not all go."

"Very true; and if I had my will, we should.  But my dear
Lydia, I don't at all like your going such a way off.  Must
it be so?"

"Oh, lord! yes;--there is nothing in that.  I shall like it
of all things.  You and papa, and my sisters, must come down
and see us.  We shall be at Newcastle all the winter, and I
dare say there will be some balls, and I will take care to get
good partners for them all."

"I should like it beyond anything!" said her mother.

"And then when you go away, you may leave one or two of my
sisters behind you; and I dare say I shall get husbands for
them before the winter is over."

"I thank you for my share of the favour," said Elizabeth;
"but I do not particularly like your way of getting husbands."

Their visitors were not to remain above ten days with them.
Mr. Wickham had received his commission before he left London,
and he was to join his regiment\index{regiment} at the end of a fortnight\index{fortnight}.

No one but Mrs. Bennet regretted that their stay would be so
short; and she made the most of the time by visiting about with
her daughter, and having very frequent parties at home.  These
parties were acceptable to all; to avoid a family circle was
even more desirable to such as did think, than such as did not.

Wickham's affection for Lydia was just what Elizabeth had
expected to find it; not equal to Lydia's for him.  She had
scarcely needed her present observation to be satisfied, from
the reason of things, that their elopement had been brought on
by the strength of her love, rather than by his; and she would
have wondered why, without violently caring for her, he chose
to elope\index{elope} with her at all, had she not felt certain that his
flight was rendered\index{rendered} necessary by distress of circumstances;
and if that were the case, he was not the young man to resist
an opportunity of having a companion.

Lydia was exceedingly fond of him.  He was her dear Wickham on
every occasion; no one was to be put in competition with him.
He did every thing best in the world; and she was sure he would
kill more birds on the first of September, than any body else
in the country.

One morning, soon after their arrival, as she was sitting with
her two elder sisters, she said to Elizabeth:

"Lizzy, I never gave \emph{you} an account of my wedding, I believe.
You were not by, when I told mamma and the others all about it.
Are not you curious to hear how it was managed?"

"No really," replied Elizabeth; "I think there cannot be too
little said on the subject\index{subject}."

"La!  You are so strange!  But I must tell you how it went off.
We were married, you know, at St. Clement's, because Wickham's
lodgings were in that parish\index{parish}.  And it was settled that we
should all be there by eleven o'clock.  My uncle and aunt and
I were to go together; and the others were to meet us at the
church.  Well, Monday morning came, and I was in such a fuss!
I was so afraid, you know, that something would happen to put
it off, and then I should have gone quite \textbf{distracted}\index{distracted}\stepcounter{footnote}\footnotetext{
\textbf{distracted} \emph{adj.}  (his-TRAK-tehd)
having one's attention diverted especially because of anxiety.}.  And
there was my aunt, all the time I was dressing, preaching and
talking away just as if she was reading a sermon.  However, I
did not hear above one word in ten, for I was thinking, you may
suppose, of my dear Wickham.  I longed to know whether he would
be married in his blue coat."

"Well, and so we breakfasted at ten as usual; I thought it
would never be over; for, by the bye, you are to understand,
that my uncle and aunt were horrid unpleasant all the time I
was with them.  If you'll believe me, I did not once put my
foot out of doors, though I was there a fortnight\index{fortnight}.  Not one
party, or scheme, or anything.  To be sure London was rather
thin, but, however, the Little Theatre was open.  Well, and so
just as the carriage came to the door, my uncle was called away
upon business to that horrid man Mr. Stone.  And then, you
know, when once they get together, there is no end of it.
Well, I was so frightened I did not know what to do, for my
uncle was to give me away; and if we were beyond the hour, we
could not be married all day.  But, luckily, he came back again
in ten minutes\index{minutes}' time, and then we all set out.  However, I
recollected afterwards that if he had been prevented going,
the wedding need not be put off, for Mr. Darcy might have done
as well."

"Mr. Darcy!" repeated Elizabeth, in utter amazement.

"Oh, yes!--he was to come there with Wickham, you know.  But
gracious me!  I quite forgot!  I ought not to have said a word
about it.  I promised them so faithfully!  What will Wickham
say?  It was to be such a secret!"

"If it was to be secret," said Jane, "say not another word on
the subject\index{subject}.  You may depend upon my seeking no further."

"Oh! certainly," said Elizabeth, though burning with curiosity;
"we will ask you no questions."

"Thank you," said Lydia, "for if you did, I should certainly
tell you all, and then Wickham would be angry."

On such encouragement to ask, Elizabeth was forced to put it
out of her power, by running away.

But to live in ignorance on such a point was impossible; or at
least it was impossible not to try for information.  Mr. Darcy
had been at her sister's wedding.  It was exactly a scene, and
exactly among people, where he had apparently least to do, and
least temptation to go.  Conjectures\index{Conjectures} as to the meaning of it,
rapid\index{rapid} and wild, hurried into her brain; but she was satisfied
with none.  Those that best pleased her, as placing his conduct
in the noblest light, seemed most improbable\index{improbable}.  She could not
bear such suspense; and hastily seizing a sheet of paper, wrote
a short letter to her aunt, to request an explanation of what
Lydia had dropt, if it were compatible with the secrecy which
had been intended.

"You may readily\index{readily} comprehend\index{comprehend}," she added, "what my curiosity
must be to know how a person unconnected with any of us, and
(comparatively\index{comparatively} speaking) a stranger to our family, should have
been amongst you at such a time.  Pray write instantly, and
let\index{let} me understand it--unless it is, for very \textbf{cogent}\index{cogent}\stepcounter{footnote}\footnotetext{
\textbf{cogent} \emph{adj.}  (KOH-juhnt)
  powerfully persuasive.} reasons,
to remain in the secrecy which Lydia seems to think necessary;
and then I must endeavour\index{endeavour} to be satisfied with ignorance."

"Not that I \emph{shall}, though," she added to herself, as she
finished the letter; "and my dear aunt, if you do not tell me
in an honourable manner, I shall certainly be reduced to tricks
and stratagems to find it out."

Jane's delicate sense of honour would not allow her to speak to
Elizabeth privately\index{privately} of what Lydia had let\index{let} fall; Elizabeth was
glad of it;--till it appeared whether her inquiries would
receive any satisfaction, she had rather be without a \textbf{confidante}\index{confidante}\stepcounter{footnote}\footnotetext{
\textbf{confidante} \emph{n.}  (kahn-fih-DAHNT)
a female confidant; a woman trusted with secrets.}.



\chapter{52}


Elizabeth had the satisfaction of receiving an answer to her
letter as soon as she possibly could.  She was no sooner in
possession of it than, hurrying into the little copse\index{copse}, where
she was least likely to be interrupted, she sat down on one of
the benches and prepared to be happy; for the length of the
letter convinced her that it did not contain a denial.

"Gracechurch street, Sept. 6.

"\emph{my} \emph{dear} \emph{niece},

"I have just received your letter, and shall devote this whole
morning to answering it, as I foresee that a \emph{little} writing
will not comprise what I have to tell you.  I must confess
myself surprised by your application\index{application}; I did not expect it from
\emph{you}.  Don't think me angry, however, for I only mean to let\index{let}
you know that I had not imagined such inquiries to be necessary
on \emph{your} side.  If you do not choose to understand me, forgive
my impertinence\index{impertinence}.  Your uncle is as much surprised as I am--and
nothing but the belief of your being a party concerned would
have allowed him to act as he has done.  But if you are really
innocent and ignorant, I must be more explicit\index{explicit}.

"On the very day of my coming home from Longbourn, your uncle had a
most unexpected visitor.  Mr. Darcy called, and was shut up with him
several hours.  It was all over before I arrived; so my curiosity was
not so dreadfully \textbf{racked}\index{racked}\stepcounter{footnote}\footnotetext{
\textbf{rack} \emph{n.} \textbf{-ed}  (RAK)
} as \emph{your}'S seems to have been.  He came to
tell Mr. Gardiner that he had found out where your sister and
Mr. Wickham were, and that he had seen and talked with them both;
Wickham repeatedly, Lydia once.  From what I can collect, he left
Derbyshire only one day after ourselves, and came to town with the
resolution\index{resolution} of hunting for them.  The motive professed was his
conviction\index{conviction} of its being owing to himself that Wickham's worthlessness
had not been so well known as to make it impossible for any young
woman of character to love or confide in him.  He generously imputed\index{imputed}
the whole to his mistaken pride, and confessed that he had before
thought it beneath him to lay his private\index{private} actions open to the world.
His character was to speak for itself.  He called it, therefore, his
duty to step forward\index{forward}, and endeavour\index{endeavour} to remedy an evil which had been
brought on by himself.  If he \emph{had} \emph{another} motive, I am sure it would
never disgrace him.  He had been some days in town, before he was able
to discover them; but he had something to direct his search, which was
more than \emph{we} had; and the consciousness of this was another reason for
his resolving to follow us.

"There is a lady, it seems, a Mrs. Younge, who was some time ago
governess to Miss Darcy, and was dismissed from her charge on some
cause of disapprobation\index{disapprobation}, though he did not say what.  She then took a
large house in Edward-street, and has since maintained herself by
letting lodgings.  This Mrs. Younge was, he knew, intimately
acquainted with Wickham; and he went to her for intelligence of him as
soon as he got to town.  But it was two or three days before he could
get from her what he wanted.  She would not betray her trust, I
suppose, without bribery and corruption, for she really did know where
her friend was to be found.  Wickham indeed had gone to her on their
first arrival in London, and had she been able to receive them into
her house, they would have taken up their abode\index{abode} with her.  At length,
however, our kind friend procured\index{procured} the wished-for direction.  They were
in ---- street.  He saw Wickham, and afterwards insisted on seeing
Lydia.  His first object with her, he acknowledged, had been to
persuade her to quit her present disgraceful situation, and return to
her friends as soon as they could be prevailed\index{prevailed} on to receive her,
offering his assistance, as far as it would go.  But he found Lydia
absolutely resolved on remaining where she was.  She cared for none of
her friends; she wanted no help of his; she would not hear of leaving
Wickham.  She was sure they should be married some time or other, and
it did not much signify when.  Since such were her feelings, it only
remained, he thought, to secure and \textbf{expedite}\index{expedite}\stepcounter{footnote}\footnotetext{
\textbf{expedite} \emph{tr.v.}  (EHX-pih-dyt)
speed up the progress of; facilitate.} a marriage, which, in his
very first conversation with Wickham, he easily learnt had never been
\emph{his} design.  He confessed himself obliged to leave the regiment\index{regiment}, on
account of some debts of honour, which were very pressing; and
scrupled\index{scrupled} not to lay all the ill-consequences\index{consequences} of Lydia's flight on her
own folly\index{folly} alone.  He meant to resign his commission immediately; and
as to his future situation, he could conjecture\index{conjecture} very little about it.
He must go somewhere, but he did not know where, and he knew he should
have nothing to live on.

"Mr. Darcy asked him why he had not married your sister at once.
Though Mr. Bennet was not imagined to be very rich, he would have been
able to do something for him, and his situation must have been
benefited by marriage.  But he found, in reply to this question, that
Wickham still cherished the hope of more effectually\index{effectually} making his
fortune by marriage in some other country.  Under such circumstances,
however, he was not likely to be proof against the temptation of
immediate relief.

"They met several times, for there was much to be discussed.
Wickham of course wanted more than he could get; but at length
was reduced to be reasonable.

"Every thing being settled between \emph{them}, Mr. Darcy's next step was to
make your uncle acquainted with it, and he first called in Gracechurch
street the evening before I came home.  But Mr. Gardiner could not be
seen, and Mr. Darcy found, on further inquiry, that your father was
still with him, but would quit town the next morning.  He did not
judge your father to be a person whom he could so properly consult as
your uncle, and therefore readily\index{readily} postponed seeing him till after the
departure of the former.  He did not leave his name, and till the next
day it was only known that a gentleman had called on business.

"On Saturday he came again.  Your father was gone, your uncle at home,
and, as I said before, they had a great deal of talk together.

"They met again on Sunday, and then \emph{I} saw him too.  It was not all
settled before Monday: as soon as it was, the express was sent off to
Longbourn.  But our visitor was very obstinate\index{obstinate}.  I fancy, Lizzy, that
obstinacy\index{obstinacy} is the real defect of his character, after all.  He has been
accused of many faults at different times, but \emph{this} is the true one.
Nothing was to be done that he did not do himself; though I am sure
(and I do not speak it to be thanked, therefore say nothing about it),
your uncle would most readily\index{readily} have settled the whole.

"They battled it together for a long time, which was more than either
the gentleman or lady concerned in it deserved.  But at last your
uncle was forced to yield, and instead of being allowed to be of use
to his niece, was forced to put up with only having the probable\index{probable}
credit of it, which went sorely against the grain; and I really
believe your letter this morning gave him great pleasure, because it
required an explanation that would rob him of his borrowed feathers,
and give the praise where it was due.  But, Lizzy, this must go no
farther than yourself, or Jane at most.

"You know pretty well, I suppose, what has been done for the young
people.  His debts are to be paid, amounting, I believe, to
considerably more than a thousand pounds, another thousand in addition
to her own settled upon \emph{her}, and his commission purchased.  The reason
why all this was to be done by him alone, was such as I have given
above.  It was owing to him, to his reserve and want of proper
consideration, that Wickham's character had been so misunderstood, and
consequently that he had been received and noticed as he was.  Perhaps
there was some truth in \emph{this}; though I doubt whether \emph{his} reserve, or
\emph{anybody}'S reserve, can be answerable for the event.  But in spite of
all this fine talking, my dear Lizzy, you may rest perfectly assured
that your uncle would never have yielded, if we had not given him
credit for \emph{another} \emph{interest} in the affair.

"When all this was resolved on, he returned again to his friends, who
were still staying at Pemberley; but it was agreed that he should be
in London once more when the wedding took place, and all money matters
were then to receive the last finish.

"I believe I have now told you every thing.  It is a relation which you
tell me is to give you great surprise; I hope at least it will not
afford you any displeasure.  Lydia came to us; and Wickham had
constant admission to the house.  \emph{he} was exactly what he had been,
when I knew him in Hertfordshire; but I would not tell you how little
I was satisfied with her behaviour while she staid with us, if I had
not perceived\index{perceived}, by Jane's letter last Wednesday, that her conduct on
coming home was exactly of a piece with it, and therefore what I now
tell you can give you no fresh pain.  I talked to her repeatedly in
the most serious manner, representing to her all the wickedness of
what she had done, and all the unhappiness she had brought on her
family.  If she heard me, it was by good luck, for I am sure she did
not listen.  I was sometimes quite provoked\index{provoked}, but then I recollected my
dear Elizabeth and Jane, and for their sakes had patience with her.

"Mr. Darcy was punctual\index{punctual} in his return, and as Lydia informed you,
attended the wedding.  He dined with us the next day, and was to leave
town again on Wednesday or Thursday.  Will you be very angry with me,
my dear Lizzy, if I take this opportunity of saying (what I was never
bold enough to say before) how much I like him.  His behaviour to us
has, in every respect, been as pleasing as when we were in Derbyshire.
His understanding and opinions all please me; he wants nothing but a
little more liveliness, and \emph{that}, if he marry \emph{prudently\index{prudently}}, his wife may
teach him.  I thought him very sly;--he hardly ever mentioned your
name.  But slyness seems the fashion\index{fashion}.

"Pray forgive me if I have been very presuming, or at least do not
punish me so far as to exclude me from P.  I shall never be quite
happy till I have been all round the park.  A low phaeton\index{phaeton}, with a nice
little pair of ponies, would be the very thing.

"But I must write no more.  The children have been wanting me this half
hour.

"Yours, very sincerely,

"M. \emph{gardiner}."

The contents of this letter threw Elizabeth into a flutter
of spirits, in which it was difficult to determine whether
pleasure or pain bore the greatest share.  The vague and
unsettled suspicions which uncertainty had produced of what
Mr. Darcy might have been doing to forward\index{forward} her sister's match,
which she had feared to encourage as an exertion\index{exertion} of goodness
too great to be probable\index{probable}, and at the same time dreaded to be
just, from the pain of obligation, were proved beyond their
greatest extent to be true!  He had followed them purposely to
town, he had taken on himself all the trouble and mortification\index{mortification}
attendant on such a research; in which \textbf{supplication}\index{supplication}\stepcounter{footnote}\footnotetext{
\textbf{supplicate} \emph{v.}  (SUHP-plih-kayt)
to humbly request help, especially in prayer.} had been
necessary to a woman whom he must \textbf{abominate}\index{abominate}\stepcounter{footnote}\footnotetext{
\textbf{abominate} \emph{tr.v.}  (uh-BAH-mih-nate)
to find repugnant, especially in a moral sense.} and despise, and
where he was reduced to meet, frequently meet, reason with,
persuade, and finally bribe, the man whom he always most wished
to avoid, and whose very name it was punishment to him to
pronounce.  He had done all this for a girl whom he could
neither regard nor esteem.  Her heart did whisper that he had
done it for her.  But it was a hope shortly checked by other
considerations, and she soon felt that even her vanity\index{vanity} was
insufficient, when required to depend on his affection for her
--for a woman who had already refused him--as able to overcome
a sentiment\index{sentiment} so natural as abhorrence\index{abhorrence} against relationship with
Wickham.  Brother-in-law of Wickham!  Every kind of pride must
revolt from the connection.  He had, to be sure, done much.
She was ashamed to think how much.  But he had given a reason
for his interference, which asked no extraordinary stretch of
belief.  It was reasonable that he should feel he had been wrong;
he had liberality\index{liberality}, and he had the means of exercising it; and
though she would not place herself as his principal\index{principal} inducement\index{inducement},
she could, perhaps, believe that remaining partiality\index{partiality} for her
might assist his endeavours\index{endeavours} in a cause where her peace of mind
must be materially concerned.  It was painful, exceedingly
painful, to know that they were under obligations to a person
who could never receive a return.  They owed the restoration of
Lydia, her character, every thing, to him.  Oh! how heartily\index{heartily}
did she grieve\index{grieve} over every ungracious sensation she had ever
encouraged, every \textbf{saucy}\index{saucy}\stepcounter{footnote}\footnotetext{
\textbf{saucy} \emph{adj.}  (SAW-see)
 improperly forward or bold.} speech she had ever directed towards him.
For herself she was humbled; but she was proud of him.  Proud
that in a cause of compassion and honour, he had been able to get
the better of himself.  She read over her aunt's commendation\index{commendation}
of him again and again.  It was hardly enough; but it pleased
her.  She was even sensible of some pleasure, though mixed with
regret, on finding how steadfastly\index{steadfastly} both she and her uncle had
been persuaded that affection and confidence\index{confidence} subsisted\index{subsisted} between
Mr. Darcy and herself.

She was roused from her seat, and her reflections, by some
one's approach; and before she could strike into another path,
she was overtaken by Wickham.

"I am afraid I interrupt your solitary\index{solitary} ramble\index{ramble}, my dear sister?"
said he, as he joined her.

"You certainly do," she replied with a smile; "but it does not
follow that the interruption must be unwelcome."

"I should be sorry indeed, if it were.  We were always good
friends; and now we are better."

"True.  Are the others coming out?"

"I do not know.  Mrs. Bennet and Lydia are going in the
carriage to Meryton.  And so, my dear sister, I find, from
our uncle and aunt, that you have actually seen Pemberley."

She replied in the affirmative\index{affirmative}.

"I almost envy you the pleasure, and yet I believe it would
be too much for me, or else I could take it in my way to
Newcastle.  And you saw the old housekeeper, I suppose?  Poor
Reynolds, she was always very fond of me.  But of course she
did not mention my name to you."

"Yes, she did."

"And what did she say?"

"That you were gone into the army, and she was afraid had
--not turned out well.  At such a distance as \emph{that}, you
know, things are strangely misrepresented."

"Certainly," he replied, biting his lips.  Elizabeth hoped she
had silenced him; but he soon afterwards said:

"I was surprised to see Darcy in town last month.  We passed
each other several times.  I wonder what he can be doing
there."

"Perhaps preparing for his marriage with Miss de Bourgh," said
Elizabeth.  "It must be something particular, to take him there
at this time of year."

"Undoubtedly.  Did you see him while you were at Lambton?
I thought I understood from the Gardiners that you had."

"Yes; he introduced us to his sister."

"And do you like her?"

"Very much."

"I have heard, indeed, that she is uncommonly improved within
this year or two.  When I last saw her, she was not very
promising.  I am very glad you liked her.  I hope she will turn
out well."

"I dare say she will; she has got over the most trying age."

"Did you go by the village of Kympton?"

"I do not recollect that we did."

"I mention it, because it is the living which I ought to have
had.  A most delightful place!--Excellent Parsonage House!
It would have suited me in every respect."

"How should you have liked making sermons?"

"Exceedingly well.  I should have considered it as part of my
duty, and the exertion\index{exertion} would soon have been nothing.  One ought
not to repine\index{repine};--but, to be sure, it would have been such a
thing for me!  The quiet, the retirement of such a life would
have answered all my ideas of happiness!  But it was not to be.
Did you ever hear Darcy mention the circumstance, when you were
in Kent?"

"I have heard from authority, which I thought \emph{as} \emph{good},
that it was left you conditionally only, and at the will of the
present patron."

"You have.  Yes, there was something in \emph{that}; I told you so
from the first, you may remember."

"I \emph{did} hear, too, that there was a time, when sermon-making
was not so \textbf{palatable}\index{palatable}\stepcounter{footnote}\footnotetext{
\textbf{palatable} \emph{adj.}  (puh-LAT-uh-bl)
acceptable to the taste or mind; eatible.} to you as it seems to be at present; that
you actually declared your resolution\index{resolution} of never taking orders,
and that the business had been compromised accordingly."

"You did! and it was not wholly\index{wholly} without foundation.  You may
remember what I told you on that point, when first we talked
of it."

They were now almost at the door of the house, for she
had walked fast to get rid of him; and unwilling, for her
sister's sake, to provoke\index{provoke} him, she only said in reply, with
a good-humoured smile:

"Come, Mr. Wickham, we are brother and sister, you know.
Do not let\index{let} us quarrel about the past.  In future, I hope we
shall be always of one mind."

She held out her hand; he kissed it with affectionate gallantry\index{gallantry},
though he hardly knew how to look, and they entered the house.



\chapter{53}


Mr. Wickham was so perfectly satisfied with this conversation
that he never again distressed himself, or provoked\index{provoked} his dear
sister Elizabeth, by introducing the subject\index{subject} of it; and she was
pleased to find that she had said enough to keep him quiet.

The day of his and Lydia's departure soon came, and Mrs. Bennet
was forced to submit to a separation, which, as her husband by
no means entered into her scheme of their all going to
Newcastle, was likely to continue at least a twelvemonth.

"Oh! my dear Lydia," she cried, "when shall we meet again?"

"Oh, lord! I don't know.  Not these two or three years,
perhaps."

"Write to me very often, my dear."

"As often as I can.  But you know married women have never much
time for writing.  My sisters may write to \emph{me}.  They will
have nothing else to do."

Mr. Wickham's adieus\index{adieus} were much more affectionate than his
wife's.  He smiled, looked handsome, and said many pretty
things.

"He is as fine a fellow," said Mr. Bennet, as soon as they were
out of the house, "as ever I saw.  He simpers\index{simpers}, and smirks, and
makes love to us all.  I am prodigiously\index{prodigiously} proud of him.  I defy
even Sir William Lucas himself to produce a more valuable
son-in-law."

The loss of her daughter made Mrs. Bennet very dull for
several days.

"I often think," said she, "that there is nothing so bad as
parting with one's friends.  One seems so \textbf{forlorn}\index{forlorn}\stepcounter{footnote}\footnotetext{
\textbf{forlorn} \emph{adj.}  (for-LORN)
 hopelessness; a pitiable feeling of abandonment.} without
them."

"This is the consequence\index{consequence}, you see, Madam, of marrying a daughter,"
said Elizabeth.  "It must make you better satisfied that your
other four are single."

"It is no such thing.  Lydia does not leave me because she is
married, but only because her husband's regiment\index{regiment} happens to be
so far off.  If that had been nearer, she would not have gone
so soon."

But the spiritless condition which this event threw her into
was shortly relieved, and her mind opened again to the
agitation of hope, by an article of news which then began to be
in circulation\index{circulation}.  The housekeeper at Netherfield had received
orders to prepare for the arrival of her master, who was coming
down in a day or two, to shoot there for several weeks.
Mrs. Bennet was quite in the fidgets.  She looked at Jane, and
smiled and shook her head by turns.

"Well, well, and so Mr. Bingley is coming down, sister,"
(for Mrs. Phillips first brought her the news).  "Well, so
much the better.  Not that I care about it, though.  He is
nothing to us, you know, and I am sure \emph{I} never want to
see him again.  But, however, he is very welcome to come
to Netherfield, if he likes it.  And who knows what \emph{may}
happen?  But that is nothing to us.  You know, sister, we
agreed long ago never to mention a word about it.  And so,
is it quite certain he is coming?"

"You may depend on it," replied the other, "for Mrs. Nicholls
was in Meryton last night; I saw her passing by, and went out
myself on purpose to know the truth of it; and she told me that
it was certain true.  He comes down on Thursday at the latest,
very likely on Wednesday.  She was going to the butcher's, she
told me, on purpose to order in some meat on Wednesday, and she
has got three couple of ducks just fit to be killed."

Miss Bennet had not been able to hear of his coming without
changing colour.  It was many months since she had mentioned
his name to Elizabeth; but now, as soon as they were alone
together, she said:

"I saw you look at me to-day, Lizzy, when my aunt told us of
the present report; and I know I appeared distressed.  But
don't imagine it was from any silly cause.  I was only confused
for the moment, because I felt that I \emph{should} be looked at.
I do assure you that the news does not affect me either with
pleasure or pain.  I am glad of one thing, that he comes alone;
because we shall see the less of him.  Not that I am afraid of
\emph{myself}, but I dread other people's remarks."

Elizabeth did not know what to make of it.  Had she not seen
him in Derbyshire, she might have supposed him capable of
coming there with no other view than what was acknowledged; but
she still thought him partial\index{partial} to Jane, and she wavered as to
the greater probability of his coming there \emph{with} his friend's
permission, or being bold enough to come without it.

"Yet it is hard," she sometimes thought, "that this poor man
cannot come to a house which he has legally hired, without
raising all this \textbf{speculation}\index{speculation}\stepcounter{footnote}\footnotetext{
\textbf{speculation} \emph{n.}  (spek-yoo-LAY-shun)
an opinion based on incomplete evidence.}!  I \emph{will} leave him to himself."

In spite of what her sister declared, and really believed to be
her feelings in the expectation of his arrival, Elizabeth could
easily perceive\index{perceive} that her spirits were affected by it.  They
were more disturbed, more unequal, than she had often seen them.

The subject\index{subject} which had been so warmly canvassed\index{canvassed} between their
parents, about a twelvemonth ago, was now brought forward\index{forward}
again.

"As soon as ever Mr. Bingley comes, my dear," said Mrs. Bennet,
"you will wait on him of course."

"No, no.  You forced me into visiting him last year, and
promised, if I went to see him, he should marry one of my
daughters.  But it ended in nothing, and I will not be sent on
a fool's errand again."

His wife represented to him how absolutely necessary such an
attention would be from all the neighbouring gentlemen, on his
returning to Netherfield.

"'Tis an etiquette\index{etiquette} I despise," said he.  "If he wants our
society, let\index{let} him seek it.  He knows where we live.  I will not
spend my hours in running after my neighbours every time they
go away and come back again."

"Well, all I know is, that it will be abominably rude if you do
not wait on him.  But, however, that shan't prevent my asking
him to dine here, I am determined.  We must have Mrs. Long and
the Gouldings soon.  That will make thirteen with ourselves, so
there will be just room at table for him."

Consoled\index{Consoled} by this resolution\index{resolution}, she was the better able to bear
her husband's incivility\index{incivility}; though it was very mortifying\index{mortifying} to know
that her neighbours might all see Mr. Bingley, in consequence\index{consequence}
of it, before \emph{they} did.  As the day of his arrival drew near:

"I begin to be sorry that he comes at all," said Jane to her
sister.  "It would be nothing; I could see him with perfect
indifference\index{indifference}, but I can hardly bear to hear it thus perpetually\index{perpetually}
talked of.  My mother means well; but she does not know, no one
can know, how much I suffer from what she says.  Happy shall I
be, when his stay at Netherfield is over!"

"I wish I could say anything to comfort you," replied Elizabeth;
"but it is wholly\index{wholly} out of my power.  You must feel it; and the
usual satisfaction of preaching patience to a sufferer is denied
me, because you have always so much."

Mr. Bingley arrived.  Mrs. Bennet, through the assistance of
servants, contrived\index{contrived} to have the earliest tidings\index{tidings} of it, that
the period of anxiety and fretfulness on her side might be as
long as it could.  She counted the days that must intervene
before their invitation could be sent; hopeless of seeing
him before.  But on the third morning after his arrival in
Hertfordshire, she saw him, from her dressing-room window,
enter the paddock\index{paddock} and ride towards the house.

Her daughters were eagerly called to partake\index{partake} of her joy.  Jane
resolutely\index{resolutely} kept her place at the table; but Elizabeth, to
satisfy her mother, went to the window--she looked,--she
saw Mr. Darcy with him, and sat down again by her sister.

"There is a gentleman with him, mamma," said Kitty; "who can it be?"

"Some acquaintance or other, my dear, I suppose; I am sure I
do not know."

"La!" replied Kitty, "it looks just like that man that used to
be with him before.  Mr. what's-his-name.  That tall, proud
man."

"Good gracious!  Mr. Darcy!--and so it does, I vow.  Well,
any friend of Mr. Bingley's will always be welcome here, to be
sure; but else I must say that I hate the very sight of him."

Jane looked at Elizabeth with surprise and concern.  She knew
but little of their meeting in Derbyshire, and therefore felt
for the awkwardness which must attend her sister, in seeing him
almost for the first time after receiving his explanatory
letter.  Both sisters were uncomfortable enough.  Each felt
for the other, and of course for themselves; and their mother
talked on, of her dislike of Mr. Darcy, and her resolution\index{resolution}
to be civil to him only as Mr. Bingley's friend, without
being heard by either of them.  But Elizabeth had sources of
uneasiness which could not be suspected by Jane, to whom she
had never yet had courage to shew Mrs. Gardiner's letter, or
to relate her own change of sentiment\index{sentiment} towards him.  To Jane,
he could be only a man whose proposals she had refused, and
whose merit she had undervalued; but to her own more extensive
information, he was the person to whom the whole family were
indebted for the first of benefits, and whom she regarded
herself with an interest, if not quite so tender, at least
as reasonable and just as what Jane felt for Bingley.  Her
astonishment at his coming--at his coming to Netherfield, to
Longbourn, and voluntarily seeking her again, was almost equal
to what she had known on first witnessing his altered behaviour
in Derbyshire.

The colour which had been driven from her face, returned for
half a minute\index{minute} with an additional glow, and a smile of delight
added \textbf{lustre}\index{lustre}\stepcounter{footnote}\footnotetext{
\textbf{lustre} \emph{n.}  (LUS-ter)
the shine of reflected light, especially on porcelain. Also 'luster'.} to her eyes, as she thought for that space of time
that his affection and wishes must still be unshaken.  But she
would not be secure.

"Let\index{Let} me first see how he behaves," said she; "it will then be
early enough for expectation."

She sat intently at work, striving to be composed, and without
daring to lift up her eyes, till anxious curiosity carried them
to the face of her sister as the servant was approaching the
door.  Jane looked a little paler than usual, but more sedate\index{sedate}
than Elizabeth had expected.  On the gentlemen's appearing, her
colour increased; yet she received them with tolerable\index{tolerable} ease,
and with a propriety\index{propriety} of behaviour equally free from any symptom
of resentment\index{resentment} or any unnecessary complaisance\index{complaisance}.

Elizabeth said as little to either as civility\index{civility} would allow, and
sat down again to her work, with an eagerness which it did not
often command.  She had ventured only one glance at Darcy.  He
looked serious, as usual; and, she thought, more as he had been
used to look in Hertfordshire, than as she had seen him at
Pemberley.  But, perhaps he could not in her mother's presence
be what he was before her uncle and aunt.  It was a painful,
but not an improbable\index{improbable}, conjecture\index{conjecture}.

Bingley, she had likewise seen for an instant, and in that
short period saw him looking both pleased and embarrassed.  He
was received by Mrs. Bennet with a degree of civility\index{civility} which
made her two daughters ashamed, especially when contrasted with
the cold and ceremonious\index{ceremonious} politeness of her curtsey and address\index{address}
to his friend.

Elizabeth, particularly, who knew that her mother owed to the
latter the preservation of her favourite daughter from
\textbf{irremediable}\index{irremediable}\stepcounter{footnote}\footnotetext{
\textbf{irremediable} \emph{adj.}  (ihr-ruh-MEE-dih-yuhl)
impossible to remedy, correct, or redress.} infamy\index{infamy}, was hurt and distressed to a most painful
degree by a distinction so ill applied.

Darcy, after inquiring of her how Mr. and Mrs. Gardiner did, a
question which she could not answer without confusion, said
scarcely anything.  He was not seated by her; perhaps that
was the reason of his silence; but it had not been so in
Derbyshire.  There he had talked to her friends, when he could
not to herself.  But now several minutes\index{minutes} elapsed without
bringing the sound of his voice; and when occasionally, unable
to resist the impulse of curiosity, she raised he eyes to his
face, she as often found him looking at Jane as at herself, and
frequently on no object but the ground.  More thoughtfulness
and less anxiety to please, than when they last met, were
plainly expressed.  She was disappointed, and angry with
herself for being so.

"Could I expect it to be otherwise!" said she.  "Yet why did
he come?"

She was in no humour for conversation with anyone but himself;
and to him she had hardly courage to speak.

She inquired after his sister, but could do no more.

"It is a long time, Mr. Bingley, since you went away," said
Mrs. Bennet.

He readily\index{readily} agreed to it.

"I began to be afraid you would never come back again.  People
\emph{did} say you meant to quit the place entirely at Michaelmas\index{Michaelmas};
but, however, I hope it is not true.  A great many changes have
happened in the neighbourhood, since you went away.  Miss Lucas
is married and settled.  And one of my own daughters.  I suppose
you have heard of it; indeed, you must have seen it in the
papers.  It was in The Times and The Courier, I know; though
it was not put in as it ought to be.  It was only said, 'Lately,
George Wickham, Esq. to Miss Lydia Bennet,' without there being
a syllable said of her father, or the place where she lived, or
anything.  It was my brother Gardiner's drawing up too, and I
wonder how he came to make such an awkward business of it.  Did
you see it?"

Bingley replied that he did, and made his congratulations.
Elizabeth dared not lift up her eyes.  How Mr. Darcy looked,
therefore, she could not tell.

"It is a delightful thing, to be sure, to have a daughter
well married," continued her mother, "but at the same time,
Mr. Bingley, it is very hard to have her taken such a way from
me.  They are gone down to Newcastle, a place quite northward,
it seems, and there they are to stay I do not know how long.
His regiment\index{regiment} is there; for I suppose you have heard of his
leaving the ----shire, and of his being gone into the regulars\index{regulars}.
Thank Heaven! he has \emph{some} friends, though perhaps not so
many as he deserves."

Elizabeth, who knew this to be levelled at Mr. Darcy, was
in such misery of shame, that she could hardly keep her seat.
It drew from her, however, the exertion\index{exertion} of speaking, which
nothing else had so effectually\index{effectually} done before; and she asked
Bingley whether he meant to make any stay in the country at
present.  A few weeks, he believed.

"When you have killed all your own birds, Mr. Bingley,"
said her mother, "I beg you will come here, and shoot as
many as you please on Mr. Bennet's manor.  I am sure he
will be \textbf{vastly}\index{vastly}\stepcounter{footnote}\footnotetext{
\textbf{vast} \emph{adj.} \textbf{-ly}  (VAST)
 huge, expansive, very large.} happy to oblige you, and will save all the
best of the \textbf{covies}\index{covies}\stepcounter{footnote}\footnotetext{
\textbf{} \emph{}  ()
} for you."

Elizabeth's misery increased, at such unnecessary, such
officious\index{officious} attention!  Were the same fair prospect\index{prospect} to arise at
present as had flattered them a year ago, every thing, she was
persuaded, would be hastening to the same vexatious\index{vexatious} conclusion.
At that instant, she felt that years of happiness could not
make Jane or herself amends\index{amends} for moments of such painful
confusion.

"The first wish of my heart," said she to herself, "is never
more to be in company with either of them.  Their society can
afford no pleasure that will atone\index{atone} for such wretchedness as
this!  Let\index{Let} me never see either one or the other again!"

Yet the misery, for which years of happiness were to offer no
compensation, received soon afterwards material relief, from
observing how much the beauty of her sister re-kindled the
admiration of her former lover.  When first he came in, he had
spoken to her but little; but every five minutes\index{minutes} seemed to be
giving her more of his attention.  He found her as handsome as
she had been last year; as good natured, and as unaffected\index{unaffected},
though not quite so chatty.  Jane was anxious that no difference
should be perceived\index{perceived} in her at all, and was really persuaded that
she talked as much as ever.  But her mind was so busily engaged,
that she did not always know when she was silent.

When the gentlemen rose to go away, Mrs. Bennet was mindful of
her intended civility\index{civility}, and they were invited and engaged to
dine at Longbourn in a few days time.

"You are quite a visit in my debt, Mr. Bingley," she added,
"for when you went to town last winter, you promised to take
a family dinner with us, as soon as you returned.  I have not
forgot, you see; and I assure you, I was very much disappointed
that you did not come back and keep your engagement."

Bingley looked a little silly at this reflection, and said
something of his concern at having been prevented by business.
They then went away.

Mrs. Bennet had been strongly inclined to ask them to stay and
dine there that day; but, though she always kept a very good
table, she did not think anything less than two courses could
be good enough for a man on whom she had such anxious designs,
or satisfy the appetite and pride of one who had ten thousand a
year.



\chapter{54}


As soon as they were gone, Elizabeth walked out to recover
her spirits; or in other words, to dwell without interruption
on those subjects\index{subjects} that must deaden them more.  Mr. Darcy's
behaviour astonished and vexed\index{vexed} her.

"Why, if he came only to be silent, grave\index{grave}, and indifferent\index{indifferent},"
said she, "did he come at all?"

She could settle it in no way that gave her pleasure.

"He could be still amiable\index{amiable}, still pleasing, to my uncle and
aunt, when he was in town; and why not to me?  If he fears me,
why come hither?  If he no longer cares for me, why silent?
Teasing, teasing, man!  I will think no more about him."

Her resolution\index{resolution} was for a short time involuntarily kept by
the approach of her sister, who joined her with a cheerful
look, which showed her better satisfied with their visitors,
than Elizabeth.

"Now," said she, "that this first meeting is over, I feel
perfectly easy.  I know my own strength, and I shall never be
embarrassed again by his coming.  I am glad he dines here on
Tuesday.  It will then be publicly seen that, on both sides,
we meet only as common and indifferent\index{indifferent} acquaintance."

"Yes, very indifferent\index{indifferent} indeed," said Elizabeth, laughingly.
"Oh, Jane, take care."

"My dear Lizzy, you cannot think me so weak, as to be in danger
now?"

"I think you are in very great danger of making him as much
in love with you as ever."



They did not see the gentlemen again till Tuesday; and
Mrs. Bennet, in the meanwhile, was giving way to all the
happy schemes, which the good humour and common politeness
of Bingley, in half an hour's visit, had revived.

On Tuesday there was a large party assembled at Longbourn;
and the two who were most anxiously expected, to the credit of
their punctuality as sportsmen, were in very good time.  When
they repaired to the dining-room, Elizabeth eagerly watched to
see whether Bingley would take the place, which, in all their
former parties, had belonged to him, by her sister.  Her
prudent\index{prudent} mother, occupied by the same ideas, forbore to invite
him to sit by herself.  On entering the room, he seemed to
hesitate; but Jane happened to look round, and happened to
smile: it was decided.  He placed himself by her.

Elizabeth, with a triumphant\index{triumphant} sensation, looked towards his
friend.  He bore it with noble indifference\index{indifference}, and she would have
imagined that Bingley had received his sanction\index{sanction} to be happy,
had she not seen his eyes likewise turned towards Mr. Darcy,
with an expression of half-laughing alarm.

His behaviour to her sister was such, during dinner time, as
showed an admiration of her, which, though more guarded than
formerly, persuaded Elizabeth, that if left wholly\index{wholly} to himself,
Jane's happiness, and his own, would be speedily secured.
Though she dared not depend upon the consequence\index{consequence}, she yet
received pleasure from observing his behaviour.  It gave her
all the animation that her spirits could boast; for she was in
no cheerful humour.  Mr. Darcy was almost as far from her as
the table could divide them.  He was on one side of her mother.
She knew how little such a situation would give pleasure to
either, or make either appear to advantage.  She was not near
enough to hear any of their discourse\index{discourse}, but she could see how
seldom they spoke to each other, and how formal and cold was
their manner whenever they did.  Her mother's ungraciousness,
made the sense of what they owed him more painful to Elizabeth's
mind; and she would, at times, have given anything to be
privileged to tell him that his kindness was neither unknown
nor unfelt by the whole of the family.

She was in hopes that the evening would afford some opportunity
of bringing them together; that the whole of the visit would
not pass away without enabling them to enter into something
more of conversation than the mere ceremonious\index{ceremonious} salutation
attending his entrance.  Anxious and uneasy, the period which
passed in the drawing-room, before the gentlemen came, was
wearisome and dull to a degree that almost made her uncivil\index{uncivil}.
She looked forward\index{forward} to their entrance as the point on which all
her chance of pleasure for the evening must depend.

"If he does not come to me, \emph{then}," said she, "I shall give
him up for ever."

The gentlemen came; and she thought he looked as if he would
have answered her hopes; but, alas! the ladies had crowded
round the table, where Miss Bennet was making tea, and
Elizabeth pouring out the coffee, in so close a \textbf{confederacy}\index{confederacy}\stepcounter{footnote}\footnotetext{
\textbf{} \emph{}  ()
}
that there was not a single vacancy near her which would admit
of a chair.  And on the gentlemen's approaching, one of the
girls moved closer to her than ever, and said, in a whisper:

"The men shan't come and part us, I am determined.  We want
none of them; do we?"

Darcy had walked away to another part of the room.  She
followed him with her eyes, envied everyone to whom he spoke,
had scarcely patience enough to help anybody to coffee; and
then was enraged against herself for being so silly!

"A man who has once been refused!  How could I ever be foolish
enough to expect a renewal of his love?  Is there one among the
sex, who would not protest against such a weakness as a second
proposal to the same woman?  There is no indignity\index{indignity} so \textbf{abhorrent}\index{abhorrent}\stepcounter{footnote}\footnotetext{
\textbf{abhorrent} \emph{adj.}  (ab-HAWR-rent)
offensive to the mind; horrible or disgusting.}
to their feelings!"

She was a little revived, however, by his bringing back his
coffee cup himself; and she seized the opportunity of saying:

"Is your sister at Pemberley still?"

"Yes, she will remain there till Christmas."

"And quite alone?  Have all her friends left her?"

"Mrs. Annesley is with her.  The others have been gone on to
Scarborough, these three weeks."

She could think of nothing more to say; but if he wished to
converse with her, he might have better success.  He stood by
her, however, for some minutes\index{minutes}, in silence; and, at last, on
the young lady's whispering to Elizabeth again, he walked away.

When the tea-things were removed, and the card-tables placed,
the ladies all rose, and Elizabeth was then hoping to be soon
joined by him, when all her views were overthrown by seeing him
fall a victim to her mother's \textbf{rapacity}\index{rapacity}\stepcounter{footnote}\footnotetext{
\textbf{rapacious} \emph{adj.}  (ruh-PAY-shuhs)
insatiable desire for wealth.} for whist\index{whist} players, and
in a few moments after seated with the rest of the party.  She
now lost every expectation of pleasure.  They were confined for
the evening at different tables, and she had nothing to hope,
but that his eyes were so often turned towards her side of the
room, as to make him play as unsuccessfully as herself.

Mrs. Bennet had designed to keep the two Netherfield gentlemen
to supper; but their carriage was unluckily ordered before any
of the others, and she had no opportunity of detaining them.

"Well girls," said she, as soon as they were left to themselves,
"What say you to the day?  I think every thing has passed off
uncommonly well, I assure you.  The dinner was as well dressed
as any I ever saw.  The \textbf{venison}\index{venison}\stepcounter{footnote}\footnotetext{
\textbf{venison} \emph{n.}  (VEH-nih-suhn)
meat from a deer used as food.} was roasted to a turn--and
everybody said they never saw so fat a \textbf{haunch}\index{haunch}\stepcounter{footnote}\footnotetext{
\textbf{haunch} \emph{n.}  (HAWNCH)
the hip, buttock, and upper thigh.}.  The soup was
fifty times better than what we had at the Lucases' last week;
and even Mr. Darcy acknowledged, that the \textbf{partridges}\index{partridges}\stepcounter{footnote}\footnotetext{
\textbf{partridge} \emph{n.}  (PAR-tridge)
Any of several plump European game birds.} were
remarkably well done; and I suppose he has two or three French
cooks at least.  And, my dear Jane, I never saw you look in
greater beauty.  Mrs. Long said so too, for I asked her whether
you did not.  And what do you think she said besides? 'Ah! Mrs.
Bennet, we shall have her at Netherfield at last.'  She did
indeed.  I do think Mrs. Long is as good a creature as ever
lived--and her nieces are very pretty behaved girls, and not
at all handsome: I like them prodigiously\index{prodigiously}."

Mrs. Bennet, in short, was in very great spirits; she had seen
enough of Bingley's behaviour to Jane, to be convinced that she
would get him at last; and her expectations of advantage to her
family, when in a happy humour, were so far beyond reason, that
she was quite disappointed at not seeing him there again the
next day, to make his proposals.

"It has been a very agreeable day," said Miss Bennet to
Elizabeth.  "The party seemed so well selected, so suitable
one with the other.  I hope we may often meet again."

Elizabeth smiled.

"Lizzy, you must not do so.  You must not suspect me.  It
mortifies me.  I assure you that I have now learnt to enjoy
his conversation as an agreeable and sensible young man,
without having a wish beyond it.  I am perfectly satisfied,
from what his manners now are, that he never had any design
of engaging my affection.  It is only that he is blessed
with greater sweetness of address\index{address}, and a stronger desire of
generally pleasing, than any other man."

"You are very cruel," said her sister, "you will not let\index{let} me
smile, and are provoking me to it every moment."

"How hard it is in some cases to be believed!"

"And how impossible in others!"

"But why should you wish to persuade me that I feel more than I
acknowledge?"

"That is a question which I hardly know how to answer.  We all
love to instruct, though we can teach only what is not worth
knowing.  Forgive me; and if you persist in indifference\index{indifference}, do
not make me your confidante\index{confidante}."



\chapter{55}


A few days after this visit, Mr. Bingley called again, and
alone.  His friend had left him that morning for London, but
was to return home in ten days time.  He sat with them above an
hour, and was in remarkably good spirits.  Mrs. Bennet invited
him to dine with them; but, with many expressions of concern,
he confessed himself engaged elsewhere.

"Next time you call," said she, "I hope we shall be more
lucky."

He should be particularly happy at any time, etc. etc.; and if
she would give him leave, would take an early opportunity of
waiting on them.

"Can you come to-morrow?"

Yes, he had no engagement at all for to-morrow; and her
invitation was accepted with alacrity\index{alacrity}.

He came, and in such very good time that the ladies were none
of them dressed.  In ran Mrs. Bennet to her daughter's room, in
her dressing gown, and with her hair half finished, crying out:

"My dear Jane, make haste and hurry down.  He is come--Mr.
Bingley is come.  He is, indeed.  Make haste, make haste.
Here, Sarah, come to Miss Bennet this moment, and help her
on with her gown.  Never mind Miss Lizzy's hair."

"We will be down as soon as we can," said Jane; "but I dare say
Kitty is forwarder than either of us, for she went up stairs
half an hour ago."

"Oh! hang Kitty! what has she to do with it?  Come be quick,
be quick!  Where is your sash, my dear?"

But when her mother was gone, Jane would not be prevailed\index{prevailed} on to
go down without one of her sisters.

The same anxiety to get them by themselves was visible again
in the evening.  After tea, Mr. Bennet retired to the library,
as was his custom, and Mary went up stairs to her instrument.
Two obstacles of the five being thus removed, Mrs. Bennet
sat looking and winking at Elizabeth and Catherine for a
considerable time, without making any impression on them.
Elizabeth would not observe her; and when at last Kitty did,
she very innocently said, "What is the matter mamma?  What do
you keep winking at me for?  What am I to do?"

"Nothing child, nothing.  I did not wink at you."  She then sat
still five minutes\index{minutes} longer; but unable to waste such a precious
occasion, she suddenly got up, and saying to Kitty, "Come here,
my love, I want to speak to you," took her out of the room.
Jane instantly gave a look at Elizabeth which spoke her
distress at such \textbf{premeditation}\index{premeditation}\stepcounter{footnote}\footnotetext{
\textbf{premeditation} \emph{n.}  (prih-meh-dih-TAY-shun)
planning or plotting in advance of acting.}, and her entreaty\index{entreaty} that \emph{she}
would not give in to it.  In a few minutes\index{minutes}, Mrs. Bennet
half-opened the door and called out:

"Lizzy, my dear, I want to speak with you."

Elizabeth was forced to go.

"We may as well leave them by themselves you know;" said her
mother, as soon as she was in the hall.  "Kitty and I are going
upstairs to sit in my dressing-room."

Elizabeth made no attempt to reason with her mother, but
remained quietly in the hall, till she and Kitty were out of
sight, then returned into the drawing-room.

Mrs. Bennet's schemes for this day were \textbf{ineffectual}\index{ineffectual}\stepcounter{footnote}\footnotetext{
\textbf{ineffectual} \emph{adj.}  (ihn-ee-FEHK-choo-uhl)
not producing an intended effect.}.  Bingley
was every thing that was charming, except the professed lover
of her daughter.  His ease and cheerfulness rendered\index{rendered} him a
most agreeable addition to their evening party; and he bore
with the ill-judged officiousness of the mother, and heard all
her silly remarks with a forbearance\index{forbearance} and command of countenance\index{countenance}
particularly grateful to the daughter.

He scarcely needed an invitation to stay supper; and before he
went away, an engagement was formed, chiefly through his own
and Mrs. Bennet's means, for his coming next morning to shoot
with her husband.

After this day, Jane said no more of her indifference\index{indifference}.
Not a word passed between the sisters concerning Bingley;
but Elizabeth went to bed in the happy belief that all must
speedily be concluded, unless Mr. Darcy returned within the
stated time.  Seriously, however, she felt tolerably persuaded
that all this must have taken place with that gentleman's
concurrence\index{concurrence}.

Bingley was punctual\index{punctual} to his appointment; and he and Mr. Bennet
spent the morning together, as had been agreed on.  The latter
was much more agreeable than his companion expected.  There was
nothing of presumption\index{presumption} or folly\index{folly} in Bingley that could provoke\index{provoke}
his ridicule, or disgust him into silence; and he was more
communicative, and less \textbf{eccentric}\index{eccentric}\stepcounter{footnote}\footnotetext{
\textbf{} \emph{}  ()
}, than the other had ever seen
him.  Bingley of course returned with him to dinner; and in the
evening Mrs. Bennet's invention was again at work to get every
body away from him and her daughter.  Elizabeth, who had a
letter to write, went into the breakfast room for that purpose
soon after tea; for as the others were all going to sit down to
cards, she could not be wanted to counteract her mother's
schemes.

But on returning to the drawing-room, when her letter was
finished, she saw, to her infinite surprise, there was
reason to fear that her mother had been too ingenious\index{ingenious} for
her.  On opening the door, she perceived\index{perceived} her sister and
Bingley standing together over the \textbf{hearth}\index{hearth}\stepcounter{footnote}\footnotetext{
\textbf{hearth} \emph{n.}  (HARTH)
 the floor of a fireplace, often extending out into the room.}, as if engaged in
earnest\index{earnest} conversation; and had this led to no suspicion, the
faces of both, as they hastily turned round and moved away
from each other, would have told it all.  Their situation
was awkward enough; but \emph{her}'S she thought was still worse.
Not a syllable was uttered by either; and Elizabeth was on
the point of going away again, when Bingley, who as well as
the other had sat down, suddenly rose, and whispering a few
words to her sister, ran out of the room.

Jane could have no reserves from Elizabeth, where confidence\index{confidence}
would give pleasure; and instantly embracing her, acknowledged,
with the liveliest emotion, that she was the happiest creature
in the world.

"'Tis too much!" she added, "by far too much.  I do not
deserve it.  Oh! why is not everybody as happy?"

Elizabeth's congratulations were given with a sincerity, a
warmth, a delight, which words could but poorly express.  Every
sentence of kindness was a fresh source of happiness to Jane.
But she would not allow herself to stay with her sister, or say
half that remained to be said for the present.

"I must go instantly to my mother;" she cried.  "I would not on
any account trifle with her affectionate solicitude\index{solicitude}; or allow
her to hear it from anyone but myself.  He is gone to my
father already.  Oh! Lizzy, to know that what I have to relate
will give such pleasure to all my dear family! how shall I
bear so much happiness!"

She then hastened away to her mother, who had purposely broken
up the card party, and was sitting up stairs with Kitty.

Elizabeth, who was left by herself, now smiled at the rapidity
and ease with which an affair was finally settled, that had
given them so many previous months of suspense and vexation\index{vexation}.

"And this," said she, "is the end of all his friend's anxious
circumspection\index{circumspection}! of all his sister's falsehood and contrivance\index{contrivance}!
the happiest, wisest, most reasonable end!"

In a few minutes\index{minutes} she was joined by Bingley, whose conference
with her father had been short and to the purpose.

"Where is your sister?" said he hastily, as he opened the door.

"With my mother up stairs.  She will be down in a moment,
I dare say."

He then shut the door, and, coming up to her, claimed the good
wishes and affection of a sister.  Elizabeth honestly and
heartily\index{heartily} expressed her delight in the prospect\index{prospect} of their
relationship.  They shook hands with great cordiality; and
then, till her sister came down, she had to listen to all he
had to say of his own happiness, and of Jane's perfections;
and in spite of his being a lover, Elizabeth really believed
all his expectations of felicity\index{felicity} to be rationally\index{rationally} founded,
because they had for basis the excellent understanding, and
super-excellent disposition\index{disposition} of Jane, and a general similarity
of feeling and taste between her and himself.

It was an evening of no common delight to them all; the
satisfaction of Miss Bennet's mind gave a glow of such sweet
animation to her face, as made her look handsomer than ever.
Kitty \textbf{simpered}\index{simpered}\stepcounter{footnote}\footnotetext{
\textbf{simper} \emph{v.} \textbf{-ed}  (SIM-per)
 A constrained, self-conscious, or silly smile; a smirk.} and smiled, and hoped her turn was coming soon.
Mrs. Bennet could not give her consent or speak her approbation\index{approbation}
in terms warm enough to satisfy her feelings, though she talked
to Bingley of nothing else for half an hour; and when Mr.
Bennet joined them at supper, his voice and manner plainly
showed how really happy he was.

Not a word, however, passed his lips in allusion\index{allusion} to it, till
their visitor took his leave for the night; but as soon as he
was gone, he turned to his daughter, and said:

"Jane, I congratulate you.  You will be a very happy woman."

Jane went to him instantly, kissed him, and thanked him for his
goodness.

"You are a good girl;" he replied, "and I have great pleasure
in thinking you will be so happily settled.  I have not a doubt
of your doing very well together.  Your tempers are by no means
unlike.  You are each of you so complying\index{complying}, that nothing will
ever be resolved on; so easy, that every servant will cheat
you; and so generous, that you will always exceed your income."

"I hope not so.  Imprudence\index{Imprudence} or thoughtlessness in money matters
would be unpardonable in me."

"Exceed their income!  My dear Mr. Bennet," cried his wife,
"what are you talking of?  Why, he has four or five thousand a
year, and very likely more."  Then addressing\index{addressing} her daughter,
"Oh! my dear, dear Jane, I am so happy!  I am sure I shan't
get a wink of sleep all night.  I knew how it would be.  I
always said it must be so, at last.  I was sure you could not
be so beautiful for nothing!  I remember, as soon as ever I saw
him, when he first came into Hertfordshire last year, I thought
how likely it was that you should come together.  Oh! he is
the handsomest young man that ever was seen!"

Wickham, Lydia, were all forgotten.  Jane was beyond competition
her favourite child.  At that moment, she cared for no other.  Her
younger sisters soon began to make interest with her for objects
of happiness which she might in future be able to dispense.

Mary petitioned\index{petitioned} for the use of the library at Netherfield; and
Kitty begged very hard for a few balls there every winter.

Bingley, from this time, was of course a daily visitor at
Longbourn; coming frequently before breakfast, and always
remaining till after supper; unless when some barbarous\index{barbarous}
neighbour, who could not be enough detested, had given him
an invitation to dinner which he thought himself obliged to
accept.

Elizabeth had now but little time for conversation with her
sister; for while he was present, Jane had no attention to
bestow\index{bestow} on anyone else; but she found herself considerably
useful to both of them in those hours of separation that must
sometimes occur.  In the absence of Jane, he always attached
himself to Elizabeth, for the pleasure of talking of her; and
when Bingley was gone, Jane constantly sought the same means
of relief.

"He has made me so happy," said she, one evening, "by telling
me that he was totally ignorant of my being in town last
spring!  I had not believed it possible."

"I suspected as much," replied Elizabeth.  "But how did he
account for it?"

"It must have been his sister's doing.  They were certainly no
friends to his acquaintance with me, which I cannot wonder at,
since he might have chosen so much more advantageously in many
respects.  But when they see, as I trust they will, that their
brother is happy with me, they will learn to be contented, and
we shall be on good terms again; though we can never be what we
once were to each other."

"That is the most unforgiving speech," said Elizabeth, "that
I ever heard you utter.  Good girl!  It would vex\index{vex} me, indeed,
to see you again the dupe\index{dupe} of Miss Bingley's pretended regard."

"Would you believe it, Lizzy, that when he went to town last
November, he really loved me, and nothing but a persuasion of
\emph{my} being indifferent\index{indifferent} would have prevented his coming down
again!"

"He made a little mistake to be sure; but it is to the credit
of his modesty."

This naturally introduced a panegyric\index{panegyric} from Jane on his
diffidence\index{diffidence}, and the little value he put on his own good
qualities.  Elizabeth was pleased to find that he had not
betrayed the interference of his friend; for, though Jane had
the most generous and forgiving heart in the world, she knew
it was a circumstance which must prejudice\index{prejudice} her against him.

"I am certainly the most fortunate creature that ever existed!"
cried Jane.  "Oh! Lizzy, why am I thus singled from my family,
and blessed above them all!  If I could but see \emph{you} as happy!
If there \emph{were} but such another man for you!"

"If you were to give me forty such men, I never could be so
happy as you.  Till I have your disposition\index{disposition}, your goodness,
I never can have your happiness.  No, no, let\index{let} me shift for
myself; and, perhaps, if I have very good luck, I may meet
with another Mr. Collins in time."

The situation of affairs in the Longbourn family could not be
long a secret.  Mrs. Bennet was privileged to whisper it to
Mrs. Phillips, and she ventured, without any permission, to do
the same by all her neighbours in Meryton.

The Bennets were speedily pronounced to be the luckiest family
in the world, though only a few weeks before, when Lydia had
first run away, they had been generally proved to be marked out
for misfortune.



\chapter{56}


One morning, about a week after Bingley's engagement with
Jane had been formed, as he and the females of the family
were sitting together in the dining-room, their attention
was suddenly drawn to the window, by the sound of a carriage;
and they perceived\index{perceived} a chaise\index{chaise} and four driving up the lawn.
It was too early in the morning for visitors, and besides, the
\textbf{equipage}\index{equipage}\stepcounter{footnote}\footnotetext{
\textbf{equipage} \emph{n.}  (ih-KWIH-pehj)
a wheeled vehicle drawn by horses.} did not answer to that of any of their neighbours.
The horses were post\index{post}; and neither the carriage, nor the livery\index{livery}
of the servant who preceded it, were familiar to them.  As it
was certain, however, that somebody was coming, Bingley
instantly prevailed\index{prevailed} on Miss Bennet to avoid the confinement of
such an intrusion, and walk away with him into the shrubbery\index{shrubbery}.
They both set off, and the conjectures\index{conjectures} of the remaining three
continued, though with little satisfaction, till the door was
thrown open and their visitor entered.  It was Lady Catherine
de Bourgh.

They were of course all intending to be surprised; but their
astonishment was beyond their expectation; and on the part of
Mrs. Bennet and Kitty, though she was perfectly unknown to
them, even inferior\index{inferior} to what Elizabeth felt.

She entered the room with an air more than usually ungracious,
made no other reply to Elizabeth's salutation than a slight
inclination\index{inclination} of the head, and sat down without saying a word.
Elizabeth had mentioned her name to her mother on her ladyship's
entrance, though no request of introduction had been made.

Mrs. Bennet, all amazement, though flattered by having a
guest of such high importance, received her with the utmost
politeness.  After sitting for a moment in silence, she said
very stiffly to Elizabeth,

"I hope you are well, Miss Bennet.  That lady, I suppose,
is your mother."

Elizabeth replied very concisely that she was.

"And \emph{that} I suppose is one of your sisters."

"Yes, madam," said Mrs. Bennet, delighted to speak to a Lady
Catherine.  "She is my youngest girl but one.  My youngest of
all is lately married, and my eldest is somewhere about the
grounds, walking with a young man who, I believe, will soon
become a part of the family."

"You have a very small park here," returned Lady Catherine
after a short silence.

"It is nothing in comparison of Rosings, my lady, I dare say;
but I assure you it is much larger than Sir William Lucas's."

"This must be a most inconvenient sitting room for the evening,
in summer; the windows are full west."

Mrs. Bennet assured her that they never sat there after dinner,
and then added:

"May I take the liberty of asking your ladyship whether you
left Mr. and Mrs. Collins well."

"Yes, very well.  I saw them the night before last."

Elizabeth now expected that she would produce a letter for
her from Charlotte, as it seemed the only probable\index{probable} motive for
her calling.  But no letter appeared, and she was completely
puzzled.

Mrs. Bennet, with great civility\index{civility}, begged her ladyship to take
some refreshment; but Lady Catherine very resolutely\index{resolutely}, and not
very politely, declined eating anything; and then, rising up,
said to Elizabeth,

"Miss Bennet, there seemed to be a prettyish kind of a little
wilderness on one side of your lawn.  I should be glad to take
a turn in it, if you will favour me with your company."

"Go, my dear," cried her mother, "and show her ladyship about
the different walks.  I think she will be pleased with the
\textbf{hermitage}\index{hermitage}\stepcounter{footnote}\footnotetext{
\textbf{Hermitage} \emph{n.}  (HER-mih-tehdge)
A full-bodied wine made in southeastern France.}."

Elizabeth obeyed, and running into her own room for her
parasol, attended her noble guest downstairs.  As they passed
through the hall, Lady Catherine opened the doors into the
dining-parlour and drawing-room, and pronouncing them, after
a short survey, to be decent looking rooms, walked on.

Her carriage remained at the door, and Elizabeth saw that her
waiting-woman was in it.  They proceeded in silence along the
gravel walk that led to the copse\index{copse}; Elizabeth was determined to
make no effort for conversation with a woman who was now more
than usually insolent\index{insolent} and disagreeable.

"How could I ever think her like her nephew?" said she, as she
looked in her face.

As soon as they entered the copse\index{copse}, Lady Catherine began in the
following manner:--

"You can be at no loss, Miss Bennet, to understand the reason
of my journey hither.  Your own heart, your own conscience\index{conscience},
must tell you why I come."

Elizabeth looked with unaffected\index{unaffected} astonishment.

"Indeed, you are mistaken, Madam.  I have not been at all able
to account for the honour of seeing you here."

"Miss Bennet," replied her ladyship, in an angry tone, "you
ought to know, that I am not to be trifled with.  But however
insincere \emph{you} may choose to be, you shall not find \emph{me} so.
My character has ever been celebrated for its sincerity and
frankness, and in a cause of such moment as this, I shall
certainly not depart from it.  A report of a most alarming
nature reached me two days ago.  I was told that not only your
sister was on the point of being most advantageously married,
but that you, that Miss Elizabeth Bennet, would, in all
likelihood, be soon afterwards united to my nephew, my own
nephew, Mr. Darcy.  Though I \emph{know} it must be a scandalous\index{scandalous}
falsehood, though I would not injure him so much as to suppose
the truth of it possible, I instantly resolved on setting off
for this place, that I might make my sentiments\index{sentiments} known to you."

"If you believed it impossible to be true," said Elizabeth,
colouring with astonishment and disdain, "I wonder you took the
trouble of coming so far.  What could your ladyship propose by
it?"

"At once to insist upon having such a report universally
contradicted."

"Your coming to Longbourn, to see me and my family," said
Elizabeth coolly, "will be rather a confirmation of it; if,
indeed, such a report is in existence."

"If!  Do you then pretend to be ignorant of it?  Has it not
been \textbf{industriously}\index{industriously}\stepcounter{footnote}\footnotetext{
\textbf{industrious} \emph{adj.} \textbf{-ly}  (ihn-DUH-tree-uhs)
busy, dilligent, hard working, and productive.} circulated by yourselves?  Do you not know
that such a report is spread abroad?"

"I never heard that it was."

"And can you likewise declare, that there is no foundation
for it?"

"I do not pretend to possess equal frankness with your ladyship.
You may ask questions which I shall not choose to answer."

"This is not to be borne\index{borne}.  Miss Bennet, I insist on being
satisfied.  Has he, has my nephew, made you an offer of
marriage?"

"Your ladyship has declared it to be impossible."

"It ought to be so; it must be so, while he retains the use of
his reason.  But your arts and \textbf{allurements}\index{allurements}\stepcounter{footnote}\footnotetext{
\textbf{} \emph{}  ()
} may, in a moment
of infatuation, have made him forget what he owes to himself
and to all his family.  You may have drawn him in."

"If I have, I shall be the last person to confess it."

"Miss Bennet, do you know who I am?  I have not been accustomed
to such language as this.  I am almost the nearest relation he has
in the world, and am entitled to know all his dearest concerns."

"But you are not entitled to know mine; nor will such behaviour
as this, ever induce\index{induce} me to be explicit\index{explicit}."

"Let\index{Let} me be rightly understood.  This match, to which you have
the presumption\index{presumption} to \textbf{aspire}\index{aspire}\stepcounter{footnote}\footnotetext{
\textbf{aspire} \emph{intr.v.}  (as-SPIRE)
to have an ambitious plan or a lofty goal.}, can never take place.  No, never.
Mr. Darcy is engaged to my daughter.  Now what have you to say?"

"Only this; that if he is so, you can have no reason to suppose
he will make an offer to me."

Lady Catherine hesitated for a moment, and then replied:

"The engagement between them is of a peculiar\index{peculiar} kind.  From their
infancy, they have been intended for each other.  It was the
favourite wish of \emph{his} mother, as well as of her's.  While in
their cradles, we planned the union: and now, at the moment
when the wishes of both sisters would be accomplished in their
marriage, to be prevented by a young woman of inferior\index{inferior} birth,
of no importance in the world, and wholly\index{wholly} unallied to the
family!  Do you pay no regard to the wishes of his friends?
To his \textbf{tacit}\index{tacit}\stepcounter{footnote}\footnotetext{
\textbf{tacit} \emph{adj.}  (TAS-siht)
communicated but not said.} engagement with Miss de Bourgh?  Are you lost to
every feeling of propriety\index{propriety} and delicacy\index{delicacy}?  Have you not heard
me say that from his earliest hours he was destined for his
cousin?"

"Yes, and I had heard it before.  But what is that to me?  If
there is no other objection to my marrying your nephew, I shall
certainly not be kept from it by knowing that his mother and
aunt wished him to marry Miss de Bourgh.  You both did as much
as you could in planning the marriage.  Its completion depended
on others.  If Mr. Darcy is neither by honour nor inclination\index{inclination}
confined to his cousin, why is not he to make another choice?
And if I am that choice, why may not I accept him?"

"Because honour, decorum\index{decorum}, prudence\index{prudence}, nay, interest, forbid it.
Yes, Miss Bennet, interest; for do not expect to be noticed
by his family or friends, if you wilfully act against the
inclinations\index{inclinations} of all.  You will be censured\index{censured}, slighted, and
despised, by everyone connected with him.  Your alliance\index{alliance} will
be a disgrace; your name will never even be mentioned by any
of us."

"These are heavy misfortunes," replied Elizabeth.  "But the
wife of Mr. Darcy must have such extraordinary sources of
happiness necessarily attached to her situation, that she
could, upon the whole, have no cause to repine\index{repine}."

"Obstinate\index{Obstinate}, headstrong\index{headstrong} girl!  I am ashamed of you!  Is this
your gratitude for my attentions to you last spring?  Is
nothing due to me on that score?  Let\index{Let} us sit down.  You are to
understand, Miss Bennet, that I came here with the determined
resolution\index{resolution} of carrying my purpose; nor will I be dissuaded\index{dissuaded}
from it.  I have not been used to submit to any person's whims.
I have not been in the habit of \textbf{brooking}\index{brooking}\stepcounter{footnote}\footnotetext{
\textbf{brook} \emph{tr.v.} \textbf{-ing}  (BROOHK)
} disappointment."

"\emph{that} will make your ladyship's situation at present more
pitiable; but it will have no effect on me."

"I will not be interrupted.  Hear me in silence.  My daughter
and my nephew are formed for each other.  They are descended,
on the maternal side, from the same noble line; and, on the
father's, from respectable, honourable, and ancient--though
untitled--families.  Their fortune on both sides is splendid.
They are destined for each other by the voice of every member
of their respective houses; and what is to divide them?
The upstart pretensions\index{pretensions} of a young woman without family,
connections, or fortune.  Is this to be endured!  But it
must not, shall not be.  If you were sensible of your own
good, you would not wish to quit the sphere in which you
have been brought up."

"In marrying your nephew, I should not consider myself as
quitting that sphere.  He is a gentleman; I am a gentleman's
daughter; so far we are equal."

"True.  You \emph{are} a gentleman's daughter.  But who was your
mother?  Who are your uncles and aunts?  Do not imagine me
ignorant of their condition."

"Whatever my connections may be," said Elizabeth, "if your
nephew does not object to them, they can be nothing to \emph{you}."

"Tell me once for all, are you engaged to him?"

Though Elizabeth would not, for the mere purpose of obliging
Lady Catherine, have answered this question, she could not but
say, after a moment's deliberation\index{deliberation}:

"I am not."

Lady Catherine seemed pleased.

"And will you promise me, never to enter into such an engagement?"

"I will make no promise of the kind."

"Miss Bennet I am shocked and astonished.  I expected to find a
more reasonable young woman.  But do not deceive yourself into
a belief that I will ever \textbf{recede}\index{recede}\stepcounter{footnote}\footnotetext{
\textbf{recede} \emph{tr.v.}  (ree-SEED)
to move back and away from; to cede or give something back.}.  I shall not go away till you
have given me the assurance I require."

"And I certainly \emph{never} shall give it.  I am not to be intimidated
into anything so wholly\index{wholly} unreasonable.  Your ladyship wants
Mr. Darcy to marry your daughter; but would my giving you the
wished-for promise make their marriage at all more probable\index{probable}?
Supposing him to be attached to me, would my refusing to accept
his hand make him wish to bestow\index{bestow} it on his cousin?  Allow me to
say, Lady Catherine, that the arguments with which you have
supported this extraordinary application\index{application} have been as frivolous\index{frivolous}
as the application\index{application} was ill-judged.  You have widely mistaken my
character, if you think I can be worked on by such persuasions
as these.  How far your nephew might approve of your interference
in his affairs, I cannot tell; but you have certainly no right
to concern yourself in mine.  I must beg, therefore, to be
importuned\index{importuned} no farther on the subject\index{subject}."

"Not so hasty, if you please.  I have by no means done.
To all the objections I have already urged, I have still
another to add.  I am no stranger to the particulars of your
youngest sister's infamous\index{infamous} elopement.  I know it all; that
the young man's marrying her was a patched-up business, at the
expence of your father and uncles.  And is such a girl to be
my nephew's sister?  Is her husband, is the son of his late
father's steward\index{steward}, to be his brother?  Heaven and earth!--of
what are you thinking?  Are the shades of Pemberley to be
thus polluted?"

"You can now have nothing further to say," she resentfully
answered.  "You have insulted me in every possible method.
I must beg to return to the house."

And she rose as she spoke.  Lady Catherine rose also, and they
turned back.  Her ladyship was highly incensed\index{incensed}.

"You have no regard, then, for the honour and credit of my
nephew!  Unfeeling, selfish girl!  Do you not consider that
a connection with you must disgrace him in the eyes of
everybody?"

"Lady Catherine, I have nothing further to say.  You know my
sentiments\index{sentiments}."

"You are then resolved to have him?"

"I have said no such thing.  I am only resolved to act in that
manner, which will, in my own opinion, \textbf{constitute}\index{constitute}\stepcounter{footnote}\footnotetext{
\textbf{constitute} \emph{tr.v.}  (KAHN-stih-tyoot)
to form or compose.} my happiness,
without reference to \emph{you}, or to any person so wholly\index{wholly} unconnected
with me."

"It is well.  You refuse, then, to oblige me.  You refuse to
obey the claims of duty, honour, and gratitude.  You are
determined to ruin him in the opinion of all his friends,
and make him the contempt\index{contempt} of the world."

"Neither duty, nor honour, nor gratitude," replied Elizabeth,
"have any possible claim on me, in the present instance.  No
principle of either would be violated by my marriage with
Mr. Darcy.  And with regard to the resentment\index{resentment} of his family, or
the indignation\index{indignation} of the world, if the former \emph{were} excited by his
marrying me, it would not give me one moment's concern--and
the world in general would have too much sense to join in the
scorn."

"And this is your real opinion!  This is your final resolve!
Very well.  I shall now know how to act.  Do not imagine, Miss
Bennet, that your ambition will ever be gratified.  I came to
try you.  I hoped to find you reasonable; but, depend upon it,
I will carry my point."

In this manner Lady Catherine talked on, till they were at the
door of the carriage, when, turning hastily round, she added,
"I take no leave of you, Miss Bennet.  I send no compliments to
your mother.  You deserve no such attention.  I am most
seriously displeased."

Elizabeth made no answer; and without attempting to persuade
her ladyship to return into the house, walked quietly into it
herself.  She heard the carriage drive away as she proceeded
up stairs.  Her mother impatiently met her at the door of the
dressing-room, to ask why Lady Catherine would not come in
again and rest herself.

"She did not choose it," said her daughter, "she would go."

"She is a very fine-looking woman! and her calling here was
prodigiously\index{prodigiously} civil! for she only came, I suppose, to tell us
the Collinses were well.  She is on her road somewhere, I dare
say, and so, passing through Meryton, thought she might as well
call on you.  I suppose she had nothing particular to say to
you, Lizzy?"

Elizabeth was forced to give into a little falsehood here;
for to acknowledge the substance of their conversation was
impossible.



\chapter{57}


The discomposure of spirits which this extraordinary visit
threw Elizabeth into, could not be easily overcome; nor
could she, for many hours, learn to think of it less than
incessantly\index{incessantly}.  Lady Catherine, it appeared, had actually taken
the trouble of this journey from Rosings, for the sole purpose
of breaking off her supposed engagement with Mr. Darcy.  It was
a rational\index{rational} scheme, to be sure!  but from what the report of
their engagement could originate, Elizabeth was at a loss to
imagine; till she recollected that \emph{his} being the intimate
friend of Bingley, and \emph{her} being the sister of Jane, was
enough, at a time when the expectation of one wedding made
everybody eager for another, to supply the idea.  She had not
herself forgotten to feel that the marriage of her sister must
bring them more frequently together.  And her neighbours at
Lucas Lodge, therefore (for through their communication with
the Collinses, the report, she concluded, had reached lady
Catherine), had only set that down as almost certain and
immediate, which she had looked forward\index{forward} to as possible at
some future time.

In revolving Lady Catherine's expressions, however, she could
not help feeling some uneasiness as to the possible consequence\index{consequence}
of her persisting in this interference.  From what she had said
of her resolution\index{resolution} to prevent their marriage, it occurred to
Elizabeth that she must meditate\index{meditate} an application\index{application} to her nephew;
and how \emph{he} might take a similar representation of the evils
attached to a connection with her, she dared not pronounce.
She knew not the exact degree of his affection for his aunt, or
his dependence on her judgment, but it was natural to suppose
that he thought much higher of her ladyship than \emph{she} could
do; and it was certain that, in enumerating the miseries of a
marriage with \emph{one}, whose immediate connections were so unequal
to his own, his aunt would address\index{address} him on his weakest side.
With his notions of dignity, he would probably feel that the
arguments, which to Elizabeth had appeared weak and ridiculous,
contained much good sense and solid reasoning.

If he had been wavering before as to what he should do, which
had often seemed likely, the advice and entreaty\index{entreaty} of so near a
relation might settle every doubt, and determine him at once to
be as happy as dignity unblemished could make him.  In that
case he would return no more.  Lady Catherine might see him in
her way through town; and his engagement to Bingley of coming
again to Netherfield must give way.

"If, therefore, an excuse for not keeping his promise should
come to his friend within a few days," she added, "I shall
know how to understand it.  I shall then give over every
expectation, every wish of his constancy.  If he is satisfied
with only regretting me, when he might have obtained my
affections and hand, I shall soon cease to regret him at all."



The surprise of the rest of the family, on hearing who their
visitor had been, was very great; but they obligingly satisfied
it, with the same kind of supposition\index{supposition} which had appeased\index{appeased}
Mrs. Bennet's curiosity; and Elizabeth was spared\index{spared} from much
teasing on the subject\index{subject}.

The next morning, as she was going downstairs, she was met by
her father, who came out of his library with a letter in his
hand.

"Lizzy," said he, "I was going to look for you; come into my
room."

She followed him thither; and her curiosity to know what he
had to tell her was heightened by the supposition\index{supposition} of its being
in some manner connected with the letter he held.  It suddenly
struck her that it might be from Lady Catherine; and she
anticipated with dismay all the consequent explanations.

She followed her father to the fire place, and they both sat
down.  He then said,

"I have received a letter this morning that has astonished me
exceedingly.  As it principally\index{principally} concerns yourself, you ought
to know its contents.  I did not know before, that I had two
daughters on the brink of matrimony\index{matrimony}.  Let\index{Let} me congratulate you
on a very important conquest."

The colour now rushed into Elizabeth's cheeks in the instantaneous
conviction\index{conviction} of its being a letter from the nephew, instead of the
aunt; and she was undetermined whether most to be pleased that
he explained himself at all, or offended that his letter was not
rather addressed\index{addressed} to herself; when her father continued:

"You look conscious.  Young ladies have great \textbf{penetration}\index{penetration}\stepcounter{footnote}\footnotetext{
\textbf{penetration} \emph{n.}  (peh-neh-TRAY-shun)
} in
such matters as these; but I think I may defy even \emph{your}
\textbf{sagacity}\index{sagacity}\stepcounter{footnote}\footnotetext{
\textbf{sagacity} \emph{n.}  (suh-GAS-sih-tee)
wisdom, good judgement, sage-like.}, to discover the name of your admirer.  This letter
is from Mr. Collins."

"From Mr. Collins! and what can \emph{he} have to say?"

"Something very much to the purpose of course.  He begins with
congratulations on the approaching nuptials\index{nuptials} of my eldest daughter,
of which, it seems, he has been told by some of the good-natured,
gossiping Lucases.  I shall not sport with your impatience, by
reading what he says on that point.  What relates to yourself, is
as follows:  'Having thus offered you the sincere congratulations
of Mrs. Collins and myself on this happy event, let\index{let} me now add
a short hint on the subject\index{subject} of another; of which we have been
advertised by the same authority.  Your daughter Elizabeth, it is
presumed, will not long bear the name of Bennet, after her elder
sister has resigned it, and the chosen partner of her fate may be
reasonably looked up to as one of the most \textbf{illustrious}\index{illustrious}\stepcounter{footnote}\footnotetext{
\textbf{illustrious} \emph{adj.}  (ih-LUHS-tree-uhs)
widely known and esteemed.} personages
in this land.'

"Can you possibly guess, Lizzy, who is meant by this?"  'This
young gentleman is blessed, in a peculiar\index{peculiar} way, with every thing
the heart of mortal can most desire,--splendid property,
noble \textbf{kindred}\index{kindred}\stepcounter{footnote}\footnotetext{
\textbf{kindred} \emph{n.}  (KIHN-drehd)
similar or related in quality or character.}, and extensive patronage\index{patronage}.  Yet in spite of all
these temptations, let\index{let} me warn my cousin Elizabeth, and
yourself, of what evils you may incur\index{incur} by a precipitate\index{precipitate} closure
with this gentleman's proposals, which, of course, you will be
inclined to take immediate advantage of.'

"Have you any idea, Lizzy, who this gentleman is?  But now it
comes out:

"'My motive for cautioning you is as follows.  We have reason to
imagine that his aunt, Lady Catherine de Bourgh, does not look
on the match with a friendly eye.'

"\emph{mr}. \emph{darcy}, you see, is the man!  Now, Lizzy, I think I
\emph{have} surprised you.  Could he, or the Lucases, have pitched
on any man within the circle of our acquaintance, whose name
would have given the lie more effectually\index{effectually} to what they related?
Mr. Darcy, who never looks at any woman but to see a blemish,
and who probably never looked at you in his life!  It is
admirable!"

Elizabeth tried to join in her father's pleasantry, but could
only force one most reluctant smile.  Never had his wit been
directed in a manner so little agreeable to her.

"Are you not diverted\index{diverted}?"

"Oh! yes.  Pray read on."

"'After mentioning the likelihood of this marriage to her ladyship
last night, she immediately, with her usual condescension\index{condescension},
expressed what she felt on the occasion; when it become apparent,
that on the score of some family objections on the part of my
cousin, she would never give her consent to what she termed so
disgraceful a match.  I thought it my duty to give the speediest
intelligence of this to my cousin, that she and her noble
admirer may be aware of what they are about, and not run
hastily into a marriage which has not been properly sanctioned\index{sanctioned}.'
Mr. Collins moreover adds, 'I am truly rejoiced that my cousin
Lydia's sad business has been so well hushed up, and am only
concerned that their living together before the marriage took
place should be so generally known.  I must not, however,
neglect the duties of my station, or refrain from declaring my
amazement at hearing that you received the young couple into
your house as soon as they were married.  It was an encouragement
of vice\index{vice}; and had I been the rector\index{rector} of Longbourn, I should very
strenuously have opposed it.  You ought certainly to forgive them,
as a Christian, but never to admit them in your sight, or allow
their names to be mentioned in your hearing.'  That is his notion
of Christian forgiveness!  The rest of his letter is only about
his dear Charlotte's situation, and his expectation of a young
olive-branch.  But, Lizzy, you look as if you did not enjoy it.
You are not going to be \emph{\textbf{missish}\index{missish}\stepcounter{footnote}\footnotetext{
\textbf{} \emph{}  ()
}}, I hope, and pretend to be
affronted\index{affronted} at an idle report.  For what do we live, but to make
sport for our neighbours, and laugh at them in our turn?"

"Oh!" cried Elizabeth, "I am excessively diverted\index{diverted}.  But it is
so strange!"

"Yes--\emph{that} is what makes it amusing.  Had they fixed on any other
man it would have been nothing; but \emph{his} perfect indifference\index{indifference},
and \emph{your} pointed dislike, make it so delightfully absurd!  Much
as I abominate\index{abominate} writing, I would not give up Mr. Collins's
correspondence for any consideration.  Nay, when I read a letter
of his, I cannot help giving him the preference even over Wickham,
much as I value the impudence\index{impudence} and \textbf{hypocrisy}\index{hypocrisy}\stepcounter{footnote}\footnotetext{
\textbf{hypocrisy} \emph{n.}  (hih-PAH-krih-see)
professing beliefs, values, or sentiments one does not truly believe or live by; insincerity; falseness.} of my son-in-law.
And pray, Lizzy, what said Lady Catherine about this report?
Did she call to refuse her consent?"

To this question his daughter replied only with a laugh; and
as it had been asked without the least suspicion, she was not
distressed by his repeating it.  Elizabeth had never been
more at a loss to make her feelings appear what they were not.
It was necessary to laugh, when she would rather have cried.
Her father had most cruelly mortified her, by what he said of
Mr. Darcy's indifference\index{indifference}, and she could do nothing but wonder
at such a want of penetration\index{penetration}, or fear that perhaps, instead
of his seeing too little, she might have fancied too much.



\chapter{58}


Instead of receiving any such letter of excuse from his friend,
as Elizabeth half expected Mr. Bingley to do, he was able to
bring Darcy with him to Longbourn before many days had passed
after Lady Catherine's visit.  The gentlemen arrived early;
and, before Mrs. Bennet had time to tell him of their having
seen his aunt, of which her daughter sat in momentary dread,
Bingley, who wanted to be alone with Jane, proposed their
all walking out.  It was agreed to.  Mrs. Bennet was not in
the habit of walking; Mary could never spare time; but the
remaining five set off together.  Bingley and Jane, however,
soon allowed the others to outstrip them.  They lagged behind,
while Elizabeth, Kitty, and Darcy were to entertain each other.
Very little was said by either; Kitty was too much afraid of
him to talk; Elizabeth was secretly forming a desperate
resolution\index{resolution}; and perhaps he might be doing the same.

They walked towards the Lucases, because Kitty wished to call
upon Maria; and as Elizabeth saw no occasion for making it a
general concern, when Kitty left them she went boldly on with
him alone.  Now was the moment for her resolution\index{resolution} to be
executed, and, while her courage was high, she immediately
said:

"Mr. Darcy, I am a very selfish creature; and, for the sake of
giving relief to my own feelings, care not how much I may be
wounding your's.  I can no longer help thanking you for your
unexampled kindness to my poor sister.  Ever since I have
known it, I have been most anxious to acknowledge to you how
gratefully I feel it.  Were it known to the rest of my family,
I should not have merely my own gratitude to express."

"I am sorry, exceedingly sorry," replied Darcy, in a tone of
surprise and emotion, "that you have ever been informed of what
may, in a mistaken light, have given you uneasiness.  I did not
think Mrs. Gardiner was so little to be trusted."

"You must not blame my aunt.  Lydia's thoughtlessness first
betrayed to me that you had been concerned in the matter; and,
of course, I could not rest till I knew the particulars.  Let\index{Let}
me thank you again and again, in the name of all my family,
for that generous compassion which induced\index{induced} you to take so much
trouble, and bear so many mortifications\index{mortifications}, for the sake of
discovering them."

"If you \emph{will} thank me," he replied, "let\index{let} it be for yourself
alone.  That the wish of giving happiness to you might add
force to the other inducements\index{inducements} which led me on, I shall not
attempt to deny.  But your \emph{family} owe me nothing.  Much as
I respect them, I believe I thought only of \emph{you}."

Elizabeth was too much embarrassed to say a word.  After a
short pause, her companion added, "You are too generous to
trifle with me.  If your feelings are still what they were
last April, tell me so at once.  \emph{my} affections and wishes
are unchanged, but one word from you will silence me on this
subject\index{subject} for ever."

Elizabeth, feeling all the more than common awkwardness and
anxiety of his situation, now forced herself to speak; and
immediately, though not very fluently, gave him to understand
that her sentiments\index{sentiments} had undergone so material a change, since
the period to which he alluded\index{alluded}, as to make her receive with
gratitude and pleasure his present assurances.  The happiness
which this reply produced, was such as he had probably never
felt before; and he expressed himself on the occasion as
sensibly and as warmly as a man violently in love can be
supposed to do.  Had Elizabeth been able to encounter his
eye, she might have seen how well the expression of heartfelt
delight, diffused\index{diffused} over his face, became him; but, though she
could not look, she could listen, and he told her of feelings,
which, in proving of what importance she was to him, made his
affection every moment more valuable.

They walked on, without knowing in what direction.  There was
too much to be thought, and felt, and said, for attention to
any other objects.  She soon learnt that they were indebted
for their present good understanding to the efforts of his
aunt, who did call on him in her return through London,
and there relate her journey to Longbourn, its motive, and
the substance of her conversation with Elizabeth; dwelling
emphatically on every expression of the latter which, in her
ladyship's apprehension\index{apprehension}, peculiarly\index{peculiarly} denoted her perverseness
and assurance; in the belief that such a relation must assist
her endeavours\index{endeavours} to obtain that promise from her nephew which
she had refused to give.  But, unluckily for her ladyship,
its effect had been exactly \textbf{contrariwise}\index{contrariwise}\stepcounter{footnote}\footnotetext{
\textbf{contrariwise} \emph{adv.}  (kuhn-TRAIR-ree-wyz)
}.

"It taught me to hope," said he, "as I had scarcely ever allowed
myself to hope before.  I knew enough of your disposition\index{disposition} to
be certain that, had you been absolutely, \textbf{irrevocably}\index{irrevocably}\stepcounter{footnote}\footnotetext{
\textbf{irrevocable} \emph{adj.} \textbf{-ly}  (ihr-reh-VOH-kuh-bl)
impossible to retract, take back, nullify, or cancle.} decided
against me, you would have acknowledged it to Lady Catherine,
frankly and openly."

Elizabeth coloured and laughed as she replied, "Yes, you know
enough of my frankness to believe me capable of \emph{that}.
After abusing you so abominably to your face, I could have no
scruple\index{scruple} in abusing you to all your relations."

"What did you say of me, that I did not deserve?  For, though
your accusations were ill-founded, formed on mistaken premises,
my behaviour to you at the time had merited the severest
reproof\index{reproof}.  It was unpardonable.  I cannot think of it without
abhorrence\index{abhorrence}."

"We will not quarrel for the greater share of blame \textbf{annexed}\index{annexed}\stepcounter{footnote}\footnotetext{
\textbf{annex} \emph{tr.v.} \textbf{-ed}  (uhn-NEHX)
 to attach to something larger or more important.} to
that evening," said Elizabeth.  "The conduct of neither, if
strictly examined, will be irreproachable; but since then, we
have both, I hope, improved in civility\index{civility}."

"I cannot be so easily reconciled\index{reconciled} to myself.  The recollection\index{recollection}
of what I then said, of my conduct, my manners, my expressions
during the whole of it, is now, and has been many months,
inexpressibly painful to me.  Your reproof\index{reproof}, so well applied, I
shall never forget: 'had you behaved in a more gentlemanlike
manner.' Those were your words.  You know not, you can
scarcely conceive, how they have tortured me;--though it was
some time, I confess, before I was reasonable enough to allow
their justice."

"I was certainly very far from expecting them to make so strong
an impression.  I had not the smallest idea of their being ever
felt in such a way."

"I can easily believe it.  You thought me then devoid of
every proper feeling, I am sure you did.  The turn of your
countenance\index{countenance} I shall never forget, as you said that I could
not have addressed\index{addressed} you in any possible way that would induce\index{induce}
you to accept me."

"Oh! do not repeat what I then said.  These recollections\index{recollections}
will not do at all.  I assure you that I have long been most
heartily\index{heartily} ashamed of it."

Darcy mentioned his letter.  "Did it," said he, "did it soon
make you think better of me?  Did you, on reading it, give any
credit to its contents?"

She explained what its effect on her had been, and how gradually
all her former prejudices\index{prejudices} had been removed.

"I knew," said he, "that what I wrote must give you pain,
but it was necessary.  I hope you have destroyed the letter.
There was one part especially, the opening of it, which I
should dread your having the power of reading again.  I can
remember some expressions which might justly make you hate me."

"The letter shall certainly be burnt, if you believe it
essential to the preservation of my regard; but, though we have
both reason to think my opinions not entirely unalterable, they
are not, I hope, quite so easily changed as that implies."

"When I wrote that letter," replied Darcy, "I believed myself
perfectly calm and cool, but I am since convinced that it was
written in a dreadful bitterness of spirit."

"The letter, perhaps, began in bitterness, but it did not end
so.  The adieu\index{adieu} is charity itself.  But think no more of the
letter.  The feelings of the person who wrote, and the person
who received it, are now so widely different from what they
were then, that every unpleasant circumstance attending it
ought to be forgotten.  You must learn some of my philosophy\index{philosophy}.
Think only of the past as its remembrance\index{remembrance} gives you pleasure."

"I cannot give you credit for any philosophy\index{philosophy} of the kind.
Your \textbf{retrospections}\index{retrospections}\stepcounter{footnote}\footnotetext{
\textbf{} \emph{}  ()
} must be so totally \textbf{void}\index{void}\stepcounter{footnote}\footnotetext{
\textbf{void} \emph{adj.}  (VOYD)
 an empty area or space; nonexistence.} of reproach\index{reproach}, that
the contentment\index{contentment} arising from them is not of philosophy\index{philosophy}, but,
what is much better, of innocence.  But with me, it is not
so.  Painful recollections\index{recollections} will intrude which cannot, which
ought not, to be repelled.  I have been a selfish being all my
life, in practice, though not in principle.  As a child I was
taught what was right, but I was not taught to correct my
temper.  I was given good principles, but left to follow them
in pride and conceit\index{conceit}.  Unfortunately an only son (for many
years an only child), I was spoilt by my parents, who, though
good themselves (my father, particularly, all that was
\textbf{benevolent}\index{benevolent}\stepcounter{footnote}\footnotetext{
\textbf{benevolent} \emph{adj.}  (buh-NEH-vuh-lent)
intending or showing kindness; generous.} and amiable\index{amiable}), allowed, encouraged, almost taught me
to be selfish and overbearing; to care for none beyond my own
family circle; to think meanly of all the rest of the world;
to wish at least to think meanly of their sense and worth
compared with my own.  Such I was, from eight to eight and
twenty; and such I might still have been but for you, dearest,
loveliest Elizabeth!  What do I not owe you!  You taught me a
lesson, hard indeed at first, but most advantageous.  By you,
I was properly humbled.  I came to you without a doubt of my
reception.  You showed me how insufficient were all my
pretensions\index{pretensions} to please a woman worthy of being pleased."

"Had you then persuaded yourself that I should?"

"Indeed I had.  What will you think of my vanity\index{vanity}?  I believed
you to be wishing, expecting my addresses."

"My manners must have been in fault, but not intentionally,
I assure you.  I never meant to deceive you, but my spirits
might often lead me wrong.  How you must have hated me after
\emph{that} evening?"

"Hate you!  I was angry perhaps at first, but my anger soon
began to take a proper direction."

"I am almost afraid of asking what you thought of me, when we
met at Pemberley.  You blamed me for coming?"

"No indeed; I felt nothing but surprise."

"Your surprise could not be greater than \emph{mine} in being
noticed by you.  My conscience\index{conscience} told me that I deserved no
extraordinary politeness, and I confess that I did not expect
to receive \emph{more} than my due."

"My object then," replied Darcy, "was to show you, by every
civility\index{civility} in my power, that I was not so mean as to resent the
past; and I hoped to obtain your forgiveness, to lessen your
ill opinion, by letting you see that your reproofs\index{reproofs} had been
attended to.  How soon any other wishes introduced themselves
I can hardly tell, but I believe in about half an hour after
I had seen you."

He then told her of Georgiana's delight in her acquaintance,
and of her disappointment at its sudden interruption; which
naturally leading to the cause of that interruption, she soon
learnt that his resolution\index{resolution} of following her from Derbyshire in
quest of her sister had been formed before he quitted the inn,
and that his gravity\index{gravity} and thoughtfulness there had arisen from
no other struggles than what such a purpose must comprehend\index{comprehend}.

She expressed her gratitude again, but it was too painful a
subject\index{subject} to each, to be dwelt on farther.

After walking several miles in a leisurely manner, and too busy
to know anything about it, they found at last, on examining
their watches, that it was time to be at home.

"What could become of Mr. Bingley and Jane!" was a wonder
which introduced the discussion of their affairs.  Darcy
was delighted with their engagement; his friend had given
him the earliest information of it.

"I must ask whether you were surprised?" said Elizabeth.

"Not at all.  When I went away, I felt that it would soon
happen."

"That is to say, you had given your permission.  I guessed as
much."  And though he exclaimed at the term, she found that it
had been pretty much the case.

"On the evening before my going to London," said he, "I made a
confession to him, which I believe I ought to have made long
ago.  I told him of all that had occurred to make my former
interference in his affairs absurd and impertinent\index{impertinent}.  His
surprise was great.  He had never had the slightest suspicion.
I told him, moreover, that I believed myself mistaken in
supposing, as I had done, that your sister was indifferent\index{indifferent} to
him; and as I could easily perceive\index{perceive} that his attachment to her
was \textbf{unabated}\index{unabated}\stepcounter{footnote}\footnotetext{
\textbf{unabated} \emph{adj.}  (uhn-uh-BAY-tuhd)
continuing at full strength or intensity with no decrease.}, I felt no doubt of their happiness together."

Elizabeth could not help smiling at his easy manner of
directing his friend.

"Did you speak from your own observation," said she, "when
you told him that my sister loved him, or merely from my
information last spring?"

"From the former.  I had narrowly observed her during the two
visits which I had lately made here; and I was convinced of her
affection."

"And your assurance of it, I suppose, carried immediate
conviction\index{conviction} to him."

"It did.  Bingley is most unaffectedly\index{unaffectedly} modest.  His diffidence\index{diffidence}
had prevented his depending on his own judgment in so anxious
a case, but his reliance on mine made every thing easy.  I
was obliged to confess one thing, which for a time, and not
unjustly, offended him.  I could not allow myself to conceal
that your sister had been in town three months last winter,
that I had known it, and purposely kept it from him.  He was
angry.  But his anger, I am persuaded, lasted no longer than
he remained in any doubt of your sister's sentiments\index{sentiments}.  He has
heartily\index{heartily} forgiven me now."

Elizabeth longed to observe that Mr. Bingley had been a most
delightful friend; so easily guided that his worth was
invaluable\index{invaluable}; but she checked herself.  She remembered that he
had yet to learn to be laughed at, and it was rather too early
to begin.  In anticipating the happiness of Bingley, which of
course was to be inferior\index{inferior} only to his own, he continued the
conversation till they reached the house.  In the hall they
parted.



\chapter{59}


"My dear Lizzy, where can you have been walking to?" was a
question which Elizabeth received from Jane as soon as she
entered their room, and from all the others when they sat down
to table.  She had only to say in reply, that they had wandered
about, till she was beyond her own knowledge.  She coloured as
she spoke; but neither that, nor anything else, awakened a
suspicion of the truth.

The evening passed quietly, unmarked by anything extraordinary.
The acknowledged lovers talked and laughed, the unacknowledged
were silent.  Darcy was not of a disposition\index{disposition} in which happiness
overflows in mirth\index{mirth}; and Elizabeth, agitated and confused, rather
\emph{knew} that she was happy than \emph{felt} herself to be so; for, besides
the immediate embarrassment, there were other evils before her.
She anticipated what would be felt in the family when her
situation became known; she was aware that no one liked him but
Jane; and even feared that with the others it was a dislike
which not all his fortune and consequence\index{consequence} might do away.

At night she opened her heart to Jane.  Though suspicion was
very far from Miss Bennet's general habits, she was absolutely
incredulous\index{incredulous} here.

"You are joking, Lizzy.  This cannot be!--engaged to Mr. Darcy!
No, no, you shall not deceive me.  I know it to be impossible."

"This is a wretched\index{wretched} beginning indeed!  My sole dependence was
on you; and I am sure nobody else will believe me, if you do
not.  Yet, indeed, I am in earnest\index{earnest}.  I speak nothing but the
truth.  He still loves me, and we are engaged."

Jane looked at her doubtingly.  "Oh, Lizzy! it cannot be.
I know how much you dislike him."

"You know nothing of the matter.  \emph{that} is all to be forgot.
Perhaps I did not always love him so well as I do now.  But in
such cases as these, a good memory is unpardonable.  This is
the last time I shall ever remember it myself."

Miss Bennet still looked all amazement.  Elizabeth again, and
more seriously assured her of its truth.

"Good Heaven! can it be really so!  Yet now I must believe
you," cried Jane.  "My dear, dear Lizzy, I would--I do
congratulate you--but are you certain? forgive the question
--are you quite certain that you can be happy with him?"

"There can be no doubt of that.  It is settled between us
already, that we are to be the happiest couple in the world.
But are you pleased, Jane?  Shall you like to have such a
brother?"

"Very, very much.  Nothing could give either Bingley or
myself more delight.  But we considered it, we talked of it as
impossible.  And do you really love him quite well enough?
Oh, Lizzy! do anything rather than marry without affection.
Are you quite sure that you feel what you ought to do?"

"Oh, yes!  You will only think I feel \emph{more} than I ought to
do, when I tell you all."

"What do you mean?"

"Why, I must confess that I love him better than I do Bingley.
I am afraid you will be angry."

"My dearest sister, now \emph{be} serious.  I want to talk very
seriously.  Let\index{Let} me know every thing that I am to know, without
delay.  Will you tell me how long you have loved him?"

"It has been coming on so gradually, that I hardly know when it
began.  But I believe I must date it from my first seeing his
beautiful grounds at Pemberley."

Another entreaty\index{entreaty} that she would be serious, however, produced
the desired effect; and she soon satisfied Jane by her solemn\index{solemn}
assurances of attachment.  When convinced on that article, Miss
Bennet had nothing further to wish.

"Now I am quite happy," said she, "for you will be as happy as
myself.  I always had a value for him.  Were it for nothing but
his love of you, I must always have esteemed him; but now, as
Bingley's friend and your husband, there can be only Bingley
and yourself more dear to me.  But Lizzy, you have been very
sly, very reserved with me.  How little did you tell me of what
passed at Pemberley and Lambton!  I owe all that I know of it
to another, not to you."

Elizabeth told her the motives of her secrecy.  She had been
unwilling to mention Bingley; and the unsettled state of her
own feelings had made her equally avoid the name of his friend.
But now she would no longer conceal from her his share in
Lydia's marriage.  All was acknowledged, and half the night
spent in conversation.



"Good gracious!" cried Mrs. Bennet, as she stood at a window
the next morning, "if that disagreeable Mr. Darcy is not coming
here again with our dear Bingley!  What can he mean by being so
tiresome as to be always coming here?  I had no notion but he
would go a-shooting, or something or other, and not disturb us
with his company.  What shall we do with him?  Lizzy, you must
walk out with him again, that he may not be in Bingley's way."

Elizabeth could hardly help laughing at so convenient a
proposal; yet was really vexed\index{vexed} that her mother should be
always giving him such an \textbf{epithet}\index{epithet}\stepcounter{footnote}\footnotetext{
\textbf{epithet} \emph{n.}  (EH-pih-theht)
}.

As soon as they entered, Bingley looked at her so expressively,
and shook hands with such warmth, as left no doubt of his good
information; and he soon afterwards said aloud, "Mrs. Bennet,
have you no more lanes hereabouts in which Lizzy may lose her
way again to-day?"

"I advise Mr. Darcy, and Lizzy, and Kitty," said Mrs. Bennet,
"to walk to Oakham Mount this morning.  It is a nice long walk,
and Mr. Darcy has never seen the view."

"It may do very well for the others," replied Mr. Bingley; "but
I am sure it will be too much for Kitty.  Won't it, Kitty?"
Kitty owned that she had rather stay at home.  Darcy professed
a great curiosity to see the view from the Mount, and Elizabeth
silently consented.  As she went up stairs to get ready,
Mrs. Bennet followed her, saying:

"I am quite sorry, Lizzy, that you should be forced to have
that disagreeable man all to yourself.  But I hope you will not
mind it: it is all for Jane's sake, you know; and there is no
occasion for talking to him, except just now and then.  So, do
not put yourself to inconvenience."

During their walk, it was resolved that Mr. Bennet's consent
should be asked in the course of the evening.  Elizabeth
reserved to herself the application\index{application} for her mother's.  She
could not determine how her mother would take it; sometimes
doubting whether all his wealth and grandeur\index{grandeur} would be enough
to overcome her abhorrence\index{abhorrence} of the man.  But whether she were
violently set against the match, or violently delighted with
it, it was certain that her manner would be equally ill adapted
to do credit to her sense; and she could no more bear that
Mr. Darcy should hear the first raptures\index{raptures} of her joy, than the
first \textbf{vehemence}\index{vehemence}\stepcounter{footnote}\footnotetext{
\textbf{vehement} \emph{adj.}  (VEE-huh-mehnt)
intensity or forcefulness of expression.} of her disapprobation\index{disapprobation}.



In the evening, soon after Mr. Bennet withdrew to the library,
she saw Mr. Darcy rise also and follow him, and her agitation
on seeing it was extreme.  She did not fear her father's
opposition, but he was going to be made unhappy; and that it
should be through her means--that \emph{she}, his favourite child,
should be distressing him by her choice, should be filling him
with fears and regrets in disposing of her--was a wretched\index{wretched}
reflection, and she sat in misery till Mr. Darcy appeared
again, when, looking at him, she was a little relieved by his
smile.  In a few minutes\index{minutes} he approached the table where she was
sitting with Kitty; and, while pretending to admire her work
said in a whisper, "Go to your father, he wants you in the
library."  She was gone directly.

Her father was walking about the room, looking grave\index{grave} and
anxious.  "Lizzy," said he, "what are you doing?  Are you out
of your senses, to be accepting this man?  Have not you always
hated him?"

How earnestly\index{earnestly} did she then wish that her former opinions had
been more reasonable, her expressions more moderate!  It would
have spared\index{spared} her from explanations and professions which it was
exceedingly awkward to give; but they were now necessary, and
she assured him, with some confusion, of her attachment to
Mr. Darcy.

"Or, in other words, you are determined to have him.  He is
rich, to be sure, and you may have more fine clothes and fine
carriages than Jane.  But will they make you happy?"

"Have you any other objection," said Elizabeth, "than your
belief of my indifference\index{indifference}?"

"None at all.  We all know him to be a proud, unpleasant sort
of man; but this would be nothing if you really liked him."

"I do, I do like him," she replied, with tears in her eyes,
"I love him.  Indeed he has no improper pride.  He is perfectly
amiable\index{amiable}.  You do not know what he really is; then pray do not
pain me by speaking of him in such terms."

"Lizzy," said her father, "I have given him my consent.
He is the kind of man, indeed, to whom I should never dare
refuse anything, which he condescended\index{condescended} to ask.  I now give it
to \emph{you}, if you are resolved on having him.  But let\index{let} me advise
you to think better of it.  I know your disposition\index{disposition}, Lizzy.
I know that you could be neither happy nor respectable, unless
you truly esteemed your husband; unless you looked up to him
as a superior\index{superior}.  Your lively talents would place you in the
greatest danger in an unequal marriage.  You could scarcely
escape discredit and misery.  My child, let\index{let} me not have the
grief of seeing \emph{you} unable to respect your partner in life.
You know not what you are about."

Elizabeth, still more affected, was earnest\index{earnest} and solemn\index{solemn} in her
reply; and at length, by repeated assurances that Mr. Darcy was
really the object of her choice, by explaining the gradual
change which her estimation of him had undergone, relating her
absolute certainty that his affection was not the work of a
day, but had stood the test of many months suspense, and
enumerating with energy all his good qualities, she did conquer
her father's incredulity\index{incredulity}, and reconcile\index{reconcile} him to the match.

"Well, my dear," said he, when she ceased speaking, "I have no
more to say.  If this be the case, he deserves you.  I could
not have parted with you, my Lizzy, to anyone less worthy."

To complete the favourable impression, she then told him what
Mr. Darcy had voluntarily done for Lydia.  He heard her with
astonishment.

"This is an evening of wonders, indeed!  And so, Darcy did
every thing; made up the match, gave the money, paid the
fellow's debts, and got him his commission!  So much the
better.  It will save me a world of trouble and economy\index{economy}.
Had it been your uncle's doing, I must and \emph{would} have paid
him; but these violent young lovers carry every thing their
own way.  I shall offer to pay him to-morrow; he will rant
and storm about his love for you, and there will be an end
of the matter."

He then recollected her embarrassment a few days before, on his
reading Mr. Collins's letter; and after laughing at her some
time, allowed her at last to go--saying, as she quitted the
room, "If any young men come for Mary or Kitty, send them in,
for I am quite at leisure."

Elizabeth's mind was now relieved from a very heavy weight;
and, after half an hour's quiet reflection in her own room,
she was able to join the others with tolerable\index{tolerable} composure\index{composure}.
Every thing was too recent for gaiety\index{gaiety}, but the evening passed
tranquilly\index{tranquilly} away; there was no longer anything material to
be dreaded, and the comfort of ease and familiarity would
come in time.

When her mother went up to her dressing-room at night, she
followed her, and made the important communication.  Its effect
was most extraordinary; for on first hearing it, Mrs. Bennet
sat quite still, and unable to utter a syllable.  Nor was it
under many, many minutes\index{minutes} that she could comprehend\index{comprehend} what she
heard; though not in general backward to credit what was for
the advantage of her family, or that came in the shape of a
lover to any of them.  She began at length to recover, to
fidget about in her chair, get up, sit down again, wonder,
and bless herself.

"Good gracious!  Lord bless me!  only think!  dear me!
Mr. Darcy!  Who would have thought it!  And is it really true?
Oh! my sweetest Lizzy! how rich and how great you will be!
What pin-money, what jewels, what carriages you will have!
Jane's is nothing to it--nothing at all.  I am so pleased--so
happy.  Such a charming man!--so handsome! so tall!--Oh, my
dear Lizzy! pray apologise for my having disliked him so much
before.  I hope he will overlook it.  Dear, dear Lizzy.  A house
in town!  Every thing that is charming!  Three daughters
married!  Ten thousand a year!  Oh, Lord!  What will become of
me.  I shall go distracted\index{distracted}."

This was enough to prove that her approbation\index{approbation} need not be
doubted: and Elizabeth, rejoicing that such an effusion\index{effusion} was
heard only by herself, soon went away.  But before she had
been three minutes\index{minutes} in her own room, her mother followed her.

"My dearest child," she cried, "I can think of nothing else!
Ten thousand a year, and very likely more!  'Tis as good as a
Lord!  And a special licence.  You must and shall be married
by a special licence.  But my dearest love, tell me what dish
Mr. Darcy is particularly fond of, that I may have it to-morrow."

This was a sad omen of what her mother's behaviour to the
gentleman himself might be; and Elizabeth found that, though in
the certain possession of his warmest affection, and secure of
her relations' consent, there was still something to be wished
for.  But the morrow passed off much better than she expected;
for Mrs. Bennet luckily stood in such awe\index{awe} of her intended
son-in-law that she ventured not to speak to him, unless it was
in her power to offer him any attention, or mark her deference\index{deference}
for his opinion.

Elizabeth had the satisfaction of seeing her father taking
pains to get acquainted with him; and Mr. Bennet soon assured
her that he was rising every hour in his esteem.

"I admire all my three sons-in-law highly," said he.  "Wickham,
perhaps, is my favourite; but I think I shall like \emph{your} husband
quite as well as Jane's."



\chapter{60}


Elizabeth's spirits soon rising to playfulness again, she
wanted Mr. Darcy to account for his having ever fallen in love
with her.  "How could you begin?" said she.  "I can comprehend\index{comprehend}
your going on charmingly, when you had once made a beginning;
but what could set you off in the first place?"

"I cannot fix on the hour, or the spot, or the look, or the
words, which laid the foundation.  It is too long ago.  I was
in the middle before I knew that I \emph{had} begun."

"My beauty you had early withstood, and as for my manners--my
behaviour to \emph{you} was at least always bordering on the uncivil\index{uncivil},
and I never spoke to you without rather wishing to give you pain
than not.  Now be sincere; did you admire me for my impertinence\index{impertinence}?"

"For the liveliness of your mind, I did."

"You may as well call it impertinence\index{impertinence} at once.  It was very
little less.  The fact is, that you were sick of civility\index{civility}, of
deference\index{deference}, of officious\index{officious} attention.  You were disgusted with
the women who were always speaking, and looking, and thinking
for \emph{your} approbation\index{approbation} alone.  I roused, and interested you,
because I was so unlike \emph{them}.  Had you not been really
amiable\index{amiable}, you would have hated me for it; but in spite of the
pains you took to disguise yourself, your feelings were always
noble and just; and in your heart, you thoroughly despised the
persons who so assiduously\index{assiduously} courted you.  There--I have saved
you the trouble of accounting for it; and really, all things
considered, I begin to think it perfectly reasonable.  To be
sure, you knew no actual good of me--but nobody thinks of
\emph{that} when they fall in love."

"Was there no good in your affectionate behaviour to Jane while
she was ill at Netherfield?"

"Dearest Jane! who could have done less for her?  But make a
virtue\index{virtue} of it by all means.  My good qualities are under your
protection, and you are to exaggerate them as much as possible;
and, in return, it belongs to me to find occasions for teasing
and quarrelling with you as often as may be; and I shall begin
directly by asking you what made you so unwilling to come to
the point at last.  What made you so shy of me, when you first
called, and afterwards dined here?  Why, especially, when you
called, did you look as if you did not care about me?"

"Because you were grave\index{grave} and silent, and gave me no encouragement."

"But I was embarrassed."

"And so was I."

"You might have talked to me more when you came to dinner."

"A man who had felt less, might."

"How unlucky that you should have a reasonable answer to give,
and that I should be so reasonable as to admit it!  But I
wonder how long you \emph{would} have gone on, if you had been left
to yourself.  I wonder when you \emph{would} have spoken, if I
had not asked you!  My resolution\index{resolution} of thanking you for your
kindness to Lydia had certainly great effect.  \emph{too} \emph{much}, I am
afraid; for what becomes of the moral, if our comfort springs
from a breach\index{breach} of promise?  for I ought not to have mentioned
the subject\index{subject}.  This will never do."

"You need not distress yourself.  The moral will be perfectly
fair.  Lady Catherine's unjustifiable endeavours\index{endeavours} to separate us
were the means of removing all my doubts.  I am not indebted
for my present happiness to your eager desire of expressing
your gratitude.  I was not in a humour to wait for any opening
of your's.  My aunt's intelligence had given me hope, and I was
determined at once to know every thing."

"Lady Catherine has been of infinite use, which ought to make
her happy, for she loves to be of use.  But tell me, what did
you come down to Netherfield for?  Was it merely to ride to
Longbourn and be embarrassed? or had you intended any more
serious consequence\index{consequence}?"

"My real purpose was to see \emph{you}, and to judge, if I could,
whether I might ever hope to make you love me.  My avowed\index{avowed} one,
or what I avowed\index{avowed} to myself, was to see whether your sister were
still partial\index{partial} to Bingley, and if she were, to make the
confession to him which I have since made."

"Shall you ever have courage to announce to Lady Catherine
what is to befall her?"

"I am more likely to want more time than courage, Elizabeth.
But it ought to done, and if you will give me a sheet of paper,
it shall be done directly."

"And if I had not a letter to write myself, I might sit by you
and admire the evenness of your writing, as another young lady
once did.  But I have an aunt, too, who must not be longer
neglected."

From an unwillingness to confess how much her intimacy with
Mr. Darcy had been over-rated, Elizabeth had never yet
answered Mrs. Gardiner's long letter; but now, having \emph{that}
to communicate which she knew would be most welcome, she was
almost ashamed to find that her uncle and aunt had already lost
three days of happiness, and immediately wrote as follows:

"I would have thanked you before, my dear aunt, as I ought
to have done, for your long, kind, satisfactory, detail of
particulars; but to say the truth, I was too cross to write.
You supposed more than really existed.  But \emph{now} suppose as
much as you choose; give a loose rein to your fancy, indulge your
imagination in every possible flight which the subject\index{subject} will
afford, and unless you believe me actually married, you cannot
greatly err.  You must write again very soon, and praise him a
great deal more than you did in your last.  I thank you, again
and again, for not going to the Lakes.  How could I be so silly
as to wish it!  Your idea of the ponies is delightful.  We will
go round the Park every day.  I am the happiest creature in the
world.  Perhaps other people have said so before, but not one
with such justice.  I am happier even than Jane; she only
smiles, I laugh.  Mr. Darcy sends you all the love in the world
that he can spare from me.  You are all to come to Pemberley at
Christmas.  Yours, etc."

Mr. Darcy's letter to Lady Catherine was in a different style;
and still different from either was what Mr. Bennet sent to
Mr. Collins, in reply to his last.

"\emph{dear} \emph{sir},

"I must trouble you once more for congratulations.  Elizabeth
will soon be the wife of Mr. Darcy.  Console\index{Console} Lady Catherine
as well as you can.  But, if I were you, I would stand by the
nephew.  He has more to give.

"Yours sincerely, etc."

Miss Bingley's congratulations to her brother, on his approaching
marriage, were all that was affectionate and insincere.  She
wrote even to Jane on the occasion, to express her delight, and
repeat all her former professions of regard. Jane was not
deceived, but she was affected; and though feeling no reliance
on her, could not help writing her a much kinder answer than
she knew was deserved.

The joy which Miss Darcy expressed on receiving similar
information, was as sincere as her brother's in sending it.
Four sides of paper were insufficient to contain all her
delight, and all her earnest\index{earnest} desire of being loved by her
sister.

Before any answer could arrive from Mr. Collins, or any
congratulations to Elizabeth from his wife, the Longbourn
family heard that the Collinses were come themselves to Lucas
Lodge.  The reason of this sudden removal was soon evident.
Lady Catherine had been rendered\index{rendered} so exceedingly angry by
the contents of her nephew's letter, that Charlotte, really
rejoicing in the match, was anxious to get away till the
storm was blown over.  At such a moment, the arrival of
her friend was a sincere pleasure to Elizabeth, though in
the course of their meetings she must sometimes think the
pleasure dearly bought, when she saw Mr. Darcy exposed to all
the parading and obsequious\index{obsequious} civility\index{civility} of her husband.  He bore
it, however, with admirable calmness.  He could even listen to
Sir William Lucas, when he complimented him on carrying away
the brightest jewel of the country, and expressed his hopes of
their all meeting frequently at St. James's, with very decent
composure\index{composure}.  If he did shrug his shoulders, it was not till Sir
William was out of sight.

Mrs. Phillips's \textbf{vulgarity}\index{vulgarity}\stepcounter{footnote}\footnotetext{
\textbf{vulgarity} \emph{n.}  (vuhl-GAIR-ih-tee)
lacking taste and refinement.} was another, and perhaps a greater,
tax on his forbearance\index{forbearance}; and though Mrs. Phillips, as well as
her sister, stood in too much awe\index{awe} of him to speak with the
familiarity which Bingley's good humour encouraged, yet,
whenever she \emph{did} speak, she must be vulgar\index{vulgar}.  Nor was her
respect for him, though it made her more quiet, at all likely
to make her more elegant.  Elizabeth did all she could to
shield him from the frequent notice of either, and was ever
anxious to keep him to herself, and to those of her family with
whom he might converse without mortification\index{mortification}; and though the
uncomfortable feelings arising from all this took from the
season of courtship much of its pleasure, it added to the hope
of the future; and she looked forward\index{forward} with delight to the time
when they should be removed from society so little pleasing to
either, to all the comfort and elegance of their family party
at Pemberley.



\chapter{61}


Happy for all her maternal feelings was the day on which
Mrs. Bennet got rid of her two most deserving daughters.
With what delighted pride she afterwards visited Mrs. Bingley,
and talked of Mrs. Darcy, may be guessed.  I wish I could say,
for the sake of her family, that the accomplishment of her
earnest\index{earnest} desire in the establishment of so many of her children
produced so happy an effect as to make her a sensible, amiable\index{amiable},
well-informed woman for the rest of her life; though perhaps it
was lucky for her husband, who might not have relished domestic\index{domestic}
felicity\index{felicity} in so unusual a form, that she still was occasionally
nervous and invariably silly.

Mr. Bennet missed his second daughter exceedingly; his
affection for her drew him oftener from home than anything
else could do.  He delighted in going to Pemberley, especially
when he was least expected.

Mr. Bingley and Jane remained at Netherfield only a twelvemonth.
So near a vicinity to her mother and Meryton relations was not
desirable even to \emph{his} easy temper, or \emph{her} affectionate heart.
The darling wish of his sisters was then gratified; he bought
an estate in a neighbouring county to Derbyshire, and Jane and
Elizabeth, in addition to every other source of happiness, were
within thirty miles of each other.

Kitty, to her very material advantage, spent the chief of her
time with her two elder sisters.  In society so superior\index{superior} to
what she had generally known, her improvement was great.  She
was not of so ungovernable a temper as Lydia; and, removed from
the influence of Lydia's example, she became, by proper
attention and management, less irritable, less ignorant, and
less insipid\index{insipid}.  From the further disadvantage of Lydia's society
she was of course carefully kept, and though Mrs. Wickham
frequently invited her to come and stay with her, with the
promise of balls and young men, her father would never consent
to her going.

Mary was the only daughter who remained at home; and she was
necessarily drawn from the pursuit of accomplishments by
Mrs. Bennet's being quite unable to sit alone.  Mary was
obliged to mix more with the world, but she could still
moralize over every morning visit; and as she was no longer
mortified by comparisons between her sisters' beauty and her
own, it was suspected by her father that she submitted to
the change without much reluctance.

As for Wickham and Lydia, their characters suffered no
\textbf{revolution}\index{revolution}\stepcounter{footnote}\footnotetext{
\textbf{revolution} \emph{n.}  (reh-vuh-LOO-shun)
the overthrow of a government.} from the marriage of her sisters.  He bore with
philosophy\index{philosophy} the conviction\index{conviction} that Elizabeth must now become
acquainted with whatever of his ingratitude and falsehood
had before been unknown to her; and in spite of every thing,
was not wholly\index{wholly} without hope that Darcy might yet be prevailed\index{prevailed}
on to make his fortune.  The congratulatory letter which
Elizabeth received from Lydia on her marriage, explained to
her that, by his wife at least, if not by himself, such a
hope was cherished.  The letter was to this effect:

"\emph{my} \emph{dear} \emph{lizzy},

"I wish you joy.  If you love Mr. Darcy half as well as I do my
dear Wickham, you must be very happy.  It is a great comfort to
have you so rich, and when you have nothing else to do, I hope
you will think of us.  I am sure Wickham would like a place at
court very much, and I do not think we shall have quite money
enough to live upon without some help.  Any place would do, of
about three or four hundred a year; but however, do not speak
to Mr. Darcy about it, if you had rather not.

"Yours, etc."

As it happened that Elizabeth had \emph{much} rather not, she
endeavoured\index{endeavoured} in her answer to put an end to every entreaty\index{entreaty}
and expectation of the kind.  Such relief, however, as it
was in her power to afford, by the practice of what might be
called economy\index{economy} in her own private\index{private} expences, she frequently
sent them.  It had always been evident to her that such an
income as theirs, under the direction of two persons so
extravagant in their wants, and heedless of the future, must
be very insufficient to their support; and whenever they
changed their quarters, either Jane or herself were sure of
being applied to for some little assistance towards discharging
their bills.  Their manner of living, even when the restoration
of peace dismissed them to a home, was unsettled in the
extreme.  They were always moving from place to place in quest
of a cheap situation, and always spending more than they ought.
His affection for her soon sunk into indifference\index{indifference}; her's lasted
a little longer; and in spite of her youth and her manners, she
retained all the claims to reputation which her marriage had
given her.

Though Darcy could never receive \emph{him} at Pemberley, yet, for
Elizabeth's sake, he assisted him further in his profession.
Lydia was occasionally a visitor there, when her husband was
gone to enjoy himself in London or Bath; and with the Bingleys
they both of them frequently staid so long, that even Bingley's
good humour was overcome, and he proceeded so far as to talk
of giving them a hint to be gone.

Miss Bingley was very deeply mortified by Darcy's marriage; but
as she thought it advisable to retain the right of visiting at
Pemberley, she dropt all her resentment\index{resentment}; was fonder than ever
of Georgiana, almost as attentive to Darcy as heretofore, and
paid off every \textbf{arrear}\index{arrear}\stepcounter{footnote}\footnotetext{
\textbf{} \emph{}  ()
} of civility\index{civility} to Elizabeth.

Pemberley was now Georgiana's home; and the attachment of the
sisters was exactly what Darcy had hoped to see.  They were able
to love each other even as well as they intended.  Georgiana had
the highest opinion in the world of Elizabeth; though at first
she often listened with an astonishment bordering on alarm at
her lively, sportive, manner of talking to her brother.  He, who
had always inspired in herself a respect which almost overcame
her affection, she now saw the object of open pleasantry.  Her
mind received knowledge which had never before fallen in her way.
By Elizabeth's instructions, she began to comprehend\index{comprehend} that a woman
may take liberties with her husband which a brother will not
always allow in a sister more than ten years younger than himself.

Lady Catherine was extremely indignant\index{indignant} on the marriage of her
nephew; and as she gave way to all the genuine frankness of her
character in her reply to the letter which announced its
arrangement, she sent him language so very abusive, especially
of Elizabeth, that for some time all intercourse\index{intercourse} was at an end.
But at length, by Elizabeth's persuasion, he was prevailed\index{prevailed} on
to overlook the offence, and seek a reconciliation\index{reconciliation}; and, after
a little further resistance on the part of his aunt, her
resentment\index{resentment} gave way, either to her affection for him, or her
curiosity to see how his wife conducted herself; and she
condescended\index{condescended} to wait on them at Pemberley, in spite of that
pollution which its woods had received, not merely from the
presence of such a mistress, but the visits of her uncle and
aunt from the city.

With the Gardiners, they were always on the most intimate
terms.  Darcy, as well as Elizabeth, really loved them; and
they were both ever sensible of the warmest gratitude towards
the persons who, by bringing her into Derbyshire, had been the
means of uniting them.




\backmatter
\newpage\textbf{APPENDIX}



\textbf{affectation} \emph{n.}   a deliberate pretense or exaggerated display.

\textbf{implacable} \emph{adj.}   impossible to appease, sooth, or placate.

\textbf{Michaelmas} \emph{n.}   A Christian feast honoring the archangel Michael.

\textbf{prejudice} \emph{n.}   an opinion that is preconceived and (usually) unfavorable.

\textbf{resignation} \emph{n.}   acceptance of despair; surrender.

\textbf{abatement} \emph{n.}   an interruption in the intensity or amount of something.

\textbf{abhorrence} \emph{n.}   hate coupled with disgust.

\textbf{abhorrent} \emph{adj.}   offensive to the mind; horrible or disgusting.

\textbf{ablution} \emph{n.}   washing or imersing one's body in water, especially as a religious act.

\textbf{abode} \emph{v.}    a person's legal place of residence.

\textbf{abominable} \emph{adj.} \textbf{-ly}   unequivocally detestable; morally repugnant.

\textbf{abominate} \emph{tr.v.}   to find repugnant, especially in a moral sense.

\textbf{abound} \emph{intr.v.}    to exist in large quantities.

\textbf{accost} \emph{tr.v.} \textbf{-ed}   to approach and verbally attack, affront, or make demand of.

\textbf{acquiesce} \emph{intr.v.}   to agree or express agreement; to consent.

\textbf{acquiescence} \emph{n.}   passive acceptance or agreement.

\textbf{acquit} \emph{tr.v.}    to pronounce not guilty of criminal charges.

\textbf{acrimony} \emph{n.}   sharp and bitter speech or behavior.

\textbf{actuate} \emph{tr.v.} \textbf{-ed}   to activate or cause to function.

\textbf{acute} \emph{adj.} \textbf{-ly}    having a rapid onset

\textbf{adept} \emph{adj.}    highly skilled; someone highly skilled.

\textbf{adhere} \emph{v.} \textbf{-ing}   to stick to (like glue); to be resolute in an idea or ideology.

\textbf{affable} \emph{adj.}    friendly and approachable; easy to talk to.

\textbf{affinity} \emph{n.}   a natural attraction or feeling of kinship.

\textbf{afflict} \emph{tr.v.} \textbf{-ed}    to cause pain or suffering.

\textbf{affliction} \emph{n.}   great suffering, distress, or pain.

\textbf{affront} \emph{tr.v.} \textbf{-ed}   a deliberately offensive act or statement.

\textbf{alacrity} \emph{n.}   liveliness and eagerness.

\textbf{alienate} \emph{tr.v.} \textbf{-ed}   to estrange; to disassociate with.

\textbf{allay} \emph{tr.v.} \textbf{-ed}   To relieve or reduce in intensity.

\textbf{alleviate} \emph{tr.v.} \textbf{-ed}   to provide physical relief, as from pain; make easier.

\textbf{alliance} \emph{n.}   an organization of people (or countries) in a pact or treaty.

\textbf{allusion} \emph{n.}   an indirect reference.

\textbf{aloof} \emph{adj.}    emotionally reserved or remote.

\textbf{amendment} \emph{n.}   a statement that revises, adds to, or improves a proposal or document.

\textbf{amends} \emph{pl.n.}   financial or other compensation for an injury or insult.

\textbf{amiable} \emph{adj.}   good-natured and likable.

\textbf{amiss} \emph{adj.}    not functioning properly; out of order.

\textbf{ample} \emph{adj.} \textbf{-ly}    more than enough in size, scope, or capacity.

\textbf{anecdote} \emph{n.}   short account of an incident (especially biographical).

\textbf{annex} \emph{tr.v.} \textbf{-ed}    to attach to something larger or more important.

\textbf{antagonist} \emph{n.}   someone who offers opposition; an adversary.

\textbf{antechamber} \emph{n.}   an entrance or reception room or area.

\textbf{apace} \emph{adv.}    quick, or with rapid movement.

\textbf{apothecary} \emph{n.}   a health professional who prepares and dispenses drugs; a pharmacist.

\textbf{appertain} \emph{intr.v.}   be a part or attribute of.

\textbf{application} \emph{n.}   a formal request for assistance, employment, or admission.

\textbf{apprehend} \emph{v.} \textbf{-ing}   To take into custody; to arrest.

\textbf{apprehension} \emph{n.}   fearful expectation or anticipation.

\textbf{apprehensive} \emph{adj.}   worried, anxious, or fearful.

\textbf{approbation} \emph{n.}   official recognition or approval.

\textbf{ardent} \emph{adj.} \textbf{-ly}    with strong emotion and intense enthusiasm.

\textbf{ascertain} \emph{tr.v.} \textbf{-ing}   to understand, discover, or clarify with certainty.

\textbf{aspire} \emph{intr.v.}   to have an ambitious plan or a lofty goal.

\textbf{assent} \emph{intr.v.} \textbf{-ed}   To agree with a proposal or statement.

\textbf{assertion} \emph{n.}   a declaration made as if no supporting evidence were necessary.

\textbf{assiduous} \emph{adj.} \textbf{-ly}   marked by care and persistent effort.

\textbf{atone} \emph{v.} \textbf{-ed}    to correct or make right a sin or transgression.

\textbf{augment} \emph{v.}   to enlarge, increase, or intensify.

\textbf{austerity} \emph{n.}   the trait of great self-denial (especially from worldly pleasures).

\textbf{authoritative} \emph{adj.}   of recognized authority or excellence.

\textbf{avail} \emph{v.}    to be useful to; to help.

\textbf{avarice} \emph{n.}   insatiable desire for wealth; greed.

\textbf{avowal} \emph{n.}    a sincere acknowledgment or assertion.

\textbf{avow} \emph{tr.v.} \textbf{-ed}    to boldly and openly declare.

\textbf{awe} \emph{n.} \textbf{-ed}    a state inspired by fearful wonderment or reverence.

\textbf{balm} \emph{n.}    any substance used for healing and soothing; ointment.

\textbf{barbarous} \emph{adj.} \textbf{-ly}   disposed to inflicting pain or suffering.

\textbf{beau} \emph{n.} \textbf{-x}    A boyfriend.

\textbf{beneficence} \emph{n.}   quality of being kind, good, or charitable.

\textbf{benevolence} \emph{n.}   an inclination toward kind or charitable acts.

\textbf{benevolent} \emph{adj.}   intending or showing kindness; generous.

\textbf{bequest} \emph{n.}   (legal) a gift of personal property by will.

\textbf{bestow} \emph{tr.v.} \textbf{-ed}    to give something formally or officially.

\textbf{bewail} \emph{tr.v.} \textbf{-ed}   to lament or cry intensely over.

\textbf{bewitch} \emph{tr.v.} \textbf{-ed}   to put under a spell as if by magic.

\textbf{Boulanger} \emph{1}   French music teacher of several modern American composers, including Virgil Thomson and Aaron Copland.

\textbf{breach} \emph{n.}   an opening (especially a gap in a dike or fortification).

\textbf{brevity} \emph{n.}   being short or brief in time; short and concise in expression.

\textbf{candid} \emph{adj.}   blunt directness in manner or speech; without subtlety or evasion.

\textbf{capital} \emph{n.}   assets available for use in the production of further assets; money currently in position (not credit).

\textbf{caprice} \emph{n.}   A whimsical change of mind; an impulsive desire.

\textbf{casino} \emph{n.}   a public room for gambling.

\textbf{celerity} \emph{n.}   speedy rapid action.

\textbf{censure} \emph{n.} \textbf{-ed}    A strong expression of criticism or disapproval.

\textbf{ceremonious} \emph{adj.}   rigidly formal or bound by convention or ritual.

\textbf{cessation} \emph{n.}   a stopping.

\textbf{chagrin} \emph{n.}   strong feelings of embarrassment or hurt pride.

\textbf{chaise} \emph{n.}    a two-wheeled carriage drawn by one horse, often with a collapsible hood.

\textbf{chaperon} \emph{n.}   someone who supervises a young woman or gatherings of young people.

\textbf{circulation} \emph{n.}   opening to public discussion and debate; distribution of a publication.

\textbf{circumspect} \emph{adj.}   heedful of potential consequences; prudent.

\textbf{circumspection} \emph{n.}   knowing how to avoid embarrassment or distress; prudent.

\textbf{civility} \emph{n.}   formal or perfunctory politeness.

\textbf{clamorous} \emph{adj.}   a sustained din as if from a crowd; offensively loud.

\textbf{coax} \emph{v.} \textbf{-ed}    to persuade, cajole, or convince by flattery.

\textbf{cogent} \emph{adj.}     powerfully persuasive.

\textbf{combat} \emph{v.} \textbf{-ed}    to fight in battle.

\textbf{commend} \emph{tr.v.} \textbf{-ed}    to praise as admirable; to reccomend.

\textbf{commendation} \emph{n.}   An act of praise, bestowing honor and public recognition.

\textbf{comparative} \emph{adj.}   something's significance in relation to something else.

\textbf{compass} \emph{n.}    an area in which something operates or has power or control.

\textbf{complacency} \emph{n.}   a feeling of ease or satisfaction, often causing unawareness of danger.

\textbf{complaisance} \emph{n.}   a willingness or tendency to yield to the will of others; amiability.

\textbf{compliance} \emph{n.}   the act of submitting; acquiescence.

\textbf{comply} \emph{intr.v.} \textbf{-ing}   to follow another person's order; to acquiesce or act in accord.

\textbf{composure} \emph{n.}   steadiness of mind under stress.

\textbf{comprehend} \emph{tr.v.} \textbf{-ed}   to fully understand or grasp.

\textbf{conceit} \emph{n.} \textbf{-ed}    false pride; an exaggerated sense of self-importance.

\textbf{concerto} \emph{n.}   a musical composition for orchestra and a soloist.

\textbf{conciliate} \emph{v.} \textbf{-ing}   overcoming animosity or hostility; to appease.

\textbf{condescend} \emph{intr.v.} \textbf{-ed}   do something considered beneath one's dignity; to lower oneself.

\textbf{condescension} \emph{n.}   a disrespecetful communication that patronizes the recipient.

\textbf{condole} \emph{intr.v.}    to express sympathy for another's grief.

\textbf{condolence} \emph{n.}   an expression of sympathy with another's grief.

\textbf{confidante} \emph{n.}   a female confidant; a woman trusted with secrets.

\textbf{confidence} \emph{n.}   a feeling of trust (in someone or something).

\textbf{congenial} \emph{adj.}   suitable to your needs, taste, or nature; warm and friendly.

\textbf{conjecture} \emph{n.} \textbf{-ed}   Inference from incomplete evidence; a logical guess.

\textbf{conjugal} \emph{adj.}   relating to marriage.

\textbf{conjunction} \emph{n.}   something that joins or connects.

\textbf{connivance} \emph{n.}   agreement on a secret plot.

\textbf{connubial} \emph{adj.}   related to marriage; in conjugal or familliar manner as if married.

\textbf{conscience} \emph{n.}   a feeling of shame when you do something immoral.

\textbf{conscientious} \emph{adj.} \textbf{-ly}   careful or meticulous.

\textbf{consequence} \emph{n.}   the natural or unavoidable result of an action.

\textbf{consign} \emph{v.} \textbf{-ed}   give over to another for care or safekeeping.

\textbf{consolation} \emph{n.}   the act of consoling; giving relief to someone in affliction.

\textbf{console} \emph{tr.v.}   To give comfort or solace to.

\textbf{consternation} \emph{n.}   a combination of fear, amazement, and dismay in face of danger.

\textbf{constitute} \emph{tr.v.}   to form or compose.

\textbf{construe} \emph{v.} \textbf{-ing}   to make sense of; to interpret.

\textbf{contemplation} \emph{n.}   a deep thinking session about a particular topic; a long consideration.

\textbf{contempt} \emph{n.}   open disrespect for a person or thing.

\textbf{contentment} \emph{n.}   satisfaction; happiness with one's situation in life.

\textbf{contrivance} \emph{n.}   a device that is useful for a particular job; any improvised arrangement for temporary use.

\textbf{conviction} \emph{n.}   an unshakable belief not needing proof or evidence.

\textbf{coppice} \emph{n.}   a dense growth of bushes.

\textbf{copse} \emph{n.}    a dense growth of bushes.

\textbf{coquetry} \emph{n.}   playful behavior intended to arouse sexual interest.

\textbf{cordial} \emph{adj.} \textbf{-ly}   friendly, warm, or sincere; a strong and highly flavored sweet liquor usually drunk after a meal.

\textbf{corps} \emph{n.}    an army unit usually consisting of two or more divisions.

\textbf{corroborate} \emph{tr.v.}   to support or establish by evidence or proof.

\textbf{countenance} \emph{n.}   facial appearance or expression; to give permission.

\textbf{culprit} \emph{n.}   someone who perpetrates a crime or wrongdoing.

\textbf{cultivation} \emph{n.}   socialization through training and education.

\textbf{cunning} \emph{adj.}   shrewdness as shown by being skilled in deception.

\textbf{curricle} \emph{n.}   An open, two-wheeled carriage, drawn by two horses side by side.

\textbf{curtail} \emph{tr.v.} \textbf{-ed}    to cut short.

\textbf{dawdle} \emph{v.} \textbf{-ed}    To take more time than necessary.

\textbf{decamp} \emph{intr.v.} \textbf{-ing}   To leave or depart suddenly or secretly.

\textbf{decorum} \emph{n.}   propriety in manners or conduct.

\textbf{deduce} \emph{tr.v.} \textbf{-ed}   to reach a conclusion logically.

\textbf{default} \emph{n.}   failing to meet a financial obligation.

\textbf{deference} \emph{n.}   a tendency to yield to the will of others.

\textbf{defer} \emph{v.} \textbf{-ed}    To postpone or put off.

\textbf{defiance} \emph{n.}   brazen resistance to authority.

\textbf{degenerate} \emph{adj.}   a person who deviates from acceptable norms, especially in sexual behavior.

\textbf{degradation} \emph{n.}   changing to a lower, or less respected, state or status.

\textbf{deign} \emph{v.} \textbf{-ed}    To condescend to give or bestow; to grant.

\textbf{dejection} \emph{n.}   a state of melancholy depression; feeling down.

\textbf{deliberation} \emph{n.}   careful consideration.

\textbf{delicacy} \emph{n.}   something considered choice to eat, often rare.

\textbf{demean} \emph{tr.v.}    to debase in character; to disgrace or dishonor.

\textbf{deportment} \emph{n.}   behavior or personal conduct.

\textbf{depravity} \emph{n.}   moral perversion; a degenerate act or practice.

\textbf{depreciate} \emph{v.}   to lower the price or value of something; to belittle.

\textbf{derange} \emph{tr.v.} \textbf{-ed}   to make insane.

\textbf{derision} \emph{n.}   exposing someone to laughter or ridicule.

\textbf{descent} \emph{n.}   a downward slope.

\textbf{despond} \emph{intr.v.} \textbf{-ing}   to become discouraged, gloomy, or feel let down.

\textbf{destitute} \emph{adj.}   poor enough to need help from others.

\textbf{deter} \emph{v.}    to persuade someone not to do something.

\textbf{dialog} \emph{n.} \textbf{-ue}   a conversation between two people.

\textbf{dictatorial} \emph{adj.}   domineering; characteristic of an absolute ruler.

\textbf{diffidence} \emph{n.}   a lack of self-confidence.

\textbf{diffident} \emph{adj.}   lacking self-confidence.

\textbf{diffuse} \emph{v.} \textbf{-ed}    spread out through a wide area or large volume.

\textbf{digression} \emph{n.}   an act of wandering from the main path or subject matter.

\textbf{dilatory} \emph{adj.}   inclined to waste time and lag behind; intentionally delaying.

\textbf{diligence} \emph{n.}   conscientiousness and persistent in perform a task.

\textbf{diminution} \emph{n.}   a decrease or reduction; an act of making smaller.

\textbf{disapprobation} \emph{n.}   an expression of strong disapproval.

\textbf{disarm} \emph{v.}   remove offensive capability from; to take weapons away from.

\textbf{discernible} \emph{adj.}   perceptible by the senses or intellect.

\textbf{discernment} \emph{n.}   the trait of judging wisely and objectively.

\textbf{discompose} \emph{tr.v.} \textbf{-ed}   To disturb a person's composure; to ruffle or disturb.

\textbf{disconcert} \emph{tr.v.} \textbf{-ed}   to upset one's self-possession or self-bearing; to alarm.

\textbf{discretion} \emph{n.}   knowing how to avoid embarrassment or distress; tact

\textbf{dishearten} \emph{tr.v.} \textbf{-ed}   to reduce in hope or enthusiasm.

\textbf{disinclination} \emph{n.}   a certain degree of unwillingness; reluctance.

\textbf{disoblige} \emph{tr.v.} \textbf{-ing}   to be intentionally unaccommodating.

\textbf{despatch} \emph{v.} \textbf{-ed}   to send off or away; same as 'dispatch'.

\textbf{dispirit} \emph{tr.v.} \textbf{-ed}   showing no enthusiasm; forlorn or dejected.

\textbf{dispose} \emph{v.} \textbf{-ed}    a personal inclination (usually followed by `to').

\textbf{disposition} \emph{n.}   One's usual mood; temperament.

\textbf{dissemble} \emph{v.}   to hide under a false appearance.

\textbf{dissuade} \emph{tr.v.}   to turn away from by persuasion.

\textbf{distracted} \emph{adj.}   having one's attention diverted especially because of anxiety.

\textbf{divert} \emph{v.} \textbf{-ed}    to turn aside; to turn away from.

\textbf{doleful} \emph{adj.}   filled with or evoking sadness.

\textbf{domestic} \emph{adj.}   relating to the home or family; a household servant.

\textbf{draught} \emph{n.}    same as draft.

\textbf{droop} \emph{v.} \textbf{-ing}    hanging down (as from exhaustion or weakness).

\textbf{ductile} \emph{adj.}   Easily formed into wire or hammered into thin sheets.

\textbf{dupe} \emph{n.} \textbf{-ed}   to trick or deceive.

\textbf{duplicity} \emph{n.}   a fraudulent or deceptive representation.

\textbf{earnest} \emph{adj.} \textbf{-ly}    sincerity or firm belief in the validity of one's opinions.

\textbf{economy} \emph{n.}   the efficient use of resources.

\textbf{ecstasy} \emph{n.}    a state of elated bliss.

\textbf{efficacy} \emph{n.}   capacity or power to produce a desired effect or result.

\textbf{effusion} \emph{n.}   an unrestrained expression of emotion; an outburst.

\textbf{elope} \emph{intr.v.} \textbf{-ment}    the act of running away with a lover (usually to get married).

\textbf{eloquence} \emph{n.}   powerful and effective language.

\textbf{eloquent} \emph{adj.}   expressing oneself readily, clearly, and effectively.

\textbf{elude} \emph{tr.v.} \textbf{-ed}    To evade or escape from persuers; to avoid capture.

\textbf{embargo} \emph{n.}    a government order imposing a trade barrier.

\textbf{eminence} \emph{n.}   high status importance owing to marked superiority.

\textbf{eminent} \emph{adj.}   standing above others in quality or position.

\textbf{emphasis} \emph{n.}   special importance or significance.

\textbf{emphatic} \emph{adj.} \textbf{-ly}   forceful and definite in expression or action.

\textbf{encroach} \emph{intr.v.} \textbf{-ing}   to gradually intrude or take over something without right or permision.

\textbf{encumbrance} \emph{n.}   a burden or obstacle.

\textbf{endeavour} \emph{n.}   a project or undertaking; to try; variant of endeavor.

\textbf{endurable} \emph{adj.}   capable of being tolerated although unpleasant.

\textbf{engross} \emph{tr.v.} \textbf{-ed}   to enclose or envelop completely, as if by swallowing; to fully occupy.

\textbf{ensign} \emph{n.}   an emblem displaying as symbol of nationality, often with insignia of a specific military unit.

\textbf{entail} \emph{tr.v.} \textbf{-ed}    To have as a necessary clause, condition, or consequence; to include.

\textbf{entreat} \emph{v.} \textbf{-ed}    to make a forthright, sincere request of.

\textbf{entreaty} \emph{n.}   a sincere plea or request.

\textbf{enumerate} \emph{tr.v.}   to list or count off item by item.

\textbf{environ} \emph{tr.v.}    the area in which something exists or lives, especially in a city.

\textbf{equipage} \emph{n.}   a wheeled vehicle drawn by horses.

\textbf{equivocal} \emph{adj.}   open to two or more interpretations; (often) intended to mislead.

\textbf{estimable} \emph{adj.}   may be computed or estimated.

\textbf{etiquette} \emph{n.}   rules governing socially acceptable behavior.

\textbf{exasperate} \emph{tr.v.}   to make worse; to make very impatient or annoyed.

\textbf{execution} \emph{n.}   completion of an order or plan.

\textbf{exertion} \emph{n.}   strenuous use of physical or mental energy; hard work.

\textbf{exigence} \emph{n.}   a pressing situation requiring immediate attention.

\textbf{expedite} \emph{tr.v.}   speed up the progress of; facilitate.

\textbf{expeditious} \emph{adj.} \textbf{-ly}   in an efficient manner.

\textbf{explicit} \emph{adj.} \textbf{-ly}   absolutely clear and understood; leaving nothing to implication.

\textbf{expostulate} \emph{intr.v.}   an expression of opposition to a course of action.

\textbf{extenuate} \emph{tr.v.} \textbf{-ing}   to lessen or to try to lessen the seriousness or extent of.

\textbf{extort} \emph{tr.v.} \textbf{-ed}   obtain by coercion or intimidation.

\textbf{extract} \emph{tr.v.}   To draw or pull out; a thing in concentrate form.

\textbf{exuberance} \emph{n.}   joyfully overflowing with enthusiasm.

\textbf{exuberant} \emph{adj.}   feeling unrestrained joy.

\textbf{exultation} \emph{n.}   a feeling of extreme joy.

\textbf{faculty} \emph{n.}   An inherent power or ability.

\textbf{fashion} \emph{n.}    to make out of components, often in an improvising manner.

\textbf{fastidious} \emph{adj.}   careful attention to detail; overly concerned with cleanliness.

\textbf{felicitation} \emph{n.}   Congratulations.

\textbf{felicity} \emph{n.}   state of positive emotions ranging from contentment to intense joy.

\textbf{fervent} \emph{adj.} \textbf{-ly}   with passionate zeal or fervor.

\textbf{filial} \emph{adj.}    relating to a son or daughtor.

\textbf{folly} \emph{n.}    a foolish act lacking judgement or foresight.

\textbf{forbearance} \emph{n.}   good-natured tolerance of delay or incompetence.

\textbf{forbear} \emph{v.}   to restrain oneself or to refrain from.

\textbf{forlorn} \emph{adj.}    hopelessness; a pitiable feeling of abandonment.

\textbf{formidable} \emph{adj.}   extremely impressive in strength or excellence; inspiring fear.

\textbf{fortitude} \emph{n.}   strength of mind that enables one to endure adversity with courage.

\textbf{fortnight} \emph{n.}   a period of fourteen consecutive days.

\textbf{forward} \emph{adj.}    lacking restraint or modesty.

\textbf{fret} \emph{n.}    agitation resulting from active worry.

\textbf{fretful} \emph{adj.} \textbf{-ness}    an irritable or petulant feeling; troubled through nagging anxiety.

\textbf{frivolous} \emph{adj.}   not serious in content, attitude, or behavior; trivial.

\textbf{gaiety} \emph{n.}   a merry, festive feeling.

\textbf{gallantry} \emph{n.}   heroic courage in face of danger; courtesy towards women.

\textbf{gaudy} \emph{adj.}    tastelessly showy.

\textbf{genteel} \emph{adj.}   marked by refinement in taste and manners; well-bred and polite.

\textbf{glen} \emph{n.}    A narrow, secluded valley (in the mountains).

\textbf{godson} \emph{n.}   A male godchild.

\textbf{grandeur} \emph{n.}   the quality of being magnificent, splendid, or grand.

\textbf{grave} \emph{adj.} \textbf{-ly}    of great gravity or crucial import; requiring serious thought.

\textbf{gravity} \emph{n.}    a serious and solemn manner.

\textbf{grievance} \emph{n.}   a situation or curcumstance viewed as reason for complaint.

\textbf{grievous} \emph{adj.}   shockingly brutal or cruel; causing great anguish.

\textbf{habitual} \emph{adj.}   made a norm, custom, or habit.

\textbf{hack} \emph{v.}    cough spasmodically.

\textbf{hackneyed} \emph{adj.}   Overfamiliar through overuse; trite.

\textbf{haughty} \emph{adj.}   showing arrogant superiority to and disdain of those one views as inferior.

\textbf{haunch} \emph{n.}   the hip, buttock, and upper thigh.

\textbf{hauteur} \emph{n.}   Arrogance or haughtiness.

\textbf{headstrong} \emph{adj.}   habitually disposed to disobedience and opposition.

\textbf{hearth} \emph{n.}    the floor of a fireplace, often extending out into the room.

\textbf{heinous} \emph{adj.}    shockingly brutal or cruel.

\textbf{heir} \emph{n.}    a person who inherits.

\textbf{Hermitage} \emph{n.}   A full-bodied wine made in southeastern France.

\textbf{hitherto} \emph{th}   before; until this time.

\textbf{humility} \emph{n.}   the quality of being humble or modest; without pride or arrogance.

\textbf{hypocrisy} \emph{n.}   professing beliefs, values, or sentiments one does not truly believe or live by; insincerity; falseness.

\textbf{hypocritical} \emph{adj.}   professing feelings or virtues one does not have.

\textbf{illiberal} \emph{adj.}   narrow-minded or bigoted about one's opinions.

\textbf{illiterate} \emph{adj.}   not able to read or write .

\textbf{illustrious} \emph{adj.}   widely known and esteemed.

\textbf{immoral} \emph{adj.}   not adhering to ethical or moral principles; impure; licentious; wicked.

\textbf{impartial} \emph{adj.} \textbf{-ity}   unprejudiced; free from undue bias or preconceived opinions.

\textbf{impassable} \emph{adj.}   impossible to pass.

\textbf{impel} \emph{tr.v.} \textbf{-ed}    to urge or force to action through moral pressure.

\textbf{impertinence} \emph{n.}   the trait of being rude and insolent; inappropriate playfullness.

\textbf{impertinent} \emph{adj.}   improperly forward or bold.

\textbf{imperturbable} \emph{adj.} \textbf{-ly}   Cool and unshakable demeanor.

\textbf{impetuous} \emph{adj.}   undue haste; impulsive and suddenly forceful.

\textbf{implicit} \emph{adj.} \textbf{-ness}   implied; inherent in the nature of something.

\textbf{impolitic} \emph{adj.}    Socially unwise; indiscreet; inexpedient; undiplomatic.

\textbf{import} \emph{v.} \textbf{-ing}   comercially bringing in goods from a foreign country, usually for resale.

\textbf{importune} \emph{v.} \textbf{-ed}   to beg persistently and urgently.

\textbf{impose} \emph{v.} \textbf{-ed}   to establish as obligatory.

\textbf{improbable} \emph{adj.}   not likely to happen or be true.

\textbf{imprudence} \emph{n.}   a lack of caution in practical affairs

\textbf{imprudent} \emph{adj.}   lacking wise self-restraint; indiscreet.

\textbf{impudence} \emph{n.}   offensive boldness; rude and impertinent.

\textbf{impudent} \emph{adj.}   improperly forward or bold; insolent.

\textbf{impunity} \emph{n.}   exemption from punishment or loss.

\textbf{impurity} \emph{n.}   Contamination, pollution, or a foreign substance.

\textbf{impute} \emph{tr.v.} \textbf{-ing}    to assign fault or responsibility to.

\textbf{incense} \emph{tr.v.} \textbf{-ed}   to cause anger or a strong feeling of injustice.

\textbf{incessant} \emph{adj.} \textbf{-ly}   constant without interruption.

\textbf{incivility} \emph{n.}   deliberate discourtesy.

\textbf{inclination} \emph{n.}   an attitude that favors one alternative over others.

\textbf{incredulity} \emph{n.}   doubt about the truth of something; disbelief.

\textbf{incredulous} \emph{adj.}   not willing to believe; skeptical.

\textbf{incur} \emph{tr.v.} \textbf{-ed}    to make oneself subject to (something bad); to sustain (ie. damage).

\textbf{indecorum} \emph{n.}   an act of impropriety or undue intimacy.

\textbf{indifference} \emph{n.}   the trait of lacking enthusiasm or interest; apathy.

\textbf{indifferent} \emph{adj.}   (one with) a lack of interest; apathetic.

\textbf{indignant} \emph{adj.}   angered at something unjust or wrong.

\textbf{indignation} \emph{n.}   a feeling of righteous anger.

\textbf{indignity} \emph{n.}   insulting, humiliating, or degrading treatment.

\textbf{indispose} \emph{tr.v.} \textbf{-ed}   (usually followed by `to') strongly opposed; disinclined.

\textbf{indolence} \emph{n.}   laziness as a lifestyle; sloth.

\textbf{indolent} \emph{adj.}   lazy as a lifestyle.

\textbf{induce} \emph{tr.v.} \textbf{-ed}   to cause or bring about; to influence or persuade.

\textbf{inducement} \emph{n.}   a positive motivational influence.

\textbf{industrious} \emph{adj.} \textbf{-ly}   busy, dilligent, hard working, and productive.

\textbf{ineffectual} \emph{adj.}   not producing an intended effect.

\textbf{infamous} \emph{adj.}   Having an exceedingly bad reputation; notorious.

\textbf{infamy} \emph{n.}   Evil fame or reputation.

\textbf{inference} \emph{n.}   a logical conclusion drawn from factual information.

\textbf{inferior} \emph{adj.} \textbf{-ity}   lower in rank, importance, or quality.

\textbf{infliction} \emph{n.}   the act of imposing something unpleasant.

\textbf{ingenious} \emph{adj.}   marked by brilliant intellect, imagination, skill, or creativity.

\textbf{ingenuity} \emph{n.}   cleverness; the power of creative imagination.

\textbf{inhumanity} \emph{n.}   an act of atrocious cruelty.

\textbf{iniquitous} \emph{adj.}   wicked; violating the rights of others; nefarious; criminal.

\textbf{injunction} \emph{n.}   a formal command or admonition.

\textbf{injustice} \emph{n.}   a Violation of a person's rights; an unjust act.

\textbf{insipid} \emph{adj.} \textbf{-ity}    Lacking flavor or taste; trite, dull, unoriginal, or hackneyed.

\textbf{insolence} \emph{n.}   the trait of being rude and impertinent; inclined to take liberties.

\textbf{insolent} \emph{adj.}   marked by casual, arrogant disrespect.

\textbf{insufferable} \emph{adj.} \textbf{-ly}   hard to endure, intolerable.

\textbf{insupportable} \emph{adj.}   incapable of being justified or explained; intolerable.

\textbf{integrity} \emph{n.}   moral soundness.

\textbf{intercourse} \emph{n.}   communication between individuals.

\textbf{intermarry} \emph{intr.v.}   To marry a member of another group.

\textbf{intermission} \emph{n.}   a time of temporarily suspended activity.

\textbf{interpose} \emph{v.}   to insert between other parts.

\textbf{intolerable} \emph{adj.}   impossible to tolerate or endure; completely unacceptable.

\textbf{intrepid} \emph{adj.} \textbf{-ity}   resolute courageousness.

\textbf{enure} \emph{v.} \textbf{-ed}    to get used to something unpleasant; same as 'inure'.

\textbf{invalid} \emph{n.}   someone rendered non-functional by disability or other condition.

\textbf{invaluable} \emph{adj.}   priceless; having incalculable monetary worth.

\textbf{invariable} \emph{adj.} \textbf{-ly}   constant; not changing.

\textbf{invective} \emph{n.}   abusive language; words intended to discredit or denounce.

\textbf{involuntary} \emph{adj.} \textbf{-ly}   done against or without one's will.

\textbf{irksome} \emph{adj.}   Something that is annoying, boring, bothersome or tedious.

\textbf{irreligious} \emph{adj.}   hostile or indifferent to religion.

\textbf{irremediable} \emph{adj.}   impossible to remedy, correct, or redress.

\textbf{irrevocable} \emph{adj.} \textbf{-ly}   impossible to retract, take back, nullify, or cancle.

\textbf{jilt} \emph{tr.v.}    To deceive or drop (a lover) suddenly or callously, especially on the wedding day.

\textbf{kindred} \emph{n.}   similar or related in quality or character.

\textbf{laconic} \emph{adj.}    brief and to the point; concise.

\textbf{laity} \emph{n.}    everyone except the clergy.

\textbf{lament} \emph{v.} \textbf{-ing}   a cry of sorrow and grief.

\textbf{lamentation} \emph{n.}   a cry of sorrow and grief.

\textbf{languor} \emph{n.}    a feeling of lack of interest or energy.

\textbf{laudable} \emph{adj.}   worthy of high praise.

\textbf{legacy} \emph{n.}   a gift of personal property by will or inheritance.

\textbf{let} \emph{v.}    grant use or occupation of under terms of a contract.

\textbf{licentious} \emph{adj.} \textbf{-ness}   being lewd and lascivious especially in sexual conduct.

\textbf{lieu} \emph{n.}    in place of a position or function normally served by another.

\textbf{livery} \emph{n.}   a uniform worn by a menservant or chauffeurs.

\textbf{loo} \emph{n.}   a toilet in England.

\textbf{lustre} \emph{n.}   the shine of reflected light, especially on porcelain. Also 'luster'.

\textbf{magistrate} \emph{n.}   a public official authorized to decide questions bought before a court of justice.

\textbf{malice} \emph{n.}   hate or extreme ill will.

\textbf{malicious} \emph{adj.}   wishing evil to others; deliberately harmful.

\textbf{manifold} \emph{adj.}   Many and varied.

\textbf{matrimony} \emph{n.}   The state of being married.

\textbf{mayoralty} \emph{n.}   the position of mayor.

\textbf{melancholy} \emph{n.}   Sadness or depression.

\textbf{mien} \emph{n.}    bearing of mannor or conduct; dignified.

\textbf{militia} \emph{n.}    an army composed of civilians, usually for defense only.

\textbf{milliner} \emph{n.}   someone who makes and sells hats.

\textbf{mirth} \emph{n.}    great merriment, laughing, and gaiety.

\textbf{mischance} \emph{n.}   a mishap; bad luck.

\textbf{miserly} \emph{adj.}   avaricious or lacking generosity.

\textbf{monotonous} \emph{adj.}   tediously repetitious or lacking in variety; an unvarying tone or sound.

\textbf{morality} \emph{n.}   acting in line with right or good conduct.

\textbf{mortification} \emph{n.}   strong feelings of embarrassment.

\textbf{mortify} \emph{v.} \textbf{-ing}    to make someone feel great shame or humiliation.

\textbf{nettle} \emph{n.} \textbf{-ed}   to arouse impatience or anger.

\textbf{nominal} \emph{adj.} \textbf{-ly}    in name only.

\textbf{nonsensical} \emph{adj.}   so lacking in intelligence or good sense as to be laughable.

\textbf{novelty} \emph{n.}    originality by being new and surprising.

\textbf{nuptial} \emph{adj.}    the social event associated with a wedding ceremony.

\textbf{obeisance} \emph{n.}   a body movement, such as a bow or curtsy, intended to show respect or submission.

\textbf{obsequious} \emph{adj.} \textbf{-ness}   showing servile compliance; fawning.

\textbf{obstinacy} \emph{n.}   resolute adherence to one's own ideas; stubborn.

\textbf{obstinate} \emph{adj.}   stubborn.

\textbf{obtrude} \emph{v.} \textbf{-ed}   to forcibly impose (oneself or one's opinion) on others.

\textbf{oddity} \emph{n.}   something unusual, perhaps worthy of collecting.

\textbf{odious} \emph{adj.}   utterly detestable and hateful.

\textbf{officious} \emph{adj.} \textbf{-ness}   a strong inclination to advance oneself or one's ideas when unwanted.

\textbf{ordination} \emph{n.}   formal acceptance into the clergy.

\textbf{ostentation} \emph{n.}   a pretentious, showy, or vulgar display meant to impress others.

\textbf{ostentatious} \emph{adj.}   intended to attract notice and impress others; tawdry or vulgar.

\textbf{overture} \emph{n.}   A musical introduction to a larger work, such as an opera.

\textbf{palatable} \emph{adj.}   acceptable to the taste or mind; eatible.

\textbf{palliate} \emph{tr.v.}   to lessen the seriousness or extent of; to extenuate; to mitigate.

\textbf{paltry} \emph{adj.}    not worth considering; contemptibly small in amount.

\textbf{panegyric} \emph{n.}   formal public expression of praise.

\textbf{parish} \emph{n.}    a local church community.

\textbf{partake} \emph{v.}    to take or receive a share of; to participate.

\textbf{partial} \emph{adj.}   having a strong preference or liking for.

\textbf{partiality} \emph{n.}   a predisposition to like something.

\textbf{partridge} \emph{n.}   Any of several plump European game birds.

\textbf{patronage} \emph{n.}   customers seen as collective body.

\textbf{patroness} \emph{n.}   a woman who is a sponsor or benefactor.

\textbf{peculiar} \emph{adj.} \textbf{-ly}   strange, eccentric, or odd.

\textbf{pecuniary} \emph{adj.}   relating to or involving money.

\textbf{pedantic} \emph{adj.}   marked by a narrow, often trivial display of book learning.

\textbf{peevish} \emph{adj.}    easily irritated or annoyed.

\textbf{penance} \emph{n.}    remorse for past misconduct, often accompanied by self-inflicted suffering meant to atone.

\textbf{penitent} \emph{adj.}   feeling or expressing remorse for one's misdeeds.

\textbf{perceive} \emph{tr.v.}   to become aware of through the senses.

\textbf{perforce} \emph{adv.}   by necessity; by force of circumstance.

\textbf{perpetual} \emph{adj.} \textbf{-ly}   continuing forever or indefinitely.

\textbf{perplexity} \emph{n.}   the state of being puzzled or confused.

\textbf{perseverance} \emph{n.}   steadfast commitment to an action or toward a goal.

\textbf{perturbation} \emph{n.}   an unhappy and worried mental state.

\textbf{perturb} \emph{tr.v.} \textbf{-ed}   to agitate; to disturb deeply; to make uneasy.

\textbf{peruse} \emph{tr.v.}    to read or examine carefully, as if to remember.

\textbf{perverse} \emph{adj.} \textbf{-ness}   deliberate unruliness, often in a sexual or violent sense.

\textbf{petition} \emph{n.}   a formal request of an authority.

\textbf{petulant} \emph{adj.}   Overly irritable or peevish; easily annoyed.

\textbf{phaeton} \emph{n.}    a four-wheeled open carriage, usually drawn by two horses.

\textbf{philosophical} \emph{adj.}   meeting trouble with level-headed detachment.

\textbf{pianoforte} \emph{n.}   a musical instrument somewhat resembling a harpsichord.

\textbf{picturesque} \emph{adj.}   suitable for a picture; pretty as a picture.

\textbf{pique} \emph{n.} \textbf{-ed}    a feeling of wounded pride.

\textbf{piquet} \emph{n.}    a card game for two players using a reduced pack of 32 cards.

\textbf{plague} \emph{n.}   a serious and widespread disease epidemic, often seen as divine retribution.

\textbf{plantation} \emph{n.}   an estate where cash crops are grown on a large scale.

\textbf{pliant} \emph{adj.}   Easily bent; easily adaptable.

\textbf{pompous} \emph{adj.}   puffed up with vanity or exaggerated dignity.

\textbf{posterity} \emph{n.}   Future generations.

\textbf{precipitancy} \emph{n.}   the quality of happening with headlong haste or without warning.

\textbf{precipitate} \emph{v.}   to bring about abruptly.

\textbf{premeditate} \emph{v.} \textbf{-ed}   to preconsider or plan in advance.

\textbf{premeditation} \emph{n.}   planning or plotting in advance of acting.

\textbf{prepossession} \emph{n.}   an opinion formed without adequate evidence.

\textbf{presumption} \emph{n.}   audacious (even arrogant) behavior that is offensive.

\textbf{pretence} \emph{n.}   a false or unsupportable quality; same as pretense.

\textbf{pretense} \emph{n.}   pretending with intention to deceive.

\textbf{pretension} \emph{n.}   a false or unsupportable pretext or quality.

\textbf{prevail} \emph{intr.v.} \textbf{-ed}   to be stronger or able to subdue; to win over or convince.

\textbf{prevailing} \emph{adj.}   most frequent or common.

\textbf{principal} \emph{adj.} \textbf{-ly}   initial money invested, not the income derived from it.

\textbf{private} \emph{adj.}    an enlisted soldier of the lowest rank.

\textbf{probable} \emph{adj.}   likely but not certain

\textbf{probity} \emph{n.}   complete and confirmed integrity.

\textbf{procure} \emph{v.} \textbf{-ed,-ing}   to obtain or acquire.

\textbf{prodigious} \emph{adj.} \textbf{-ly}   exceptional in size or extent; huge.

\textbf{profuse} \emph{adj.}   produced or growing in extreme abundance; plentiful.

\textbf{profusion} \emph{n.}   extremely abundant.

\textbf{prognostic} \emph{adj.}   a prediction, forcast, or indication as to the future.

\textbf{propensity} \emph{n.}   an inclination or tendency.

\textbf{propitious} \emph{adj.}   presenting favorable circumstances.

\textbf{proportionate} \emph{adj.}   agreeing or right in amount, magnitude, or degree.

\textbf{proprietor} \emph{n.}   someone who owns (is legal possessor of) a business.

\textbf{propriety} \emph{n.}   correct or appropriate behavior.

\textbf{prospect} \emph{n.}   the possibility of future success.

\textbf{proverb} \emph{n.}    A short saying that expresses a basic truth, value, or fact.

\textbf{provision} \emph{n.}   a stipulated condition.

\textbf{provocation} \emph{n.}   unfriendly behavior that aggravates or irritates someone; an incitement.

\textbf{provoke} \emph{tr.v.} \textbf{-ing}   to intentionally provoke anger or ill-will.

\textbf{proxy} \emph{n.}    a person authorized to act for another; an agent.

\textbf{prudence} \emph{n.}   discretion; knowing how to avoid embarrassment or distress.

\textbf{prudent} \emph{adj.} \textbf{-ly}    exercising good sense, tact, and judgement.

\textbf{prudential} \emph{adj.}   good judgment, or common sense, and the ability to avoid risk or embarassment.

\textbf{punctual} \emph{adj.} \textbf{-ity}   on time; prompt.

\textbf{purport} \emph{tr.v.}    the evident, but often false, meaning of a communication.

\textbf{querulous} \emph{adj.}   constantly complaining.

\textbf{ragout} \emph{n.}   well-seasoned meat or fish stew with vegetables.

\textbf{rail} \emph{n.}    to complain bitterly; to criticize severely.

\textbf{rally} \emph{v.} \textbf{-ed}    to assemble or join together for a common purpose.

\textbf{ramble} \emph{intr.v.} \textbf{-ed}    to move or speak aimlessly; to drone.

\textbf{rapacious} \emph{adj.}   insatiable desire for wealth.

\textbf{rapid} \emph{adj.} \textbf{-ity}    in quick succession or flow.

\textbf{rapture} \emph{n.}   extacy; bliss, especially from a religious experience.

\textbf{rapturous} \emph{adj.} \textbf{-ly}   3

\textbf{rash} \emph{adj.} \textbf{-ness}    haste or recklessness in face of danger.

\textbf{readily} \emph{adv.}    without difficulty; immediately.

\textbf{recede} \emph{tr.v.}   to move back and away from; to cede or give something back.

\textbf{recital} \emph{n.}    a performance of music or dance especially by soloists.

\textbf{recollection} \emph{n.}   something recalled to the mind.

\textbf{reconcile} \emph{v.} \textbf{-ed}   come to terms; to reestablish a relationship.

\textbf{reconciliation} \emph{n.}   reestablishing friendly relations.

\textbf{rectitude} \emph{n.}   uprightness as a consequence of being honorable and honest; righteousness.

\textbf{rector} \emph{n.}    A clergyman presiding over a parish in the Protestant Episcopal Church.

\textbf{rectory} \emph{n.}   the official residence of a parson provided by a church.

\textbf{redress} \emph{tr.v.}   to make reparations or amends for.

\textbf{reel} \emph{n.}    a lively dance of Scottish highlanders; marked by circular moves and gliding steps.

\textbf{refute} \emph{tr.v.}   to prove to be false or incorrect.

\textbf{regiment} \emph{n.}   army unit smaller than a division.

\textbf{regulation} \emph{n.}   an authoritative rule or law.

\textbf{remonstrance} \emph{n.}   an expression of opposition to a course of action.

\textbf{repel} \emph{v.}    to force or drive back; to ward off.

\textbf{repine} \emph{intr.v.} \textbf{-ing}   to wish discontentedly; to dejectedly complain.

\textbf{reprehensible} \emph{adj.}   deserving severe rebuke or censure.

\textbf{repress} \emph{v.} \textbf{-ed}   to put down by force or intimidation.

\textbf{reproach} \emph{tr.v.} \textbf{-ed}   to express criticism, disapproval, or to rebuke.

\textbf{reproof} \emph{n.}   an expression of criticism and censure; to scold or rebuke.

\textbf{repugnance} \emph{n.}   intense aversion or dislike.

\textbf{repugnant} \emph{adj.}   offensive to the mind; repulsive.

\textbf{repute} \emph{tr.v.} \textbf{-ed}   to ascribe a reputation to.

\textbf{requisite} \emph{adj.}   indispensable, required, or essential.

\textbf{requite} \emph{tr.v.} \textbf{-ed}    To make repayment or return in kind.

\textbf{resentment} \emph{n.}   a feeling of deep and bitter anger or ill-will.

\textbf{resolute} \emph{adj.} \textbf{-ly}   firm in purpose or belief; determined.

\textbf{retail} \emph{n.}    the selling of goods directly to consumers, usually in small quantities.

\textbf{retaliate} \emph{v.}   strike back or take revenge; to respond 'tit-for-tat'.

\textbf{retort} \emph{n.}   to reply to a question or remark (especially a witty or critical reply).

\textbf{retrospective} \emph{adj.}   Looking back on or thinking about the past.

\textbf{reverie} \emph{n.}   absent-minded or fanciful musing while awake; daydreaming.

\textbf{revolution} \emph{n.}   the overthrow of a government.

\textbf{sagacity} \emph{n.}   wisdom, good judgement, sage-like.

\textbf{sanction} \emph{n.} \textbf{-ed}   given authoritative approval.

\textbf{sarcastic} \emph{adj.}   expressing ridicule that wounds; marked by sarcasm.

\textbf{satirical} \emph{adj.}   exposing human folly to ridicule; relating to satire.

\textbf{saucy} \emph{adj.}    improperly forward or bold.

\textbf{savour} \emph{n.}    to taste with appreciation or pleasure.

\textbf{scandalous} \emph{adj.}   giving offense to moral sensibilities and injurious to reputation.

\textbf{scarcity} \emph{n.}   a small and inadequate amount.

\textbf{scruple} \emph{n.} \textbf{-ed}    a moral, principal, or ethic that guides a person's actions.

\textbf{scrutiny} \emph{n.}   A careful examination (as for mistakes).

\textbf{sedate} \emph{tr.v.} \textbf{-ness}    characterized by dignity, propriety, and calm.

\textbf{self-complacent} \emph{adj.}   Self-satisfied.

\textbf{self-imposed} \emph{adj.}   voluntarily assumed or endured.

\textbf{seminary} \emph{n.}   A school, especially a one for religious instruction.

\textbf{seniority} \emph{n.}   higher rank than others by reason of longer service.

\textbf{sentiment} \emph{n.}   A thought or attitude based mostly on emotion.

\textbf{sentinel} \emph{n.}   a guard or sentry.

\textbf{serenity} \emph{n.}   the absence of stress or anxiety; peaceful calm; tranquility.

\textbf{servile} \emph{adj.}    abject or slavish submissiveness.

\textbf{severe} \emph{adj.}    very bad in degree or extent.

\textbf{shrubbery} \emph{n.}   a collection of shrubs or bushes growing together.

\textbf{simper} \emph{v.} \textbf{-ed}    A constrained, self-conscious, or silly smile; a smirk.

\textbf{simpleton} \emph{n.}   a person lacking intelligence or common sense.

\textbf{singular} \emph{adj.}   the single one of its kind; individual.

\textbf{slighting} \emph{adj.} \textbf{-ly}   in a disparaging manner; belittling.

\textbf{solace} \emph{n.} \textbf{-ed}   the act of comforting in distress; consolation.

\textbf{solely} \emph{adv.}    alone; without any others involved.

\textbf{solemn} \emph{adj.}    dignified and somber in manner or character.

\textbf{solemnity} \emph{n.}   the trait of dignified seriousness.

\textbf{solicitude} \emph{n.}   care or concern; excessive concern.

\textbf{solidity} \emph{n.}   the state of being solid.

\textbf{solitary} \emph{adj.}   single and isolated from others; remote or secluded.

\textbf{solitude} \emph{n.}   the state of being alone or isolated from others.

\textbf{speculation} \emph{n.}   an opinion based on incomplete evidence.

\textbf{spleen} \emph{n.}    a large dark-red organ on the left side of the body.

\textbf{splendour} \emph{n.}   magnificent, splendid, or grand; same as splendor.

\textbf{spurn} \emph{v.} \textbf{-ed}    to rebuff (by a lover) without warning; to scorn or reject.

\textbf{squeamish} \emph{adj.}   easily sickened or disgusted.

\textbf{stately} \emph{adj.}    refined or imposing in manner or appearance; dignified.

\textbf{steadfast} \emph{adj.} \textbf{-ly}   with resolute determination.

\textbf{steady} \emph{adj.}   fixed or unmoving in place or nature; consistent or reliable.

\textbf{steward} \emph{n.}   someone who manages property or other affairs for someone else.

\textbf{stile} \emph{n.}    A vertical part of a panel or frame, as in a door or window sash.

\textbf{stout} \emph{adj.} \textbf{-ly}    in a resolute manner; having determination or boldness.

\textbf{stricture} \emph{n.}   A restriction, limit, or rule.

\textbf{studious} \emph{adj.} \textbf{-ly}   marked by care and effort; diligent in study.

\textbf{subjoin} \emph{tr.v.}   To add to the end; to append.

\textbf{succession} \emph{n.}   the action of following in order.

\textbf{successor} \emph{n.}   a person who follows another person in office or position.

\textbf{summons} \emph{n.}   a call in an official matter, like to appear in court.

\textbf{supercilious} \emph{adj.} \textbf{-ness}   feeling or expressing contempt or disdain.

\textbf{superintend} \emph{tr.v.} \textbf{-ence}   To oversee and direct; to supervise

\textbf{superior} \emph{adj.} \textbf{-ity}   better than something or someone else in quality or rank.

\textbf{superlative} \emph{adj.} \textbf{-ly}   the best; superior.

\textbf{supplicate} \emph{v.}   to humbly request help, especially in prayer.

\textbf{supposition} \emph{n.}   an opinion based on incomplete evidence.

\textbf{suppress} \emph{tr.v.} \textbf{-ed}   to put down or subdue by force or authority.

\textbf{surmise} \emph{v.}    To infer something based on inconclusive evidence.

\textbf{surmount} \emph{tr.v.}   To overcome.

\textbf{susceptibility} \emph{n.}   easily affected or influenced.

\textbf{tacit} \emph{adj.}   communicated but not said.

\textbf{taciturn} \emph{adj.}   habitually reserved and uncommunicative; untalkative.

\textbf{tedious} \emph{adj.}    so lacking in interest as to cause mental weariness.

\textbf{tte--tte} \emph{adv.}   a private conversation between two people.

\textbf{tiding} \emph{n.}    new information about specific and timely events; news.

\textbf{tithe} \emph{th}    Ten percent of one's income given to chairity or cause, especially to a church.

\textbf{tolerable} \emph{adj.} \textbf{-ly}   able to be tolerated or endured.

\textbf{tractable} \emph{adj.}   Easily managed or controlled.

\textbf{tranquil} \emph{adj.} \textbf{-ly}   without emotional agitation; calm.

\textbf{tranquillity} \emph{n.}   a serene state of peace and quiet.

\textbf{trepidation} \emph{n.}   a feeling of alarm or dread.

\textbf{trifling} \emph{n.}   not worth considering because of little value or importance.

\textbf{triumph} \emph{intr.v.}   a successful ending of a struggle or contest; a win.

\textbf{triumphant} \emph{adj.} \textbf{-ly}   victorious.

\textbf{tumult} \emph{n.}    a state of commotion and loud confused noise.

\textbf{unabated} \emph{adj.}   continuing at full strength or intensity with no decrease.

\textbf{unaffected} \emph{adj.} \textbf{-ly}   free or artificiality; sincere and genuine.

\textbf{unalloyed} \emph{adj.}   Not mixed with other metals.

\textbf{unassuming} \emph{adj.}   not arrogant or presuming; modest in conduct.

\textbf{unavailing} \emph{adj.}   producing no result or effect.

\textbf{uncivil} \emph{adj.}   lacking decency or good manners.

\textbf{unconscious} \emph{adj.} \textbf{-ly}   without awareness; lacking sensory perception; not conscious.

\textbf{unshackle} \emph{tr.v.} \textbf{-ed}   To free from shackles or chains.

\textbf{upbraid} \emph{tr.v.} \textbf{-ed}    to give a severe scolding.

\textbf{vast} \emph{adj.} \textbf{-ly}    huge, expansive, very large.

\textbf{vehement} \emph{adj.}   intensity or forcefulness of expression.

\textbf{veneration} \emph{n.}   a profound awe inspired by a deity.

\textbf{venison} \emph{n.}   meat from a deer used as food.

\textbf{veracity} \emph{n.}   truthfulness.

\textbf{verdure} \emph{n.}    lush green plant growth and foliage.

\textbf{verily} \emph{adv.}    (archaic) truly or confidentially.

\textbf{vestibule} \emph{n.}   a large entrance or reception room or area.

\textbf{vexation} \emph{n.}   something or someone that causes anxiety; a source of unhappiness.

\textbf{vexatious} \emph{adj.}   causing irritation or annoyance.

\textbf{vex} \emph{tr.v.} \textbf{-ed}    to trouble or bother, especially with petty annoyances.

\textbf{villainous} \emph{adj.}   extremely wicked.

\textbf{virtue} \emph{n.}   morality with respect to sexual relations.

\textbf{vivacity} \emph{n.}   characterized by high spirits and animation.

\textbf{vogue} \emph{n.}    the popular taste, style, or fashion at a given time.

\textbf{void} \emph{adj.}    an empty area or space; nonexistence.

\textbf{volatile} \emph{adj.}   being easily excited.

\textbf{voluble} \emph{adj.}    fluent, ready, and easy in speach.

\textbf{vouch} \emph{v.}    to give personal assurance; to guarantee.

\textbf{vouchsafe} \emph{tr.v.} \textbf{-ed}   to grant or agree to in a condescending manner.

\textbf{vulgar} \emph{adj.}    tastelessly indecent; lacking refinement or cultivation.

\textbf{vulgarity} \emph{n.}   lacking taste and refinement.

\textbf{wanton} \emph{adj.} \textbf{-ly}    in a lewd, immoral, or promiscuous manner.

\textbf{whimsical} \emph{adj.}   determined by impulse or whim rather than necessity or reason.

\textbf{whist} \emph{n.}    a card game for four people, played in partnetships, using 52 cards.

\textbf{witticism} \emph{n.}   A witty, clever, or amusing remark or saying.

\textbf{wretched} \emph{adj.} \textbf{-ness}   a state affliction or misfortune; miserable.

\printindex
\end{document}
